---
title: "Module 2: Hypothesis Testing"
subtitle: "The Logic of Randomization and What P-Values Actually Tell You"
---

## Overview

Null hypothesis significance testing is everywhere in management research, yet its logic is widely misunderstood. This module builds from the ground up: why does randomization justify causal claims? What does a p-value actually mean? And what are the hidden assumptions — about units, stimuli, and assignment mechanisms — that most researchers ignore?

We cover the permutation logic of p-values, Type I and Type II error, power, the multiple comparisons problem, researcher degrees of freedom, and the design features (between vs. within, stimulus sampling, Latin squares) that change what you can conclude.

---

## Learning Goals

By the end of this module you should be able to:

- Derive the sampling distribution of a test statistic from the randomization distribution, not from asymptotic theory
- Explain the difference between a two-sided p-value and a posterior probability
- Calculate and interpret statistical power, and explain why most social science studies are underpowered
- Identify researcher degrees of freedom and explain how they inflate false-positive rates
- Explain the rationale for stimulus sampling and design studies that treat stimuli as random effects
- Construct a Latin square design and explain when it is preferable to a fully crossed factorial

---

## Recommended Reading

| Paper | Why it matters |
|---|---|
| Cohen (1994) — *The Earth is Round (p < .05)* | The classic critique of NHST, still the best starting point |
| Simmons, Nelson & Simonsohn (2011) — *False-Positive Psychology* | Demonstrates how researcher degrees of freedom inflate false positives |
| Simonsohn, Nelson & Simmons (2014) — *P-curve* | A tool for detecting publication bias and assessing evidential value |
| Wells & Windschitl (1999) — *Stimulus Sampling* | Why treating stimuli as fixed effects is a validity threat |
| Bahník & Vranka (2017) | A worked example of stimulus sampling in practice |
| Andre (2022) — *How (Not) to Exclude Outliers* | Researcher degrees of freedom in outlier exclusion |
| Smaldino (2016) — *Models are Stupid, and We Need More of Them* | Why formal models help you think clearly about hypothesis testing |

All papers are available in the `Module 2 - Recommended Reading/` folder.

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  fig.width  = 9,
  fig.height = 5,
  fig.path   = "figs/",
  dev        = "svg"
)
dir.create("figs", recursive = TRUE, showWarnings = FALSE)
```

```{=html}
<style>
/* ── Floating bottom control bar ──────────────────────────────────────────── */
#global-output-toggle {
  position: fixed; bottom: 0; left: 0; right: 0; z-index: 9999;
  display: flex; flex-wrap: wrap; gap: 10px;
  justify-content: center; align-items: center;
  padding: 11px 24px;
  background: #1e3a5f; border-top: 3px solid #4a90d9;
  box-shadow: 0 -3px 16px rgba(0,0,0,0.35);
}
#global-output-toggle span {
  color: #c8d8ed; font-size: 0.85em; font-weight: 700;
  letter-spacing: 0.6px; text-transform: uppercase; margin-right: 6px;
}
body { padding-bottom: 70px !important; }
.gtbtn {
  padding: 9px 24px; color: #ffffff; border-radius: 7px;
  font-size: 1em; font-weight: 700; cursor: pointer;
  letter-spacing: 0.2px; transition: background 0.15s, border-color 0.15s;
  white-space: nowrap; border: 2px solid transparent;
}
/* Code button — blue */
.gtbtn.code-btn   { background: #2563a8; border-color: #5b9bd5; }
.gtbtn.code-btn:hover { background: #1a4a80; border-color: #90bfe8; }
/* Output button — teal/green */
.gtbtn.output-btn { background: #2d6a4f; border-color: #52b788; }
.gtbtn.output-btn:hover { background: #1b4332; border-color: #74c69d; }
/* ── Per-output collapsible wrappers ──────────────────────────────────────── */
details.out-fold { margin: 6px 0 14px 0; }
summary.out-fold-btn {
  display: inline-flex; align-items: center; gap: 7px;
  padding: 8px 20px; background: #eef4fb;
  border: 2px solid #4a90d9; border-radius: 7px; cursor: pointer;
  font-size: 0.94em; font-weight: 700; color: #1a3a6b;
  user-select: none; list-style: none; transition: background 0.12s;
}
summary.out-fold-btn::-webkit-details-marker { display: none; }
summary.out-fold-btn:hover { background: #cfe2f7; }
details[open].out-fold > summary.out-fold-btn { background: #d6ecff; }
summary.out-fold-btn .fold-icon { font-size: 1em; }
details[open] .fold-icon-closed { display: none; }
details:not([open]) .fold-icon-open  { display: none; }
</style>

<script>
document.addEventListener('DOMContentLoaded', function () {
  var outputSels = ['.cell-output-stdout','.cell-output-stderr','.cell-output-display'];
  document.querySelectorAll(outputSels.join(',')).forEach(function (el) {
    if (el.closest('.callout') || el.closest('details.out-fold')) return;
    var hasImg = !!el.querySelector('img, svg, canvas');
    var hasTbl = !!el.querySelector('table');
    var label  = hasImg ? '📊  Show plot' : hasTbl ? '📋  Show table' : '📄  Show output';
    var det = document.createElement('details');
    det.className = 'out-fold';
    var sum = document.createElement('summary');
    sum.className = 'out-fold-btn';
    sum.innerHTML =
      '<span class="fold-icon fold-icon-closed">▶</span>' +
      '<span class="fold-icon fold-icon-open">▼</span>' + label;
    el.parentNode.insertBefore(det, el);
    det.appendChild(sum); det.appendChild(el);
  });
  var bar = document.createElement('div');
  bar.id = 'global-output-toggle';
  bar.innerHTML =
    '<span>Quick controls:</span>' +
    '<button id="code-toggle-btn"   class="gtbtn code-btn"   onclick="toggleCode()">\u25bc Expand All Code</button>' +
    '<button id="output-toggle-btn" class="gtbtn output-btn" onclick="toggleOutput()">\u25bc Expand All Output</button>';
  document.body.appendChild(bar);
});

var codeOpen   = false;
var outputOpen = false;

function toggleCode() {
  codeOpen = !codeOpen;
  document.querySelectorAll('details').forEach(function (d) {
    if (!d.classList.contains('out-fold') && d.querySelector('.sourceCode')) {
      d.open = codeOpen;
    }
  });
  var btn = document.getElementById('code-toggle-btn');
  if (btn) btn.textContent = codeOpen ? '\u25b2 Collapse All Code' : '\u25bc Expand All Code';
}

function toggleOutput() {
  outputOpen = !outputOpen;
  document.querySelectorAll('details.out-fold').forEach(function (d) {
    d.open = outputOpen;
  });
  var btn = document.getElementById('output-toggle-btn');
  if (btn) btn.textContent = outputOpen ? '\u25b2 Collapse All Output' : '\u25bc Expand All Output';
}
</script>
```

## What This Tutorial Is About

There is a single thread running through this tutorial and its predecessor:

> **Your observable may not accurately reflect the latent construct you need.**

In Module 1, that problem lived in your *outcome variable Y* — your scale picked up variance from constructs you never intended to measure. Here, the same problem appears in **two places at once**: in your *treatment variable X*, and in the act of *randomization* that is supposed to make X interpretable.

This tutorial covers three ideas:

- **Part 1 — P-values:** What they actually mean, where their inferential power comes from, and what assumptions they require — demonstrated by building null distributions from scratch using permutation.

- **Part 2 — The Exclusion Restriction:** Just as a measured outcome Y can fail discriminant validity by absorbing multiple constructs, a manipulated treatment X can inject multiple independent signals into the system simultaneously. When this happens, your "experiment" is really a quasi-experiment — and the single coefficient on treatment conflates several distinct causal pathways.

- **Part 3 — Randomization:** Why "clicking randomize" (the observable) is not the same as "achieving randomization" (the latent property), and why this gap grows as the construct Y you are studying becomes broader and more complex.

::: {.callout-note}
## Learning Objectives

By the end of this tutorial, you will be able to:

- Explain what a p-value is — and is **not** — using simulation rather than formulas
- Build a null distribution from scratch using label permutation
- Identify the three ingredients that give a p-value inferential power
- Explain why a treatment manipulation can violate discriminant validity in the same way a measurement scale can
- Use open-ended text to map the constructs injected by an experimental manipulation and those driving an outcome
- Articulate the difference between **observable randomization** and **latent randomization** using the construct-validity language from Module 1
- Calculate the minimum observations needed for approximate orthogonality given the complexity of Y
- Run diagnostic checks to assess whether your observed randomization achieved approximate orthogonality
:::

---

```{r packages}
#| code-summary: "▶ Load required packages"
# install.packages(c("tidyverse","ggplot2","knitr","scales","patchwork"))

library(tidyverse)
library(ggplot2)
library(knitr)
library(scales)
library(patchwork)
if (!requireNamespace("lavaan", quietly = TRUE)) install.packages("lavaan")
library(lavaan)

set.seed(2025)
```

---
