---
title: "Part 5: Design Solutions and Assignment Strategies"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  fig.width  = 9,
  fig.height = 5,
  fig.path   = "figs/",
  dev        = "svg"
)
dir.create("figs", recursive = TRUE, showWarnings = FALSE)
```

```{r packages}
#| code-summary: "▶ Load required packages"
library(tidyverse); library(ggplot2); library(knitr); library(scales); library(patchwork)
if (!requireNamespace("lavaan", quietly = TRUE)) install.packages("lavaan")
library(lavaan)
set.seed(2025)
```

# Part 5: Experimental Design Strategies — Using Structure to Buy Efficiency {#sec-designs}

## The Core Tension

Parts 3 and 4 showed a fundamental tension:

- **More dimensions of Y** → larger design space → more N needed for orthogonality
- **More N** → more expensive → temptation to reduce sample → back to poor orthogonality

Section 4 offered two partial design-based solutions (stratification, clustering). Here we go further: rather than trying to *achieve* orthogonality with brute-force N, we ask whether **the experimental design itself** can reduce the size of the design space you need to fill.

## Completely Random, Factorial, and Latin Square Designs

The simplest way to see the tradeoffs is to contrast three design types on a common example.

**Scenario:** You want to test the effect of:

- **Factor A:** Eco-certification badge (present / absent) — 2 levels
- **Factor B:** Price level (low / medium / high) — 3 levels
- **Factor C:** Message framing (individual benefit / social benefit) — 2 levels

```{r design-comparison, fig.height=6, fig.width=11}
#| code-summary: "▶ Visualize: CRD vs. Factorial vs. Latin Square cell structures"

# ── Fully factorial: all A×B×C combinations ──────────────────────────────────
fact_df <- expand.grid(
  EcoBadge  = c("No Badge", "Eco Badge"),
  PriceLevel = c("Low", "Medium", "High"),
  Framing    = c("Individual\nBenefit", "Social\nBenefit")
) |>
  mutate(cell = paste(EcoBadge, PriceLevel, Framing, sep = "\n"),
         design = "Fully Factorial\n(12 cells, 12N total)")

# ── Latin square: Factor C varied across rows and Factor A×B as column×row
# 2×3 Latin Square (accommodate unequal levels by using 3×3 and dropping one row)
ls_grid <- expand.grid(Row = 1:2, Col = 1:3) |>
  mutate(
    EcoBadge   = ifelse(Col %in% c(1, 3), "No Badge", "Eco Badge"),
    PriceLevel = c("Low","Medium","High","High","Low","Medium"),
    Framing    = c("Individual\nBenefit","Social\nBenefit","Individual\nBenefit",
                   "Social\nBenefit","Individual\nBenefit","Social\nBenefit"),
    design     = "Latin Square\n(6 cells, 6N total)"
  )

# ── CRD: just two cells (eco vs. no eco, collapsing B and C) ─────────────────
crd_df <- data.frame(
  EcoBadge   = c("No Badge", "Eco Badge"),
  PriceLevel = "Pooled\n(B,C varied\nbut uncontrolled)",
  Framing    = "Pooled",
  design     = "Simple A/B (CRA)\n(2 cells, N total)",
  Row = 1, Col = 1:2
)

# ── Plot factorial and latin square side by side ──────────────────────────────
p_fact <- ggplot(fact_df, aes(x = PriceLevel, y = Framing, fill = EcoBadge)) +
  geom_tile(color = "white", linewidth = 1.2, width = 0.95, height = 0.95) +
  geom_text(aes(label = paste0("n\u2265", ifelse(EcoBadge == "Eco Badge", "1", "1"))),
            size = 3.5, fontface = "bold", color = "gray20") +
  scale_fill_manual(values = c("No Badge" = "#d6ecff", "Eco Badge" = "#2d6a4f"),
                    name = "Eco Badge") +
  labs(title = "Fully Factorial (12 conditions)",
       subtitle = "Every combination tested  \u00b7  Min N = 12\nTests A, B, C main effects AND interactions",
       x = "Price Level", y = "Message Framing") +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(), legend.position = "top")

p_ls <- ggplot(ls_grid, aes(x = factor(Col), y = factor(Row), fill = EcoBadge)) +
  geom_tile(color = "white", linewidth = 1.5, width = 0.92, height = 0.92) +
  geom_text(aes(label = paste0(EcoBadge, "\n", PriceLevel, "\n", Framing)),
            size = 2.9, lineheight = 0.85, color = "gray10") +
  scale_fill_manual(values = c("No Badge" = "#d6ecff", "Eco Badge" = "#2d6a4f"),
                    name = "Eco Badge") +
  labs(title = "Latin Square (6 conditions)",
       subtitle = "Each factor level appears exactly once per row and column\nMin N = 6  \u00b7  Tests main effects only (no interactions)",
       x = "Column (order slot / price)", y = "Row (participant block)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(), legend.position = "top",
        axis.text = element_text(size = 9))

p_fact + p_ls
```

```{r design-efficiency-table}
#| code-summary: "▶ Table: design comparison on key dimensions"

design_tab <- data.frame(
  Design = c("Simple A/B (CRA)", "Fully Factorial", "Latin Square"),
  `Conditions` = c(2, 12, 6),
  `Min N (1 obs/cell)` = c(2, 12, 6),
  `Main effects testable` = c("A only", "A, B, C", "A, B, C"),
  `Interactions testable` = c("None", "A\u00d7B, A\u00d7C, B\u00d7C, A\u00d7B\u00d7C", "None"),
  `Orthogonality coverage` = c("Only A vs. B,C confounded", "Perfect within cells", "Approximate — by design"),
  check.names = FALSE
)

kable(design_tab,
      caption = "Design comparison: efficiency, testability, and orthogonality coverage")
```

::: {.callout-note}
## Connecting to Section 3: how designs reduce the effective design space

Recall from Section 3.6 that the required N for approximate orthogonality grows with the number of dimensions D and falls as $r_{\max}$ increases. A Latin Square design directly reduces the effective D by **structurally balancing** factors across the design matrix rather than relying on chance:

- A fully factorial 2×3×2 design with N = 48 (4 per cell) achieves exact balance on A, B, and C by construction.
- A Latin Square of the same factors with N = 24 achieves approximate balance by ensuring each factor level appears an equal number of times across the other factors' levels.
- A simple A/B study with N = 48 leaves B and C entirely to chance — back to the orthogonality problem from Section 3.

The Latin Square is a direct structural solution to the Section 3 problem: **it buys you orthogonality through design structure rather than through sample size.**
:::

## Interactive Latin Square Design Builder

Use the controls below to design your own Latin Square experiment. Set the number of factors, give each one a name and the number of levels you need, and specify your participant budget per cell. The grid updates instantly — colors show which level of the primary factor each cell receives, labels inside the cells show the other factors.

```{ojs}
//| echo: false
//| code-fold: false

viewof ls_config = Inputs.form({
  n_fac:      Inputs.range([2, 5], {step: 1, value: 3,
                label: "Number of factors to study"}),
  a_name:     Inputs.text({label: "Factor A — name",   value: "Eco badge",    width: 160}),
  a_levels:   Inputs.select([2, 3, 4], {label: "Factor A — levels", value: 2}),
  b_name:     Inputs.text({label: "Factor B — name",   value: "Price level",  width: 160}),
  b_levels:   Inputs.select([2, 3, 4], {label: "Factor B — levels", value: 3}),
  c_name:     Inputs.text({label: "Factor C — name",   value: "Framing",      width: 160}),
  c_levels:   Inputs.select([2, 3, 4], {label: "Factor C — levels", value: 2}),
  d_name:     Inputs.text({label: "Factor D — name",   value: "Source cred.", width: 160}),
  d_levels:   Inputs.select([2, 3, 4], {label: "Factor D — levels", value: 2}),
  e_name:     Inputs.text({label: "Factor E — name",   value: "Ad copy",      width: 160}),
  e_levels:   Inputs.select([2, 3, 4], {label: "Factor E — levels", value: 2}),
  n_per_cell: Inputs.range([1, 30], {step: 1, value: 5,
                label: "Participants per cell"}),
})
```

```{ojs}
//| echo: false
//| code-fold: false
//| output: false

ls_state = {
  const nf   = ls_config.n_fac
  const defs = [
    {name: ls_config.a_name || "Factor A", levels: +ls_config.a_levels},
    {name: ls_config.b_name || "Factor B", levels: +ls_config.b_levels},
    {name: ls_config.c_name || "Factor C", levels: +ls_config.c_levels},
    {name: ls_config.d_name || "Factor D", levels: +ls_config.d_levels},
    {name: ls_config.e_name || "Factor E", levels: +ls_config.e_levels},
  ]
  const facs = defs.slice(0, nf)
  const maxL = Math.max(...facs.map(f => f.levels))

  const cells = []
  for (let r = 0; r < maxL; r++) {
    for (let c = 0; c < maxL; c++) {
      const cell = {row: r + 1, col: c + 1}
      facs.forEach((fac, k) => {
        cell["lv" + k]    = ((r + c + k) % fac.levels) + 1
        cell["short" + k] = (fac.name || ("F" + (k+1))).substring(0, 4) + "\u202f" +
                             (((r + c + k) % fac.levels) + 1)
      })
      cell.fa_label     = "Level\u00a0" + (((r + c) % facs[0].levels) + 1)
      cell.other_labels = facs.slice(1)
        .map((f, i) => (f.name || ("F" + (i+2))).substring(0, 4) + "\u202f" +
                        (((r + c + i + 1) % f.levels) + 1))
        .join("  ")
      cells.push(cell)
    }
  }

  const full_factorial = facs.reduce((acc, f) => acc * f.levels, 1)
  const ls_cells       = maxL * maxL
  const n_per_cell     = +ls_config.n_per_cell

  return {facs, cells, maxL, full_factorial, ls_cells,
          total_n: ls_cells * n_per_cell, n_per_cell, nf}
}
```

```{ojs}
//| echo: false
//| code-fold: false

{
  const {facs, cells, maxL} = ls_state

  const palettes = {
    2: ["#4e79a7","#f28e2b"],
    3: ["#4e79a7","#f28e2b","#59a14f"],
    4: ["#4e79a7","#f28e2b","#59a14f","#e15759"]
  }
  const pal    = palettes[Math.min(facs[0].levels, 4)]
  const cellPx = Math.max(80, Math.min(150, 540 / maxL))

  return Plot.plot({
    width:       cellPx * maxL + 110,
    height:      cellPx * maxL + 90,
    marginLeft:  85, marginBottom: 52, marginTop: 36, marginRight: 10,
    x: { domain: Array.from({length: maxL}, (_, i) => i + 1),
         tickFormat: d => "Pos. " + d, label: "Order position \u2192" },
    y: { domain: Array.from({length: maxL}, (_, i) => i + 1),
         tickFormat: d => "Block " + d, label: "\u2190 Participant block" },
    color: {
      domain: Array.from({length: facs[0].levels}, (_, i) => "Level\u00a0" + (i+1)),
      range:  pal, legend: true,
      label:  (facs[0].name || "Factor A") + " (cell color)"
    },
    marks: [
      Plot.cell(cells, {
        x: "col", y: "row", fill: "fa_label",
        stroke: "white", strokeWidth: 3, rx: 6
      }),
      Plot.text(cells, {
        x: "col", y: "row",
        text: d => (facs[0].name || "A").substring(0, 4) + "\u202f" + d["lv0"],
        fill: "white", fontWeight: "bold", fontSize: 13,
        dy: facs.length > 1 ? -9 : 0
      }),
      ...(facs.length > 1 ? [Plot.text(cells, {
        x: "col", y: "row",
        text: "other_labels",
        fill: "white", fontSize: 9.5, opacity: 0.92, dy: 9
      })] : [])
    ]
  })
}
```

```{ojs}
//| echo: false
//| code-fold: false

{
  const {facs, cells, ls_cells, full_factorial, total_n, n_per_cell, maxL} = ls_state
  const pct   = Math.round(ls_cells / full_factorial * 100)
  const saved = full_factorial - ls_cells

  const balance = facs.map((fac, k) => {
    const expected = maxL / fac.levels
    let ok = true
    for (let r = 1; r <= maxL; r++) {
      for (let lv = 1; lv <= fac.levels; lv++) {
        const n = cells.filter(c => c.row === r && c["lv" + k] === lv).length
        if (Math.abs(n - expected) > 0.49) ok = false
      }
    }
    return {name: fac.name, levels: fac.levels, expected: expected.toFixed(1), balanced: ok}
  })

  return html`
  <div style="display:flex;gap:14px;flex-wrap:wrap;margin:16px 0 10px">
    <div style="flex:1;min-width:110px;background:#e8f4f8;border-radius:10px;
                padding:13px 16px;text-align:center;border-left:4px solid #457b9d">
      <div style="font-size:2em;font-weight:700;color:#457b9d;line-height:1">${ls_cells}</div>
      <div style="font-size:0.78em;color:#555;margin-top:4px">Latin Square<br>conditions</div>
    </div>
    <div style="flex:1;min-width:110px;background:#fff8e1;border-radius:10px;
                padding:13px 16px;text-align:center;border-left:4px solid #f9a825">
      <div style="font-size:2em;font-weight:700;color:#f9a825;line-height:1">${full_factorial}</div>
      <div style="font-size:0.78em;color:#555;margin-top:4px">Full factorial<br>conditions</div>
    </div>
    <div style="flex:1;min-width:110px;background:#e8f5e9;border-radius:10px;
                padding:13px 16px;text-align:center;border-left:4px solid #388e3c">
      <div style="font-size:2em;font-weight:700;color:#388e3c;line-height:1">${pct}%</div>
      <div style="font-size:0.78em;color:#555;margin-top:4px">Design efficiency<br>(vs. full factorial)</div>
    </div>
    <div style="flex:1;min-width:110px;background:#fce4ec;border-radius:10px;
                padding:13px 16px;text-align:center;border-left:4px solid #c62828">
      <div style="font-size:2em;font-weight:700;color:#c62828;line-height:1">${total_n}</div>
      <div style="font-size:0.78em;color:#555;margin-top:4px">Total participants<br>(${n_per_cell}/cell)</div>
    </div>
  </div>
  ${saved > 0
    ? html`<p style="color:#388e3c;font-size:0.88em;margin:6px 0 10px">
        <strong>&#10003; Savings:</strong> Testing all ${facs.length} factors with a Latin Square
        uses <strong>${ls_cells} conditions</strong> instead of ${full_factorial} &mdash;
        saving ${saved} conditions (${Math.round(saved/full_factorial*100)}% fewer cells)
        without losing any main-effect estimation power.
      </p>`
    : html`<p style="color:#666;font-size:0.88em;margin:6px 0 10px">
        All factors have the same number of levels, so the Latin Square equals
        the full factorial here. Try mixing levels (e.g., 2 and 3) to see the efficiency gain.
      </p>`
  }
  <details style="margin-top:6px">
    <summary style="cursor:pointer;color:#457b9d;font-weight:600;font-size:0.88em">
      &#9654; Balance check &mdash; how many times does each factor level appear per row?
    </summary>
    <table style="border-collapse:collapse;font-size:0.83em;margin-top:8px;width:100%">
      <thead>
        <tr style="background:#f0f4f8">
          <th style="padding:5px 10px;text-align:left;border:1px solid #ddd">Factor</th>
          <th style="padding:5px 10px;text-align:center;border:1px solid #ddd">Levels</th>
          <th style="padding:5px 10px;text-align:center;border:1px solid #ddd">Appearances per row</th>
          <th style="padding:5px 10px;text-align:center;border:1px solid #ddd">Balanced?</th>
        </tr>
      </thead>
      <tbody>
        ${balance.map(b => html`<tr>
          <td style="padding:5px 10px;border:1px solid #ddd"><strong>${b.name}</strong></td>
          <td style="padding:5px 10px;text-align:center;border:1px solid #ddd">${b.levels}</td>
          <td style="padding:5px 10px;text-align:center;border:1px solid #ddd">${b.expected}</td>
          <td style="padding:5px 10px;text-align:center;border:1px solid #ddd">
            ${b.balanced
              ? html`<span style="color:#388e3c;font-weight:bold">&#10003; Perfect</span>`
              : html`<span style="color:#f9a825;font-weight:bold">&#8776; Approximate</span>`}
          </td>
        </tr>`)}
      </tbody>
    </table>
  </details>`
}
```

::: {.callout-warning}
## Latin Squares cannot estimate interactions
A Latin Square tests each factor's **main effect** using the full sample, but it **cannot** detect whether factors interact (e.g., whether the eco-badge effect depends on price level). If an interaction is theoretically important, use a full factorial design for those two factors and apply the Latin Square only to the remaining ones.
:::

### Analysis code for Latin Square designs

```{r latin-analysis-code, eval=FALSE}
#| code-summary: "▶ Analysis template: fit Latin Square model"

# Replace Y, Factor_A, Factor_B, Factor_C with your actual variable names.
# Row and Column are nuisance blocking variables — always include them.

# Two-factor model
fit2 <- lm(Y ~ Factor_A + Factor_B + Row + Column, data = your_df)
emmeans(fit2, pairwise ~ Factor_A, adjust = "holm")$contrasts
emmeans(fit2, pairwise ~ Factor_B, adjust = "holm")$contrasts

# Three-factor model
fit3 <- lm(Y ~ Factor_A + Factor_B + Factor_C + Row + Column, data = your_df)
emmeans(fit3, pairwise ~ Factor_A, adjust = "holm")$contrasts
emmeans(fit3, pairwise ~ Factor_B, adjust = "holm")$contrasts
emmeans(fit3, pairwise ~ Factor_C, adjust = "holm")$contrasts
```

---

## Metastudies: Pooling Power Across Multiple Factors

The Latin Square is efficient because it uses design structure to study multiple factors simultaneously. A related idea scales this up to the level of a research program: the **metastudy** (DeKay, Rubinchik, Li, & De Boeck, 2022, *Perspectives on Psychological Science*).

A metastudy is a single, pre-registered study that simultaneously tests many experimental factors — what DeKay et al. call *facets* — by crossing them in a balanced factorial design. Each participant sees one randomly sampled combination of facet levels (a *microstudy*); the researcher estimates each facet's main effect by contrasting all participants who received the high level of that facet against all who received the low level, averaged across all other facets.

**Why this provides more power per hypothesis than running separate A/B tests:**

The key insight is about how the fixed participant budget is allocated. In a metastudy, all N participants are exposed to all K factors simultaneously. When estimating the effect of any one factor, the *entire* sample contributes to that estimate — N/2 participants in each "level" of that factor — regardless of how many other factors are also being tested. This is because the factors are orthogonally assigned, so each factor's effect is estimated independently of the others using the full N.

By contrast, if you run K separate A/B tests and divide your budget equally, each test gets only N/K participants. Power for each test depends on N/K, which shrinks as K grows.

**The comparison with fixed total N:** if you have N = 400 participants and want to test K hypotheses:
- **Metastudy:** all 400 test all K factors. Each factor is estimated using all 400 participants → power stays constant.
- **Separate A/B tests:** 400/K participants per factor → as K grows, per-test sample shrinks → power falls.

At K = 1, the two designs are *identical*: a single A/B test with 400 participants. As K grows, they diverge sharply. As DeKay et al. (2022) put it, metastudies "intentionally and completely sacrifice power at the level of the individual microstudies to achieve greater power for testing the effects of interest at the aggregate level" — a counterintuitive but mathematically principled trade-off.

```{r metastudy-power, fig.height=5.5, fig.width=12}
#| code-summary: "▶ Simulate: metastudy maintains power as K grows; separate A/B tests collapse"

set.seed(2026)
K_vec   <- 1:7       # 1 to 7 factors (stop at K = 7)
N_total <- 400       # fixed total budget (same for both designs)
d       <- 0.30      # true effect size per factor
n_sim   <- 1000      # simulations per cell

# ── Metastudy: all N=400 test all K factors simultaneously (orthogonal OLS) ────
# Key: each factor is assigned orthogonally, so estimating Factor 1 uses all N
# participants (N/2 per level), regardless of K. SE ≈ 2*sigma/sqrt(N) = constant.
# At K=1 this is identical to a single A/B test with N participants.
sim_meta_power <- function(N, K, d, n_sim = 500) {
  mean(replicate(n_sim, {
    # Random ±0.5 assignment to each of K orthogonal factors
    X   <- matrix(sample(c(-0.5, 0.5), N * K, replace = TRUE), nrow = N)
    Y   <- X %*% rep(d, K) + rnorm(N, 0, 1)
    fit <- lm(Y ~ X)
    # p-value for Factor 1 (all factors symmetric by design)
    coef(summary(fit))[2, 4] < 0.05
  }))
}

# ── Separate A/B tests: total N = 400 split EQUALLY among K tests ─────────────
# Each of the K factors gets its own study with N/K participants.
# At K=1: N_per_test = 400 → identical to metastudy (K=1 case) → equal power.
# As K grows: N_per_test = 400/K shrinks → power falls.
sim_ab_power <- function(N_per_test, d, n_sim = 500) {
  n_t <- ceiling(N_per_test / 2)
  n_c <- floor(N_per_test / 2)
  mean(replicate(n_sim, {
    obs <- mean(rnorm(n_t,  d / 2, 1)) - mean(rnorm(n_c, -d / 2, 1))
    se  <- sqrt(1/n_t + 1/n_c)
    abs(obs / se) > qnorm(0.975)
  }))
}

meta_powers <- sapply(K_vec, function(K) sim_meta_power(N_total, K, d, n_sim))
ab_powers   <- sapply(K_vec, function(K) sim_ab_power(floor(N_total / K), d, n_sim))

power_df <- data.frame(
  n_factors = rep(K_vec, 2),
  power     = c(meta_powers, ab_powers),
  design    = rep(c("Metastudy (total N = 400; all factors tested simultaneously)",
                    "Separate A/B tests (N = 400 / K per test)"),
                 each = length(K_vec))
)

ggplot(power_df, aes(x = n_factors, y = power,
                     color = design, group = design)) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 3.5) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "gray40") +
  annotate("text", x = 7.05, y = 0.815, label = "80% power",
           color = "gray40", size = 3.2, hjust = 0) +
  scale_color_manual(
    values = c("Metastudy (total N = 400; all factors tested simultaneously)" = "#2d6a4f",
               "Separate A/B tests (N = 400 / K per test)"                   = "#e63946"),
    name = NULL) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
  scale_x_continuous(breaks = K_vec) +
  labs(x = "Number of factors tested (K)",
       y = "Power per factor (d = 0.30)",
       title = "Metastudy Maintains Power as K Grows; Separate A/B Tests Collapse",
       subtitle = paste0(
         "Both designs use the same total N = 400  \u00b7  d = 0.30  \u00b7  At K=1 both designs are identical (one A/B test, N=400)\n",
         "Metastudy: all 400 contribute to each factor\u2019s estimate \u2192 power stays constant  \u00b7  ",
         "Separate A/B: 400/K per test \u2192 power falls")) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(), legend.position = "top",
        legend.text = element_text(size = 9),
        plot.margin = margin(5, 30, 5, 5))
```

**Reading the plot:** at K = 1, both designs are literally the same study — one A/B test with 400 participants — and achieve the same power (~85%). As K grows, the designs diverge. The metastudy *maintains* near-constant high power because all 400 participants contribute to each factor's estimate; the orthogonal design means Factor 1's SE is always ≈ 2/√N regardless of how many other factors are also being tested. Separate A/B tests collapse because the budget is divided: at K = 7, each test gets only 400/7 ≈ 57 participants, pushing power below 20%. The metastudy tests 7× as many hypotheses at the same power level using exactly the same resources.

**The connection to Section 3:** a metastudy also directly reduces the Section 3 problem. By deliberately varying many potential confounders as factors in the design (rather than leaving them to chance), you ensure they are orthogonally distributed across conditions *by construction* — transforming unobserved confounders into *measured experimental factors* and extracting their effects cleanly.

---

## Within-Participant Designs: Same Observations, More Power

### The Intuition

In all the designs above, we compared *different people* in different conditions. A within-participant (repeated-measures) design instead exposes *the same person* to multiple conditions, measuring the outcome each time.

**Why does this increase power?** Consider the source of variance in your data:

$$\text{Total variance in Y} = \underbrace{\text{Between-person variance}}_{\text{individual differences}} + \underbrace{\text{Within-person variance}}_{\text{response to conditions}} + \underbrace{\text{Measurement error}}_{\text{noise}}$$

In a between-participant design, your treatment effect estimate must be separated from *both* between-person variance and measurement error. Between-person variance is typically the largest component — people simply differ from each other a lot, regardless of condition.

In a within-participant design, between-person variance **cancels out** because each person serves as their own control. You are estimating the effect within each person and averaging those within-person effects. The only remaining noise is within-person variance and measurement error — which are much smaller.

```{r within-between-power, fig.height=5.5, fig.width=10}
#| code-summary: "▶ Simulate: between vs. within participant power comparison"

set.seed(2026)

# Parameters
true_effect <- 0.40   # true within-person effect (SD units)
icc_range   <- seq(0.10, 0.90, by = 0.10)  # ICC = between / total variance

power_comparison <- expand.grid(
  icc    = icc_range,
  design = c("Between-participant", "Within-participant")
) |>
  mutate(
    # For fixed total N = 100, n_per_arm = 50 between, N = 50 for within
    N        = 100,
    n_arm    = 50,
    # Between: variance includes between-person component (inflates denominator)
    sd_eff_b = sqrt(1),             # full SD
    # Within: SD of the DIFFERENCE scores = sqrt(2*(1-icc)) * total_sd
    sd_eff_w = sqrt(2 * (1 - icc)),
    power_val = mapply(function(des, sd_e) {
      tryCatch(
        power.t.test(n = 50, delta = true_effect,
                     sd = sd_e, sig.level = 0.05,
                     type = ifelse(des == "Between-participant",
                                   "two.sample", "paired"))$power,
        error = function(e) NA_real_
      )
    }, design, ifelse(design == "Between-participant", sd_eff_b, sd_eff_w))
  )

ggplot(power_comparison, aes(x = icc, y = power_val,
                              color = design, group = design)) +
  geom_line(linewidth = 1.4) +
  geom_point(size = 3.2) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "gray40") +
  annotate("text", x = 0.91, y = 0.82, label = "80% power",
           color = "gray40", size = 3.2) +
  scale_color_manual(values = c("Between-participant" = "#e63946",
                                 "Within-participant"  = "#2d6a4f"),
                     name = NULL) +
  scale_x_continuous(labels = function(x) sprintf("ICC = %.2f\n(%.0f%% between-person)", x, x*100)) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(x = "Intraclass correlation (ICC = proportion of variance between people)",
       y = "Statistical power",
       title = "Within-Participant Designs Have Greater Power — Especially When ICC Is High",
       subtitle = paste0("Fixed N = 100 participants  \u00b7  True effect = d = ", true_effect,
                         "\nWithin-participant design removes between-person variance from the error term")) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(), legend.position = "top")
```

::: {.callout-note}
## Why ICC determines how much you gain

The **intraclass correlation (ICC)** measures what fraction of total variance in Y is due to stable between-person differences. When ICC is high:

- People differ a lot from each other (but are consistent with themselves across conditions)
- Between-participant designs "waste" most of their variance budget on individual differences
- Within-participant designs remove all of this, leaving only the small within-person residual

When ICC is low:
- People are not very consistent — their responses vary randomly from occasion to occasion
- Within-participant designs offer less of an advantage because the between-person variance that cancels out was small to begin with

**Practical benchmark:** for most psychological constructs (attitudes, preferences, perceptions), ICC values between 0.40 and 0.70 are common. In this range, within-participant designs typically require **half to one-third** the sample size of between-participant designs for equivalent power.
:::

### The connection to Section 3: within-participant designs and the design space

Within-participant designs also address the Section 3 problem in a specific and powerful way. Because each person sees *multiple* conditions, the entire design space for that person's dimensions is held constant across conditions — **by construction**. Between-participant designs rely on random assignment to balance the design space across people; within-participant designs remove the need for that balance entirely within a person's data.

The tradeoff is well known: **carryover effects** (condition A changes how the person responds to condition B), **demand characteristics** (participants figure out the hypothesis), and **fatigue or practice effects**. These are most severe when:

- The treatment induces lasting attitude change (hard to reverse)
- The conditions are highly transparent (easy to guess the hypothesis)
- The outcome requires learning or practice (performance tasks)

For short perceptual or preference tasks — including most WTP and product evaluation paradigms — within-participant designs are often feasible and substantially more efficient.

---

::: {.callout-note}
## Module 2 — the complete thread

Every concept in this module is a direct extension of Module 1, applied to the treatment variable and the act of randomization rather than the outcome variable and the act of measurement:

| Module 1 (Measurement of Y) | Module 2 (Manipulation of X and Randomization) |
|---|---|
| Scale items are observables for a latent construct | Treatment assignment is an observable for latent randomization |
| Discriminant validity: does Y measure *only* what it should? | Exclusion restriction: does X activate *only* the intended pathway? |
| Construct breadth determines how many items you need | Construct breadth of Y determines how much N you need for orthogonality |
| Correlated subscales: efficient but can hide distinct pathways | Correlated latent dimensions: can ease orthogonality OR cascade failures |
| Open-ended text reveals the nomological net of Y | Open-ended text identifies unmeasured confounders and latent dimensions |
| Stratified measurement (subscales by domain) | Stratified randomization (block by measured covariates) |
| Multi-level data (participants nested in groups) | Clustered randomization (interventions at group level, ICC inflates required N) |
| Narrow constructs: few items, easy reliability | Narrow Y: small D, easy orthogonality, A/B suffices |
| Broad constructs: many items, hard to achieve discriminant validity | Broad Y: large D, orthogonality requires large N or structural design (Latin Square, metastudy) |

**The common lesson** — across both modules — is that observables are imperfect indicators of latent properties, and the gap between the observable and the latent grows with construct complexity. Researchers who map the latent space carefully (Module 1), design their manipulations and sampling deliberately (Module 2), and use the right statistical corrections when design solutions are insufficient (Module 2, Part 4) give themselves the best chance of making valid causal claims. Module 3 will formalize all of this in the language of causal inference.
:::

---

# Design Solutions: CRA, Stratified, and Clustered Assignment {#sec-design-solutions}

The Latin Square and metastudy designs are *pre-experimental* solutions that achieve orthogonality through structure. Sometimes you cannot redesign the study — the assignment mechanism is fixed. The three designs below represent progressively stronger responses to assignment constraints.

## Experimental Design Solutions: Maximizing Orthogonality Before You Collect Data

### Completely Random Assignment — The Baseline

In completely random assignment (CRA), each participant is independently assigned to a condition with equal probability. This is what most researchers do when they check "randomize" in Qualtrics.

CRA is unbiased in expectation: on average across infinitely many experiments, it achieves balance. But for any single experiment, balance is a matter of chance — and as Part 3 showed, that chance is often disappointingly low when Y is complex.

### Stratified (Blocked) Random Assignment

**The idea:** divide participants into groups (strata or blocks) based on measured covariates that you expect to relate to Y, then randomize *within* each block.

**Concrete example:** In the eco-coffee study, you know from Module 1's analysis that "environmental values" is a strong predictor of WTP (β ≈ 0.8 SD). Before running the study, add a short pre-screening item to measure environmental values. Split participants into high vs. low environmental values strata. Then assign 50% to eco-label and 50% to control *within each stratum*.

**Why it works:** By randomizing within strata, you guarantee that each condition contains equal proportions of high- and low-environmental-values participants. This eliminates one source of latent imbalance — the most important one — by design rather than by chance.

```{r stratified-random, fig.height=5, fig.width=10}
#| code-summary: "▶ Simulate: CRA vs. stratified assignment for 500 experiments"

set.seed(2026)
N_str  <- 200
D_str  <- 3    # three binary covariates to stratify on

n_sims <- 500

# Helper: absolute SMD for a covariate
abs_smd <- function(x, trt) {
  m1 <- mean(x[trt == 1]); m0 <- mean(x[trt == 0])
  s  <- sqrt((var(x[trt == 1]) + var(x[trt == 0])) / 2)
  if (s == 0) return(0)
  abs(m1 - m0) / s
}

# Simulate one run under CRA and one under stratified
one_run <- function() {
  # Participant covariates (binary)
  X1 <- rbinom(N_str, 1, 0.45)  # environmental values
  X2 <- rbinom(N_str, 1, 0.50)  # income
  X3 <- rbinom(N_str, 1, 0.40)  # health focus

  # CRA
  trt_cra  <- sample(rep(0:1, N_str / 2))

  # Stratified: randomize within each of 2^3 = 8 strata
  strata   <- paste(X1, X2, X3, sep = "-")
  trt_strat <- integer(N_str)
  for (s in unique(strata)) {
    idx <- which(strata == s)
    n_s <- length(idx)
    trt_strat[idx] <- sample(c(rep(0, ceiling(n_s / 2)),
                                rep(1, floor(n_s / 2)))[seq_len(n_s)])
  }

  data.frame(
    method = c(rep("CRA", 3L), rep("Stratified", 3L)),
    cov    = rep(c("Env. Values", "Income", "Health Focus"), 2L),
    smd    = c(abs_smd(X1, trt_cra),  abs_smd(X2, trt_cra),  abs_smd(X3, trt_cra),
               abs_smd(X1, trt_strat), abs_smd(X2, trt_strat), abs_smd(X3, trt_strat))
  )
}

set.seed(42)
strat_sims <- bind_rows(replicate(n_sims, one_run(), simplify = FALSE))

strat_summary <- strat_sims |>
  group_by(method, cov) |>
  summarise(mean_smd = mean(smd), p90_smd = quantile(smd, 0.90), .groups = "drop") |>
  mutate(method = factor(method, levels = c("CRA", "Stratified")))

ggplot(strat_summary, aes(x = cov, y = mean_smd, fill = method,
                           ymin = 0, ymax = p90_smd)) +
  geom_col(position = position_dodge(0.7), width = 0.6, alpha = 0.85) +
  geom_errorbar(aes(ymin = mean_smd, ymax = p90_smd),
                position = position_dodge(0.7), width = 0.25, linewidth = 0.8) +
  geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray40", linewidth = 0.8) +
  annotate("text", x = 3.45, y = 0.105, label = "|SMD| = 0.10\nthreshold",
           size = 3.0, color = "gray40", hjust = 0, lineheight = 0.85) +
  scale_fill_manual(values = c("CRA" = "#e63946", "Stratified" = "#52b788"), name = NULL) +
  labs(x = "Covariate", y = "Mean |SMD| (bar) + 90th percentile (error bar)",
       title = "Stratified vs. Completely Random Assignment: Balance on Stratified Covariates",
       subtitle = paste0("N = ", N_str, "  \u00b7  500 simulated experiments  \u00b7  ",
                         "Bars = mean |SMD|  \u00b7  Error bars = 90th percentile worst case")) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(), legend.position = "top")
```

Stratification dramatically reduces both the mean and worst-case imbalance on the stratified covariates. The 90th-percentile bar for CRA often exceeds the 0.10 threshold; the stratified version almost never does.

**The limitation** is the same as in Module 1's measurement section: you can only stratify on what you have *measured*. Unstratified dimensions remain at the mercy of chance — which returns us to the need to map the construct Y first.

### Clustered Random Assignment

**The idea:** sometimes participants come in natural groups (classrooms, teams, branches, households), and the intervention must be assigned at the group level. In this case, you assign entire clusters to conditions.

**Concrete example:** You want to test whether providing eco-label information changes café purchasing behavior. You cannot assign individual customers within a café to different menu conditions simultaneously — contamination would occur. Instead, you assign *cafés* to conditions.

**The tradeoff:** clustering increases statistical efficiency when you have to intervene at the group level, but it also introduces **intraclass correlation (ICC)**. If people within the same café are more similar to each other (they self-select into similar cafés), then your N cafés provide less independent information than N individuals would.

```{r clustered-random, fig.height=4.5, fig.width=9}
#| code-summary: "▶ Illustrate: effective N under clustered assignment for different ICCs"

icc_vals     <- c(0.00, 0.05, 0.10, 0.20, 0.30, 0.50)
cluster_sizes <- c(5, 10, 20, 40)

deff_df <- expand.grid(icc = icc_vals, m = cluster_sizes) |>
  mutate(
    DEFF    = 1 + (m - 1) * icc,        # design effect
    Neff_80 = ceiling(
      power.t.test(delta = 0.40, sd = 1, sig.level = 0.05, power = 0.80,
                   type = "two.sample")$n * DEFF
    ),
    m_label = paste0("Cluster size = ", m)
  )

ggplot(deff_df, aes(x = icc, y = Neff_80,
                    color = factor(m), group = factor(m))) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.8) +
  geom_hline(yintercept = power.t.test(delta = 0.40, sd = 1, sig.level = 0.05,
                                        power = 0.80, type = "two.sample")$n,
             linetype = "dashed", color = "gray40", linewidth = 0.7) +
  annotate("text", x = 0.51, y = power.t.test(delta = 0.40, sd = 1, sig.level = 0.05,
                                                power = 0.80, type = "two.sample")$n + 3,
           label = "Required N\nif individual\nassignment",
           size = 2.8, color = "gray40", hjust = 0, lineheight = 0.85) +
  scale_color_manual(values = c("5" = "#4a90d9", "10" = "#52b788",
                                 "20" = "#f4a261", "40" = "#e63946"),
                     name = "Cluster size (m)") +
  scale_x_continuous(labels = function(x) sprintf("ICC = %.2f", x)) +
  scale_y_continuous(labels = comma) +
  labs(x = "Intraclass correlation (ICC)",
       y = "Required total N (individuals)",
       title = "Clustered Assignment: How ICC Inflates Required Sample Size",
       subtitle = "DEFF = 1 + (m\u22121)\u00d7ICC  \u00b7  Design effect multiplies into required N  \u00b7  Target: d = 0.40, 80% power") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

::: {.callout-note}
## Stratified vs. Clustered — the key contrast

| Feature | Stratified | Clustered |
|---|---|---|
| Unit of randomization | Individual (within strata) | Group / cluster |
| Effect on balance | **Improves** balance on stratified vars | No improvement (can worsen) |
| Effect on required N | Reduces N slightly | **Increases** N via design effect |
| When to use | When measured covariates predict Y strongly | When intervention must be applied at group level |
| Analogy (Module 1) | Pre-specifying subscales before CFA | Multi-level data with non-independent observations |

The core message: stratification and clustering both depart from simple CRA, but in opposite directions for power. Use stratification proactively when you can; account for clustering carefully when you must.
:::

---
