---
title: "Part 2: The Exclusion Restriction"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  fig.width  = 9,
  fig.height = 5,
  fig.path   = "figs/",
  dev        = "svg"
)
dir.create("figs", recursive = TRUE, showWarnings = FALSE)
```

```{r packages}
#| code-summary: "▶ Load required packages"
# install.packages(c("tidyverse","ggplot2","knitr","scales","patchwork"))

library(tidyverse)
library(ggplot2)
library(knitr)
library(scales)
library(patchwork)
if (!requireNamespace("lavaan", quietly = TRUE)) install.packages("lavaan")
library(lavaan)

set.seed(2025)
```

<iframe src="https://liveeur-my.sharepoint.com/personal/58542dsc_eur_nl/_layouts/15/Doc.aspx?sourcedoc={e76285ab-242d-4ae8-9861-fa2d25c86f2c}&action=embedview&wdAr=1.7777777777777777" width="100%" height="500px" frameborder="0" allowfullscreen title="Module 2 Part 2: Exclusion Restriction"></iframe>

# Part 2: The Exclusion Restriction — What Else Is in Your Treatment? {#sec-exclusion}

## One Button, Many Signals

In Part 1 we treated the eco-label manipulation as a clean, unidimensional intervention: it either increases WTP (H₁) or it does not (H₀). This framing is convenient, but it hides a problem that is exactly analogous to what Module 1 revealed about *measuring* Y.

Recall the core lesson from Module 1:

> A scale item labeled "Green Purchase Intention" often captures several distinct latent constructs simultaneously — environmental values, social signaling, perceived quality, price expectations, and more. The scale *appears* to measure one thing while actually measuring many. This is a violation of **discriminant validity**.

Now ask the same question about your treatment variable:

> When a participant sees an eco-certification badge on a coffee package, what does that label actually *do*?

It does not inject a single, clean signal of "eco-friendliness." It activates — simultaneously and heterogeneously — a cluster of inferences, associations, and motivations that vary across people. This is a violation of **the exclusion restriction**: the assumption that the treatment variable affects the outcome *only* through the intended pathway.

::: {.callout-important}
## The Exclusion Restriction — the single-mechanism assumption

The exclusion restriction is the assumption that X affects Y through **exactly one latent pathway** — the one the researcher intends. Every field has a name for what happens when this assumption breaks:

| Field | Term for the violation |
|---|---|
| Psychology / Marketing | **Confound** — the manipulation activates an unintended construct alongside the intended one |
| Econometrics | **Endogeneity** — the regressor correlates with the error term via unmodelled channels |
| Epidemiology | **Confounding** — exposure and outcome share a common cause |
| Statistics | **Omitted variable bias** — a variable that belongs in the model is left out |
| Causal inference (DAGs) | **Backdoor path** — a non-causal association flows through an unmeasured variable |

All of these are the same problem viewed through different disciplinary lenses: **there is more than one latent link between X and Y**.

If X injects additional unintended signals that also causally influence Y — and if those signals vary across participants — then the estimated "treatment effect" is a weighted mixture of several distinct causal effects, not a single clean estimate of any one of them.

**This is the fundamental identification problem in causal inference**, and it is the experimental analogue of discriminant invalidity in measurement. In Module 1 you saw that **measured Y can measure too much**. Here you will see that **manipulated X can inject too much**.
:::

## Simulating a Heterogeneous Treatment

We now simulate the coffee experiment with a realistic data-generating process. Each participant has nine binary latent characteristics that determine both *what signals they extract from the eco-label* and *how much they are willing to pay*.

```{r part2-setup}
#| code-summary: "▶ Simulate 120 participants with heterogeneous traits and WTP"

set.seed(2026)
N_p2  <- 120
half2 <- N_p2 / 2

# ── Participant-level binary traits ───────────────────────────────────────────
ptrait <- function(p) rbinom(N_p2, 1, p)

traits <- data.frame(
  env_values      = ptrait(0.45),  # cares about environment
  quality_seek    = ptrait(0.50),  # focuses on quality / taste
  price_sensitive = ptrait(0.50),  # price-conscious
  social_identity = ptrait(0.40),  # identity / values-driven purchasing
  health_focus    = ptrait(0.45),  # health / natural-ingredient preference
  brand_trust     = ptrait(0.55),  # responds to certification cues
  guilt_prone     = ptrait(0.35),  # guilt-reducing / compensatory consumption
  hedonic         = ptrait(0.50),  # pleasure-driven; taste is paramount
  habitual        = ptrait(0.40),  # habitual buyer; resists stretching budget
  demand_aware    = ptrait(0.40)   # susceptible to experimenter demand / social desirability cues
)

# ── Equal random assignment ───────────────────────────────────────────────────
condition_p2 <- rep(c("Control", "Eco-label"), each = half2)

# ── WTP simulation (in dollars, clamped to $1–$10) ────────────────────────────
# Key feature: eco-label ACTIVATES traits 1, 2, 4, 5, 6, 7 differentially.
# Traits 3 (price sensitivity), 8 (hedonic), 9 (habitual) operate in both conditions.
WTP_p2 <- with(traits, {
  eco <- as.integer(condition_p2 == "Eco-label")
  v   <- 4.50 +
    1.20 * env_values      * eco        +  # env concern → eco premium
    0.80 * quality_seek    * (0.35 + 0.65 * eco)  +  # quality matters more in eco
   -0.70 * price_sensitive            +  # price sensitivity suppresses WTP
    0.90 * social_identity * eco        +  # identity-driven eco premium
    0.60 * health_focus    * eco        +  # health focus eco premium
    0.50 * brand_trust     * eco        +  # certification trust premium
    0.35 * guilt_prone     * eco        +  # guilt-offset premium
    0.30 * hedonic                      +  # hedonic adds WTP in both
   -0.45 * habitual                    +  # habitual buyers resist stretching
    0.55 * demand_aware    * eco        +  # demand-aware participants inflate WTP in eco context
    rnorm(N_p2, 0, 0.75)
  pmax(pmin(round(v, 2), 10), 1)
})

df_p2 <- data.frame(traits, condition = condition_p2, WTP = WTP_p2)

# ── Summary by condition ──────────────────────────────────────────────────────
df_p2 |>
  group_by(condition) |>
  summarise(n = n(), mean_WTP = round(mean(WTP), 2),
            sd_WTP = round(sd(WTP), 2), .groups = "drop") |>
  kable(col.names = c("Condition", "n", "Mean WTP", "SD WTP"),
        align   = c("l", "c", "c", "c"),
        caption = "WTP summary by condition")
```

The overall treatment effect looks straightforward. But the figure below reveals the heterogeneity hiding underneath it:

```{r het-effects-plot, fig.height=5}
#| code-summary: "▶ Plot: treatment effect varies by participant trait profile"

# Compute individual-level treatment contrasts via matched simulation
# For illustration: show eco-label premium by number of "activated" traits
df_p2 <- df_p2 |>
  mutate(
    n_eco_traits = env_values + quality_seek + social_identity +
                   health_focus + brand_trust + guilt_prone,
    eco_group = cut(n_eco_traits, breaks = c(-1, 1, 3, 6),
                    labels = c("0\u20131 eco-responsive traits",
                               "2\u20133 eco-responsive traits",
                               "4\u20136 eco-responsive traits"))
  )

ggplot(df_p2, aes(x = condition, y = WTP, fill = condition, color = condition)) +
  geom_jitter(width = 0.14, alpha = 0.4, size = 2) +
  geom_boxplot(alpha = 0.20, outlier.shape = NA, width = 0.45, linewidth = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 4,
               fill = "white", color = "black", stroke = 1.3) +
  facet_wrap(~eco_group, nrow = 1) +
  scale_fill_manual(values  = c("Control" = "#4a90d9", "Eco-label" = "#2d6a4f")) +
  scale_color_manual(values = c("Control" = "#2b6cb0", "Eco-label" = "#1b4332")) +
  scale_y_continuous(labels = function(x) paste0("$", x)) +
  labs(x = NULL, y = "WTP ($)",
       title = "The Same Treatment Has Very Different Effects on Different People",
       subtitle = "Diamond = group mean  \u00b7  Eco-responsive traits: environmental values, quality-seeking, social identity, health, trust, guilt") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none", panel.grid.minor = element_blank(),
        strip.text = element_text(size = 9, face = "bold"))
```

The eco-label produces a large positive effect for participants with several eco-responsive traits, a small effect for moderately responsive participants, and near-zero (or even negative) effects for participants whose WTP is driven primarily by price sensitivity, habit, or hedonic taste expectations.

The average treatment effect you would report obscures all of this variation — and misses entirely *which* mechanism is driving WTP upward.

## What Participants Say: Template-Based Open-Ended Responses

To make the multi-signal nature of the treatment visible, we generate short open-ended responses from each simulated participant. These represent **(a)** what they infer about the product and **(b)** why they chose their WTP. We use template-based generation — each participant's active traits and condition determine which phrases they contribute.

> **Note:** We are using template-based text here to keep everything reproducible without an LLM API. In a real study you would collect actual open-ended responses. The patterns illustrated below — differential concept activation by condition, and concept–WTP correlations — would be recovered from those real responses using the same analysis pipeline.

```{r generate-text}
#| code-summary: "▶ Generate template-based open-ended text for each participant"

set.seed(2026)

# ── Product perception text ───────────────────────────────────────────────────
gen_product_text <- function(i) {
  cond <- condition_p2[i]
  ev   <- traits$env_values[i]
  qs   <- traits$quality_seek[i]
  ps   <- traits$price_sensitive[i]
  si   <- traits$social_identity[i]
  hf   <- traits$health_focus[i]
  bt   <- traits$brand_trust[i]
  gp   <- traits$guilt_prone[i]
  hed  <- traits$hedonic[i]
  da   <- traits$demand_aware[i]

  parts <- character(0)

  if (cond == "Control") {
    parts <- c(parts, sample(c(
      "This looks like a standard coffee blend.",
      "A fairly typical coffee package — nothing unusual.",
      "The packaging is clean and straightforward."
    ), 1))
    if (qs  == 1) parts <- c(parts, "The blend seems like decent, everyday quality.")
    if (ps  == 1) parts <- c(parts, "The price point appears reasonable and fair.")
    if (bt  == 1) parts <- c(parts, "The packaging gives a professional, trustworthy impression.")
    if (hed == 1) parts <- c(parts, "I am curious how the flavor compares to my usual brand.")
    if (length(parts) == 1)
      parts <- c(parts, "Nothing in particular stands out about this product.")
  } else {
    parts <- c(parts, "The eco-certification badge is immediately noticeable.")
    if (ev == 1) parts <- c(parts, "The sustainability label signals that this coffee is responsibly and environmentally sourced.")
    if (qs == 1) parts <- c(parts, "Eco-certified products suggest a higher standard of care in production, which I associate with better quality and taste.")
    if (ps == 1) parts <- c(parts, "Certifications usually carry a price premium, so I expect this to cost more than a standard option.")
    if (si == 1) parts <- c(parts, "This brand's values align with mine — I actively look for and support products like this.")
    if (hf == 1) parts <- c(parts, "Eco-certified means fewer pesticides and more natural farming practices, which I find very appealing.")
    if (bt == 1) parts <- c(parts, "A third-party certification badge means independent accountability — that increases my trust in the product significantly.")
    if (gp == 1) parts <- c(parts, "Buying a certified sustainable product lets me offset some of my everyday environmental footprint.")
    if (da == 1) parts <- c(parts, "The research context makes it clear this study is about eco-labeling, which makes me more conscious of how much I should value green products.")
    if (length(parts) == 1)
      parts <- c(parts, "The label catches my eye, though I am not sure how much weight I give eco-certifications in practice.")
  }

  paste(sample(parts, min(2L, length(parts))), collapse = " ")
}

# ── WTP justification text ────────────────────────────────────────────────────
gen_wtp_text <- function(i) {
  cond <- condition_p2[i]
  ev   <- traits$env_values[i]
  qs   <- traits$quality_seek[i]
  ps   <- traits$price_sensitive[i]
  si   <- traits$social_identity[i]
  hf   <- traits$health_focus[i]
  hed  <- traits$hedonic[i]
  hab  <- traits$habitual[i]
  da   <- traits$demand_aware[i]

  parts <- character(0)

  if (hab == 1 & ps == 1)
    parts <- c(parts, "I am fairly budget-conscious about coffee and set my limit based on what I can realistically afford each week.")
  if (qs == 1)
    parts <- c(parts, "I am willing to pay more when I believe the quality genuinely delivers on taste and enjoyment.")
  if (ev == 1 & cond == "Eco-label")
    parts <- c(parts, "I factored in the environmental benefits — supporting sustainable production is worth a modest premium to me.")
  if (si == 1 & cond == "Eco-label")
    parts <- c(parts, "I would pay more for a product that represents my values, even if cheaper alternatives are available.")
  if (hf == 1 & cond == "Eco-label")
    parts <- c(parts, "Cleaner and more natural sourcing justifies a higher price — I think of it as an investment in my health.")
  if (ps == 1)
    parts <- c(parts, "Price matters a great deal to me — I try not to overpay regardless of how appealing the product looks.")
  if (hab == 1)
    parts <- c(parts, "My spending on coffee tends to stay fairly consistent — habit guides my budget more than any single product feature.")
  if (hed == 1)
    parts <- c(parts, "If I expect genuinely exceptional taste, I am happy to stretch my usual budget.")
  if (da == 1 & cond == "Eco-label")
    parts <- c(parts, "Being in a study about eco products made me want to demonstrate my environmental awareness, which may have inflated my stated price slightly.")
  if (length(parts) == 0)
    parts <- "I set my price based on the overall value I expect — quality relative to cost."

  paste(sample(parts, min(2L, length(parts))), collapse = " ")
}

product_text <- sapply(seq_len(N_p2), gen_product_text)
wtp_text     <- sapply(seq_len(N_p2), gen_wtp_text)

# Preview a few responses
preview <- data.frame(
  Condition        = condition_p2[c(1, half2 + 1)],
  `Product text`   = product_text[c(1, half2 + 1)],
  `WTP text`       = wtp_text[c(1, half2 + 1)],
  check.names = FALSE
)
kable(preview, caption = "Example responses (one control, one eco-label participant)")
```

## Text Analysis: How Concept Composition Shifts Across Conditions

We now run a simple concept-count analysis — no API, no black-box model. Each concept is defined by a set of keywords; a participant "mentions" a concept if any keyword appears in their text.

```{r text-analysis}
#| code-summary: "▶ Count concept mentions and compare across conditions"

# ── Concept dictionary (regex patterns) ──────────────────────────────────────
concept_dict <- c(
  sustainability  = "sustainab|eco.certif|environmental|responsib",
  quality_taste   = "quality|taste|flavor|blend",
  price_cost      = "price|premium|afford|budget|overpay",
  social_values   = "values|align|represent|support products",
  health_natural  = "health|natural|pesticide|farming|cleaner",
  trust_cert      = "certif|accountab|trustworth",
  guilt_offset    = "footprint|guilt|offset",
  demand_effect   = "research context|study.*eco|should value|demonstrate.*environ|inflated.*stated|awareness.*study",
  hedonic         = "exceptional|happy to stretch|enjoyment",
  habit_routine   = "habit|consistent|weekly"
)

# Concept presence (1/0) per participant
prod_conc <- as.data.frame(
  lapply(concept_dict, function(pat) as.integer(grepl(pat, tolower(product_text)))))

wtp_conc <- as.data.frame(
  lapply(concept_dict, function(pat) as.integer(grepl(pat, tolower(wtp_text)))))

prod_conc$condition <- condition_p2
wtp_conc$condition  <- condition_p2
wtp_conc$WTP        <- WTP_p2

# ── Log-odds: how much more often is each concept in eco-label vs control? ───
log_odds <- function(x, cond, target = "Eco-label") {
  p_t <- (sum(x[cond == target])       + 0.5) / (sum(cond == target)       + 1)
  p_c <- (sum(x[cond != target])       + 0.5) / (sum(cond != target)       + 1)
  log(p_t / (1 - p_t)) - log(p_c / (1 - p_c))
}

lo_prod <- sapply(names(concept_dict), function(v)
  log_odds(prod_conc[[v]], prod_conc$condition))

lo_wtp  <- sapply(names(concept_dict), function(v)
  log_odds(wtp_conc[[v]],  wtp_conc$condition))

# ── Correlation of concept mentions with WTP ──────────────────────────────────
wtp_cor_prod <- sapply(names(concept_dict), function(v)
  cor(prod_conc[[v]], WTP_p2, method = "spearman"))

wtp_cor_wtp <- sapply(names(concept_dict), function(v) {
  x <- wtp_conc[[v]]
  # If concept never appears in WTP text, correlation is undefined → treat as 0
  if (sd(x) < 1e-9) 0 else cor(x, WTP_p2, method = "spearman")
})

# ── Summary table ─────────────────────────────────────────────────────────────
concept_summary <- data.frame(
  Concept      = names(concept_dict),
  pct_ctrl     = round(colMeans(prod_conc[prod_conc$condition == "Control",
                                          names(concept_dict)]) * 100, 0),
  pct_eco      = round(colMeans(prod_conc[prod_conc$condition == "Eco-label",
                                          names(concept_dict)]) * 100, 0),
  log_odds_col = round(lo_prod, 2),
  r_prod       = round(wtp_cor_prod, 2),
  r_wtp        = round(wtp_cor_wtp, 2)
)

kable(concept_summary,
      col.names = c("Concept", "Control (%)", "Eco-label (%)",
                    "Log-odds (product text)", "r with WTP (product)", "r with WTP (WTP text)"),
      caption = "Concept prevalence and associations by condition (product perception text)")
```

```{r concept-plot, fig.height=5.5, fig.width=10}
#| code-summary: "▶ Plot: log-odds ratios and WTP correlations"

lo_df <- data.frame(
  concept  = rep(names(concept_dict), 2),
  log_odds = c(lo_prod, lo_wtp),
  source   = rep(c("Product perception text", "WTP justification text"), each = length(concept_dict))
) |>
  mutate(concept = factor(concept, levels = names(concept_dict)[order(lo_prod)]))

ggplot(lo_df, aes(x = log_odds, y = concept, fill = source)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6, alpha = 0.85) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.8) +
  scale_fill_manual(values = c("Product perception text" = "#4a90d9",
                                "WTP justification text"  = "#2d6a4f"),
                    name = NULL) +
  labs(x = "Log-odds ratio (Eco-label vs. Control)",
       y = NULL,
       title = "How the Eco-Label Shifts the Concepts People Reach For",
       subtitle = "Positive = concept mentioned more in eco-label condition") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank())
```

The eco-label dramatically increases mentions of sustainability, quality, social values, health, and trust in the *product perception* text. The effect bleeds through — though more weakly — into the *WTP justification* text. Note that `habit_routine` and `hedonic` show no shift at all: they are driven by participant characteristics, not by the experimental manipulation.

## What's in X? What's in Y? {#sec-whats-in-xy}

::: {.callout-important}
## This is the most important figure in Section 2 — read it carefully

The diagram below is *not* just a summary of our coffee simulation. It represents the fundamental challenge of experimental interpretation.

**The left column (X)** shows every latent construct injected into the system by the eco-label manipulation. Only **one** of these — environmental responsibility — is the researcher's *account of interest* (the intended mechanism). Every other entry in the X column is a **confounder**: a pathway that the treatment activates alongside the intended one.

**The right column (Y)** shows every latent construct that causally drives WTP. Some overlap with X (shared constructs = blue lines). Some do not (orange boxes).

**The blue lines are the problem.** Each blue line is a latent pathway from X to Y that exists *in addition to* the one the researcher intends. When any of these additional pathways carry variance, the treatment coefficient in a regression is not "the effect of eco-labeling on environmental value". It is a conflation of all connected pathways, weighted by how strongly each latent construct in Y is activated by X.

**The researcher's responsibility** is to:

1. *Before* running the study: use the measurement tools from Module 1 (pilot text analysis, CFA, construct mapping) to enumerate the latent space of both X and Y. Every construct in that space that is common to both columns is a pathway you must account for — either by measuring it (so you can check balance) or by designing the manipulation to deactivate it.

2. *After* data collection: check balance on every measured common construct. Acknowledge that unmeasured common constructs remain a threat.

The question "What's in the X?" is the discriminant validity question asked about a *manipulation* rather than a *scale*. The tools are the same; the object of inquiry has shifted from the outcome variable to the treatment variable.
:::

The analysis above lets us ask the same question we posed in Module 1 about outcome measurement, but now about the treatment:

> **"What's in the X?" and "What's in the Y?" — and where do they overlap?**

The diagram below maps the constructs injected by the eco-label manipulation (left column) against the constructs that drive WTP (right column). **Blue lines are the confounds** — shared constructs that create additional latent pathways from X to Y beyond the intended one. Orange boxes mark constructs that are unique to one side.

```{r whats-in-xy, fig.height=8.5, fig.width=10}
#| code-summary: "▶ Plot: construct overlap diagram (What's in X? What's in Y?)"

x_topics <- c(
  "Environmental\nresponsibility\n[account of interest]",
  "Premium quality\nperception",
  "Price expectation\n(eco = premium)",
  "Social identity\n& values alignment",
  "Health &\nnaturalness",
  "Certification\ntrust",
  "Demand effects\n(context signals\n'eco is expected')",
  "Guilt\nreduction"
)

y_topics <- c(
  "Personal\nenvironmental values",
  "Perceived product\nquality",
  "Price sensitivity\n& budget",
  "Social identity",
  "Health concern",
  "Brand trust",
  "Social desirability\n& demand compliance",
  "Hedonic taste\nexpectations",
  "Habitual spending\npatterns"
)

# First 7 x-topics are shared with first 7 y-topics; last in each column are unique
nx <- length(x_topics)
ny <- length(y_topics)

plot_x <- data.frame(
  topic  = x_topics,
  xpos   = 0.15,
  ypos   = seq(nx, 1, by = -1L),
  shared = c(rep(TRUE, 7L), FALSE),
  stringsAsFactors = FALSE
)

plot_y <- data.frame(
  topic  = y_topics,
  xpos   = 0.85,
  ypos   = seq(nx, nx / ny, length.out = ny),
  shared = c(rep(TRUE, 7L), FALSE, FALSE),
  stringsAsFactors = FALSE
)

conn_df <- data.frame(
  x1 = 0.265, x2 = 0.735,
  y1 = plot_x$ypos[plot_x$shared],
  y2 = plot_y$ypos[plot_y$shared],
  is_demand = c(rep(FALSE, 6L), TRUE)   # flag the demand-effect link
)

ggplot() +
  # Non-demand links — standard blue
  geom_segment(data = conn_df[!conn_df$is_demand, ],
               aes(x = x1, xend = x2, y = y1, yend = y2),
               color = "#4a90d9", linewidth = 1.4, alpha = 0.55) +
  # Demand-effect link — highlighted red-orange
  geom_segment(data = conn_df[conn_df$is_demand, ],
               aes(x = x1, xend = x2, y = y1, yend = y2),
               color = "#d7301f", linewidth = 2.0, alpha = 0.80,
               arrow = arrow(ends = "both", length = unit(0.08, "inches"),
                             type = "open")) +
  geom_label(data = plot_x,
             aes(x = xpos, y = ypos, label = topic, fill = shared),
             size = 2.7, fontface = "bold", lineheight = 0.88,
             label.padding = unit(0.38, "lines"),
             label.r       = unit(0.15, "lines"), hjust = 0.5) +
  geom_label(data = plot_y,
             aes(x = xpos, y = ypos, label = topic, fill = shared),
             size = 2.7, fontface = "bold", lineheight = 0.88,
             label.padding = unit(0.38, "lines"),
             label.r       = unit(0.15, "lines"), hjust = 0.5) +
  # Column header boxes
  annotate("rect", xmin = 0.02, xmax = 0.28, ymin = 8.55, ymax = 9.20,
           fill = "#1e3a5f", color = NA) +
  annotate("text", x = 0.15, y = 8.88,
           label = "What's in X?\n(Eco-label Manipulation)",
           color = "white", fontface = "bold", size = 3.5, lineheight = 0.9) +
  annotate("rect", xmin = 0.72, xmax = 0.98, ymin = 8.55, ymax = 9.20,
           fill = "#2d6a4f", color = NA) +
  annotate("text", x = 0.85, y = 8.88,
           label = "What's in Y?\n(Willingness to Pay)",
           color = "white", fontface = "bold", size = 3.5, lineheight = 0.9) +
  # Demand-effect annotation
  annotate("label", x = 0.50, y = plot_x$ypos[7] + 0.30,
           label = "Demand effect confounder",
           fill = "#fff0ed", color = "#d7301f", fontface = "bold",
           size = 2.6, label.padding = unit(0.25, "lines")) +
  scale_fill_manual(
    values = c("TRUE" = "#d6ecff", "FALSE" = "#ffe0b2"),
    labels = c("Shared construct (confound)", "Unique to this variable"),
    name   = NULL
  ) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0.4, 9.5)) +
  labs(
    title    = "Construct Overlap Between Treatment (X) and Outcome (Y)",
    subtitle = paste0(
      "Blue lines = shared constructs (confounds)  \u00b7  Red line = demand effect  \u00b7  Orange boxes = unique to one side\n",
      "The account of interest is ONE pathway. Every other shared link is a confounder the researcher must account for."
    )
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5, size = 14,
                                 margin = margin(b = 6)),
    plot.subtitle = element_text(hjust = 0.5, color = "#555555", size = 10,
                                 lineheight = 1.3),
    legend.position = "bottom",
    plot.margin = margin(15, 15, 15, 15)
  )
```

::: {.callout-important}
## The parallel with Module 1: discriminant validity for manipulations

In Module 1, we said that a *scale* violates discriminant validity when it correlates too strongly with constructs it is not supposed to measure. The scale "measures too much."

Here, the *eco-label manipulation* violates an analogous principle: it injects multiple independent constructs simultaneously — not just "environmental responsibility" but also quality expectations, price priors, social identity signals, health inferences, certification trust, and guilt-offset motivation. The manipulation "does too much."

**Constructs unique to X (not in Y):** Guilt reduction. This signal is activated by the eco-label but does not translate into WTP variance in this simulation.

**Constructs unique to Y (not activated by X):** Hedonic taste expectations and habitual spending patterns. These explain variance in WTP in both conditions but are not triggered by the eco-label.

**The demand effect:** This is a particularly insidious shared pathway. The eco-label condition signals to demand-aware participants that the study is *about* eco-labeling — and they respond by inflating their stated WTP to comply with what they perceive the researcher expects. This is a confound that lives entirely in the experimental context itself: it is not driven by any genuine product attribute, but by the act of conducting the experiment. Notice that this pathway appears in *both* the treatment (X injects the cue) *and* the outcome (Y picks up the compliance). In the real world, researchers cannot remove this pathway by controlling the manipulation — they must measure it explicitly and either partial it out or interpret the treatment effect as "eco-label + demand context" rather than "eco-label alone."

**The practical implication:** Your treatment coefficient in a regression is a weighted average across all the shared constructs, confounded by whatever imbalance exists in the unshared ones. This is not a clean estimate of "the effect of eco-labeling" — it is the effect of a bundle of simultaneous signals, filtered through a population of heterogeneous responders.
:::

### How to Map the Construct Space in Practice: Nomological Net Scanning

The diagram above gives the theoretical picture — the full overlap between X and Y. In practice, you rarely have complete knowledge of this overlap at the start of a study. **Open-ended text responses** from a small pilot give you an empirical window into both sides of the diagram.

The approach: ask a small pilot sample (n ≈ 20–30) to write briefly about their response to the stimulus (for X) and about their reasoning for the outcome variable (for Y). Then use text analysis — either keyword-based or unsupervised — to extract which constructs are active.

**This is the nomological net scan.** It serves three purposes simultaneously:

1. **Enumeration:** it tells you *how many* latent pathways from X to Y exist — the practical answer to "What's in X? What's in Y?"
2. **Measurement guidance:** every concept the text reveals is a mediator you should measure quantitatively in the main study (using the scale construction tools from Module 1)
3. **Identification strategy:** once you have measured mediators, you can decompose the total treatment effect into its component channels via structural models (which we address in Part 3 and in detail in Module 3)

::: {.callout-important}
## The identification problem and its resolution

In Module 1 we showed that a scale conflates multiple latent constructs (discriminant invalidity) — and the fix was to *measure* those constructs separately and check that they can be discriminated.

Here, the treatment X activates multiple causal pathways to Y. The fix is identical: *measure* those pathways as mediators, then use a structural model to identify which channels carry the effect. Without measurement, you are left with an identification problem — no statistical model can tell you which pathway generated the observed X→Y association.

**The nomological net scan gives you the list. Quantitative measurement gives you the data. Structural modeling gives you the decomposition.**
:::

```{r nomological-net-scan}
#| code-summary: "▶ Use WTP text to scan for unmeasured Y dimensions (pilot)"

# Recall: we already have wtp_text and wtp_conc from earlier in Part 2.
# Simulate a "pre-study pilot" where only 30 participants write about their WTP.
# The goal: identify which concepts are active and should become measured mediators.

set.seed(999)
pilot_idx  <- sample(seq_len(N_p2), 30)
pilot_text <- wtp_text[pilot_idx]

# Concept frequency in the pilot
pilot_freq <- sapply(concept_dict, function(pat)
  mean(grepl(pat, tolower(pilot_text))))

pilot_freq_df <- data.frame(
  Concept    = names(pilot_freq),
  Prevalence = round(pilot_freq * 100, 0)
) |>
  arrange(desc(Prevalence)) |>
  mutate(
    include = ifelse(Prevalence >= 25, "\u2713 Yes (prevalent)",
                     "\u2715 No (rare in this sample)"),
    Concept = factor(Concept, levels = Concept)
  )

kable(pilot_freq_df,
      col.names = c("Y-dimension concept", "% pilot participants mentioning it",
                    "Recommendation"),
      caption   = "Pilot text scan: which sub-dimensions of WTP are active in this sample?")
```

```{r nom-net-plot, fig.height=4.5, fig.width=9}
#| code-summary: "▶ Plot: active Y-dimensions from pilot scan"

ggplot(pilot_freq_df, aes(x = Prevalence, y = Concept, fill = include)) +
  geom_col(alpha = 0.85, color = "white") +
  geom_vline(xintercept = 25, linetype = "dashed", color = "gray40", linewidth = 0.8) +
  annotate("text", x = 26, y = 1.0, label = "25% threshold",
           color = "gray40", hjust = 0, size = 3.2) +
  scale_fill_manual(values = c("\u2713 Yes (prevalent)" = "#2d6a4f",
                               "\u2715 No (rare in this sample)" = "#b0c4de"),
                    name = NULL) +
  labs(x = "% of pilot participants mentioning this concept",
       y = NULL,
       title = "Scanning the Nomological Net of WTP via Open-Ended Text",
       subtitle = "Prevalent concepts become measured mediators in the main study") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(), legend.position = "top")
```

### Alternative: LDA Topic Modeling (Unsupervised Discovery)

The keyword approach above requires you to specify patterns in advance. **Latent Dirichlet Allocation (LDA)** flips this: it automatically discovers clusters of co-occurring words across responses, with no keywords required. Each cluster is a *topic* — a statistical summary of what participants tended to write about together.

**The intuition:** Imagine sorting all WTP responses by word similarity. Responses mentioning "environment", "sustainable", "planet" cluster together naturally. Responses mentioning "organic", "healthy", "fresh" form a different cluster. LDA finds these clusters by asking: *what grouping of words best explains which responses sound similar?* The number of topics `k` is chosen by you — start with k ≈ number of expected sub-dimensions.

```{r lda-topic-model, fig.height=5, fig.width=10}
#| code-summary: "▶ LDA topic model: unsupervised discovery of Y-dimensions from text"

if (!requireNamespace("topicmodels", quietly = TRUE)) install.packages("topicmodels")
if (!requireNamespace("tidytext",    quietly = TRUE)) install.packages("tidytext")
library(topicmodels)
library(tidytext)

# ── Tokenize pilot text → document-term matrix ────────────────────────────────
pilot_tidy <- data.frame(
  doc_id = seq_along(pilot_text),
  text   = pilot_text
) |>
  unnest_tokens(word, text) |>
  anti_join(stop_words, by = "word") |>        # remove "the", "and", etc.
  filter(nchar(word) > 2) |>                   # drop 1-2 character tokens
  count(doc_id, word, sort = TRUE)

pilot_dtm <- pilot_tidy |>
  cast_dtm(doc_id, word, n)                    # convert to document-term matrix

# ── Fit LDA with k = 5 topics ─────────────────────────────────────────────────
# k = 5 is a reasonable start; in a real pilot you might try k = 3 to 8 and
# inspect coherence scores (e.g. via ldatuning package) to pick the best k.
set.seed(42)
lda_fit <- LDA(pilot_dtm, k = 5, control = list(seed = 42))

# ── Extract top words per topic ────────────────────────────────────────────────
top_words <- tidy(lda_fit, matrix = "beta") |>      # beta = P(word | topic)
  group_by(topic) |>
  slice_max(beta, n = 8, with_ties = FALSE) |>
  ungroup() |>
  mutate(topic_label = paste0("Topic ", topic))

# ── Assign human-readable labels (inspect top words to name each topic) ───────
# In a real analysis YOU read the top words and assign a substantive label.
# Here we auto-label by the single highest-probability word for illustration.
topic_labels <- top_words |>
  group_by(topic) |>
  slice_max(beta, n = 1) |>
  mutate(label = paste0("Topic ", topic, "\n(e.g. '", term, "')")) |>
  dplyr::select(topic, label)

top_words <- top_words |>
  left_join(topic_labels, by = "topic")

ggplot(top_words, aes(x = reorder_within(term, beta, label),
                       y = beta, fill = factor(topic))) +
  geom_col(show.legend = FALSE, alpha = 0.85) +
  scale_x_reordered() +
  facet_wrap(~ label, nrow = 1, scales = "free_y") +
  coord_flip() +
  scale_fill_manual(values = c("#08519c","#2d6a4f","#d7301f","#f4a261","#6a0572")) +
  labs(x = NULL, y = "P(word | topic)",
       title = "LDA Topic Model: Automatically Discovered Themes in WTP Text",
       subtitle = paste0("k = 5 topics  \u00b7  Each facet = one discovered theme  \u00b7  ",
                         "Top 8 words ordered by probability")) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        axis.text.x = element_text(size = 7),
        strip.text = element_text(face = "bold", size = 8))
```

::: {.callout-note}
## Keyword search vs. LDA: when to use each

| | Keyword search | LDA topic model |
|---|---|---|
| **Prior knowledge needed?** | Yes — specify patterns | No — fully unsupervised |
| **Best for** | Confirming expected concepts | Discovering unexpected ones |
| **Output** | % of responses per concept | Probabilistic word clusters |
| **Interpretability** | High (you chose the labels) | Requires human labeling |
| **Use in practice** | When you have a theory | Early pilot, no prior theory |

In practice, use **LDA first** to spot concepts you hadn't anticipated, then use **keyword patterns** to measure their prevalence precisely in the full sample.
:::

### From Text to Numbers: Measuring and Modeling the Nomological Net

The text analysis above identifies *which* constructs are active in the causal chain from X to Y. The next step is to **measure** those constructs quantitatively and **model** their pathways structurally. Together, these three stages — text scan → quantitative measurement → structural modeling — form the complete pipeline for dealing with exclusion restriction violations.

#### Step 2: Generate Quantitative Mediator Scores

The text scan identified which mediator constructs are active. Now we simulate what it would look like if we had *measured* those constructs quantitatively — as short scales administered alongside the main WTP item. Each score is generated from the underlying participant traits plus noise, reflecting what a properly validated scale would produce.

```{r quant-mediators}
#| code-summary: "▶ Step 2: Simulate quantitative scores for each text-identified mediator"

set.seed(999)

# ── Treatment assignment first — eco_latent is CAUSED by the condition ────────
# Seeing an eco-label activates eco-mindedness: it raises the salience of
# sustainability, health/naturalness, social values, and certification trust.
# The condition effect on all four eco indicators flows ENTIRELY through eco_latent;
# eco_arm has NO direct path to any indicator. This makes the DGP's causal
# structure exactly match the SEM's measurement model (eco_factor ~ condition).
eco_arm <- as.integer(condition_p2 == "Eco-label")

eco_latent    <- 0.80 * eco_arm + rnorm(N_p2, 0, 1.0)   # condition → eco-mindedness
frugal_latent <- rnorm(N_p2)                               # frugality: stable trait, unaffected by condition
social_latent <- rnorm(N_p2)                               # social proof: stable trait, unaffected by condition

# ── Correlated mediators: high within-cluster loading on eco_latent ──────────
# The 4 eco indicators share eco_latent as a common cause (loadings 0.84–0.90),
# producing pairwise correlations of r ≈ 0.65–0.93 among indicators in this DGP.
# This severe multicollinearity means parallel mediation's individual b-paths are
# unstable and their attenuated sum underestimates the true latent-factor indirect effect.

# ── Correlated ERRORS within indicator pairs ──────────────────────────────────
# Beyond eco_latent, two pairs share a response-style method factor (loading 0.55):
#   sustainability & health_natural: "natural/organic framing"
#   social_values & trust_cert: "social legitimacy/institutional approval"
# Both method factors are INDEPENDENT of eco_latent and eco_arm → pure residual covariance.
# SEM models both pairs via ~~ constraints; parallel mediation ignores them entirely.
method_sus_health <- rnorm(N_p2, 0, 1.0)   # natural/organic framing: sustainability ↔ health_natural
method_soc_trust  <- rnorm(N_p2, 0, 1.0)   # social legitimacy:       social_values  ↔ trust_cert

# ── Eco-cluster indicators: caused by eco_latent + method factors ONLY ────────
# eco_arm has NO direct effect on any indicator — its influence flows entirely
# through eco_latent. Consequences:
#   (a) The CFA (eco_factor =~ indicators) is the correct measurement model for this DGP.
#   (b) SEM's structural path (eco_factor ~ condition) recovers the true a-path
#       (condition → eco_latent), correctly weighting all 4 indicators and correcting
#       for measurement-error attenuation in each.
#   (c) Parallel mediation treats 4 highly correlated indicators (r ≈ 0.65–0.93) as
#       independent predictors → severe multicollinearity + attenuation → individual
#       b-paths unstable; summed eco-cluster indirect effect underestimates the truth.
#   (d) The method factors add within-pair residual correlations (loading 0.55) that
#       parallel mediation ignores; the SEM models them via ~~ constraints.
med_scores <- data.frame(
  sustainability  = 0.90 * eco_latent + 0.55 * method_sus_health + rnorm(N_p2, 3.5, 0.25),
  social_values   = 0.88 * eco_latent + 0.55 * method_soc_trust  + rnorm(N_p2, 3.0, 0.25),
  health_natural  = 0.86 * eco_latent + 0.55 * method_sus_health + rnorm(N_p2, 3.2, 0.25),
  trust_cert      = 0.84 * eco_latent + 0.55 * method_soc_trust  + rnorm(N_p2, 3.5, 0.25),
  # Frugal cluster: moderately reflective of frugal_latent
  price_cost      = 0.70 * frugal_latent - 0.15 * eco_arm + rnorm(N_p2, 3.0, 0.55),
  habit_routine   = 0.68 * frugal_latent                  + rnorm(N_p2, 3.0, 0.55),
  # Social/demand cluster: moderately reflective of social_latent
  demand_effect   = 0.70 * social_latent + 0.45 * eco_arm + rnorm(N_p2, 2.0, 0.55),
  guilt_offset    = 0.68 * social_latent + 0.30 * eco_arm + rnorm(N_p2, 2.5, 0.55),
  # Standalone mediators
  quality_taste   = 0.40 * eco_latent   + 0.35 * eco_arm + rnorm(N_p2, 3.5, 0.60),
  hedonic         = 0.20 * eco_latent   + 0.15 * eco_arm + rnorm(N_p2, 3.8, 0.65)
)

# Standardize mediators to z-scores
med_z <- as.data.frame(scale(med_scores))
med_z$condition <- eco_arm

# ── WTP: depends on LATENT FACTORS, not individual indicators ────────────────
# This is the key DGP design: WTP reflects eco_latent (the shared construct
# behind the 4 correlated eco items), NOT each indicator separately.
# Parallel mediation is doubly misspecified: (1) treats 4 highly correlated
# indicators (r ≈ 0.65–0.93) as independent predictors → multicollinearity;
# (2) uses noisy proxies for eco_latent → attenuation bias in the b-paths.
# SEM is correctly specified — it models eco_latent explicitly via CFA,
# corrects for attenuation, and models the residual covariances via ~~.
WTP_med <- 5.0 +
  0.50 * eco_arm       +   # direct condition effect
  1.20 * eco_latent    +   # main indirect path (eco-mindedness → WTP)
 -0.55 * frugal_latent +   # frugality lowers WTP
  0.50 * social_latent +   # social proof boosts WTP
  0.40 * (med_scores$quality_taste - mean(med_scores$quality_taste)) /
          sd(med_scores$quality_taste) +
  0.15 * (med_scores$hedonic - mean(med_scores$hedonic)) /
          sd(med_scores$hedonic) +
  rnorm(N_p2, 0, 1.5)
med_z$WTP <- scale(WTP_med)[, 1]

cat("Mediator score means by condition (standardized; should differ for eco-activated dims):\n")
med_scores |>
  mutate(condition = condition_p2) |>
  group_by(condition) |>
  summarise(across(everything(), ~ round(mean(.), 2)), .groups = "drop") |>
  kable(caption = "Mean mediator scores by condition (raw scale)")
```

#### Step 3: Parallel Mediation Model

A parallel mediation model regresses each mediator on treatment (the *a* paths) and regresses WTP on treatment + all mediators simultaneously (the *b* paths + direct effect *c'*). The indirect effects (a × b) decompose the total X→Y effect into specific causal channels.

This model assumes the mediators are **independent** of each other (no correlated residuals). This is the standard Hayes PROCESS approach. **We will address mediation models in much greater detail in Module 3**, where we formalize the causal assumptions required for valid mediation inference and introduce more flexible structural equation modeling approaches. Here, we use parallel mediation as a first-pass decomposition tool to illustrate the concept — and to show where it breaks down.

```{r parallel-mediation}
#| code-summary: "▶ Step 3: Parallel mediation via lavaan (fits model; no standalone plot)"

med_names <- c("sustainability", "quality_taste", "price_cost", "social_values",
               "health_natural", "trust_cert", "demand_effect", "guilt_offset",
               "hedonic", "habit_routine")

# Build lavaan model string dynamically
a_paths <- paste0(med_names, " ~ a", seq_along(med_names), " * condition",
                  collapse = "\n  ")
b_paths <- paste0("WTP ~ ",
                  paste0("b", seq_along(med_names), " * ", med_names, collapse = " + "),
                  " + cp * condition")
ind_defs <- paste0("ind", seq_along(med_names), " := a", seq_along(med_names),
                   " * b", seq_along(med_names), collapse = "\n  ")
total_ind <- paste0("total_indirect := ",
                    paste0("ind", seq_along(med_names), collapse = " + "))

parallel_model <- paste("# a paths\n ", a_paths,
                        "\n# outcome\n ", b_paths,
                        "\n# indirect effects\n ", ind_defs,
                        "\n ", total_ind)

fit_parallel <- sem(parallel_model, data = med_z, se = "bootstrap", bootstrap = 200)

# Extract indirect effect estimates
param_par <- parameterEstimates(fit_parallel, boot.ci.type = "bca.simple") |>
  filter(op == ":=") |>
  mutate(
    label_clean = c(paste0("via ", gsub("_", " ", med_names)), "Total indirect"),
    sig = ifelse(pvalue < 0.05, "*", "ns"),
    model = "Parallel mediation"
  )
cat("Parallel mediation fitted. Total indirect:",
    round(param_par$est[param_par$label_clean == "Total indirect"], 3), "\n")
```

#### Step 4: SEM with Latent Factors

**Why parallel mediation underestimates the eco-cluster effect:** The eco-label activates a *cluster* of correlated constructs — sustainability, social values, health/natural, and trust in certification — all driven by the same underlying eco-mindedness latent factor. Regressing WTP on all 10 mediators simultaneously creates **multicollinearity** among those four; the regression splits their shared variance arbitrarily across four rows. Each individual eco-cluster pathway appears small and noisy.

**The SEM solution:** instead of treating each indicator as an independent mediator, model the shared variance explicitly as a *latent factor* using a CFA measurement model. The latent `eco_factor` extracts the common variance from its four indicators, accounts for measurement error in each, and serves as a single, well-identified mediator in the structural part of the model. This is precisely the link to Module 1: the CFA measurement model IS a discriminant/convergent validity check applied directly to the mediator set.

```{r sem-latent, fig.height=5.5, fig.width=9}
#| code-summary: "▶ Step 4: SEM with latent eco-factor (CFA + structural mediation)"

# ── Build cluster data: standardize cluster means so all effects are comparable ─
# eco_cluster, frugal_cluster, social_cluster = means of z-scored items → already
# near-zero mean, but SD ≈ 0.5–0.8 (smaller than individual items because averaging
# reduces noise).  We scale them to SD = 1 so oracle, parallel, and SEM indirect
# effects are all in standardised-WTP-per-unit-condition units.
cl_names <- c("eco_cluster","frugal_cluster","social_cluster","quality_taste","hedonic")

eco_raw    <- rowMeans(med_z[, c("sustainability","social_values","health_natural","trust_cert")])
frugal_raw <- rowMeans(med_z[, c("price_cost","habit_routine")])
social_raw <- rowMeans(med_z[, c("demand_effect","guilt_offset")])

med_cl <- data.frame(
  eco_cluster    = scale(eco_raw)[,1],      # standardised cluster means
  frugal_cluster = scale(frugal_raw)[,1],
  social_cluster = scale(social_raw)[,1],
  quality_taste  = med_z$quality_taste,     # already z-scored
  hedonic        = med_z$hedonic,
  condition      = med_z$condition,
  WTP            = med_z$WTP
)

# Extend with raw eco indicators for the CFA measurement model
# (indicators are already z-scored from med_z)
med_sem <- cbind(
  med_cl,
  sustainability = med_z$sustainability,
  social_values  = med_z$social_values,
  health_natural = med_z$health_natural,
  trust_cert     = med_z$trust_cert
)

# ── Latent SEM model ──────────────────────────────────────────────────────────
# Measurement model (CFA): eco_factor is a latent variable measured by 4 indicators.
# std.lv = TRUE standardises the latent variable to variance = 1, putting the SEM's
# eco_factor on the same scale as eco_latent_std in the oracle (both SD = 1).
sem_model_txt <- '
  # Measurement model: eco_factor = latent eco-mindedness from 4 z-scored indicators
  eco_factor =~ sustainability + social_values + health_natural + trust_cert

  # Residual covariances: pairs of indicators share method variance in the DGP
  # (natural/organic framing for sus↔health_natural; social legitimacy for soc↔trust_cert).
  # These parameters are IN the DGP but completely ignored by parallel mediation.
  sustainability ~~ health_natural
  social_values  ~~ trust_cert

  # a-paths: condition predicts each mediator / latent factor
  eco_factor     ~ a1 * condition
  frugal_cluster ~ a2 * condition
  social_cluster ~ a3 * condition
  quality_taste  ~ a4 * condition
  hedonic        ~ a5 * condition

  # b-paths + direct effect on standardised WTP
  WTP ~ b1 * eco_factor + b2 * frugal_cluster + b3 * social_cluster +
        b4 * quality_taste + b5 * hedonic + cp * condition

  # Indirect effects (all in standardised WTP units)
  ind_eco     := a1 * b1
  ind_frugal  := a2 * b2
  ind_social  := a3 * b3
  ind_quality := a4 * b4
  ind_hedonic := a5 * b5
  total_ind   := ind_eco + ind_frugal + ind_social + ind_quality + ind_hedonic
'

# std.lv = TRUE: latent eco_factor standardised to variance 1 → comparable scale
fit_sem <- sem(sem_model_txt, data = med_sem,
               std.lv = TRUE, se = "bootstrap", bootstrap = 200)

param_sem <- parameterEstimates(fit_sem, boot.ci.type = "bca.simple") |>
  filter(op == ":=") |>
  mutate(
    label_clean = c("via eco cluster", "via frugal cluster", "via social cluster",
                    "via quality taste", "via hedonic", "Total indirect"),
    model = "SEM (latent eco-factor)"
  )

cat("SEM fitted. Total indirect:",
    round(param_sem$est[param_sem$label_clean == "Total indirect"], 3), "\n")

# ── Aggregate parallel mediation within clusters for comparison ───────────────
# Sum individual a×b products within each theoretical cluster.
# Also sum CI bounds (approximate: assumes within-cluster paths are independent).
par_agg <- param_par |>
  filter(label_clean != "Total indirect") |>
  mutate(cluster = case_when(
    label_clean %in% c("via sustainability","via social values",
                       "via health natural","via trust cert") ~ "via eco cluster",
    label_clean %in% c("via price cost","via habit routine")  ~ "via frugal cluster",
    label_clean %in% c("via demand effect","via guilt offset") ~ "via social cluster",
    TRUE ~ label_clean
  )) |>
  group_by(cluster) |>
  summarise(
    est      = sum(est),
    ci.lower = sum(ci.lower),
    ci.upper = sum(ci.upper),
    .groups  = "drop"
  ) |>
  rename(label_clean = cluster) |>
  mutate(model = "Parallel (summed within cluster)")

cat("SEM and parallel mediation estimates computed. Proceeding to DGP recovery comparison.\n")
```

#### Step 5: How Well Does Each Approach Recover the True DGP?

The simulation tells us exactly how WTP was generated (from the latent factors). We can compare what each modelling strategy estimates against that known ground truth.

```{r dgp-recovery, fig.height=5.5, fig.width=12}
#| code-summary: "▶ Step 5: Compare recovered effects to true DGP"

# ── Oracle true indirect effects using the LATENT factors from the DGP ────────
# The TRUE causal chain runs through eco_latent, frugal_latent, and social_latent —
# not through their noisy observed composites.  Using the latent factors directly
# (a) eliminates attenuation bias from measurement error and (b) reflects the actual
# DGP coefficients.  A well-specified SEM should recover these values; parallel
# mediation, which enters noisy observed indicators, cannot.

wtp_std_vec <- scale(WTP_med)[, 1]   # standardised WTP from the DGP

# Standardise latent factors to SD = 1 (matches SEM std.lv = TRUE scale)
eco_latent_std    <- scale(eco_latent)[, 1]
frugal_latent_std <- scale(frugal_latent)[, 1]
social_latent_std <- scale(social_latent)[, 1]
quality_std       <- med_z$quality_taste        # already z-scored
hedonic_std_vec   <- med_z$hedonic              # already z-scored

# True a-paths: OLS of each latent factor on condition
true_a_eco     <- coef(lm(eco_latent_std ~ eco_arm))["eco_arm"]    # ≈ +0.74
true_a_frugal  <- coef(lm(frugal_latent_std ~ eco_arm))["eco_arm"] # ≈  0
true_a_social  <- coef(lm(social_latent_std ~ eco_arm))["eco_arm"] # ≈  0
true_a_quality <- coef(lm(quality_std ~ eco_arm))["eco_arm"]
true_a_hedonic <- coef(lm(hedonic_std_vec ~ eco_arm))["eco_arm"]

# True b-paths: partial OLS of WTP_std on all latent factors simultaneously
# (avoids attenuation bias from using noisy observed composites as proxies)
oracle_true_lm <- lm(wtp_std_vec ~ eco_latent_std + frugal_latent_std +
                       social_latent_std + quality_std + hedonic_std_vec)
true_b_eco     <- coef(oracle_true_lm)["eco_latent_std"]
true_b_frugal  <- coef(oracle_true_lm)["frugal_latent_std"]
true_b_social  <- coef(oracle_true_lm)["social_latent_std"]
true_b_quality <- coef(oracle_true_lm)["quality_std"]
true_b_hedonic <- coef(oracle_true_lm)["hedonic_std_vec"]

# True indirect effects (latent-factor oracle; no CIs — these ARE the DGP truth)
true_ind_oracle <- setNames(
  c(true_a_eco * true_b_eco,
    true_a_frugal * true_b_frugal,
    true_a_social * true_b_social,
    true_a_quality * true_b_quality,
    true_a_hedonic * true_b_hedonic),
  cl_names
)

cat("Oracle (latent-factor) indirect effects:\n")
print(round(true_ind_oracle, 3))

# ── Build comparison data frame with 95% bootstrap CIs ───────────────────────
# Oracle: point estimate only, no CI (it IS the DGP truth for this sample)
# par_agg and param_sem carry ci.lower / ci.upper from the bootstrap runs in Step 3/4
oracle_rows <- data.frame(
  label_clean = paste0("via ", gsub("_", " ", cl_names)),
  effect      = round(true_ind_oracle, 3),
  ci_lo       = NA_real_,
  ci_hi       = NA_real_,
  source      = "True DGP (oracle)"
)

par_rows <- par_agg |>
  filter(label_clean != "Total indirect") |>
  transmute(label_clean,
            effect = est, ci_lo = ci.lower, ci_hi = ci.upper,
            source = "Parallel (summed within cluster)")

sem_rows <- param_sem |>
  filter(label_clean != "Total indirect") |>
  transmute(label_clean,
            effect = est, ci_lo = ci.lower, ci_hi = ci.upper,
            source = "SEM (latent eco-factor)")

recovery_df <- bind_rows(oracle_rows, par_rows, sem_rows) |>
  mutate(source = factor(source,
                          levels = c("True DGP (oracle)",
                                     "Parallel (summed within cluster)",
                                     "SEM (latent eco-factor)")))

# Order y-axis by oracle effect size
oracle_order <- oracle_rows |> arrange(effect) |> pull(label_clean)
recovery_df  <- recovery_df |>
  mutate(label_clean = factor(label_clean, levels = oracle_order))

ggplot(recovery_df, aes(x = effect, y = label_clean,
                         color = source, shape = source)) +
  geom_vline(xintercept = 0, color = "gray40", linewidth = 0.7) +
  geom_errorbarh(aes(xmin = ci_lo, xmax = ci_hi),
                 position = position_dodge(width = 0.65),
                 height = 0.22, linewidth = 0.6, na.rm = TRUE) +
  geom_point(position = position_dodge(width = 0.65), size = 3.5) +
  scale_color_manual(values = c("True DGP (oracle)"                = "#d7301f",
                                 "Parallel (summed within cluster)" = "#4a90d9",
                                 "SEM (latent eco-factor)"          = "#2d6a4f"),
                     name = NULL) +
  scale_shape_manual(values = c("True DGP (oracle)"                = 18,
                                 "Parallel (summed within cluster)" = 16,
                                 "SEM (latent eco-factor)"          = 15),
                     name = NULL) +
  labs(x = "Indirect effect (standardized)",
       y = NULL,
       title = "Model Recovery vs. True DGP",
       subtitle = paste0("Red diamonds = oracle (latent-factor truth, no CI)  \u00b7  ",
                         "Error bars = 95% bootstrap CI\n",
                         "SEM (green) corrects for attenuation + residual covariances; ",
                         "parallel mediation (blue) attenuated by multicollinearity")) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), legend.position = "top")
```

::: {.callout-note}
## What the recovery plot shows

**The SEM (green) closely tracks the true oracle values (red), while parallel mediation (blue) systematically underestimates the eco-cluster pathway.** The oracle is computed from the true latent factors (eco_latent, frugal_latent, social_latent) used to generate the data — the unattenuated causal quantities in the DGP. Parallel mediation deviates from this truth for two compounding reasons:

1. **Attenuation bias.** Each eco indicator is a noisy proxy for eco_latent (loading ≈ 0.84–0.90 plus a method factor, plus idiosyncratic noise). Entering noisy proxies in a regression shrinks the b-path estimates toward zero — the classic errors-in-variables attenuation. Even when the four attenuated b-paths are summed across the eco cluster, the resulting indirect effect underestimates the true latent-factor effect.

2. **Multicollinearity.** The four eco indicators are highly intercorrelated (r ≈ 0.65–0.93) because they share both eco_latent and a response-style method factor. Entering all four simultaneously in the WTP regression creates severe multicollinearity (VIF >> 5), making individual b-path estimates unstable and inflating their variance — compounding the attenuation problem.

By contrast, the SEM explicitly defines a latent `eco_factor` via a CFA measurement model. The CFA correctly partitions eco_latent signal from measurement noise and method-factor variance, yielding an **unattenuated** b-path estimate from eco_factor to WTP. The residual covariance constraints (`sustainability ~~ health_natural`; `social_values ~~ trust_cert`) further correct for the within-pair shared method variance that parallel mediation ignores. The result is a structural path estimate that closely tracks the true latent-factor oracle.

Neither model is perfect. The key insight is that **any structural model is only as good as the list of mediators you feed into it**. The text analysis gave us that list. Without the text analysis, you would have entered the study with an incomplete nomological map — and the "treatment effect" coefficient would have remained an uninterpretable mixture of all active pathways.

This is the direct parallel to Module 1's discriminant validity problem: just as measuring only some sub-dimensions of Y produces a biased estimate of your latent construct, measuring only some mediators of the X→Y relationship produces a biased estimate of the causal effect. In both cases, the solution is the same: map the latent space carefully, measure what you can, and report the gap transparently.
:::

**5. Two Further Threats to Mediation Inference — and Their Remedies**

The parallel mediation and SEM above address *which* mediators to include and *how* to model them. Two additional threats remain regardless of model specification: **non-normal residuals / influential outliers** (which can corrupt OLS *b*-path estimates) and **unmeasured M–Y confounders** which we will disucss in Module 3.

---

::: {.callout-note}
## Tying it all together: the Module 1–2 thread

Every concept in this tutorial maps directly onto a concept from Module 1:

| Module 1 (Measurement) | Module 2 (Experimentation and Identification) |
|---|---|
| Measured Observables | Treatment conditions |
| Latent construct (e.g., Green Purchase Intention) | Latent property of randomness / independence |
| Discriminant validity: does your scale measure *only* what it should? | Exclusion restriction: does your manipulation activate *only* the pathway it should? |
| Convergent validity: do items for the same construct correlate? | Randomization check: do characteristics correlate with treatment assignment? |
| Measurement invariance: does the scale work the same across groups? | Heterogeneous treatment effects: does the manipulation work the same across participant types? |
| Latent classes: are there sub-populations with different response patterns? | Effect heterogeneity: are there sub-populations for whom the treatment operates through different mechanisms? |
| Low signal-to-noise → poor reliability | Broad, complex Y → high N required for true independence |

**The common lesson:** observables (scale scores, group assignments, p-values) are imperfect proxies for latent quantities (true constructs, true independence, true effect sizes). Researchers who confuse the observable with the latent — who treat "my scale is published" as "my measurement is valid," or "Qualtrics randomized" as "my groups are balanced" — are making the same type of error, just in different places in the research pipeline. The path forward in both cases is the same: map the latent space carefully, measure what you can, acknowledge what you cannot, and report the gap transparently.
:::

---
