---
title: "Part 1: What Does a P-value Mean?"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  fig.width  = 9,
  fig.height = 5,
  fig.path   = "figs/",
  dev        = "svg"
)
dir.create("figs", recursive = TRUE, showWarnings = FALSE)
```

```{r packages}
#| code-summary: "▶ Load required packages"
# install.packages(c("tidyverse","ggplot2","knitr","scales","patchwork"))

library(tidyverse)
library(ggplot2)
library(knitr)
library(scales)
library(patchwork)
if (!requireNamespace("lavaan", quietly = TRUE)) install.packages("lavaan")
library(lavaan)

set.seed(2025)
```

<iframe src="https://docs.google.com/presentation/d/e/2PACX-1vRg0zz3ILArjQNbjlILaUy3EvVrBqDrDrLsbzieDsJU9VcB1NBVWJufM-Oiv3EeJA/pubembed?start=false&loop=false&delayms=3000" width="100%" height="500px" frameborder="0" allowfullscreen="true" mozallowfullscreen="true" webkitallowfullscreen="true" title="Module 2 Part 1: P-values"></iframe>

# Part 1: What Does a P-value Actually Mean? {#sec-pvalues}

## The Core Idea

A p-value answers one very specific question:

> **"How surprising is my observed result, *if the null hypothesis were true*?"**

That's it. The answer is a probability: *the proportion of results at least as extreme as mine that would occur if I repeated the experiment infinitely many times under the null hypothesis.*

Before we compute a single number, we need three things:

1. A **null hypothesis** — a concrete statement about what the world would look like if there were no effect
2. A **null distribution** — the distribution of our test statistic under that null world
3. A way to locate **our observed result** within that distribution

We will build all three by simulation.

## The Study: Does Eco-Labeling Change Willingness to Pay?

A researcher studies whether eco-certification labels on coffee packages increase consumers' willingness to pay (WTP). Sixty participants were recruited and randomly assigned to one of two conditions:

| Condition | n | Stimulus |
|---|---|---|
| Control | 30 | Standard coffee package |
| Eco-label | 30 | Same package with eco-certification badge |

After viewing the package, each participant stated their maximum WTP in dollars (on a continuous $1–$10 scale). The researcher observed that the **eco-label group was willing to pay $0.80 more** than the control group, on average.

*Is this difference real — or could it be explained by chance?*

## The Null World: What "No Effect" Actually Means

The null hypothesis states:

> **H₀:** Eco-labeling has no effect on WTP. Any observed difference between groups is due to sampling variability alone.

If H₀ is true, then all 60 participants are draws from the **same underlying population** — and the group labels ("Control" vs. "Eco-label") are arbitrary. We could have assigned those labels any other way without changing anything meaningful about the data.

This is the crucial insight:

> **Under H₀, the treatment labels are *exchangeable*.**

We can exploit this directly: shuffle the labels thousands of times, recompute the group difference each time, and build up the full distribution of differences that arise *purely from chance*. This is the **null distribution**.

```{r sim-data}
#| code-summary: "▶ Simulate WTP data and compute observed difference"

set.seed(2025)

N    <- 60
mu   <- 5.50    # True population mean WTP ($)
sig  <- 1.50    # True population SD ($)

# All 60 participants drawn from the same population (null is true).
# In our specific sample the eco-label group happened to score higher — by chance.
wtp_all <- rnorm(N, mean = mu, sd = sig) |> round(2) |> pmax(1) |> pmin(10)
group   <- factor(rep(c("Control", "Eco-label"), each = N / 2))
df_coffee <- data.frame(wtp = wtp_all, group = group)

obs_diff <- mean(df_coffee$wtp[df_coffee$group == "Eco-label"]) -
            mean(df_coffee$wtp[df_coffee$group == "Control"])

cat(sprintf("Observed mean difference: $%.2f\n", obs_diff))
```

```{r raw-data-plot, fig.height=5}
#| code-summary: "▶ Plot: raw WTP by condition"

ggplot(df_coffee, aes(x = group, y = wtp, fill = group, color = group)) +
  geom_jitter(width = 0.15, alpha = 0.45, size = 2.5, shape = 16) +
  geom_boxplot(alpha = 0.20, outlier.shape = NA, width = 0.45, linewidth = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 5,
               fill = "white", color = "black", stroke = 1.5) +
  annotate("segment", x = 1, xend = 2, y = 10.4, yend = 10.4,
           color = "gray30", linewidth = 0.8) +
  annotate("text", x = 1.5, y = 10.7,
           label = sprintf("Observed \u0394 = $%.2f", obs_diff),
           size = 4, color = "gray20", fontface = "bold") +
  scale_fill_manual(values  = c("Control" = "#4a90d9", "Eco-label" = "#2d6a4f")) +
  scale_color_manual(values = c("Control" = "#2b6cb0", "Eco-label" = "#1b4332")) +
  scale_y_continuous(limits = c(1, 11), breaks = 1:10,
                     labels = function(x) paste0("$", x)) +
  labs(x = NULL, y = "Willingness to Pay ($)",
       title = "Eco-Coffee WTP by Condition",
       subtitle = "Diamonds show group means  \u00b7  Points are individual participants") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none", panel.grid.minor = element_blank())
```

**What this plot is telling you — before we touch a single formula.**

Each dot is one participant's stated WTP. The box spans the middle 50% of values within each condition; the diamond marks the group mean. The bracket at the top labels the observed mean difference: Δ = $`r sprintf("%.2f", abs(obs_diff))`$.

A few things worth noticing right now:

- **The distributions heavily overlap.** The vast majority of the WTP values in both conditions occupy the same range. There is no clean separation — only a slight shift in the center of one distribution relative to the other.
- **Individual variation swamps the group difference.** The spread *within* each group (roughly $1–$10) is many times larger than the $`r sprintf("%.2f", abs(obs_diff))`$ gap *between* them. This is almost always true in behavioral experiments: people differ from one another far more than any treatment moves them.
- **This is precisely why "significant" requires more than a single number.** The gap of $`r sprintf("%.2f", abs(obs_diff))`$ could be real — or it could have appeared by chance simply because of which 30 people happened to land in each condition. The p-value we compute next turns that intuition into a probability.

## Building the Null Distribution by Shuffling Labels

To ask "how often would chance alone produce a difference this large?", we:

1. Pool all 60 WTP values
2. Randomly re-assign 30 to "Eco-label" and 30 to "Control" (shuffle the labels)
3. Record the new group difference
4. Repeat 10,000 times

Each iteration is a *parallel universe* in which the null is true and the labels are assigned differently. The collection of differences across all these universes **is** the null distribution.

```{r permutation-test}
#| code-summary: "▶ Run permutation test (10,000 label shuffles)"

set.seed(42)
n_perm      <- 10000
n_per_group <- N / 2

null_dist <- replicate(n_perm, {
  shuffled <- sample(df_coffee$wtp)
  mean(shuffled[1:n_per_group]) - mean(shuffled[(n_per_group + 1):N])
})

p_val <- mean(abs(null_dist) >= abs(obs_diff))

cat(sprintf("Two-sided permutation p-value: %.3f\n", p_val))
cat(sprintf("95%% CI of null distribution: [$%.2f, $%.2f]\n",
            quantile(null_dist, 0.025), quantile(null_dist, 0.975)))
```

### Watching the Shuffle in Action

Before looking at the summary of 10,000 shuffles, it helps to watch what a *single* shuffle actually does. The nine panels below show the original data plus eight individual shuffles. In each panel the **WTP values are exactly the same** — only the group labels (the colors) are reassigned. Notice how the group means (crossbars) bounce around at random, sometimes positive, sometimes negative, sometimes almost zero.

```{r shuffle-panels, fig.height=7.5, fig.width=10}
#| code-summary: "▶ Plot: 9 panels — observed data + 8 individual shuffles"

set.seed(77)
n_show <- 8

# Build the observed panel
obs_pnl <- data.frame(
  wtp       = df_coffee$wtp,
  group     = df_coffee$group,
  panel_id  = 0L,
  diff_k    = obs_diff
)

# Build 8 shuffle panels
shuf_list <- lapply(seq_len(n_show), function(k) {
  lbl  <- sample(df_coffee$group)
  dk   <- mean(df_coffee$wtp[lbl == "Eco-label"]) -
          mean(df_coffee$wtp[lbl == "Control"])
  data.frame(wtp = df_coffee$wtp, group = lbl,
             panel_id = k, diff_k = dk)
})

all_pnls <- bind_rows(obs_pnl, shuf_list)

# Build ordered panel labels
panel_meta <- all_pnls |>
  distinct(panel_id, diff_k) |>
  arrange(panel_id) |>
  mutate(panel_label = c(
    sprintf("\u2605 Observed  \u0394 = $%+.2f", diff_k[1]),
    sprintf("Shuffle %d  \u0394 = $%+.2f", seq_len(n_show), diff_k[-1])
  ))

all_pnls <- left_join(all_pnls, panel_meta |> dplyr::select(panel_id, panel_label),
                      by = "panel_id") |>
  mutate(panel_label = factor(panel_label, levels = panel_meta$panel_label))

ggplot(all_pnls, aes(x = group, y = wtp, color = group)) +
  geom_jitter(width = 0.14, alpha = 0.45, size = 1.6) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.45,
               fatten = 2.5, linewidth = 0.7, color = "black") +
  facet_wrap(~panel_label, nrow = 3, ncol = 3) +
  scale_color_manual(values = c("Control" = "#4a90d9", "Eco-label" = "#2d6a4f")) +
  scale_y_continuous(limits = c(1, 10), breaks = c(2, 5, 8),
                     labels = function(x) paste0("$", x)) +
  labs(x = NULL, y = "WTP ($)",
       title = "The Labels Change — the Data Do Not",
       subtitle = "Each panel re-assigns the same 60 WTP values to groups  \u00b7  Crossbar = group mean") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none",
        strip.text = element_text(size = 8.5, face = "bold"),
        axis.text.x = element_text(size = 8),
        panel.grid.minor = element_blank(),
        panel.spacing = unit(0.8, "lines"))
```

The starred panel (★ Observed) is the only one that reflects the actual group assignment. All other panels show what chance alone could have produced. The null distribution is nothing more than the collection of Δ values across all possible such panels.

### The Full Null Distribution

```{r null-dist-plot, fig.height=5}
#| code-summary: "▶ Plot: null distribution with observed difference highlighted"

null_df <- data.frame(
  diff    = null_dist,
  extreme = abs(null_dist) >= abs(obs_diff)
)

q025 <- quantile(null_dist, 0.025)
q975 <- quantile(null_dist, 0.975)

# Shared break-points so both histogram layers align perfectly
null_breaks <- seq(
  floor(min(null_dist) * 20) / 20 - 0.025,
  ceiling(max(null_dist) * 20) / 20 + 0.025,
  by = 0.05
)

ggplot(null_df, aes(x = diff, fill = extreme)) +
  geom_histogram(aes(y = after_stat(density)),
                 breaks = null_breaks, color = "white", linewidth = 0.15) +
  geom_density(aes(x = diff), data = null_df, inherit.aes = FALSE,
               color = "#08519c", linewidth = 1.0) +
  geom_vline(xintercept =  obs_diff, color = "#d7301f", linewidth = 1.6) +
  geom_vline(xintercept = -obs_diff, color = "#d7301f", linewidth = 1.0,
             linetype = "dashed") +
  geom_vline(xintercept = q025, color = "#525252", linewidth = 0.7,
             linetype = "dotted") +
  geom_vline(xintercept = q975, color = "#525252", linewidth = 0.7,
             linetype = "dotted") +
  annotate("label", x = 0, y = max(density(null_dist)$y) * 0.82,
           label = sprintf("p = %.3f", p_val),
           fill = "#deebf7", color = "#08306b", fontface = "bold", size = 5.5,
           label.size = 0.4, label.padding = unit(0.35, "lines")) +
  annotate("text",
           x = obs_diff + sign(obs_diff) * 0.08,
           y = max(density(null_dist)$y) * 0.52,
           label = sprintf("Observed\n\u0394 = $%.2f", obs_diff),
           color = "#d7301f", fontface = "bold", size = 3.5,
           hjust = ifelse(obs_diff >= 0, 0, 1)) +
  annotate("text", x = q025 - 0.03, y = 0.16,
           label = "2.5th\npercentile", color = "#525252", size = 2.9, hjust = 1) +
  annotate("text", x = q975 + 0.03, y = 0.16,
           label = "97.5th\npercentile", color = "#525252", size = 2.9, hjust = 0) +
  scale_fill_manual(
    values = c("FALSE" = "#9ecae1", "TRUE" = "#de2d26"),
    labels = c(expression(paste("Consistent with ", H[0])),
               "As or more extreme than observed"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(-1.5, 1.5, 0.5),
                     labels = function(x) sprintf("$%.1f", x)) +
  labs(
    x = "Mean difference (Eco-label \u2212 Control) across 10,000 shuffles",
    y = "Density",
    title = "Null Distribution: What Chance Alone Produces",
    subtitle = sprintf(
      "Blue = consistent with H0  \u00b7  Red = as or more extreme (p = %.3f)  \u00b7  Dotted = 95%% CI of null",
      p_val)
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor  = element_blank(),
    panel.grid.major  = element_line(color = "#eeeeee"),
    legend.position   = "bottom",
    plot.title        = element_text(face = "bold"),
    plot.subtitle     = element_text(color = "#555555", size = 10.5)
  )
```

The p-value of **`r sprintf("%.3f", p_val)`** tells us: `r sprintf("%.1f", p_val * 100)`% of shuffled datasets produced a difference as large as $`r sprintf("%.2f", abs(obs_diff))` or larger. This difference is *not* unusual under the null — chance produces it regularly.

::: {.callout-note}
## What the p-value is and is NOT

**It IS:** The probability of a result at least this extreme, given that H₀ is true.

**It is NOT:**
- The probability that H₀ is true
- The probability that the effect is real
- A measure of the size or importance of the effect
- A guarantee that you have not made an error

The p-value is a statement about a hypothetical distribution — the null world — not a statement about reality.
:::

## Where Does a P-value Get Its Power?

A p-value becomes small (informative) only when the observed result falls far into the tail of the null distribution. Three things push results into the tail:

1. **A large true effect (δ)** — the signal is genuinely far from zero
2. **A large sample size (N)** — random variability shrinks, so even moderate signals stand out
3. **A low-noise outcome** — tight distributions make small differences more visible

### How big an effect can N = 60 actually detect?

Before running a study, it is worth asking: if the eco-label truly has effect δ, how many participants do I need to detect it 80% of the time?

```{r power-table}
#| code-summary: "▶ Compute minimum N for different true effect sizes"

true_deltas <- c(0.20, 0.40, 0.60, 0.80, 1.00, 1.50)

power_tab <- data.frame(
  true_eff  = sprintf("$%.2f", true_deltas),
  cohens_d  = round(true_deltas / sig, 2),
  n_per_grp = sapply(true_deltas, function(d) {
    ceiling(power.t.test(delta = d, sd = sig, sig.level = 0.05,
                         power = 0.80, type = "two.sample")$n)
  }),
  check.names = FALSE
) |>
  mutate(total_n    = n_per_grp * 2,
         detectable = ifelse(total_n <= 60, "\u2713 Yes", "\u2717 No"))

names(power_tab) <- c("True effect (\u03b4)", "Cohen\u2019s d",
                       "N per group (80% power)", "Total N", "Detectable at N = 60?")

kable(power_tab,
      caption = sprintf(
        "Minimum N for 80%% power at \u03b1 = .05 (SD = $%.2f, two-sided t-test)", sig))
```

```{r pval-distribution, fig.height=5}
#| code-summary: "▶ Plot: p-value distributions for different true effect sizes"

sim_pval <- function(effect, N = 60, mu = 5.5, sig = 1.5) {
  y_c <- rnorm(N / 2, mu,          sig)
  y_t <- rnorm(N / 2, mu + effect, sig)
  t.test(y_t, y_c)$p.value
}

effects <- c(0, 0.30, 0.60, 1.20)
labels  <- c("No effect (\u03b4 = 0)",
             "Small (\u03b4 = $0.30)",
             "Medium (\u03b4 = $0.60)",
             "Large (\u03b4 = $1.20)")

set.seed(42)
pval_df <- map2_dfr(effects, labels, function(eff, lbl) {
  data.frame(effect = lbl, p = replicate(5000, sim_pval(eff)))
}) |>
  mutate(effect = factor(effect, levels = labels),
         sig    = p < 0.05)

pval_df |>
  group_by(effect) |>
  summarise(power = mean(sig), .groups = "drop") |>
  mutate(power = percent(power, accuracy = 0.1)) |>
  kable(col.names = c("True Effect", "% experiments with p < .05"),
        caption = "Statistical power at \u03b1 = .05 (N = 60 per group, SD = $1.50)")
```

```{r pval-hist, fig.height=4.5}
#| code-summary: "▶ Plot: p-value histograms by effect size"

ggplot(pval_df, aes(x = p, fill = sig)) +
  geom_histogram(breaks = seq(0, 1, by = 0.025), color = "white", alpha = 0.9) +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "black", linewidth = 0.9) +
  scale_fill_manual(values = c("FALSE" = "#9ecae1", "TRUE" = "#de2d26"),
                    labels = c("p \u2265 .05", "p < .05"), name = NULL) +
  facet_wrap(~effect, nrow = 1, scales = "free_y") +
  labs(x = "p-value", y = "Count (5,000 simulated experiments)",
       title = "Distribution of P-values Across Effect Sizes",
       subtitle = "Under H\u2080 (\u03b4 = 0): uniform  \u00b7  As \u03b4 grows: pile-up near zero") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        strip.text = element_text(size = 9.5))
```

::: {.callout-important}
## The uniform distribution under H₀ is not a coincidence

When the null is exactly true, p-values are **uniformly distributed** between 0 and 1. This means that 5% of null experiments will produce p < .05, 1% will produce p < .01, and so on.

**This is the false positive rate, not a coincidence.** α is not a threshold for truth — it is the rate at which you agree to be wrong when nothing is actually going on.

A practical implication: across a literature of 100 experiments where H₀ is true in all of them, you would still expect ~5 published papers with p < .05.
:::

## The Assumptions Behind Every P-value

The permutation logic above works because of one foundational assumption:

> **Exchangeability:** Under H₀, the labels are arbitrary — any assignment of observations to groups is equally likely.

This assumption is satisfied when:
- Participants are a random sample from the population
- Treatment was assigned randomly and independently

If either fails, the null distribution we construct by shuffling labels does not correspond to the distribution under the true null. The p-value becomes meaningless — or worse, systematically biased.

**The act of randomization is what makes exchangeability plausible.** Which brings us to Parts 2 and 3.

---

## Researcher Checklist: What Does a P-Value Mean? {#sec-checklist-part1}

::: {.callout-note}
## Interpreting the p-value correctly

- [ ] **Can you state what your p-value means in a single sentence without using the word "probability" incorrectly?** The correct interpretation: "If the null hypothesis were true, the probability of observing a test statistic at least as extreme as mine is p." It does not mean "the probability the null is true" or "the probability the effect is real."
- [ ] **Does your p-value test the thing you care about?** A significant t-test tells you the means differ in your sample — it does not tell you the effect is theoretically meaningful, replicable, or practically important. Report the effect size (Cohen's d, η²) alongside the p-value.
- [ ] **Did you compute a two-sided or one-sided p-value?** If one-sided, the directional prediction must have been registered before data collection, not selected after seeing which direction the data went.
- [ ] **Is the "null world" you are testing actually the right null?** For many theoretical predictions, the interesting question is not "is the effect exactly zero?" but "is the effect larger than some theoretically meaningful threshold?" Consider equivalence testing (TOST) if you want to rule out practically significant effects.
:::

::: {.callout-note}
## The three ingredients — are they present?

- [ ] **Same data-generating process in both conditions?** If your two conditions differ in more than the intended treatment (see Part 2 on the exclusion restriction), the null distribution built by permuting labels is not valid.
- [ ] **Exchangeability under H0?** Did randomization achieve approximate independence between treatment and all potential confounders (see Part 3)? If not, the permutation null distribution is constructed under a false assumption.
- [ ] **Pre-specified test statistic?** Did you choose the test statistic and analysis before seeing the data? Post-hoc selection of the test that gives p < 0.05 among several alternatives inflates the false-positive rate dramatically.
:::

::: {.callout-note}
## Effect size and practical significance

- [ ] **Have you reported the raw effect size in meaningful units** (e.g., $$\Delta WTP = \$0.80$$, not just d = 0.53)?
- [ ] **Is the effect large enough to matter substantively?** Statistical significance with N = 800 can be achieved for effects far too small to have theoretical or managerial relevance. Apply the "smallest effect size of interest" (SESOI) criterion.
- [ ] **Did you report the confidence interval alongside the p-value?** A 95% CI that spans from near-zero to large reveals more about uncertainty than p = 0.03 alone.
:::

*The central question for this part: Is a p < 0.05 result in this study strong evidence against the null — or does it depend on assumptions about data generation, exchangeability, and analysis specification that may not hold?*
