---
title: "Module 2 Code Tutorial"
subtitle: "P-values, Permutation Inference, and the Hidden Assumptions of Random Assignment"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  fig.width  = 9,
  fig.height = 5,
  fig.path   = "figs/",
  dev        = "svg"
)
dir.create("figs", recursive = TRUE, showWarnings = FALSE)
```

```{=html}
<style>
/* ── Floating bottom control bar ──────────────────────────────────────────── */
#global-output-toggle {
  position: fixed; bottom: 0; left: 0; right: 0; z-index: 9999;
  display: flex; flex-wrap: wrap; gap: 10px;
  justify-content: center; align-items: center;
  padding: 11px 24px;
  background: #1e3a5f; border-top: 3px solid #4a90d9;
  box-shadow: 0 -3px 16px rgba(0,0,0,0.35);
}
#global-output-toggle span {
  color: #c8d8ed; font-size: 0.85em; font-weight: 700;
  letter-spacing: 0.6px; text-transform: uppercase; margin-right: 6px;
}
body { padding-bottom: 70px !important; }
.gtbtn {
  padding: 9px 24px; color: #ffffff; border-radius: 7px;
  font-size: 1em; font-weight: 700; cursor: pointer;
  letter-spacing: 0.2px; transition: background 0.15s, border-color 0.15s;
  white-space: nowrap; border: 2px solid transparent;
}
/* Code button — blue */
.gtbtn.code-btn   { background: #2563a8; border-color: #5b9bd5; }
.gtbtn.code-btn:hover { background: #1a4a80; border-color: #90bfe8; }
/* Output button — teal/green */
.gtbtn.output-btn { background: #2d6a4f; border-color: #52b788; }
.gtbtn.output-btn:hover { background: #1b4332; border-color: #74c69d; }
/* ── Per-output collapsible wrappers ──────────────────────────────────────── */
details.out-fold { margin: 6px 0 14px 0; }
summary.out-fold-btn {
  display: inline-flex; align-items: center; gap: 7px;
  padding: 8px 20px; background: #eef4fb;
  border: 2px solid #4a90d9; border-radius: 7px; cursor: pointer;
  font-size: 0.94em; font-weight: 700; color: #1a3a6b;
  user-select: none; list-style: none; transition: background 0.12s;
}
summary.out-fold-btn::-webkit-details-marker { display: none; }
summary.out-fold-btn:hover { background: #cfe2f7; }
details[open].out-fold > summary.out-fold-btn { background: #d6ecff; }
summary.out-fold-btn .fold-icon { font-size: 1em; }
details[open] .fold-icon-closed { display: none; }
details:not([open]) .fold-icon-open  { display: none; }
</style>

<script>
document.addEventListener('DOMContentLoaded', function () {
  var outputSels = ['.cell-output-stdout','.cell-output-stderr','.cell-output-display'];
  document.querySelectorAll(outputSels.join(',')).forEach(function (el) {
    if (el.closest('.callout') || el.closest('details.out-fold')) return;
    var hasImg = !!el.querySelector('img, svg, canvas');
    var hasTbl = !!el.querySelector('table');
    var label  = hasImg ? '📊  Show plot' : hasTbl ? '📋  Show table' : '📄  Show output';
    var det = document.createElement('details');
    det.className = 'out-fold';
    var sum = document.createElement('summary');
    sum.className = 'out-fold-btn';
    sum.innerHTML =
      '<span class="fold-icon fold-icon-closed">▶</span>' +
      '<span class="fold-icon fold-icon-open">▼</span>' + label;
    el.parentNode.insertBefore(det, el);
    det.appendChild(sum); det.appendChild(el);
  });
  var bar = document.createElement('div');
  bar.id = 'global-output-toggle';
  bar.innerHTML =
    '<span>Quick controls:</span>' +
    '<button id="code-toggle-btn"   class="gtbtn code-btn"   onclick="toggleCode()">\u25bc Expand All Code</button>' +
    '<button id="output-toggle-btn" class="gtbtn output-btn" onclick="toggleOutput()">\u25bc Expand All Output</button>';
  document.body.appendChild(bar);
});

var codeOpen   = false;
var outputOpen = false;

function toggleCode() {
  codeOpen = !codeOpen;
  document.querySelectorAll('details').forEach(function (d) {
    if (!d.classList.contains('out-fold') && d.querySelector('.sourceCode')) {
      d.open = codeOpen;
    }
  });
  var btn = document.getElementById('code-toggle-btn');
  if (btn) btn.textContent = codeOpen ? '\u25b2 Collapse All Code' : '\u25bc Expand All Code';
}

function toggleOutput() {
  outputOpen = !outputOpen;
  document.querySelectorAll('details.out-fold').forEach(function (d) {
    d.open = outputOpen;
  });
  var btn = document.getElementById('output-toggle-btn');
  if (btn) btn.textContent = outputOpen ? '\u25b2 Collapse All Output' : '\u25bc Expand All Output';
}
</script>
```

## What This Tutorial Is About

There is a single thread running through this tutorial and its predecessor:

> **Your observable may not accurately reflect the latent construct you need.**

In Module 1, that problem lived in your *outcome variable Y* — your scale picked up variance from constructs you never intended to measure. Here, the same problem appears in **two places at once**: in your *treatment variable X*, and in the act of *randomization* that is supposed to make X interpretable.

This tutorial covers three ideas:

- **Part 1 — P-values:** What they actually mean, where their inferential power comes from, and what assumptions they require — demonstrated by building null distributions from scratch using permutation.

- **Part 2 — The Exclusion Restriction:** Just as a measured outcome Y can fail discriminant validity by absorbing multiple constructs, a manipulated treatment X can inject multiple independent signals into the system simultaneously. When this happens, your "experiment" is really a quasi-experiment — and the single coefficient on treatment conflates several distinct causal pathways.

- **Part 3 — Randomization:** Why "clicking randomize" (the observable) is not the same as "achieving randomization" (the latent property), and why this gap grows as the construct Y you are studying becomes broader and more complex.

::: {.callout-note}
## Learning Objectives

By the end of this tutorial, you will be able to:

- Explain what a p-value is — and is **not** — using simulation rather than formulas
- Build a null distribution from scratch using label permutation
- Identify the three ingredients that give a p-value inferential power
- Explain why a treatment manipulation can violate discriminant validity in the same way a measurement scale can
- Use open-ended text to map the constructs injected by an experimental manipulation and those driving an outcome
- Articulate the difference between **observable randomization** and **latent randomization** using the construct-validity language from Module 1
- Calculate the minimum observations needed for approximate orthogonality given the complexity of Y
- Run diagnostic checks to assess whether your observed randomization achieved approximate orthogonality
:::

---

```{r packages}
#| code-summary: "▶ Load required packages"
# install.packages(c("tidyverse","ggplot2","knitr","scales","patchwork"))

library(tidyverse)
library(ggplot2)
library(knitr)
library(scales)
library(patchwork)
if (!requireNamespace("lavaan", quietly = TRUE)) install.packages("lavaan")
library(lavaan)

set.seed(2025)
```

---

# Part 1: What Does a P-value Actually Mean? {#sec-pvalues}

## The Core Idea

A p-value answers one very specific question:

> **"How surprising is my observed result, *if the null hypothesis were true*?"**

That's it. The answer is a probability: *the proportion of results at least as extreme as mine that would occur if I repeated the experiment infinitely many times under the null hypothesis.*

Before we compute a single number, we need three things:

1. A **null hypothesis** — a concrete statement about what the world would look like if there were no effect
2. A **null distribution** — the distribution of our test statistic under that null world
3. A way to locate **our observed result** within that distribution

We will build all three by simulation.

## The Study: Does Eco-Labeling Change Willingness to Pay?

A researcher studies whether eco-certification labels on coffee packages increase consumers' willingness to pay (WTP). Sixty participants were recruited and randomly assigned to one of two conditions:

| Condition | n | Stimulus |
|---|---|---|
| Control | 30 | Standard coffee package |
| Eco-label | 30 | Same package with eco-certification badge |

After viewing the package, each participant stated their maximum WTP in dollars (on a continuous $1–$10 scale). The researcher observed that the **eco-label group was willing to pay $0.80 more** than the control group, on average.

*Is this difference real — or could it be explained by chance?*

## The Null World: What "No Effect" Actually Means

The null hypothesis states:

> **H₀:** Eco-labeling has no effect on WTP. Any observed difference between groups is due to sampling variability alone.

If H₀ is true, then all 60 participants are draws from the **same underlying population** — and the group labels ("Control" vs. "Eco-label") are arbitrary. We could have assigned those labels any other way without changing anything meaningful about the data.

This is the crucial insight:

> **Under H₀, the treatment labels are *exchangeable*.**

We can exploit this directly: shuffle the labels thousands of times, recompute the group difference each time, and build up the full distribution of differences that arise *purely from chance*. This is the **null distribution**.

```{r sim-data}
#| code-summary: "▶ Simulate WTP data and compute observed difference"

set.seed(2025)

N    <- 60
mu   <- 5.50    # True population mean WTP ($)
sig  <- 1.50    # True population SD ($)

# All 60 participants drawn from the same population (null is true).
# In our specific sample the eco-label group happened to score higher — by chance.
wtp_all <- rnorm(N, mean = mu, sd = sig) |> round(2) |> pmax(1) |> pmin(10)
group   <- factor(rep(c("Control", "Eco-label"), each = N / 2))
df_coffee <- data.frame(wtp = wtp_all, group = group)

obs_diff <- mean(df_coffee$wtp[df_coffee$group == "Eco-label"]) -
            mean(df_coffee$wtp[df_coffee$group == "Control"])

cat(sprintf("Observed mean difference: $%.2f\n", obs_diff))
```

```{r raw-data-plot, fig.height=5}
#| code-summary: "▶ Plot: raw WTP by condition"

ggplot(df_coffee, aes(x = group, y = wtp, fill = group, color = group)) +
  geom_jitter(width = 0.15, alpha = 0.45, size = 2.5, shape = 16) +
  geom_boxplot(alpha = 0.20, outlier.shape = NA, width = 0.45, linewidth = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 5,
               fill = "white", color = "black", stroke = 1.5) +
  annotate("segment", x = 1, xend = 2, y = 10.4, yend = 10.4,
           color = "gray30", linewidth = 0.8) +
  annotate("text", x = 1.5, y = 10.7,
           label = sprintf("Observed \u0394 = $%.2f", obs_diff),
           size = 4, color = "gray20", fontface = "bold") +
  scale_fill_manual(values  = c("Control" = "#4a90d9", "Eco-label" = "#2d6a4f")) +
  scale_color_manual(values = c("Control" = "#2b6cb0", "Eco-label" = "#1b4332")) +
  scale_y_continuous(limits = c(1, 11), breaks = 1:10,
                     labels = function(x) paste0("$", x)) +
  labs(x = NULL, y = "Willingness to Pay ($)",
       title = "Eco-Coffee WTP by Condition",
       subtitle = "Diamonds show group means  \u00b7  Points are individual participants") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none", panel.grid.minor = element_blank())
```

**What this plot is telling you — before we touch a single formula.**

Each dot is one participant's stated WTP. The box spans the middle 50% of values within each condition; the diamond marks the group mean. The bracket at the top labels the observed mean difference: Δ = $`r sprintf("%.2f", abs(obs_diff))`$.

A few things worth noticing right now:

- **The distributions heavily overlap.** The vast majority of the WTP values in both conditions occupy the same range. There is no clean separation — only a slight shift in the center of one distribution relative to the other.
- **Individual variation swamps the group difference.** The spread *within* each group (roughly $1–$10) is many times larger than the $`r sprintf("%.2f", abs(obs_diff))`$ gap *between* them. This is almost always true in behavioral experiments: people differ from one another far more than any treatment moves them.
- **This is precisely why "significant" requires more than a single number.** The gap of $`r sprintf("%.2f", abs(obs_diff))`$ could be real — or it could have appeared by chance simply because of which 30 people happened to land in each condition. The p-value we compute next turns that intuition into a probability.

## Building the Null Distribution by Shuffling Labels

To ask "how often would chance alone produce a difference this large?", we:

1. Pool all 60 WTP values
2. Randomly re-assign 30 to "Eco-label" and 30 to "Control" (shuffle the labels)
3. Record the new group difference
4. Repeat 10,000 times

Each iteration is a *parallel universe* in which the null is true and the labels are assigned differently. The collection of differences across all these universes **is** the null distribution.

```{r permutation-test}
#| code-summary: "▶ Run permutation test (10,000 label shuffles)"

set.seed(42)
n_perm      <- 10000
n_per_group <- N / 2

null_dist <- replicate(n_perm, {
  shuffled <- sample(df_coffee$wtp)
  mean(shuffled[1:n_per_group]) - mean(shuffled[(n_per_group + 1):N])
})

p_val <- mean(abs(null_dist) >= abs(obs_diff))

cat(sprintf("Two-sided permutation p-value: %.3f\n", p_val))
cat(sprintf("95%% CI of null distribution: [$%.2f, $%.2f]\n",
            quantile(null_dist, 0.025), quantile(null_dist, 0.975)))
```

### Watching the Shuffle in Action

Before looking at the summary of 10,000 shuffles, it helps to watch what a *single* shuffle actually does. The nine panels below show the original data plus eight individual shuffles. In each panel the **WTP values are exactly the same** — only the group labels (the colors) are reassigned. Notice how the group means (crossbars) bounce around at random, sometimes positive, sometimes negative, sometimes almost zero.

```{r shuffle-panels, fig.height=7.5, fig.width=10}
#| code-summary: "▶ Plot: 9 panels — observed data + 8 individual shuffles"

set.seed(77)
n_show <- 8

# Build the observed panel
obs_pnl <- data.frame(
  wtp       = df_coffee$wtp,
  group     = df_coffee$group,
  panel_id  = 0L,
  diff_k    = obs_diff
)

# Build 8 shuffle panels
shuf_list <- lapply(seq_len(n_show), function(k) {
  lbl  <- sample(df_coffee$group)
  dk   <- mean(df_coffee$wtp[lbl == "Eco-label"]) -
          mean(df_coffee$wtp[lbl == "Control"])
  data.frame(wtp = df_coffee$wtp, group = lbl,
             panel_id = k, diff_k = dk)
})

all_pnls <- bind_rows(obs_pnl, shuf_list)

# Build ordered panel labels
panel_meta <- all_pnls |>
  distinct(panel_id, diff_k) |>
  arrange(panel_id) |>
  mutate(panel_label = c(
    sprintf("\u2605 Observed  \u0394 = $%+.2f", diff_k[1]),
    sprintf("Shuffle %d  \u0394 = $%+.2f", seq_len(n_show), diff_k[-1])
  ))

all_pnls <- left_join(all_pnls, panel_meta |> select(panel_id, panel_label),
                      by = "panel_id") |>
  mutate(panel_label = factor(panel_label, levels = panel_meta$panel_label))

ggplot(all_pnls, aes(x = group, y = wtp, color = group)) +
  geom_jitter(width = 0.14, alpha = 0.45, size = 1.6) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.45,
               fatten = 2.5, linewidth = 0.7, color = "black") +
  facet_wrap(~panel_label, nrow = 3, ncol = 3) +
  scale_color_manual(values = c("Control" = "#4a90d9", "Eco-label" = "#2d6a4f")) +
  scale_y_continuous(limits = c(1, 10), breaks = c(2, 5, 8),
                     labels = function(x) paste0("$", x)) +
  labs(x = NULL, y = "WTP ($)",
       title = "The Labels Change — the Data Do Not",
       subtitle = "Each panel re-assigns the same 60 WTP values to groups  \u00b7  Crossbar = group mean") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none",
        strip.text = element_text(size = 8.5, face = "bold"),
        axis.text.x = element_text(size = 8),
        panel.grid.minor = element_blank(),
        panel.spacing = unit(0.8, "lines"))
```

The starred panel (★ Observed) is the only one that reflects the actual group assignment. All other panels show what chance alone could have produced. The null distribution is nothing more than the collection of Δ values across all possible such panels.

### The Full Null Distribution

```{r null-dist-plot, fig.height=5}
#| code-summary: "▶ Plot: null distribution with observed difference highlighted"

null_df <- data.frame(
  diff    = null_dist,
  extreme = abs(null_dist) >= abs(obs_diff)
)

q025 <- quantile(null_dist, 0.025)
q975 <- quantile(null_dist, 0.975)

# Shared break-points so both histogram layers align perfectly
null_breaks <- seq(
  floor(min(null_dist) * 20) / 20 - 0.025,
  ceiling(max(null_dist) * 20) / 20 + 0.025,
  by = 0.05
)

ggplot(null_df, aes(x = diff, fill = extreme)) +
  geom_histogram(aes(y = after_stat(density)),
                 breaks = null_breaks, color = "white", linewidth = 0.15) +
  geom_density(aes(x = diff), data = null_df, inherit.aes = FALSE,
               color = "#08519c", linewidth = 1.0) +
  geom_vline(xintercept =  obs_diff, color = "#d7301f", linewidth = 1.6) +
  geom_vline(xintercept = -obs_diff, color = "#d7301f", linewidth = 1.0,
             linetype = "dashed") +
  geom_vline(xintercept = q025, color = "#525252", linewidth = 0.7,
             linetype = "dotted") +
  geom_vline(xintercept = q975, color = "#525252", linewidth = 0.7,
             linetype = "dotted") +
  annotate("label", x = 0, y = max(density(null_dist)$y) * 0.82,
           label = sprintf("p = %.3f", p_val),
           fill = "#deebf7", color = "#08306b", fontface = "bold", size = 5.5,
           label.size = 0.4, label.padding = unit(0.35, "lines")) +
  annotate("text",
           x = obs_diff + sign(obs_diff) * 0.08,
           y = max(density(null_dist)$y) * 0.52,
           label = sprintf("Observed\n\u0394 = $%.2f", obs_diff),
           color = "#d7301f", fontface = "bold", size = 3.5,
           hjust = ifelse(obs_diff >= 0, 0, 1)) +
  annotate("text", x = q025 - 0.03, y = 0.16,
           label = "2.5th\npercentile", color = "#525252", size = 2.9, hjust = 1) +
  annotate("text", x = q975 + 0.03, y = 0.16,
           label = "97.5th\npercentile", color = "#525252", size = 2.9, hjust = 0) +
  scale_fill_manual(
    values = c("FALSE" = "#9ecae1", "TRUE" = "#de2d26"),
    labels = c(expression(paste("Consistent with ", H[0])),
               "As or more extreme than observed"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(-1.5, 1.5, 0.5),
                     labels = function(x) sprintf("$%.1f", x)) +
  labs(
    x = "Mean difference (Eco-label \u2212 Control) across 10,000 shuffles",
    y = "Density",
    title = "Null Distribution: What Chance Alone Produces",
    subtitle = sprintf(
      "Blue = consistent with H0  \u00b7  Red = as or more extreme (p = %.3f)  \u00b7  Dotted = 95%% CI of null",
      p_val)
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor  = element_blank(),
    panel.grid.major  = element_line(color = "#eeeeee"),
    legend.position   = "bottom",
    plot.title        = element_text(face = "bold"),
    plot.subtitle     = element_text(color = "#555555", size = 10.5)
  )
```

The p-value of **`r sprintf("%.3f", p_val)`** tells us: `r sprintf("%.1f", p_val * 100)`% of shuffled datasets produced a difference as large as $`r sprintf("%.2f", abs(obs_diff))` or larger. This difference is *not* unusual under the null — chance produces it regularly.

::: {.callout-note}
## What the p-value is and is NOT

**It IS:** The probability of a result at least this extreme, given that H₀ is true.

**It is NOT:**
- The probability that H₀ is true
- The probability that the effect is real
- A measure of the size or importance of the effect
- A guarantee that you have not made an error

The p-value is a statement about a hypothetical distribution — the null world — not a statement about reality.
:::

## Where Does a P-value Get Its Power?

A p-value becomes small (informative) only when the observed result falls far into the tail of the null distribution. Three things push results into the tail:

1. **A large true effect (δ)** — the signal is genuinely far from zero
2. **A large sample size (N)** — random variability shrinks, so even moderate signals stand out
3. **A low-noise outcome** — tight distributions make small differences more visible

### How big an effect can N = 60 actually detect?

Before running a study, it is worth asking: if the eco-label truly has effect δ, how many participants do I need to detect it 80% of the time?

```{r power-table}
#| code-summary: "▶ Compute minimum N for different true effect sizes"

true_deltas <- c(0.20, 0.40, 0.60, 0.80, 1.00, 1.50)

power_tab <- data.frame(
  true_eff  = sprintf("$%.2f", true_deltas),
  cohens_d  = round(true_deltas / sig, 2),
  n_per_grp = sapply(true_deltas, function(d) {
    ceiling(power.t.test(delta = d, sd = sig, sig.level = 0.05,
                         power = 0.80, type = "two.sample")$n)
  }),
  check.names = FALSE
) |>
  mutate(total_n    = n_per_grp * 2,
         detectable = ifelse(total_n <= 60, "\u2713 Yes", "\u2717 No"))

names(power_tab) <- c("True effect (\u03b4)", "Cohen\u2019s d",
                       "N per group (80% power)", "Total N", "Detectable at N = 60?")

kable(power_tab,
      caption = sprintf(
        "Minimum N for 80%% power at \u03b1 = .05 (SD = $%.2f, two-sided t-test)", sig))
```

```{r pval-distribution, fig.height=5}
#| code-summary: "▶ Plot: p-value distributions for different true effect sizes"

sim_pval <- function(effect, N = 60, mu = 5.5, sig = 1.5) {
  y_c <- rnorm(N / 2, mu,          sig)
  y_t <- rnorm(N / 2, mu + effect, sig)
  t.test(y_t, y_c)$p.value
}

effects <- c(0, 0.30, 0.60, 1.20)
labels  <- c("No effect (\u03b4 = 0)",
             "Small (\u03b4 = $0.30)",
             "Medium (\u03b4 = $0.60)",
             "Large (\u03b4 = $1.20)")

set.seed(42)
pval_df <- map2_dfr(effects, labels, function(eff, lbl) {
  data.frame(effect = lbl, p = replicate(5000, sim_pval(eff)))
}) |>
  mutate(effect = factor(effect, levels = labels),
         sig    = p < 0.05)

pval_df |>
  group_by(effect) |>
  summarise(power = mean(sig), .groups = "drop") |>
  mutate(power = percent(power, accuracy = 0.1)) |>
  kable(col.names = c("True Effect", "% experiments with p < .05"),
        caption = "Statistical power at \u03b1 = .05 (N = 60 per group, SD = $1.50)")
```

```{r pval-hist, fig.height=4.5}
#| code-summary: "▶ Plot: p-value histograms by effect size"

ggplot(pval_df, aes(x = p, fill = sig)) +
  geom_histogram(breaks = seq(0, 1, by = 0.025), color = "white", alpha = 0.9) +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "black", linewidth = 0.9) +
  scale_fill_manual(values = c("FALSE" = "#9ecae1", "TRUE" = "#de2d26"),
                    labels = c("p \u2265 .05", "p < .05"), name = NULL) +
  facet_wrap(~effect, nrow = 1, scales = "free_y") +
  labs(x = "p-value", y = "Count (5,000 simulated experiments)",
       title = "Distribution of P-values Across Effect Sizes",
       subtitle = "Under H\u2080 (\u03b4 = 0): uniform  \u00b7  As \u03b4 grows: pile-up near zero") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        strip.text = element_text(size = 9.5))
```

::: {.callout-important}
## The uniform distribution under H₀ is not a coincidence

When the null is exactly true, p-values are **uniformly distributed** between 0 and 1. This means that 5% of null experiments will produce p < .05, 1% will produce p < .01, and so on.

**This is the false positive rate, not a coincidence.** α is not a threshold for truth — it is the rate at which you agree to be wrong when nothing is actually going on.

A practical implication: across a literature of 100 experiments where H₀ is true in all of them, you would still expect ~5 published papers with p < .05.
:::

## The Assumptions Behind Every P-value

The permutation logic above works because of one foundational assumption:

> **Exchangeability:** Under H₀, the labels are arbitrary — any assignment of observations to groups is equally likely.

This assumption is satisfied when:
- Participants are a random sample from the population
- Treatment was assigned randomly and independently

If either fails, the null distribution we construct by shuffling labels does not correspond to the distribution under the true null. The p-value becomes meaningless — or worse, systematically biased.

**The act of randomization is what makes exchangeability plausible.** Which brings us to Parts 2 and 3.

---

# Part 2: The Exclusion Restriction — What Else Is in Your Treatment? {#sec-exclusion}

## One Button, Many Signals

In Part 1 we treated the eco-label manipulation as a clean, unidimensional intervention: it either increases WTP (H₁) or it does not (H₀). This framing is convenient, but it hides a problem that is exactly analogous to what Module 1 revealed about *measuring* Y.

Recall the core lesson from Module 1:

> A scale item labeled "Green Purchase Intention" often captures several distinct latent constructs simultaneously — environmental values, social signaling, perceived quality, price expectations, and more. The scale *appears* to measure one thing while actually measuring many. This is a violation of **discriminant validity**.

Now ask the same question about your treatment variable:

> When a participant sees an eco-certification badge on a coffee package, what does that label actually *do*?

It does not inject a single, clean signal of "eco-friendliness." It activates — simultaneously and heterogeneously — a cluster of inferences, associations, and motivations that vary across people. This is a violation of **the exclusion restriction**: the assumption that the treatment variable affects the outcome *only* through the intended pathway.

::: {.callout-important}
## The Exclusion Restriction — the single-mechanism assumption

The exclusion restriction is the assumption that X affects Y through **exactly one latent pathway** — the one the researcher intends. Every field has a name for what happens when this assumption breaks:

| Field | Term for the violation |
|---|---|
| Psychology / Marketing | **Confound** — the manipulation activates an unintended construct alongside the intended one |
| Econometrics | **Endogeneity** — the regressor correlates with the error term via unmodelled channels |
| Epidemiology | **Confounding** — exposure and outcome share a common cause |
| Statistics | **Omitted variable bias** — a variable that belongs in the model is left out |
| Causal inference (DAGs) | **Backdoor path** — a non-causal association flows through an unmeasured variable |

All of these are the same problem viewed through different disciplinary lenses: **there is more than one latent link between X and Y**.

If X injects additional unintended signals that also causally influence Y — and if those signals vary across participants — then the estimated "treatment effect" is a weighted mixture of several distinct causal effects, not a single clean estimate of any one of them.

**This is the fundamental identification problem in causal inference**, and it is the experimental analogue of discriminant invalidity in measurement. In Module 1 you saw that **measured Y can measure too much**. Here you will see that **manipulated X can inject too much**.
:::

## Simulating a Heterogeneous Treatment

We now simulate the coffee experiment with a realistic data-generating process. Each participant has nine binary latent characteristics that determine both *what signals they extract from the eco-label* and *how much they are willing to pay*.

```{r part2-setup}
#| code-summary: "▶ Simulate 120 participants with heterogeneous traits and WTP"

set.seed(2026)
N_p2  <- 120
half2 <- N_p2 / 2

# ── Participant-level binary traits ───────────────────────────────────────────
ptrait <- function(p) rbinom(N_p2, 1, p)

traits <- data.frame(
  env_values      = ptrait(0.45),  # cares about environment
  quality_seek    = ptrait(0.50),  # focuses on quality / taste
  price_sensitive = ptrait(0.50),  # price-conscious
  social_identity = ptrait(0.40),  # identity / values-driven purchasing
  health_focus    = ptrait(0.45),  # health / natural-ingredient preference
  brand_trust     = ptrait(0.55),  # responds to certification cues
  guilt_prone     = ptrait(0.35),  # guilt-reducing / compensatory consumption
  hedonic         = ptrait(0.50),  # pleasure-driven; taste is paramount
  habitual        = ptrait(0.40),  # habitual buyer; resists stretching budget
  demand_aware    = ptrait(0.40)   # susceptible to experimenter demand / social desirability cues
)

# ── Equal random assignment ───────────────────────────────────────────────────
condition_p2 <- rep(c("Control", "Eco-label"), each = half2)

# ── WTP simulation (in dollars, clamped to $1–$10) ────────────────────────────
# Key feature: eco-label ACTIVATES traits 1, 2, 4, 5, 6, 7 differentially.
# Traits 3 (price sensitivity), 8 (hedonic), 9 (habitual) operate in both conditions.
WTP_p2 <- with(traits, {
  eco <- as.integer(condition_p2 == "Eco-label")
  v   <- 4.50 +
    1.20 * env_values      * eco        +  # env concern → eco premium
    0.80 * quality_seek    * (0.35 + 0.65 * eco)  +  # quality matters more in eco
   -0.70 * price_sensitive            +  # price sensitivity suppresses WTP
    0.90 * social_identity * eco        +  # identity-driven eco premium
    0.60 * health_focus    * eco        +  # health focus eco premium
    0.50 * brand_trust     * eco        +  # certification trust premium
    0.35 * guilt_prone     * eco        +  # guilt-offset premium
    0.30 * hedonic                      +  # hedonic adds WTP in both
   -0.45 * habitual                    +  # habitual buyers resist stretching
    0.55 * demand_aware    * eco        +  # demand-aware participants inflate WTP in eco context
    rnorm(N_p2, 0, 0.75)
  pmax(pmin(round(v, 2), 10), 1)
})

df_p2 <- data.frame(traits, condition = condition_p2, WTP = WTP_p2)

# ── Summary by condition ──────────────────────────────────────────────────────
df_p2 |>
  group_by(condition) |>
  summarise(n = n(), mean_WTP = round(mean(WTP), 2),
            sd_WTP = round(sd(WTP), 2), .groups = "drop") |>
  kable(col.names = c("Condition", "n", "Mean WTP", "SD WTP"),
        align   = c("l", "c", "c", "c"),
        caption = "WTP summary by condition")
```

The overall treatment effect looks straightforward. But the figure below reveals the heterogeneity hiding underneath it:

```{r het-effects-plot, fig.height=5}
#| code-summary: "▶ Plot: treatment effect varies by participant trait profile"

# Compute individual-level treatment contrasts via matched simulation
# For illustration: show eco-label premium by number of "activated" traits
df_p2 <- df_p2 |>
  mutate(
    n_eco_traits = env_values + quality_seek + social_identity +
                   health_focus + brand_trust + guilt_prone,
    eco_group = cut(n_eco_traits, breaks = c(-1, 1, 3, 6),
                    labels = c("0\u20131 eco-responsive traits",
                               "2\u20133 eco-responsive traits",
                               "4\u20136 eco-responsive traits"))
  )

ggplot(df_p2, aes(x = condition, y = WTP, fill = condition, color = condition)) +
  geom_jitter(width = 0.14, alpha = 0.4, size = 2) +
  geom_boxplot(alpha = 0.20, outlier.shape = NA, width = 0.45, linewidth = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 4,
               fill = "white", color = "black", stroke = 1.3) +
  facet_wrap(~eco_group, nrow = 1) +
  scale_fill_manual(values  = c("Control" = "#4a90d9", "Eco-label" = "#2d6a4f")) +
  scale_color_manual(values = c("Control" = "#2b6cb0", "Eco-label" = "#1b4332")) +
  scale_y_continuous(labels = function(x) paste0("$", x)) +
  labs(x = NULL, y = "WTP ($)",
       title = "The Same Treatment Has Very Different Effects on Different People",
       subtitle = "Diamond = group mean  \u00b7  Eco-responsive traits: environmental values, quality-seeking, social identity, health, trust, guilt") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none", panel.grid.minor = element_blank(),
        strip.text = element_text(size = 9, face = "bold"))
```

The eco-label produces a large positive effect for participants with several eco-responsive traits, a small effect for moderately responsive participants, and near-zero (or even negative) effects for participants whose WTP is driven primarily by price sensitivity, habit, or hedonic taste expectations.

The average treatment effect you would report obscures all of this variation — and misses entirely *which* mechanism is driving WTP upward.

## What Participants Say: Template-Based Open-Ended Responses

To make the multi-signal nature of the treatment visible, we generate short open-ended responses from each simulated participant. These represent **(a)** what they infer about the product and **(b)** why they chose their WTP. We use template-based generation — each participant's active traits and condition determine which phrases they contribute.

> **Note:** We are using template-based text here to keep everything reproducible without an LLM API. In a real study you would collect actual open-ended responses. The patterns illustrated below — differential concept activation by condition, and concept–WTP correlations — would be recovered from those real responses using the same analysis pipeline.

```{r generate-text}
#| code-summary: "▶ Generate template-based open-ended text for each participant"

set.seed(2026)

# ── Product perception text ───────────────────────────────────────────────────
gen_product_text <- function(i) {
  cond <- condition_p2[i]
  ev   <- traits$env_values[i]
  qs   <- traits$quality_seek[i]
  ps   <- traits$price_sensitive[i]
  si   <- traits$social_identity[i]
  hf   <- traits$health_focus[i]
  bt   <- traits$brand_trust[i]
  gp   <- traits$guilt_prone[i]
  hed  <- traits$hedonic[i]
  da   <- traits$demand_aware[i]

  parts <- character(0)

  if (cond == "Control") {
    parts <- c(parts, sample(c(
      "This looks like a standard coffee blend.",
      "A fairly typical coffee package — nothing unusual.",
      "The packaging is clean and straightforward."
    ), 1))
    if (qs  == 1) parts <- c(parts, "The blend seems like decent, everyday quality.")
    if (ps  == 1) parts <- c(parts, "The price point appears reasonable and fair.")
    if (bt  == 1) parts <- c(parts, "The packaging gives a professional, trustworthy impression.")
    if (hed == 1) parts <- c(parts, "I am curious how the flavor compares to my usual brand.")
    if (length(parts) == 1)
      parts <- c(parts, "Nothing in particular stands out about this product.")
  } else {
    parts <- c(parts, "The eco-certification badge is immediately noticeable.")
    if (ev == 1) parts <- c(parts, "The sustainability label signals that this coffee is responsibly and environmentally sourced.")
    if (qs == 1) parts <- c(parts, "Eco-certified products suggest a higher standard of care in production, which I associate with better quality and taste.")
    if (ps == 1) parts <- c(parts, "Certifications usually carry a price premium, so I expect this to cost more than a standard option.")
    if (si == 1) parts <- c(parts, "This brand's values align with mine — I actively look for and support products like this.")
    if (hf == 1) parts <- c(parts, "Eco-certified means fewer pesticides and more natural farming practices, which I find very appealing.")
    if (bt == 1) parts <- c(parts, "A third-party certification badge means independent accountability — that increases my trust in the product significantly.")
    if (gp == 1) parts <- c(parts, "Buying a certified sustainable product lets me offset some of my everyday environmental footprint.")
    if (da == 1) parts <- c(parts, "The research context makes it clear this study is about eco-labeling, which makes me more conscious of how much I should value green products.")
    if (length(parts) == 1)
      parts <- c(parts, "The label catches my eye, though I am not sure how much weight I give eco-certifications in practice.")
  }

  paste(sample(parts, min(2L, length(parts))), collapse = " ")
}

# ── WTP justification text ────────────────────────────────────────────────────
gen_wtp_text <- function(i) {
  cond <- condition_p2[i]
  ev   <- traits$env_values[i]
  qs   <- traits$quality_seek[i]
  ps   <- traits$price_sensitive[i]
  si   <- traits$social_identity[i]
  hf   <- traits$health_focus[i]
  hed  <- traits$hedonic[i]
  hab  <- traits$habitual[i]
  da   <- traits$demand_aware[i]

  parts <- character(0)

  if (hab == 1 & ps == 1)
    parts <- c(parts, "I am fairly budget-conscious about coffee and set my limit based on what I can realistically afford each week.")
  if (qs == 1)
    parts <- c(parts, "I am willing to pay more when I believe the quality genuinely delivers on taste and enjoyment.")
  if (ev == 1 & cond == "Eco-label")
    parts <- c(parts, "I factored in the environmental benefits — supporting sustainable production is worth a modest premium to me.")
  if (si == 1 & cond == "Eco-label")
    parts <- c(parts, "I would pay more for a product that represents my values, even if cheaper alternatives are available.")
  if (hf == 1 & cond == "Eco-label")
    parts <- c(parts, "Cleaner and more natural sourcing justifies a higher price — I think of it as an investment in my health.")
  if (ps == 1)
    parts <- c(parts, "Price matters a great deal to me — I try not to overpay regardless of how appealing the product looks.")
  if (hab == 1)
    parts <- c(parts, "My spending on coffee tends to stay fairly consistent — habit guides my budget more than any single product feature.")
  if (hed == 1)
    parts <- c(parts, "If I expect genuinely exceptional taste, I am happy to stretch my usual budget.")
  if (da == 1 & cond == "Eco-label")
    parts <- c(parts, "Being in a study about eco products made me want to demonstrate my environmental awareness, which may have inflated my stated price slightly.")
  if (length(parts) == 0)
    parts <- "I set my price based on the overall value I expect — quality relative to cost."

  paste(sample(parts, min(2L, length(parts))), collapse = " ")
}

product_text <- sapply(seq_len(N_p2), gen_product_text)
wtp_text     <- sapply(seq_len(N_p2), gen_wtp_text)

# Preview a few responses
preview <- data.frame(
  Condition        = condition_p2[c(1, half2 + 1)],
  `Product text`   = product_text[c(1, half2 + 1)],
  `WTP text`       = wtp_text[c(1, half2 + 1)],
  check.names = FALSE
)
kable(preview, caption = "Example responses (one control, one eco-label participant)")
```

## Text Analysis: How Concept Composition Shifts Across Conditions

We now run a simple concept-count analysis — no API, no black-box model. Each concept is defined by a set of keywords; a participant "mentions" a concept if any keyword appears in their text.

```{r text-analysis}
#| code-summary: "▶ Count concept mentions and compare across conditions"

# ── Concept dictionary (regex patterns) ──────────────────────────────────────
concept_dict <- c(
  sustainability  = "sustainab|eco.certif|environmental|responsib",
  quality_taste   = "quality|taste|flavor|blend",
  price_cost      = "price|premium|afford|budget|overpay",
  social_values   = "values|align|represent|support products",
  health_natural  = "health|natural|pesticide|farming|cleaner",
  trust_cert      = "certif|accountab|trustworth",
  guilt_offset    = "footprint|guilt|offset",
  demand_effect   = "research context|study.*eco|should value|demonstrate.*environ|inflated.*stated|awareness.*study",
  hedonic         = "exceptional|happy to stretch|enjoyment",
  habit_routine   = "habit|consistent|weekly"
)

# Concept presence (1/0) per participant
prod_conc <- as.data.frame(
  lapply(concept_dict, function(pat) as.integer(grepl(pat, tolower(product_text)))))

wtp_conc <- as.data.frame(
  lapply(concept_dict, function(pat) as.integer(grepl(pat, tolower(wtp_text)))))

prod_conc$condition <- condition_p2
wtp_conc$condition  <- condition_p2
wtp_conc$WTP        <- WTP_p2

# ── Log-odds: how much more often is each concept in eco-label vs control? ───
log_odds <- function(x, cond, target = "Eco-label") {
  p_t <- (sum(x[cond == target])       + 0.5) / (sum(cond == target)       + 1)
  p_c <- (sum(x[cond != target])       + 0.5) / (sum(cond != target)       + 1)
  log(p_t / (1 - p_t)) - log(p_c / (1 - p_c))
}

lo_prod <- sapply(names(concept_dict), function(v)
  log_odds(prod_conc[[v]], prod_conc$condition))

lo_wtp  <- sapply(names(concept_dict), function(v)
  log_odds(wtp_conc[[v]],  wtp_conc$condition))

# ── Correlation of concept mentions with WTP ──────────────────────────────────
wtp_cor_prod <- sapply(names(concept_dict), function(v)
  cor(prod_conc[[v]], WTP_p2, method = "spearman"))

wtp_cor_wtp <- sapply(names(concept_dict), function(v) {
  x <- wtp_conc[[v]]
  # If concept never appears in WTP text, correlation is undefined → treat as 0
  if (sd(x) < 1e-9) 0 else cor(x, WTP_p2, method = "spearman")
})

# ── Summary table ─────────────────────────────────────────────────────────────
concept_summary <- data.frame(
  Concept      = names(concept_dict),
  pct_ctrl     = round(colMeans(prod_conc[prod_conc$condition == "Control",
                                          names(concept_dict)]) * 100, 0),
  pct_eco      = round(colMeans(prod_conc[prod_conc$condition == "Eco-label",
                                          names(concept_dict)]) * 100, 0),
  log_odds_col = round(lo_prod, 2),
  r_prod       = round(wtp_cor_prod, 2),
  r_wtp        = round(wtp_cor_wtp, 2)
)

kable(concept_summary,
      col.names = c("Concept", "Control (%)", "Eco-label (%)",
                    "Log-odds (product text)", "r with WTP (product)", "r with WTP (WTP text)"),
      caption = "Concept prevalence and associations by condition (product perception text)")
```

```{r concept-plot, fig.height=5.5, fig.width=10}
#| code-summary: "▶ Plot: log-odds ratios and WTP correlations"

lo_df <- data.frame(
  concept  = rep(names(concept_dict), 2),
  log_odds = c(lo_prod, lo_wtp),
  source   = rep(c("Product perception text", "WTP justification text"), each = length(concept_dict))
) |>
  mutate(concept = factor(concept, levels = names(concept_dict)[order(lo_prod)]))

ggplot(lo_df, aes(x = log_odds, y = concept, fill = source)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6, alpha = 0.85) +
  geom_vline(xintercept = 0, color = "gray30", linewidth = 0.8) +
  scale_fill_manual(values = c("Product perception text" = "#4a90d9",
                                "WTP justification text"  = "#2d6a4f"),
                    name = NULL) +
  labs(x = "Log-odds ratio (Eco-label vs. Control)",
       y = NULL,
       title = "How the Eco-Label Shifts the Concepts People Reach For",
       subtitle = "Positive = concept mentioned more in eco-label condition") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank())
```

The eco-label dramatically increases mentions of sustainability, quality, social values, health, and trust in the *product perception* text. The effect bleeds through — though more weakly — into the *WTP justification* text. Note that `habit_routine` and `hedonic` show no shift at all: they are driven by participant characteristics, not by the experimental manipulation.

## What's in X? What's in Y? {#sec-whats-in-xy}

::: {.callout-important}
## This is the most important figure in Section 2 — read it carefully

The diagram below is *not* just a summary of our coffee simulation. It represents the fundamental challenge of experimental interpretation.

**The left column (X)** shows every latent construct injected into the system by the eco-label manipulation. Only **one** of these — environmental responsibility — is the researcher's *account of interest* (the intended mechanism). Every other entry in the X column is a **confounder**: a pathway that the treatment activates alongside the intended one.

**The right column (Y)** shows every latent construct that causally drives WTP. Some overlap with X (shared constructs = blue lines). Some do not (orange boxes).

**The blue lines are the problem.** Each blue line is a latent pathway from X to Y that exists *in addition to* the one the researcher intends. When any of these additional pathways carry variance, the treatment coefficient in a regression is not "the effect of eco-labeling on environmental value". It is a conflation of all connected pathways, weighted by how strongly each latent construct in Y is activated by X.

**The researcher's responsibility** is to:

1. *Before* running the study: use the measurement tools from Module 1 (pilot text analysis, CFA, construct mapping) to enumerate the latent space of both X and Y. Every construct in that space that is common to both columns is a pathway you must account for — either by measuring it (so you can check balance) or by designing the manipulation to deactivate it.

2. *After* data collection: check balance on every measured common construct. Acknowledge that unmeasured common constructs remain a threat.

The question "What's in the X?" is the discriminant validity question asked about a *manipulation* rather than a *scale*. The tools are the same; the object of inquiry has shifted from the outcome variable to the treatment variable.
:::

The analysis above lets us ask the same question we posed in Module 1 about outcome measurement, but now about the treatment:

> **"What's in the X?" and "What's in the Y?" — and where do they overlap?**

The diagram below maps the constructs injected by the eco-label manipulation (left column) against the constructs that drive WTP (right column). **Blue lines are the confounds** — shared constructs that create additional latent pathways from X to Y beyond the intended one. Orange boxes mark constructs that are unique to one side.

```{r whats-in-xy, fig.height=8.5, fig.width=10}
#| code-summary: "▶ Plot: construct overlap diagram (What's in X? What's in Y?)"

x_topics <- c(
  "Environmental\nresponsibility\n[account of interest]",
  "Premium quality\nperception",
  "Price expectation\n(eco = premium)",
  "Social identity\n& values alignment",
  "Health &\nnaturalness",
  "Certification\ntrust",
  "Demand effects\n(context signals\n'eco is expected')",
  "Guilt\nreduction"
)

y_topics <- c(
  "Personal\nenvironmental values",
  "Perceived product\nquality",
  "Price sensitivity\n& budget",
  "Social identity",
  "Health concern",
  "Brand trust",
  "Social desirability\n& demand compliance",
  "Hedonic taste\nexpectations",
  "Habitual spending\npatterns"
)

# First 7 x-topics are shared with first 7 y-topics; last in each column are unique
nx <- length(x_topics)
ny <- length(y_topics)

plot_x <- data.frame(
  topic  = x_topics,
  xpos   = 0.15,
  ypos   = seq(nx, 1, by = -1L),
  shared = c(rep(TRUE, 7L), FALSE),
  stringsAsFactors = FALSE
)

plot_y <- data.frame(
  topic  = y_topics,
  xpos   = 0.85,
  ypos   = seq(nx, nx / ny, length.out = ny),
  shared = c(rep(TRUE, 7L), FALSE, FALSE),
  stringsAsFactors = FALSE
)

conn_df <- data.frame(
  x1 = 0.265, x2 = 0.735,
  y1 = plot_x$ypos[plot_x$shared],
  y2 = plot_y$ypos[plot_y$shared],
  is_demand = c(rep(FALSE, 6L), TRUE)   # flag the demand-effect link
)

ggplot() +
  # Non-demand links — standard blue
  geom_segment(data = conn_df[!conn_df$is_demand, ],
               aes(x = x1, xend = x2, y = y1, yend = y2),
               color = "#4a90d9", linewidth = 1.4, alpha = 0.55) +
  # Demand-effect link — highlighted red-orange
  geom_segment(data = conn_df[conn_df$is_demand, ],
               aes(x = x1, xend = x2, y = y1, yend = y2),
               color = "#d7301f", linewidth = 2.0, alpha = 0.80,
               arrow = arrow(ends = "both", length = unit(0.08, "inches"),
                             type = "open")) +
  geom_label(data = plot_x,
             aes(x = xpos, y = ypos, label = topic, fill = shared),
             size = 2.7, fontface = "bold", lineheight = 0.88,
             label.padding = unit(0.38, "lines"),
             label.r       = unit(0.15, "lines"), hjust = 0.5) +
  geom_label(data = plot_y,
             aes(x = xpos, y = ypos, label = topic, fill = shared),
             size = 2.7, fontface = "bold", lineheight = 0.88,
             label.padding = unit(0.38, "lines"),
             label.r       = unit(0.15, "lines"), hjust = 0.5) +
  # Column header boxes
  annotate("rect", xmin = 0.02, xmax = 0.28, ymin = 8.55, ymax = 9.20,
           fill = "#1e3a5f", color = NA) +
  annotate("text", x = 0.15, y = 8.88,
           label = "What's in X?\n(Eco-label Manipulation)",
           color = "white", fontface = "bold", size = 3.5, lineheight = 0.9) +
  annotate("rect", xmin = 0.72, xmax = 0.98, ymin = 8.55, ymax = 9.20,
           fill = "#2d6a4f", color = NA) +
  annotate("text", x = 0.85, y = 8.88,
           label = "What's in Y?\n(Willingness to Pay)",
           color = "white", fontface = "bold", size = 3.5, lineheight = 0.9) +
  # Demand-effect annotation
  annotate("label", x = 0.50, y = plot_x$ypos[7] + 0.30,
           label = "Demand effect confounder",
           fill = "#fff0ed", color = "#d7301f", fontface = "bold",
           size = 2.6, label.padding = unit(0.25, "lines")) +
  scale_fill_manual(
    values = c("TRUE" = "#d6ecff", "FALSE" = "#ffe0b2"),
    labels = c("Shared construct (confound)", "Unique to this variable"),
    name   = NULL
  ) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0.4, 9.5)) +
  labs(
    title    = "Construct Overlap Between Treatment (X) and Outcome (Y)",
    subtitle = paste0(
      "Blue lines = shared constructs (confounds)  \u00b7  Red line = demand effect  \u00b7  Orange boxes = unique to one side\n",
      "The account of interest is ONE pathway. Every other shared link is a confounder the researcher must account for."
    )
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5, size = 14,
                                 margin = margin(b = 6)),
    plot.subtitle = element_text(hjust = 0.5, color = "#555555", size = 10,
                                 lineheight = 1.3),
    legend.position = "bottom",
    plot.margin = margin(15, 15, 15, 15)
  )
```

::: {.callout-important}
## The parallel with Module 1: discriminant validity for manipulations

In Module 1, we said that a *scale* violates discriminant validity when it correlates too strongly with constructs it is not supposed to measure. The scale "measures too much."

Here, the *eco-label manipulation* violates an analogous principle: it injects multiple independent constructs simultaneously — not just "environmental responsibility" but also quality expectations, price priors, social identity signals, health inferences, certification trust, and guilt-offset motivation. The manipulation "does too much."

**Constructs unique to X (not in Y):** Guilt reduction. This signal is activated by the eco-label but does not translate into WTP variance in this simulation.

**Constructs unique to Y (not activated by X):** Hedonic taste expectations and habitual spending patterns. These explain variance in WTP in both conditions but are not triggered by the eco-label.

**The demand effect:** This is a particularly insidious shared pathway. The eco-label condition signals to demand-aware participants that the study is *about* eco-labeling — and they respond by inflating their stated WTP to comply with what they perceive the researcher expects. This is a confound that lives entirely in the experimental context itself: it is not driven by any genuine product attribute, but by the act of conducting the experiment. Notice that this pathway appears in *both* the treatment (X injects the cue) *and* the outcome (Y picks up the compliance). In the real world, researchers cannot remove this pathway by controlling the manipulation — they must measure it explicitly and either partial it out or interpret the treatment effect as "eco-label + demand context" rather than "eco-label alone."

**The practical implication:** Your treatment coefficient in a regression is a weighted average across all the shared constructs, confounded by whatever imbalance exists in the unshared ones. This is not a clean estimate of "the effect of eco-labeling" — it is the effect of a bundle of simultaneous signals, filtered through a population of heterogeneous responders.
:::

### How to Map the Construct Space in Practice: Nomological Net Scanning

The diagram above gives the theoretical picture — the full overlap between X and Y. In practice, you rarely have complete knowledge of this overlap at the start of a study. **Open-ended text responses** from a small pilot give you an empirical window into both sides of the diagram.

The approach: ask a small pilot sample (n ≈ 20–30) to write briefly about their response to the stimulus (for X) and about their reasoning for the outcome variable (for Y). Then use text analysis — either keyword-based or unsupervised — to extract which constructs are active.

**This is the nomological net scan.** It serves three purposes simultaneously:

1. **Enumeration:** it tells you *how many* latent pathways from X to Y exist — the practical answer to "What's in X? What's in Y?"
2. **Measurement guidance:** every concept the text reveals is a mediator you should measure quantitatively in the main study (using the scale construction tools from Module 1)
3. **Identification strategy:** once you have measured mediators, you can decompose the total treatment effect into its component channels via structural models (which we address in Part 3 and in detail in Module 3)

::: {.callout-important}
## The identification problem and its resolution

In Module 1 we showed that a scale conflates multiple latent constructs (discriminant invalidity) — and the fix was to *measure* those constructs separately and check that they can be discriminated.

Here, the treatment X activates multiple causal pathways to Y. The fix is identical: *measure* those pathways as mediators, then use a structural model to identify which channels carry the effect. Without measurement, you are left with an identification problem — no statistical model can tell you which pathway generated the observed X→Y association.

**The nomological net scan gives you the list. Quantitative measurement gives you the data. Structural modeling gives you the decomposition.**
:::

```{r nomological-net-scan}
#| code-summary: "▶ Use WTP text to scan for unmeasured Y dimensions (pilot)"

# Recall: we already have wtp_text and wtp_conc from earlier in Part 2.
# Simulate a "pre-study pilot" where only 30 participants write about their WTP.
# The goal: identify which concepts are active and should become measured mediators.

set.seed(999)
pilot_idx  <- sample(seq_len(N_p2), 30)
pilot_text <- wtp_text[pilot_idx]

# Concept frequency in the pilot
pilot_freq <- sapply(concept_dict, function(pat)
  mean(grepl(pat, tolower(pilot_text))))

pilot_freq_df <- data.frame(
  Concept    = names(pilot_freq),
  Prevalence = round(pilot_freq * 100, 0)
) |>
  arrange(desc(Prevalence)) |>
  mutate(
    include = ifelse(Prevalence >= 25, "\u2713 Yes (prevalent)",
                     "\u2715 No (rare in this sample)"),
    Concept = factor(Concept, levels = Concept)
  )

kable(pilot_freq_df,
      col.names = c("Y-dimension concept", "% pilot participants mentioning it",
                    "Recommendation"),
      caption   = "Pilot text scan: which sub-dimensions of WTP are active in this sample?")
```

```{r nom-net-plot, fig.height=4.5, fig.width=9}
#| code-summary: "▶ Plot: active Y-dimensions from pilot scan"

ggplot(pilot_freq_df, aes(x = Prevalence, y = Concept, fill = include)) +
  geom_col(alpha = 0.85, color = "white") +
  geom_vline(xintercept = 25, linetype = "dashed", color = "gray40", linewidth = 0.8) +
  annotate("text", x = 26, y = 1.0, label = "25% threshold",
           color = "gray40", hjust = 0, size = 3.2) +
  scale_fill_manual(values = c("\u2713 Yes (prevalent)" = "#2d6a4f",
                               "\u2715 No (rare in this sample)" = "#b0c4de"),
                    name = NULL) +
  labs(x = "% of pilot participants mentioning this concept",
       y = NULL,
       title = "Scanning the Nomological Net of WTP via Open-Ended Text",
       subtitle = "Prevalent concepts become measured mediators in the main study") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(), legend.position = "top")
```

### Alternative: LDA Topic Modeling (Unsupervised Discovery)

The keyword approach above requires you to specify patterns in advance. **Latent Dirichlet Allocation (LDA)** flips this: it automatically discovers clusters of co-occurring words across responses, with no keywords required. Each cluster is a *topic* — a statistical summary of what participants tended to write about together.

**The intuition:** Imagine sorting all WTP responses by word similarity. Responses mentioning "environment", "sustainable", "planet" cluster together naturally. Responses mentioning "organic", "healthy", "fresh" form a different cluster. LDA finds these clusters by asking: *what grouping of words best explains which responses sound similar?* The number of topics `k` is chosen by you — start with k ≈ number of expected sub-dimensions.

```{r lda-topic-model, fig.height=5, fig.width=10}
#| code-summary: "▶ LDA topic model: unsupervised discovery of Y-dimensions from text"

if (!requireNamespace("topicmodels", quietly = TRUE)) install.packages("topicmodels")
if (!requireNamespace("tidytext",    quietly = TRUE)) install.packages("tidytext")
library(topicmodels)
library(tidytext)

# ── Tokenize pilot text → document-term matrix ────────────────────────────────
pilot_tidy <- data.frame(
  doc_id = seq_along(pilot_text),
  text   = pilot_text
) |>
  unnest_tokens(word, text) |>
  anti_join(stop_words, by = "word") |>        # remove "the", "and", etc.
  filter(nchar(word) > 2) |>                   # drop 1-2 character tokens
  count(doc_id, word, sort = TRUE)

pilot_dtm <- pilot_tidy |>
  cast_dtm(doc_id, word, n)                    # convert to document-term matrix

# ── Fit LDA with k = 5 topics ─────────────────────────────────────────────────
# k = 5 is a reasonable start; in a real pilot you might try k = 3 to 8 and
# inspect coherence scores (e.g. via ldatuning package) to pick the best k.
set.seed(42)
lda_fit <- LDA(pilot_dtm, k = 5, control = list(seed = 42))

# ── Extract top words per topic ────────────────────────────────────────────────
top_words <- tidy(lda_fit, matrix = "beta") |>      # beta = P(word | topic)
  group_by(topic) |>
  slice_max(beta, n = 8, with_ties = FALSE) |>
  ungroup() |>
  mutate(topic_label = paste0("Topic ", topic))

# ── Assign human-readable labels (inspect top words to name each topic) ───────
# In a real analysis YOU read the top words and assign a substantive label.
# Here we auto-label by the single highest-probability word for illustration.
topic_labels <- top_words |>
  group_by(topic) |>
  slice_max(beta, n = 1) |>
  mutate(label = paste0("Topic ", topic, "\n(e.g. '", term, "')")) |>
  dplyr::select(topic, label)

top_words <- top_words |>
  left_join(topic_labels, by = "topic")

ggplot(top_words, aes(x = reorder_within(term, beta, label),
                       y = beta, fill = factor(topic))) +
  geom_col(show.legend = FALSE, alpha = 0.85) +
  scale_x_reordered() +
  facet_wrap(~ label, nrow = 1, scales = "free_y") +
  coord_flip() +
  scale_fill_manual(values = c("#08519c","#2d6a4f","#d7301f","#f4a261","#6a0572")) +
  labs(x = NULL, y = "P(word | topic)",
       title = "LDA Topic Model: Automatically Discovered Themes in WTP Text",
       subtitle = paste0("k = 5 topics  \u00b7  Each facet = one discovered theme  \u00b7  ",
                         "Top 8 words ordered by probability")) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        axis.text.x = element_text(size = 7),
        strip.text = element_text(face = "bold", size = 8))
```

::: {.callout-note}
## Keyword search vs. LDA: when to use each

| | Keyword search | LDA topic model |
|---|---|---|
| **Prior knowledge needed?** | Yes — specify patterns | No — fully unsupervised |
| **Best for** | Confirming expected concepts | Discovering unexpected ones |
| **Output** | % of responses per concept | Probabilistic word clusters |
| **Interpretability** | High (you chose the labels) | Requires human labeling |
| **Use in practice** | When you have a theory | Early pilot, no prior theory |

In practice, use **LDA first** to spot concepts you hadn't anticipated, then use **keyword patterns** to measure their prevalence precisely in the full sample.
:::

### From Text to Numbers: Measuring and Modeling the Nomological Net

The text analysis above identifies *which* constructs are active in the causal chain from X to Y. The next step is to **measure** those constructs quantitatively and **model** their pathways structurally. Together, these three stages — text scan → quantitative measurement → structural modeling — form the complete pipeline for dealing with exclusion restriction violations.

#### Step 2: Generate Quantitative Mediator Scores

The text scan identified which mediator constructs are active. Now we simulate what it would look like if we had *measured* those constructs quantitatively — as short scales administered alongside the main WTP item. Each score is generated from the underlying participant traits plus noise, reflecting what a properly validated scale would produce.

```{r quant-mediators}
#| code-summary: "▶ Step 2: Simulate quantitative scores for each text-identified mediator"

set.seed(999)

# ── Treatment assignment first — eco_latent is CAUSED by the condition ────────
# Seeing an eco-label activates eco-mindedness: it raises the salience of
# sustainability, health/naturalness, social values, and certification trust.
# The condition effect on all four eco indicators flows ENTIRELY through eco_latent;
# eco_arm has NO direct path to any indicator. This makes the DGP's causal
# structure exactly match the SEM's measurement model (eco_factor ~ condition).
eco_arm <- as.integer(condition_p2 == "Eco-label")

eco_latent    <- 0.80 * eco_arm + rnorm(N_p2, 0, 1.0)   # condition → eco-mindedness
frugal_latent <- rnorm(N_p2)                               # frugality: stable trait, unaffected by condition
social_latent <- rnorm(N_p2)                               # social proof: stable trait, unaffected by condition

# ── Correlated mediators: high within-cluster loading on eco_latent ──────────
# The 4 eco indicators share eco_latent as a common cause (loadings 0.84–0.90),
# producing pairwise correlations of r ≈ 0.65–0.93 among indicators in this DGP.
# This severe multicollinearity means parallel mediation's individual b-paths are
# unstable and their attenuated sum underestimates the true latent-factor indirect effect.

# ── Correlated ERRORS within indicator pairs ──────────────────────────────────
# Beyond eco_latent, two pairs share a response-style method factor (loading 0.55):
#   sustainability & health_natural: "natural/organic framing"
#   social_values & trust_cert: "social legitimacy/institutional approval"
# Both method factors are INDEPENDENT of eco_latent and eco_arm → pure residual covariance.
# SEM models both pairs via ~~ constraints; parallel mediation ignores them entirely.
method_sus_health <- rnorm(N_p2, 0, 1.0)   # natural/organic framing: sustainability ↔ health_natural
method_soc_trust  <- rnorm(N_p2, 0, 1.0)   # social legitimacy:       social_values  ↔ trust_cert

# ── Eco-cluster indicators: caused by eco_latent + method factors ONLY ────────
# eco_arm has NO direct effect on any indicator — its influence flows entirely
# through eco_latent. Consequences:
#   (a) The CFA (eco_factor =~ indicators) is the correct measurement model for this DGP.
#   (b) SEM's structural path (eco_factor ~ condition) recovers the true a-path
#       (condition → eco_latent), correctly weighting all 4 indicators and correcting
#       for measurement-error attenuation in each.
#   (c) Parallel mediation treats 4 highly correlated indicators (r ≈ 0.65–0.93) as
#       independent predictors → severe multicollinearity + attenuation → individual
#       b-paths unstable; summed eco-cluster indirect effect underestimates the truth.
#   (d) The method factors add within-pair residual correlations (loading 0.55) that
#       parallel mediation ignores; the SEM models them via ~~ constraints.
med_scores <- data.frame(
  sustainability  = 0.90 * eco_latent + 0.55 * method_sus_health + rnorm(N_p2, 3.5, 0.25),
  social_values   = 0.88 * eco_latent + 0.55 * method_soc_trust  + rnorm(N_p2, 3.0, 0.25),
  health_natural  = 0.86 * eco_latent + 0.55 * method_sus_health + rnorm(N_p2, 3.2, 0.25),
  trust_cert      = 0.84 * eco_latent + 0.55 * method_soc_trust  + rnorm(N_p2, 3.5, 0.25),
  # Frugal cluster: moderately reflective of frugal_latent
  price_cost      = 0.70 * frugal_latent - 0.15 * eco_arm + rnorm(N_p2, 3.0, 0.55),
  habit_routine   = 0.68 * frugal_latent                  + rnorm(N_p2, 3.0, 0.55),
  # Social/demand cluster: moderately reflective of social_latent
  demand_effect   = 0.70 * social_latent + 0.45 * eco_arm + rnorm(N_p2, 2.0, 0.55),
  guilt_offset    = 0.68 * social_latent + 0.30 * eco_arm + rnorm(N_p2, 2.5, 0.55),
  # Standalone mediators
  quality_taste   = 0.40 * eco_latent   + 0.35 * eco_arm + rnorm(N_p2, 3.5, 0.60),
  hedonic         = 0.20 * eco_latent   + 0.15 * eco_arm + rnorm(N_p2, 3.8, 0.65)
)

# Standardize mediators to z-scores
med_z <- as.data.frame(scale(med_scores))
med_z$condition <- eco_arm

# ── WTP: depends on LATENT FACTORS, not individual indicators ────────────────
# This is the key DGP design: WTP reflects eco_latent (the shared construct
# behind the 4 correlated eco items), NOT each indicator separately.
# Parallel mediation is doubly misspecified: (1) treats 4 highly correlated
# indicators (r ≈ 0.65–0.93) as independent predictors → multicollinearity;
# (2) uses noisy proxies for eco_latent → attenuation bias in the b-paths.
# SEM is correctly specified — it models eco_latent explicitly via CFA,
# corrects for attenuation, and models the residual covariances via ~~.
WTP_med <- 5.0 +
  0.50 * eco_arm       +   # direct condition effect
  1.20 * eco_latent    +   # main indirect path (eco-mindedness → WTP)
 -0.55 * frugal_latent +   # frugality lowers WTP
  0.50 * social_latent +   # social proof boosts WTP
  0.40 * (med_scores$quality_taste - mean(med_scores$quality_taste)) /
          sd(med_scores$quality_taste) +
  0.15 * (med_scores$hedonic - mean(med_scores$hedonic)) /
          sd(med_scores$hedonic) +
  rnorm(N_p2, 0, 1.5)
med_z$WTP <- scale(WTP_med)[, 1]

cat("Mediator score means by condition (standardized; should differ for eco-activated dims):\n")
med_scores |>
  mutate(condition = condition_p2) |>
  group_by(condition) |>
  summarise(across(everything(), ~ round(mean(.), 2)), .groups = "drop") |>
  kable(caption = "Mean mediator scores by condition (raw scale)")
```

#### Step 3: Parallel Mediation Model

A parallel mediation model regresses each mediator on treatment (the *a* paths) and regresses WTP on treatment + all mediators simultaneously (the *b* paths + direct effect *c'*). The indirect effects (a × b) decompose the total X→Y effect into specific causal channels.

This model assumes the mediators are **independent** of each other (no correlated residuals). This is the standard Hayes PROCESS approach. **We will address mediation models in much greater detail in Module 3**, where we formalize the causal assumptions required for valid mediation inference and introduce more flexible structural equation modeling approaches. Here, we use parallel mediation as a first-pass decomposition tool to illustrate the concept — and to show where it breaks down.

```{r parallel-mediation}
#| code-summary: "▶ Step 3: Parallel mediation via lavaan (fits model; no standalone plot)"

med_names <- c("sustainability", "quality_taste", "price_cost", "social_values",
               "health_natural", "trust_cert", "demand_effect", "guilt_offset",
               "hedonic", "habit_routine")

# Build lavaan model string dynamically
a_paths <- paste0(med_names, " ~ a", seq_along(med_names), " * condition",
                  collapse = "\n  ")
b_paths <- paste0("WTP ~ ",
                  paste0("b", seq_along(med_names), " * ", med_names, collapse = " + "),
                  " + cp * condition")
ind_defs <- paste0("ind", seq_along(med_names), " := a", seq_along(med_names),
                   " * b", seq_along(med_names), collapse = "\n  ")
total_ind <- paste0("total_indirect := ",
                    paste0("ind", seq_along(med_names), collapse = " + "))

parallel_model <- paste("# a paths\n ", a_paths,
                        "\n# outcome\n ", b_paths,
                        "\n# indirect effects\n ", ind_defs,
                        "\n ", total_ind)

fit_parallel <- sem(parallel_model, data = med_z, se = "bootstrap", bootstrap = 200)

# Extract indirect effect estimates
param_par <- parameterEstimates(fit_parallel, boot.ci.type = "bca.simple") |>
  filter(op == ":=") |>
  mutate(
    label_clean = c(paste0("via ", gsub("_", " ", med_names)), "Total indirect"),
    sig = ifelse(pvalue < 0.05, "*", "ns"),
    model = "Parallel mediation"
  )
cat("Parallel mediation fitted. Total indirect:",
    round(param_par$est[param_par$label_clean == "Total indirect"], 3), "\n")
```

#### Step 4: SEM with Latent Factors

**Why parallel mediation underestimates the eco-cluster effect:** The eco-label activates a *cluster* of correlated constructs — sustainability, social values, health/natural, and trust in certification — all driven by the same underlying eco-mindedness latent factor. Regressing WTP on all 10 mediators simultaneously creates **multicollinearity** among those four; the regression splits their shared variance arbitrarily across four rows. Each individual eco-cluster pathway appears small and noisy.

**The SEM solution:** instead of treating each indicator as an independent mediator, model the shared variance explicitly as a *latent factor* using a CFA measurement model. The latent `eco_factor` extracts the common variance from its four indicators, accounts for measurement error in each, and serves as a single, well-identified mediator in the structural part of the model. This is precisely the link to Module 1: the CFA measurement model IS a discriminant/convergent validity check applied directly to the mediator set.

```{r sem-latent, fig.height=5.5, fig.width=9}
#| code-summary: "▶ Step 4: SEM with latent eco-factor (CFA + structural mediation)"

# ── Build cluster data: standardize cluster means so all effects are comparable ─
# eco_cluster, frugal_cluster, social_cluster = means of z-scored items → already
# near-zero mean, but SD ≈ 0.5–0.8 (smaller than individual items because averaging
# reduces noise).  We scale them to SD = 1 so oracle, parallel, and SEM indirect
# effects are all in standardised-WTP-per-unit-condition units.
cl_names <- c("eco_cluster","frugal_cluster","social_cluster","quality_taste","hedonic")

eco_raw    <- rowMeans(med_z[, c("sustainability","social_values","health_natural","trust_cert")])
frugal_raw <- rowMeans(med_z[, c("price_cost","habit_routine")])
social_raw <- rowMeans(med_z[, c("demand_effect","guilt_offset")])

med_cl <- data.frame(
  eco_cluster    = scale(eco_raw)[,1],      # standardised cluster means
  frugal_cluster = scale(frugal_raw)[,1],
  social_cluster = scale(social_raw)[,1],
  quality_taste  = med_z$quality_taste,     # already z-scored
  hedonic        = med_z$hedonic,
  condition      = med_z$condition,
  WTP            = med_z$WTP
)

# Extend with raw eco indicators for the CFA measurement model
# (indicators are already z-scored from med_z)
med_sem <- cbind(
  med_cl,
  sustainability = med_z$sustainability,
  social_values  = med_z$social_values,
  health_natural = med_z$health_natural,
  trust_cert     = med_z$trust_cert
)

# ── Latent SEM model ──────────────────────────────────────────────────────────
# Measurement model (CFA): eco_factor is a latent variable measured by 4 indicators.
# std.lv = TRUE standardises the latent variable to variance = 1, putting the SEM's
# eco_factor on the same scale as eco_latent_std in the oracle (both SD = 1).
sem_model_txt <- '
  # Measurement model: eco_factor = latent eco-mindedness from 4 z-scored indicators
  eco_factor =~ sustainability + social_values + health_natural + trust_cert

  # Residual covariances: pairs of indicators share method variance in the DGP
  # (natural/organic framing for sus↔health_natural; social legitimacy for soc↔trust_cert).
  # These parameters are IN the DGP but completely ignored by parallel mediation.
  sustainability ~~ health_natural
  social_values  ~~ trust_cert

  # a-paths: condition predicts each mediator / latent factor
  eco_factor     ~ a1 * condition
  frugal_cluster ~ a2 * condition
  social_cluster ~ a3 * condition
  quality_taste  ~ a4 * condition
  hedonic        ~ a5 * condition

  # b-paths + direct effect on standardised WTP
  WTP ~ b1 * eco_factor + b2 * frugal_cluster + b3 * social_cluster +
        b4 * quality_taste + b5 * hedonic + cp * condition

  # Indirect effects (all in standardised WTP units)
  ind_eco     := a1 * b1
  ind_frugal  := a2 * b2
  ind_social  := a3 * b3
  ind_quality := a4 * b4
  ind_hedonic := a5 * b5
  total_ind   := ind_eco + ind_frugal + ind_social + ind_quality + ind_hedonic
'

# std.lv = TRUE: latent eco_factor standardised to variance 1 → comparable scale
fit_sem <- sem(sem_model_txt, data = med_sem,
               std.lv = TRUE, se = "bootstrap", bootstrap = 200)

param_sem <- parameterEstimates(fit_sem, boot.ci.type = "bca.simple") |>
  filter(op == ":=") |>
  mutate(
    label_clean = c("via eco cluster", "via frugal cluster", "via social cluster",
                    "via quality taste", "via hedonic", "Total indirect"),
    model = "SEM (latent eco-factor)"
  )

cat("SEM fitted. Total indirect:",
    round(param_sem$est[param_sem$label_clean == "Total indirect"], 3), "\n")

# ── Aggregate parallel mediation within clusters for comparison ───────────────
# Sum individual a×b products within each theoretical cluster.
# Also sum CI bounds (approximate: assumes within-cluster paths are independent).
par_agg <- param_par |>
  filter(label_clean != "Total indirect") |>
  mutate(cluster = case_when(
    label_clean %in% c("via sustainability","via social values",
                       "via health natural","via trust cert") ~ "via eco cluster",
    label_clean %in% c("via price cost","via habit routine")  ~ "via frugal cluster",
    label_clean %in% c("via demand effect","via guilt offset") ~ "via social cluster",
    TRUE ~ label_clean
  )) |>
  group_by(cluster) |>
  summarise(
    est      = sum(est),
    ci.lower = sum(ci.lower),
    ci.upper = sum(ci.upper),
    .groups  = "drop"
  ) |>
  rename(label_clean = cluster) |>
  mutate(model = "Parallel (summed within cluster)")

cat("SEM and parallel mediation estimates computed. Proceeding to DGP recovery comparison.\n")
```

#### Step 5: How Well Does Each Approach Recover the True DGP?

The simulation tells us exactly how WTP was generated (from the latent factors). We can compare what each modelling strategy estimates against that known ground truth.

```{r dgp-recovery, fig.height=5.5, fig.width=12}
#| code-summary: "▶ Step 5: Compare recovered effects to true DGP"

# ── Oracle true indirect effects using the LATENT factors from the DGP ────────
# The TRUE causal chain runs through eco_latent, frugal_latent, and social_latent —
# not through their noisy observed composites.  Using the latent factors directly
# (a) eliminates attenuation bias from measurement error and (b) reflects the actual
# DGP coefficients.  A well-specified SEM should recover these values; parallel
# mediation, which enters noisy observed indicators, cannot.

wtp_std_vec <- scale(WTP_med)[, 1]   # standardised WTP from the DGP

# Standardise latent factors to SD = 1 (matches SEM std.lv = TRUE scale)
eco_latent_std    <- scale(eco_latent)[, 1]
frugal_latent_std <- scale(frugal_latent)[, 1]
social_latent_std <- scale(social_latent)[, 1]
quality_std       <- med_z$quality_taste        # already z-scored
hedonic_std_vec   <- med_z$hedonic              # already z-scored

# True a-paths: OLS of each latent factor on condition
true_a_eco     <- coef(lm(eco_latent_std ~ eco_arm))["eco_arm"]    # ≈ +0.74
true_a_frugal  <- coef(lm(frugal_latent_std ~ eco_arm))["eco_arm"] # ≈  0
true_a_social  <- coef(lm(social_latent_std ~ eco_arm))["eco_arm"] # ≈  0
true_a_quality <- coef(lm(quality_std ~ eco_arm))["eco_arm"]
true_a_hedonic <- coef(lm(hedonic_std_vec ~ eco_arm))["eco_arm"]

# True b-paths: partial OLS of WTP_std on all latent factors simultaneously
# (avoids attenuation bias from using noisy observed composites as proxies)
oracle_true_lm <- lm(wtp_std_vec ~ eco_latent_std + frugal_latent_std +
                       social_latent_std + quality_std + hedonic_std_vec)
true_b_eco     <- coef(oracle_true_lm)["eco_latent_std"]
true_b_frugal  <- coef(oracle_true_lm)["frugal_latent_std"]
true_b_social  <- coef(oracle_true_lm)["social_latent_std"]
true_b_quality <- coef(oracle_true_lm)["quality_std"]
true_b_hedonic <- coef(oracle_true_lm)["hedonic_std_vec"]

# True indirect effects (latent-factor oracle; no CIs — these ARE the DGP truth)
true_ind_oracle <- setNames(
  c(true_a_eco * true_b_eco,
    true_a_frugal * true_b_frugal,
    true_a_social * true_b_social,
    true_a_quality * true_b_quality,
    true_a_hedonic * true_b_hedonic),
  cl_names
)

cat("Oracle (latent-factor) indirect effects:\n")
print(round(true_ind_oracle, 3))

# ── Build comparison data frame with 95% bootstrap CIs ───────────────────────
# Oracle: point estimate only, no CI (it IS the DGP truth for this sample)
# par_agg and param_sem carry ci.lower / ci.upper from the bootstrap runs in Step 3/4
oracle_rows <- data.frame(
  label_clean = paste0("via ", gsub("_", " ", cl_names)),
  effect      = round(true_ind_oracle, 3),
  ci_lo       = NA_real_,
  ci_hi       = NA_real_,
  source      = "True DGP (oracle)"
)

par_rows <- par_agg |>
  filter(label_clean != "Total indirect") |>
  transmute(label_clean,
            effect = est, ci_lo = ci.lower, ci_hi = ci.upper,
            source = "Parallel (summed within cluster)")

sem_rows <- param_sem |>
  filter(label_clean != "Total indirect") |>
  transmute(label_clean,
            effect = est, ci_lo = ci.lower, ci_hi = ci.upper,
            source = "SEM (latent eco-factor)")

recovery_df <- bind_rows(oracle_rows, par_rows, sem_rows) |>
  mutate(source = factor(source,
                          levels = c("True DGP (oracle)",
                                     "Parallel (summed within cluster)",
                                     "SEM (latent eco-factor)")))

# Order y-axis by oracle effect size
oracle_order <- oracle_rows |> arrange(effect) |> pull(label_clean)
recovery_df  <- recovery_df |>
  mutate(label_clean = factor(label_clean, levels = oracle_order))

ggplot(recovery_df, aes(x = effect, y = label_clean,
                         color = source, shape = source)) +
  geom_vline(xintercept = 0, color = "gray40", linewidth = 0.7) +
  geom_errorbarh(aes(xmin = ci_lo, xmax = ci_hi),
                 position = position_dodge(width = 0.65),
                 height = 0.22, linewidth = 0.6, na.rm = TRUE) +
  geom_point(position = position_dodge(width = 0.65), size = 3.5) +
  scale_color_manual(values = c("True DGP (oracle)"                = "#d7301f",
                                 "Parallel (summed within cluster)" = "#4a90d9",
                                 "SEM (latent eco-factor)"          = "#2d6a4f"),
                     name = NULL) +
  scale_shape_manual(values = c("True DGP (oracle)"                = 18,
                                 "Parallel (summed within cluster)" = 16,
                                 "SEM (latent eco-factor)"          = 15),
                     name = NULL) +
  labs(x = "Indirect effect (standardized)",
       y = NULL,
       title = "Model Recovery vs. True DGP",
       subtitle = paste0("Red diamonds = oracle (latent-factor truth, no CI)  \u00b7  ",
                         "Error bars = 95% bootstrap CI\n",
                         "SEM (green) corrects for attenuation + residual covariances; ",
                         "parallel mediation (blue) attenuated by multicollinearity")) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), legend.position = "top")
```

::: {.callout-note}
## What the recovery plot shows

**The SEM (green) closely tracks the true oracle values (red), while parallel mediation (blue) systematically underestimates the eco-cluster pathway.** The oracle is computed from the true latent factors (eco_latent, frugal_latent, social_latent) used to generate the data — the unattenuated causal quantities in the DGP. Parallel mediation deviates from this truth for two compounding reasons:

1. **Attenuation bias.** Each eco indicator is a noisy proxy for eco_latent (loading ≈ 0.84–0.90 plus a method factor, plus idiosyncratic noise). Entering noisy proxies in a regression shrinks the b-path estimates toward zero — the classic errors-in-variables attenuation. Even when the four attenuated b-paths are summed across the eco cluster, the resulting indirect effect underestimates the true latent-factor effect.

2. **Multicollinearity.** The four eco indicators are highly intercorrelated (r ≈ 0.65–0.93) because they share both eco_latent and a response-style method factor. Entering all four simultaneously in the WTP regression creates severe multicollinearity (VIF >> 5), making individual b-path estimates unstable and inflating their variance — compounding the attenuation problem.

By contrast, the SEM explicitly defines a latent `eco_factor` via a CFA measurement model. The CFA correctly partitions eco_latent signal from measurement noise and method-factor variance, yielding an **unattenuated** b-path estimate from eco_factor to WTP. The residual covariance constraints (`sustainability ~~ health_natural`; `social_values ~~ trust_cert`) further correct for the within-pair shared method variance that parallel mediation ignores. The result is a structural path estimate that closely tracks the true latent-factor oracle.

Neither model is perfect. The key insight is that **any structural model is only as good as the list of mediators you feed into it**. The text analysis gave us that list. Without the text analysis, you would have entered the study with an incomplete nomological map — and the "treatment effect" coefficient would have remained an uninterpretable mixture of all active pathways.

This is the direct parallel to Module 1's discriminant validity problem: just as measuring only some sub-dimensions of Y produces a biased estimate of your latent construct, measuring only some mediators of the X→Y relationship produces a biased estimate of the causal effect. In both cases, the solution is the same: map the latent space carefully, measure what you can, and report the gap transparently.
:::

**5. Two Further Threats to Mediation Inference — and Their Remedies**

The parallel mediation and SEM above address *which* mediators to include and *how* to model them. Two additional threats remain regardless of model specification: **non-normal residuals / influential outliers** (which can corrupt OLS *b*-path estimates) and **unmeasured M–Y confounders** which we will discuss in Module 3.

---

::: {.callout-note}
## Tying it all together: the Module 1–2 thread

Every concept in this tutorial maps directly onto a concept from Module 1:

| Module 1 (Measurement) | Module 2 (Experimentation and Identification) |
|---|---|
| Measured Observables | Treatment conditions |
| Latent construct (e.g., Green Purchase Intention) | Latent property of randomness / independence |
| Discriminant validity: does your scale measure *only* what it should? | Exclusion restriction: does your manipulation activate *only* the pathway it should? |
| Convergent validity: do items for the same construct correlate? | Randomization check: do characteristics correlate with treatment assignment? |
| Measurement invariance: does the scale work the same across groups? | Heterogeneous treatment effects: does the manipulation work the same across participant types? |
| Latent classes: are there sub-populations with different response patterns? | Effect heterogeneity: are there sub-populations for whom the treatment operates through different mechanisms? |
| Low signal-to-noise → poor reliability | Broad, complex Y → high N required for true independence |

**The common lesson:** observables (scale scores, group assignments, p-values) are imperfect proxies for latent quantities (true constructs, true independence, true effect sizes). Researchers who confuse the observable with the latent — who treat "my scale is published" as "my measurement is valid," or "Qualtrics randomized" as "my groups are balanced" — are making the same type of error, just in different places in the research pipeline. The path forward in both cases is the same: map the latent space carefully, measure what you can, acknowledge what you cannot, and report the gap transparently.
:::

---

# Part 3: What Does "Random Assignment" Actually Require? {#sec-randomization}

::: {.callout-warning}
## The central message of Part 3

**Never assume that random assignment has achieved actual randomization.** Random assignment is a *procedure* — clicking "Randomize" in Qualtrics, running `sample()` in R. Actual randomization is a *property* — treatment is orthogonal to all potential confounders in the data-generating process. The procedure does not guarantee the property. Whether it does depends on (1) how complex your outcome variable Y is and (2) how many participants you have. This section develops the gap between the two and equips you to evaluate — not assume — whether randomization succeeded.
:::

## The Measurement Analogy: Observable vs. Latent

In Module 1, we introduced a distinction that runs through all of empirical research:

| | Example from Module 1 |
|---|---|
| **Observable (indicator)** | A participant's response on a 7-point Likert scale |
| **Latent construct** | Their true level of Green Purchase Intention |

The observable is what you can see. The latent construct is what you actually want to know. The entire Module 1 was about the gap between these two things: your observed scale picks up variance from the wrong latent constructs (discriminant validity), from unequal measurement properties across groups (non-invariance), and from subpopulations that behave differently (latent classes).

**Exactly the same distinction applies to randomization:**

| | In Randomization |
|---|---|
| **Observable (indicator)** | Pressing "Randomize" in Qualtrics, running `sample()` in R, using a random number table |
| **Latent construct** | The property of *true independence*: treatment assignment is orthogonal to ALL potential confounders |

Just as a 7-point scale can *look* like it measures Green Purchase Intention while actually reflecting Environmental Concern, a randomization *procedure* can look random (you clicked the button — the observable is there) while failing to achieve true independence (the latent property).

::: {.callout-important}
## The construct validity question for randomization

Ask yourself the question from Module 1, applied to your treatment variable:

> **"Does your observable act of randomization have construct validity as an indicator of latent randomization?"**

The answer depends entirely on **how complex the outcome variable Y is that you are trying to study.** We will now build the intuition for why.
:::

## A Concrete Anchor: The Playing Cards Analogy

Imagine you have a sorted deck of cards — you know exactly what order the cards are in. Before a game, you need to make the deck random. So you shuffle it once.

**Did one shuffle make the deck random?**

The answer depends on the size of the deck.

- **5 cards:** There are 5! = 120 possible orderings. One riffle shuffle covers a substantial fraction of these. The deck is close to random.
- **52 cards:** There are 52! ≈ 8 × 10⁶⁷ possible orderings. One riffle shuffle covers an astronomically small fraction of these. Significant structure from the original sorted order survives.

This is why casino dealers shuffle 7 times — a mathematical result (Bayer & Diaconis, 1992) showing that 7 GSR riffle shuffles are sufficient for a 52-card deck to approach uniformity. One shuffle is not enough.

We can visualize this directly. We simulate one-through-ten riffle shuffles for decks of different sizes and measure how much structure (correlation with the original sorted order) survives.

```{r cards-shuffle-sim, fig.height=5.5}
#| code-summary: "▶ Simulate riffle shuffles for different deck sizes"

one_riffle <- function(deck) {
  n <- length(deck)
  k <- rbinom(1, n, 0.5)
  if (k == 0L || k == n) return(deck[sample(n)])
  pile_a <- deck[1:k]; pile_b <- deck[(k + 1):n]
  result <- integer(n); idx <- 1L
  while (length(pile_a) > 0L && length(pile_b) > 0L) {
    p_a <- length(pile_a) / (length(pile_a) + length(pile_b))
    if (runif(1) < p_a) { result[idx] <- pile_a[1]; pile_a <- pile_a[-1]
    } else               { result[idx] <- pile_b[1]; pile_b <- pile_b[-1] }
    idx <- idx + 1L
  }
  if (length(pile_a) > 0L) result[idx:n] <- pile_a
  if (length(pile_b) > 0L) result[idx:n] <- pile_b
  result
}

set.seed(42)
shuffle_sim <- expand.grid(deck_size = c(5L, 10L, 20L, 52L),
                           n_shuffles = 1:10, sim = 1:400) |>
  dplyr::mutate(tau = mapply(function(n, k) {
    deck <- 1:n
    for (i in seq_len(k)) deck <- one_riffle(deck)
    cor(1:n, deck, method = "kendall")
  }, deck_size, n_shuffles))

shuffle_summary <- shuffle_sim |>
  group_by(deck_size, n_shuffles) |>
  summarise(mean_tau = mean(tau), lo = quantile(tau, 0.10),
            hi = quantile(tau, 0.90), .groups = "drop") |>
  mutate(deck_label = paste0(deck_size, " cards"))

ggplot(shuffle_summary,
       aes(x = n_shuffles, y = mean_tau,
           color = factor(deck_size), fill = factor(deck_size), group = deck_size)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.14, color = NA) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40", linewidth = 0.8) +
  annotate("text", x = 10.2, y = 0.03, label = "Random\n(\u03c4 \u2248 0)",
           size = 3.5, color = "gray40", hjust = 0) +
  scale_x_continuous(breaks = 1:10) +
  scale_color_manual(values = c("5"  = "#4a90d9", "10" = "#52b788",
                                "20" = "#f4a261", "52" = "#e63946"),
                     name = "Deck size") +
  scale_fill_manual(values  = c("5"  = "#4a90d9", "10" = "#52b788",
                                "20" = "#f4a261", "52" = "#e63946"),
                    name = "Deck size") +
  labs(x = "Number of riffle shuffles",
       y = "Kendall's \u03c4 with original sorted order\n(1 = perfectly sorted  \u00b7  0 = random)",
       title = "How Many Shuffles Are Needed to Achieve Randomness?",
       subtitle = "Ribbon = 10th\u201390th percentile across 400 simulations per deck size") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

The analogy to experiments maps directly:

| Playing Cards | Research Experiment |
|---|---|
| Cards in the deck | Participants in the study |
| Order before shuffling | Values on potential confounders |
| One shuffle | Clicking "Randomize" once in Qualtrics |
| "Truly random" deck | Treatment and control groups balanced on ALL confounders |
| **Size of the deck** | **Complexity (number of dimensions) of the outcome Y** |

**The key takeaway:** just as a larger deck needs more shuffles to achieve randomness, a more complex outcome variable Y needs more observations to achieve true independence between treatment and all potential confounders.

## The Exchangeability Assumption: When It Holds and When It Doesn't

The permutation logic in Part 1 rested on **exchangeability**: the assignment of participants to conditions could have been any other permutation without changing the distributional properties of the data.

Exchangeability holds when treatment assignment is truly independent of all potential causes of Y. In practice, researchers **assume** this is achieved by the act of randomization — but you should **never** make this assumption without empirical verification. Consider the gap:

- **What the researcher observes:** Qualtrics assigned participants to conditions using a random number generator. ✓
- **What the researcher needs (latent):** Treatment assignment is uncorrelated with *all* potential causal antecedents of Y — every dimension, every subdimension, every latent mediator in the entire data-generating process. ?

Whether the observable achieves the latent property depends on two things:
1. How many antecedents Y has (the complexity of the construct)
2. How many participants you have (the sample size)

---

## Case 1 — A Narrow Construct: Visual Fixation Time

Suppose your outcome variable Y is **total gaze duration (in milliseconds) on an eco-label during a product display**, measured via eye-tracking. This is a narrow, physiological measure driven by a small number of simple perceptual factors.

Think of **dimensions** as all the hidden moderators in the data-generating process — every property of the participant or their environment that makes them more or less likely to fixate on the label, regardless of the experimental manipulation. These are the variables that, if correlated with treatment assignment by chance, would confound your estimate of the treatment effect. In the entire data-generating process for visual fixation, suppose there are only **2 such binary dimensions**:

| Dimension | Low (0) | High (1) |
|---|---|---|
| **Screen luminance** | Dim display | Bright display |
| **Viewing distance** | Far from screen | Close to screen |

These are simple physical factors. They are not set by the researcher and could vary unsystematically across participants assigned to each condition.

Because both dimensions are binary, the complete "design space" of all possible hidden-moderator combinations is a **2 × 2 matrix** of 4 cells:

```{r narrow-matrix, fig.height=4.5, fig.width=7}
#| code-summary: "▶ Visualize: 2×2 cell matrix for narrow construct"

cells <- expand.grid(
  Luminance = c("Dim display\n(Low)", "Bright display\n(High)"),
  Distance  = c("Far from screen\n(Low)", "Close to screen\n(High)")
)
cells$label <- c("Cell A", "Cell B", "Cell C", "Cell D")

ggplot(cells, aes(x = Luminance, y = Distance)) +
  geom_tile(fill = "#d6ecff", color = "#4a90d9", linewidth = 1.2,
            width = 0.92, height = 0.92) +
  geom_text(aes(label = label), size = 5.5, fontface = "bold", color = "#1a3a6b") +
  geom_text(aes(label = "n \u2265 1"), size = 3.8, color = "#555", vjust = 3.2) +
  labs(title = "Design Space for Visual Fixation (Narrow Construct)",
       subtitle = "2 binary hidden moderators \u2192 2\u00b2 = 4 cells  \u00b7  Minimum N to populate matrix = 4",
       x = "Screen Luminance", y = "Viewing Distance") +
  theme_minimal(base_size = 13) +
  theme(panel.grid = element_blank(), axis.text = element_text(size = 10))
```

::: {.callout-note}
## Why 4 cells?

With **k binary dimensions** in the data-generating process, the total number of unique hidden-moderator combinations is **2^k**. Each cell represents one possible "type" of participant in terms of everything that drives Y. For randomization to succeed, treatment and control groups must be approximately balanced across all 2^k cells — the smaller k is, the easier this is to achieve.

For k = 2: **2² = 4 cells**. A minimal experiment of N = 4 would cover the space; N = 40 gives 10 per cell with very reliable balance.
:::

With only 4 meaningful combinations in the design space, even a modest experiment (e.g., N = 40, 10 per cell) will very likely achieve approximate balance between treatment and control. One act of randomization is almost certainly sufficient.

---

## Case 2 — A Broad Construct: Sustainable Purchase Intention

Now suppose your outcome Y is **sustainable product purchase intention** — the kind of broad, latent, multi-faceted attitude that appears throughout consumer and organizational research, and that you saw in Module 1 struggling to maintain discriminant validity across measurement attempts.

In the data-generating process, suppose there are **10 binary determinants** of this construct:

```{r broad-construct-table}
#| code-summary: "▶ Table: 10 binary dimensions of sustainable purchase intention"

dims_table <- data.frame(
  `#`       = 1:10,
  Dimension = c("Environmental values", "Household income", "Social norm salience",
                "Brand trust", "Perceived product quality", "Health motivation",
                "Peer influence", "Prior green behavior",
                "Price sensitivity", "Emotional response to sustainability messaging"),
  Low       = c("Weak", "Low income", "Low salience", "Low trust",
                "Low quality perception", "Low", "Weak influence",
                "Little prior behavior", "High sensitivity", "Low emotional response"),
  High      = c("Strong", "High income", "High salience", "High trust",
                "High quality perception", "High", "Strong influence",
                "Extensive prior behavior", "Low sensitivity", "High emotional response"),
  check.names = FALSE
)
kable(dims_table,
      col.names = c("#", "Dimension", "Level: Low (0)", "Level: High (1)"),
      caption   = "10 binary antecedents of Sustainable Purchase Intention")
```

With **k = 10 binary dimensions**, the total number of unique participant profiles in the design space is **2^10 = 1,024 cells**. Notice how quickly this grows compared to the narrow construct:

```{r combinatorial-explosion, fig.height=5}
#| code-summary: "▶ Plot: combinatorial growth of design space"

expl_df <- data.frame(k = 1:15, cells = 2^(1:15)) |>
  mutate(label = ifelse(k == 2,  "Recall\nAccuracy\n(k = 2)",
                 ifelse(k == 10, "Sustainable\nPurchase\nIntention\n(k = 10)", NA)))

ggplot(expl_df, aes(x = k, y = cells)) +
  geom_line(color = "#4a90d9", linewidth = 1.4) +
  geom_point(size = 3, color = "#4a90d9") +
  geom_point(data = filter(expl_df, k %in% c(2, 10)),
             size = 5, shape = 21, fill = "white", color = "#d7301f", stroke = 2) +
  geom_text(data = filter(expl_df, k %in% c(2, 10)),
            aes(label = label),
            nudge_x = 0.9, nudge_y = 0, hjust = 0, size = 3.2,
            color = "#d7301f", lineheight = 0.9) +
  geom_vline(xintercept = c(2, 10), linetype = "dotted", color = "#d7301f", linewidth = 0.7) +
  scale_y_log10(labels = label_comma(),
                breaks  = c(4, 32, 256, 1024, 8192, 32768)) +
  scale_x_continuous(breaks = c(1, 2, 5, 10, 15)) +
  labs(x = "Number of binary dimensions (k)",
       y = "Design space: number of unique profiles (log scale)",
       title = "How Quickly Does the Design Space Grow?",
       subtitle = "Each additional binary dimension doubles the number of distinct participant profiles (2\u1d4f)") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

### Visualizing the 1,024-Cell Matrix

With 10 binary dimensions, the full design matrix has 1,024 rows. We can visualize this as a heatmap where each column is a unique participant profile and each row is one dimension:

```{r matrix-1024, fig.height=6, fig.width=10}
#| code-summary: "▶ Plot: all 1,024 unique profiles in the 10-dimension design space"

all_profiles <- as.data.frame(
  do.call(expand.grid, replicate(10, 0:1, simplify = FALSE)))
colnames(all_profiles) <- c("Env. Values", "Income", "Social Norms",
                             "Brand Trust", "Perc. Quality", "Health Motiv.",
                             "Peer Influence", "Prior Behavior",
                             "Price Sensitiv.", "Emotional Resp.")
all_profiles$profile_id <- seq_len(nrow(all_profiles))

plot_df_m <- all_profiles |>
  pivot_longer(-profile_id, names_to = "Dimension", values_to = "Level") |>
  mutate(Dimension = factor(Dimension, levels = rev(colnames(all_profiles)[-11])))

ggplot(plot_df_m, aes(x = profile_id, y = Dimension, fill = factor(Level))) +
  geom_raster() +
  scale_fill_manual(values = c("0" = "#d6ecff", "1" = "#1a3a6b"),
                    labels = c("Low (0)", "High (1)"), name = "Level") +
  labs(x = "Profile (1 of 1,024 unique combinations)", y = NULL,
       title = "The Full Design Space for Sustainable Purchase Intention",
       subtitle = "Each column = one participant profile  \u00b7  Each row = one binary dimension  \u00b7  1,024 cells total") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.grid = element_blank(), legend.position = "top")
```

To **exactly** fill this matrix — one observation per cell — you would need **N = 1,024 participants**. That is just to have a single observation per cell, with no statistical power to spare.

---

## From Exact Coverage to Approximate Orthogonality {#sec-orthogonality}

Here is the key insight: **you do not need to fill every cell exactly**.

In a randomized experiment, what you need is not *exact* coverage, but *approximate independence* — the correlation between treatment assignment and each of the 10 dimensions should be close to zero.

How many observations do you need for that? With random assignment and N participants, the expected correlation between treatment T and any single confounder C is zero. The *actual* correlation in any given sample will deviate from zero by chance, with standard error ≈ **1 / √N**. If you want ALL D correlations to stay below some threshold $|r_{\max}|$, you must account for D simultaneous comparisons (Bonferroni):

$$N \geq \left(\frac{z_{1-\alpha/(2D)}}{r_{\max}}\right)^2$$

### What does $r_{\max}$ actually mean?

Before looking at the numbers, it is worth anchoring $r_{\max}$ concretely. If your treatment T accidentally correlates r with omitted confounder C, and C has effect β (**in SD units of Y — this is identical to Cohen's d when Y is standardized**) on the outcome, then the omitted variable bias in your estimated treatment effect is approximately **β × r**.

> **A quick reminder from the power table above:** Cohen's d expresses the treatment effect in the same SD-unit currency. An effect of d = 0.40 means the group means differ by 0.40 standard deviations of Y. The β values below use that same ruler — so a β of 0.50 means the confounder shifts Y by half a standard deviation.

```{r rmax-concrete, fig.height=5.5, fig.width=11}
#| code-summary: "▶ Plot: how r_max tolerance maps to omitted-variable bias"

rmax_vals <- c(0.05, 0.10, 0.20, 0.30, 0.40, 0.60)
beta_vals <- seq(0.1, 1.5, by = 0.1)

bias_df <- expand.grid(r_max = rmax_vals, beta = beta_vals) |>
  mutate(bias   = r_max * beta,
         r_label = sprintf("r = %.2f", r_max))

# Cohen's d reference lines
cohen_refs <- data.frame(
  d     = c(0.20, 0.50, 0.80),
  label = c("Small effect\n(d = 0.20)", "Medium effect\n(d = 0.50)", "Large effect\n(d = 0.80)")
)

ggplot(bias_df, aes(x = beta, y = bias,
                    color = factor(r_max), group = factor(r_max))) +
  geom_hline(data = cohen_refs, aes(yintercept = d),
             linetype = "dotted", color = "gray60", linewidth = 0.6) +
  geom_text(data = cohen_refs, aes(x = 1.52, y = d + 0.02, label = label),
            inherit.aes = FALSE, color = "gray50", size = 2.8, hjust = 0, lineheight = 0.85) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 2.5) +
  scale_color_manual(
    values = c("0.05" = "#08519c", "0.1" = "#4a90d9", "0.2" = "#52b788",
               "0.3"  = "#f4a261", "0.4" = "#e07b39", "0.6" = "#e63946"),
    labels = sprintf("r_max = %.2f", rmax_vals),
    name   = "Correlation\ntolerance (r_max)"
  ) +
  scale_x_continuous(breaks = seq(0.2, 1.4, 0.2),
                     labels = function(x) sprintf("\u03b2 = %.1f SD", x)) +
  scale_y_continuous(labels = function(x) sprintf("%.2f SD", x),
                     limits = c(0, 0.95)) +
  coord_cartesian(clip = "off") +
  labs(x = "Effect of omitted confounder on Y (\u03b2 = Cohen's d equivalent; SD units)",
       y = "Omitted-variable bias in your treatment estimate",
       title = "How Much Bias Does Residual Imbalance Create?",
       subtitle = paste0("Bias \u2248 r_max \u00d7 \u03b2  \u00b7  Dotted lines = small/medium/large effect benchmarks\n",
                         "Lower r_max = less bias once achieved, but requires more N to verify")) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom",
        legend.direction = "horizontal",
        plot.margin = margin(5, 50, 5, 5))
```

In the coffee WTP context (SD ≈ $1.50):

- r = 0.05 with β = 0.8 SD: bias ≈ 0.04 SD ≈ **$0.06** — negligible
- r = 0.10 with β = 0.8 SD: bias ≈ 0.08 SD ≈ **$0.12** — small but detectable
- r = 0.20 with β = 0.8 SD: bias ≈ 0.16 SD ≈ **$0.24** — now as large as a small true effect
- r = 0.40 with β = 1.0 SD → bias ≈ 0.40 SD ≈ **$0.60** — larger than many published effects
- r = 0.60 with β = 1.0 SD → bias ≈ 0.60 SD ≈ **$0.90** — could reverse the sign of your treatment estimate entirely

### Required N for Approximate Orthogonality

```{r orthogonality-table}
#| code-summary: "▶ Calculate: N required for approximate orthogonality"

required_N <- function(D, r_thresh = 0.10, alpha = 0.05) {
  z_crit <- qnorm(1 - alpha / (2 * D))
  ceiling((z_crit / r_thresh)^2)
}

dims_check   <- c(1, 2, 3, 5, 8, 10, 15, 20)
thresh_check <- c(0.05, 0.10, 0.20, 0.30, 0.40, 0.60)

ortho_table <- expand.grid(D = dims_check, r_max = thresh_check) |>
  mutate(N_req = mapply(required_N, D, r_max),
         exact_cells = 2^D) |>
  pivot_wider(names_from = r_max, values_from = N_req,
              names_prefix = "r_max = ") |>
  rename(`Dimensions (D)` = D, `Exact cells (2^D)` = exact_cells)

kable(ortho_table,
      caption = "Minimum N for approximate orthogonality at \u03b1 = .05 (Bonferroni-corrected)",
      format.args = list(big.mark = ","))
```

```{r orthogonality-plot, fig.height=5.5}
#| code-summary: "▶ Plot: N required vs. dimensions for different tolerances"

ortho_plot_df <- expand.grid(D = 1:20, r_max = thresh_check) |>
  mutate(N_required = mapply(required_N, D, r_max),
         exact_N    = 2^D,
         r_label    = paste0("r_max = ", r_max))

ggplot(ortho_plot_df, aes(x = D)) +
  geom_line(aes(y = N_required, color = r_label, group = r_label), linewidth = 1.2) +
  geom_line(aes(y = exact_N), color = "black", linetype = "dashed", linewidth = 1.0) +
  geom_vline(xintercept = c(2, 10), linetype = "dotted", color = "gray50") +
  annotate("text", x = 2.3, y = 40000, label = "Recall\nAccuracy\n(k=2)",
           size = 2.8, color = "gray40", hjust = 0) +
  annotate("text", x = 10.3, y = 40000, label = "Sust. Purchase\nIntention\n(k=10)",
           size = 2.8, color = "gray40", hjust = 0) +
  annotate("text", x = 14.5, y = 2^14 * 1.8, label = "Exact fill\n(2^D cells)",
           size = 2.8, color = "black") +
  scale_y_log10(labels = label_comma()) +
  scale_x_continuous(breaks = seq(1, 20, 2), limits = c(1, 21)) +
  scale_color_manual(
    values = c("r_max = 0.05" = "#08519c", "r_max = 0.1"  = "#4a90d9",
               "r_max = 0.2"  = "#52b788", "r_max = 0.3"  = "#f4a261",
               "r_max = 0.4"  = "#e07b39", "r_max = 0.6"  = "#e63946"),
    name = "Correlation\ntolerance"
  ) +
  coord_cartesian(clip = "off") +
  labs(x = "Number of binary dimensions of Y (D)",
       y = "Observations required (log scale)",
       title = "How Many Observations for Approximate Orthogonality?",
       subtitle = "Dashed = exact fill (2^D)  \u00b7  Colored lines = Bonferroni-corrected independence threshold") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(), legend.position = "right",
        plot.margin = margin(5, 30, 5, 5))
```

::: {.callout-warning}
## The key tradeoff — and why most experiments are implicitly operating at high $r_{\max}$

The table above reveals an uncomfortable tension that sits at the heart of experimental practice:

**Lower $r_{\max}$** = a stricter standard = less confounding once achieved = better causal inference.

**But lower $r_{\max}$ also requires far more observations** to reliably verify. N scales as $1/r_{\max}^2$, so halving your tolerance quadruples the required sample.

::: {.callout-important}
## The core punchline

This is the central tension in designing experiments for complex constructs. The researcher faces a dilemma:

- **Set a strict $r_{\max}$ (e.g., 0.05–0.10):** Your causal inferences will be clean — but you need sample sizes that few lab studies can afford.
- **Accept a lenient $r_{\max}$ (e.g., 0.30–0.40):** Your sample size is achievable — but residual confounding may be large enough to bias your estimates substantially, especially when individual confounders have strong effects (β ≈ 0.5–1.0 SD) on Y.

A typical psychology or marketing experiment with **N = 60–120** participants is, at best, achieving $r_{\max} \approx 0.30$–$0.40$ for a construct with D = 5–10 dimensions — regardless of whether Qualtrics randomized. Researchers who report a balance table with N = 80 and declare "randomization succeeded" are not wrong about the observable; they are implicitly operating at a very lenient latent $r_{\max}$, where substantial confounding remains undetected.

In Module 1's language: their randomization observable has **low construct validity** as an indicator of latent randomization, because the gap between the observable and the latent property is too wide to close at typical sample sizes.
:::

With a narrow outcome like **Recall Accuracy** (D = 2 dimensions):
- To ensure $|r| < 0.10$ with 95% confidence: N ≈ **`r required_N(2, 0.10)`** — feasible
- A study with N = 80 comfortably satisfies even a strict $r_{\max} = 0.20$

With a broad outcome like **Sustainable Purchase Intention** (D = 10 dimensions):
- To ensure $|r| < 0.10$: N ≈ **`r required_N(10, 0.10)`** — rarely achieved in practice
- To ensure $|r| < 0.20$: N ≈ **`r required_N(10, 0.20)`** — still demanding for most lab studies
- A study with N = 200 is operating at an effective $r_{\max} > 0.30$ — where a single strong confounder (β ≈ 0.8 SD) can inject bias of 0.24 SD or more into your treatment estimate

This is the direct experimental parallel to Module 1's core finding about discriminant validity. A scale with **many correlated sub-dimensions** is hard to measure cleanly; a construct with **many latent antecedents** is hard to randomize cleanly. Both problems grow with construct breadth, and both require you to map the latent space before claiming clean measurement or clean identification.
:::

---

## Simulation: Correlated Dimensions — Two Countervailing Forces {#sec-dgp-sim}

The analysis so far assumed that the 10 latent dimensions of Y are **independent of one another** in the data-generating process. But is that realistic?

Think about the constructs driving Sustainable Purchase Intention. Environmental values, health motivation, social norm salience, and peer influence tend to cluster together — the "conscious consumer" profile. Similarly, price sensitivity, income, and habitual spending patterns cluster together — the "pragmatic buyer" profile. The latent dimensions are *correlated*, not orthogonal.

### Two forces pull in opposite directions

Before running any simulation, consider the logic:

**Force 1 — fewer independent checks (sometimes helps).** If two dimensions are highly correlated (say, environmental values and social norm salience move together), then for the purpose of *counting* how many independent things can go wrong, you effectively have fewer than 10 independent dimensions. Fewer independent checks = lower probability that *any one* of them exceeds the threshold by chance = **slightly easier to achieve orthogonality** for a given N.

**Force 2 — cascade failures (always hurts).** The flip side is severe. When randomization *does* fail on one correlated dimension, it fails on all the correlated ones simultaneously. Imbalance on "environmental values" drags along imbalance on "social norms," "health motivation," and "peer influence" all at once. Each of these has its own independent β effect on Y, so the bias compounds. Instead of one confounder contributing β × r bias, you get four correlated confounders each contributing, with their effects accumulating.

### When does correlation help vs. hurt?

The key variable is **why** the dimensions are correlated:

| Scenario | Structure | Effect on randomization |
|---|---|---|
| **Helpful:** dimensions cluster around a few latent "super-factors" | e.g., all "values-based" dimensions move together as one bloc | Fewer effective checks; orthogonality easier to achieve at smaller N |
| **Hurtful:** dimensions are correlated *and* each carries independent variance in Y | e.g., environmental values, health focus, and social norms each have separate β effects on WTP | Cascade failures: one imbalanced assignment fails simultaneously on 3 dimensions with additive bias |
| **Worst case:** high correlation *and* strong independent effects on Y | The dims are correlated with each other AND each has a large β | One randomization failure contaminates every correlated dimension, compounding bias substantially |

**The concrete intuition:** imagine you are a researcher measuring balance, and you check environmental values and find it balanced (r ≈ 0). If that dimension was highly correlated with health focus and social norms in your sample, you might infer those are balanced too — and you might be right. That is the *helpful* version. But if each of those three dimensions has an independent effect on WTP that wasn't captured by environmental values alone, then any *failure* of balance cascades to all three simultaneously — that is the *hurtful* version.

This maps directly onto Module 1's concept of **correlated subscales**. When subscales are highly correlated, you gain efficiency in measurement (fewer items needed to capture the same information). But if they are discriminantly invalid — correlated *and* capturing separate variance — then mismeasuring one means mismeasuring all.

### Simulation

We run two contrasting scenarios, each featuring correlated dimensions but differing in whether the failure mode is primarily helpful or hurtful:

- **Scenario A — "Helpful" correlation (ρ = 0.75, uniform small β):** Dimensions are very strongly correlated — effectively 10 measured dimensions but only ~2–3 independent blocs. Each dimension has a similar, small β effect on Y (β = 0.15). When the correlation structure is this tight, the effective number of independent checks drops sharply, making orthogonality easier to achieve.

- **Scenario B — "Hurtful" correlation (ρ = 0.35, heterogeneous large β):** Dimensions have moderate correlation but each carries a meaningfully different and large independent effect on Y (ranging from β = 0.05 to β = 1.50). When randomization fails on one dimension, several correlated ones fail simultaneously — and because each carries its own large β, the cumulative bias explodes.

```{r dgp-simulation, fig.height=16, fig.width=10}
#| code-summary: "▶ Simulate: three DGP scenarios + Type-1 error rate (may take ~2 min)"

# ── Helper: correlated binary covariates via Cholesky ────────────────────────
corr_binary <- function(N, D, rho) {
  Sigma <- matrix(rho, D, D); diag(Sigma) <- 1
  L     <- chol(Sigma)
  z     <- matrix(rnorm(N * D), nrow = N) %*% L
  (z > 0L) * 1L
}

# ── Simulation: P(max |r| < threshold) ───────────────────────────────────────
sim_orthog <- function(N, D, rho = 0, threshold = 0.10, n_sim = 1500) {
  mean(replicate(n_sim, {
    X        <- if (rho == 0) matrix(rbinom(N * D, 1, 0.5), nrow = N)
                else          corr_binary(N, D, rho)
    T_assign <- rbinom(N, 1, 0.5)
    cors     <- apply(X, 2, function(col) abs(cor(col, T_assign)))
    max(cors) < threshold
  }))
}

# ── Simulation: cumulative bias when any failure occurs ──────────────────────
sim_bias <- function(N = 200, D = 10, rho = 0, betas, threshold = 0.10, n_sim = 1000) {
  replicate(n_sim, {
    X        <- if (rho == 0) matrix(rbinom(N * D, 1, 0.5), nrow = N)
                else          corr_binary(N, D, rho)
    T_assign <- rbinom(N, 1, 0.5)
    cors     <- apply(X, 2, function(col) cor(col, T_assign))
    sum(abs(cors[abs(cors) > threshold]) * betas[abs(cors) > threshold])
  })
}

# ── Simulation: Type-1 and Type-2 error rates, conditional on orthogonality failing ──
# The key insight: under PERFECT randomization, Type-1 is always ~5% regardless of betas.
# The interesting question is: given your experiment FAILED the orthogonality check
# (as many real experiments do), how bad are your error rates?
# Large betas (Scenario B) turn small residual imbalance into large bias → inflated Type-1.
# Tiny betas (Scenario A) produce negligible bias even when imbalance exists → Type-1 near 5%.
sim_error_rates <- function(N, D, rho, betas, r_max = 0.10,
                            true_eff = 0.40, n_sim = 2000) {
  results <- t(replicate(n_sim, {
    X        <- if (rho == 0) matrix(rbinom(N * D, 1, 0.5), nrow = N)
                else          corr_binary(N, D, rho)
    T_assign <- rbinom(N, 1, 0.5)
    cors     <- apply(X, 2, function(x) abs(cor(x, T_assign)))
    orth_ok  <- max(cors) <= r_max          # did this experiment pass the balance check?
    Y_null   <- as.vector(X %*% betas) + rnorm(N, 0, 1)
    p_null   <- t.test(Y_null[T_assign == 1], Y_null[T_assign == 0])$p.value
    Y_alt    <- true_eff * T_assign + as.vector(X %*% betas) + rnorm(N, 0, 1)
    p_alt    <- t.test(Y_alt[T_assign == 1], Y_alt[T_assign == 0])$p.value
    c(t1 = as.integer(p_null < 0.05), t2 = as.integer(p_alt > 0.05),
      orth_ok = as.integer(orth_ok))
  }))
  failed <- results[, "orth_ok"] == 0L
  passed <- results[, "orth_ok"] == 1L
  # Type-1: false-positive rate CONDITIONAL ON orthogonality FAILING (violations present)
  t1_cond  <- if (sum(failed) > 10) mean(results[failed, "t1"]) else NA_real_
  # Type-2: miss rate CONDITIONAL ON orthogonality PASSING (no violations; p-values are valid here)
  t2_valid <- if (sum(passed) > 10) mean(results[passed, "t2"]) else NA_real_
  data.frame(t1_rate = t1_cond, t2_rate = t2_valid,
             pct_fail = mean(failed))
}

set.seed(42)

N_vals <- c(50, 100, 200, 400, 800)
D      <- 10

# Scenario betas — made more extreme to amplify the contrast
betas_helpful <- rep(0.15, D)                    # uniform tiny betas
betas_hurtful <- c(1.50, 1.20, 1.00, 0.90, 0.70,
                   0.50, 0.30, 0.20, 0.10, 0.05) # large heterogeneous betas

# P(orthogonal) across N, for three DGP types
prob_df <- bind_rows(
  data.frame(N   = N_vals, DGP = "Independent (rho = 0)",
             prob = sapply(N_vals, sim_orthog, D = D, rho = 0.00)),
  data.frame(N   = N_vals, DGP = "Scenario A: Helpful (rho = 0.75, uniform small \u03b2)",
             prob = sapply(N_vals, sim_orthog, D = D, rho = 0.75)),
  data.frame(N   = N_vals, DGP = "Scenario B: Hurtful (rho = 0.35, large varied \u03b2)",
             prob = sapply(N_vals, sim_orthog, D = D, rho = 0.35))
)

scen_colors <- c(
  "Independent (rho = 0)"                                  = "#4a90d9",
  "Scenario A: Helpful (rho = 0.75, uniform small \u03b2)" = "#52b788",
  "Scenario B: Hurtful (rho = 0.35, large varied \u03b2)"  = "#e63946"
)

p_prob <- ggplot(prob_df, aes(x = N, y = prob, color = DGP, group = DGP)) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 3.5) +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "gray40") +
  annotate("text", x = 820, y = 0.957, label = "95% target",
           size = 3.0, color = "gray40", hjust = 0) +
  scale_x_continuous(breaks = N_vals) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
  scale_color_manual(values = scen_colors, name = NULL) +
  labs(x = "Sample size (N)",
       y = "P(max |r| < 0.10 across all 10 dimensions)",
       title = "Probability of Achieving Orthogonality (D = 10)",
       subtitle = paste0("Scenario A (strong clustering, tiny betas): easier — correlation reduces effective checks\n",
                         "Scenario B (moderate correlation, large varied betas): hardest — and most dangerous when it fails")) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(), legend.position = "top",
        legend.text = element_text(size = 9))

# ── Bias when randomization fails ─────────────────────────────────────────────
set.seed(42)
bias_indep   <- sim_bias(N = 200, D = D, rho = 0.00, betas = betas_hurtful)
bias_helpful <- sim_bias(N = 200, D = D, rho = 0.75, betas = betas_helpful)
bias_hurtful <- sim_bias(N = 200, D = D, rho = 0.35, betas = betas_hurtful)

bias_df <- bind_rows(
  data.frame(bias = bias_indep,   DGP = "Independent (rho = 0)"),
  data.frame(bias = bias_helpful, DGP = "Scenario A: Helpful (rho = 0.75, uniform small \u03b2)"),
  data.frame(bias = bias_hurtful, DGP = "Scenario B: Hurtful (rho = 0.35, large varied \u03b2)")
) |>
  mutate(DGP = factor(DGP, levels = c("Independent (rho = 0)",
                                       "Scenario A: Helpful (rho = 0.75, uniform small \u03b2)",
                                       "Scenario B: Hurtful (rho = 0.35, large varied \u03b2)")))

p_bias <- ggplot(bias_df |> filter(bias > 0),
                 aes(x = bias, fill = DGP)) +
  geom_histogram(bins = 35, alpha = 0.80, color = "white", linewidth = 0.2) +
  geom_vline(data = bias_df |> filter(bias > 0) |>
               group_by(DGP) |> summarise(med = median(bias), .groups = "drop"),
             aes(xintercept = med, color = DGP), linewidth = 1.1, linetype = "dashed") +
  facet_wrap(~DGP, nrow = 3, scales = "free") +
  scale_fill_manual(values = scen_colors, guide = "none") +
  scale_color_manual(values = scen_colors, guide = "none") +
  scale_x_continuous(labels = function(x) sprintf("%.2f SD", x)) +
  labs(x = "Cumulative OVB when randomization fails (SD units of Y)",
       y = "Frequency across 1,000 experiments",
       title = "When Randomization Fails: How Large Is the Bias? (N = 200)",
       subtitle = "Dashed = median cumulative bias  \u00b7  Only experiments with \u2265 1 failing dimension shown") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold", size = 8.5))

# ── Type-1 error rates across N (Type-2 intentionally excluded — see note below) ──
set.seed(42)
err_df <- bind_rows(
  lapply(N_vals, function(n) {
    cbind(N = n, DGP = "Independent (rho = 0)",
          sim_error_rates(n, D, rho = 0.00, betas = betas_hurtful))
  }),
  lapply(N_vals, function(n) {
    cbind(N = n,
          DGP = "Scenario A: Helpful (rho = 0.75, uniform small \u03b2)",
          sim_error_rates(n, D, rho = 0.75, betas = betas_helpful))
  }),
  lapply(N_vals, function(n) {
    cbind(N = n,
          DGP = "Scenario B: Hurtful (rho = 0.35, large varied \u03b2)",
          sim_error_rates(n, D, rho = 0.35, betas = betas_hurtful))
  })
) |>
  mutate(N        = as.integer(N),
         t1_rate  = as.numeric(t1_rate),
         pct_fail = as.numeric(pct_fail),
         DGP      = as.character(DGP))

# Show Type-1 error only (Type-2 is not reported — see callout below)
p_err <- ggplot(err_df |> filter(DGP != "Independent (rho = 0)"),
                aes(x = N, y = t1_rate, color = DGP, group = DGP)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3.0) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "gray40") +
  annotate("text", x = max(N_vals) * 1.02, y = 0.055, label = "5% nominal",
           color = "gray40", size = 3.2, hjust = 0) +
  scale_x_continuous(breaks = N_vals) +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     limits = c(0, NA)) +
  scale_color_manual(values = scen_colors[names(scen_colors) != "Independent (rho = 0)"],
                     name = NULL) +
  labs(x = "Sample size (N)",
       y = "Type-1 error rate (among failed experiments)",
       title = "False-Positive Rate Rises with N When Orthogonality Fails",
       subtitle = paste0(
         "Conditional on |r(T, X)| > 0.10  \u00b7  ",
         "Dashed = 5% nominal  \u00b7  ",
         "Scenario B's large \u03b2s amplify residual imbalance into detectable bias")) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), legend.position = "top",
        legend.text = element_text(size = 8.5),
        plot.margin = margin(5, 50, 5, 5))

p_prob / p_bias / p_err
```

::: {.callout-warning}
## Reading the three plots together

**Top plot — P(achieving orthogonality):** Scenario A (ρ = 0.75, tiny uniform betas) is the *easiest* case. The very strong correlation among dimensions collapses the effective number of independent checks, so a smaller N achieves the 95% target. This is the helpful force at work. Scenario B (ρ = 0.35, large varied betas) offers little advantage over independent dimensions on this metric.

**Middle plot — bias when randomization fails:** This is where the scenarios diverge dramatically. When Scenario A fails, it fails on a highly clustered bloc with tiny β values — cumulative bias remains small. When Scenario B fails, it fails on moderately correlated dimensions each with large independent β effects — the bias distribution has a fat right tail, with individual experiments easily incurring > 1.0 SD of cumulative confound. Notice how much wider the Scenario B distribution is compared to A.

**Bottom plot — Type-1 error rate (given orthogonality failed):** Conditional on at least one covariate being imbalanced (|r| > 0.10), Scenario B's large betas turn even small residual correlations into substantial bias — false positive rates rise well above 5% **and continue rising as N increases**. Scenario A stays near 5% because its tiny betas keep the omitted-variable bias small even when imbalance exists.

::: {.callout-note}
## Why is Type-2 error (power) not shown?

**Because the question of power is irrelevant when exchangeability is violated.**

Power analysis asks: "If there is a true effect, how often will my test correctly detect it?" That question is only valid when the p-value is a valid test statistic — i.e., when treatment is exchangeable with control. When orthogonality fails, the p-value is **not** a valid measure of the treatment effect. A "detection" (p < 0.05) in an experiment where orthogonality failed could mean either (a) you correctly found a real effect, or (b) the confound inflated the test statistic into significance independently of whether an effect exists. You cannot distinguish them. Power therefore has no valid interpretation under a violated exchangeability assumption — the right question is how to restore valid inference, not how often the broken test produces p < 0.05.
:::

::: {.callout-important}
## Counterintuitive danger: larger N makes invalid tests *more* dangerous, not less

Most researchers assume larger samples make experiments safer. The bottom plot reveals the opposite: **among experiments that fail the orthogonality check, the false-positive rate rises with N** — the very opposite of what researchers intuitively expect.

**The mechanism:** the plot conditions on experiments where |r(T, X)| > 0.10. Among those "failing" experiments, the residual bias (r × β) is roughly constant across N — the failing correlation barely exceeds 0.10 in all cases. What *does* change is the **standard error of the t-test**, which shrinks as 1/√N. The same small bias that was invisible noise at small N becomes a statistically detectable "effect" at large N.

**The compounding trap visualized:** The plot below shows — for Scenario B (the dangerous case) — how Type-1 error (among experiments *with* violations) and Type-2 error/miss rate (among experiments *without* violations) evolve as N grows. Point size is proportional to the frequency of orthogonality violations. Read it as a race: the red line (false-positive danger) climbs faster than the blue line (miss rate) falls.

```{r danger-overlay, fig.height=5, fig.width=11}
#| code-summary: "▶ Side-by-side: Type-1 error (violations) vs power (no violations) as N grows"

# Extract Scenario B only; compute power = 1 - miss rate; reshape to long for facets
danger_df <- err_df |>
  filter(DGP == "Scenario B: Hurtful (rho = 0.35, large varied \u03b2)") |>
  dplyr::select(N, t1_rate, t2_rate, pct_fail) |>
  mutate(power = 1 - t2_rate) |>                     # power = 1 - miss rate
  tidyr::pivot_longer(c(t1_rate, power),
                      names_to  = "metric",
                      values_to = "rate") |>
  mutate(
    panel = dplyr::case_when(
      metric == "t1_rate" ~
        "Type-1 Error Rate\n(experiments WITH orthogonality violations — p-values INVALID)",
      metric == "power" ~
        "Power (1 \u2212 Miss Rate)\n(experiments WITHOUT violations — p-values VALID)"
    ),
    panel = factor(panel, levels = c(
      "Type-1 Error Rate\n(experiments WITH orthogonality violations — p-values INVALID)",
      "Power (1 \u2212 Miss Rate)\n(experiments WITHOUT violations — p-values VALID)"
    )),
    color_val = ifelse(metric == "t1_rate", "#e63946", "#457b9d")
  )

# Reference line data: 5% for Type-1 panel, 80% for power panel
ref_lines <- data.frame(
  panel = factor(c(
    "Type-1 Error Rate\n(experiments WITH orthogonality violations — p-values INVALID)",
    "Power (1 \u2212 Miss Rate)\n(experiments WITHOUT violations — p-values VALID)"
  ), levels = levels(danger_df$panel)),
  yref  = c(0.05, 0.80),
  label = c("5% nominal \u03b1", "80% power")
)

ggplot(danger_df, aes(x = N, y = rate, group = panel)) +
  geom_hline(data = ref_lines, aes(yintercept = yref),
             linetype = "dashed", color = "gray50", linewidth = 0.8) +
  geom_text(data = ref_lines, aes(x = 820, y = yref + 0.025, label = label),
            color = "gray40", size = 3.0, hjust = 0, inherit.aes = FALSE) +
  geom_line(aes(color = panel), linewidth = 1.4) +
  geom_point(aes(size = pct_fail, color = panel), alpha = 0.88) +
  facet_wrap(~panel, nrow = 1, scales = "free_y") +
  scale_color_manual(values = c(
    "Type-1 Error Rate\n(experiments WITH orthogonality violations — p-values INVALID)" = "#e63946",
    "Power (1 \u2212 Miss Rate)\n(experiments WITHOUT violations — p-values VALID)"      = "#457b9d"
  ), guide = "none") +
  scale_size_continuous(
    name   = "Proportion of experiments\nwith |r(T,X)| > 0.10",
    range  = c(2, 8),
    labels = scales::percent_format(accuracy = 1)) +
  scale_x_continuous(breaks = N_vals) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    x        = "Sample size (N)",
    y        = NULL,
    title    = "Larger N: Type-1 Error Climbs Faster Than Power Grows — Among Experiments That Fail Orthogonality",
    subtitle = paste0(
      "Scenario B (rho = 0.35, large heterogeneous \u03b2s)  \u00b7  ",
      "Point size = proportion of experiments with at least one |r(T,X)| > 0.10\n",
      "Left: false-positive danger rises with N (invalid experiments)  \u00b7  ",
      "Right: power in valid experiments rises with N (as expected)")) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor  = element_blank(),
    legend.position   = "bottom",
    legend.direction  = "horizontal",
    legend.text       = element_text(size = 8),
    strip.text        = element_text(face = "bold", size = 9.5),
    plot.margin       = margin(5, 20, 5, 5)
  )
```

**Reading the panels:** The left panel shows Type-1 error (false-positive rate) *conditional on orthogonality having failed* — rising from ~6% at N = 50 to 20–30% at N = 800, far above the 5% nominal level. The right panel shows power *in the experiments that passed the orthogonality check* — rising as expected, because larger N does increase power for valid experiments. The **critical asymmetry**: point size (proportion of experiments with violations) shrinks as N grows, so the situation looks increasingly safe to a researcher monitoring only the balance check pass rate. Yet the rare failures that remain become dramatically more dangerous. **The compounding trap:** sample size DOES reduce the frequency of orthogonality violations — but it reduces violations *slower* than it amplifies the danger of the violations that remain.

**Larger N does not license looser balance checking — it demands stricter checking.** Tighten your orthogonality criterion (lower r_max) as N grows. Never assume you have achieved actual randomization; random assignment is a procedure, not a guarantee, and its validity must be verified empirically every time.
:::

**The practical lesson:** correlated dimensions are not uniformly good or bad. The key question is what their β values look like. If correlated dimensions all carry similar, modest effects, correlation genuinely helps by reducing effective checks. If correlated dimensions carry large, heterogeneous effects, correlation is a liability — and that liability is invisible unless you examine conditional bias and error rates, not just P(orthogonality).

This mirrors a core lesson from Module 1: a tight subscale structure looks efficient and reliable, but if the subscales carry separate predictive variance (discriminant invalidity), a failure in one cascades to all. The structural parallel is exact.
:::

---

## Checking the Observables: Randomization Diagnostics

::: {.callout-important}
## A fundamental limitation — you can only check what you measured

**You can only discover failures of randomization on dimensions you thought to measure.**

This is the direct experimental parallel to Module 1's measurement problem. In Module 1, we showed that unmeasured sub-dimensions of Y produce a biased estimate of your latent construct — you cannot diagnose measurement failure on dimensions you never operationalized. The same logic applies here: if a potential confounder was never measured as a covariate, no balance table, no Love plot, and no omnibus test will catch it.

This constraint is not a limitation of statistics — it is a fundamental epistemological constraint. Its implication is that **randomization checking is only as good as the covariate measurement you did before the study began**. The tools from Module 1 — CFA to identify sub-dimensions of Y, pilot text analysis to map what's in X and Y (see Section 2.5), discriminant validity checks — are not just measurement tools. They are *pre-registration scaffolding* that determines whether your eventual randomization check has the construct coverage to be meaningful.

**The practical lesson:** Before collecting data, use the construct-mapping tools from Module 1 to enumerate every plausible antecedent of Y. Measure them all. Then, after randomization, check balance on every one of them. The diagnostics below are only valuable if you did that groundwork first.
:::

Even though you cannot directly observe the latent dimensions, you can check the observable proxies you did measure. A balance table checks whether measured characteristics are distributed similarly across conditions — a necessary but not sufficient condition for latent randomization.

Here we simulate a study where randomization partially failed (treatment assignment was inadvertently correlated with one confounder) and show how to detect it.

```{r randomization-check-setup}
#| code-summary: "▶ Simulate study with imperfect randomization"

set.seed(42)
N_study <- 200

observed_chars <- data.frame(
  env_values     = rbinom(N_study, 1, 0.50),
  income_high    = rbinom(N_study, 1, 0.45),
  social_norms   = rbinom(N_study, 1, 0.50),
  brand_trust    = rbinom(N_study, 1, 0.55),
  quality_percep = rbinom(N_study, 1, 0.50),
  health_motiv   = rbinom(N_study, 1, 0.40),
  peer_influence = rbinom(N_study, 1, 0.50),
  prior_green    = rbinom(N_study, 1, 0.45),
  price_sensitiv = rbinom(N_study, 1, 0.50),
  emotion_resp   = rbinom(N_study, 1, 0.50)
)

# "Random" assignment — inadvertently correlated with env_values and income
# (e.g., participants self-selected into a study they saw on an eco-forum)
treatment_prob <- 0.35 +
  0.25 * observed_chars$env_values +
  0.15 * observed_chars$income_high
treatment_prob <- pmin(pmax(treatment_prob, 0.05), 0.95)
treatment <- rbinom(N_study, 1, treatment_prob)

study_df <- cbind(observed_chars,
                  treatment = factor(treatment, labels = c("Control", "Eco-label")))
```

### Balance Table

```{r balance-table}
#| code-summary: "▶ Compute balance table with chi-square tests"

balance_results <- lapply(names(observed_chars), function(var) {
  ctrl <- observed_chars[[var]][treatment == 0]
  trt  <- observed_chars[[var]][treatment == 1]
  chi  <- chisq.test(table(observed_chars[[var]], treatment), correct = FALSE)
  data.frame(
    Characteristic  = gsub("_", " ", var) |> tools::toTitleCase(),
    `Control (%)`   = round(mean(ctrl) * 100, 1),
    `Eco-label (%)` = round(mean(trt)  * 100, 1),
    Difference      = round((mean(trt) - mean(ctrl)) * 100, 1),
    p_value         = round(chi$p.value, 3),
    Flag            = ifelse(chi$p.value < 0.05, "\u26a0\ufe0f Imbalanced", "")
  )
}) |> bind_rows()

kable(balance_results,
      col.names = c("Characteristic", "Control (%)", "Eco-label (%)",
                    "Diff (pp)", "p-value (\u03c7\u00b2)", ""),
      caption = "Balance table: observed characteristics by condition")
```

### Visual Balance Check (Love Plot)

```{r balance-plot, fig.height=5}
#| code-summary: "▶ Plot: standardized differences (love plot)"

smd_df <- data.frame(
  Characteristic = gsub("_", " ", names(observed_chars)) |> tools::toTitleCase(),
  SMD = sapply(names(observed_chars), function(var) {
    ctrl <- observed_chars[[var]][treatment == 0]
    trt  <- observed_chars[[var]][treatment == 1]
    pooled_sd <- sqrt((var(trt) + var(ctrl)) / 2)
    if (pooled_sd == 0) return(0)
    (mean(trt) - mean(ctrl)) / pooled_sd
  })
) |>
  arrange(desc(abs(SMD))) |>
  mutate(Characteristic = factor(Characteristic, levels = Characteristic),
         flag           = abs(SMD) > 0.10)

ggplot(smd_df, aes(x = SMD, y = Characteristic, color = flag)) +
  geom_vline(xintercept =  0,    color = "gray40", linewidth = 0.8) +
  geom_vline(xintercept =  0.10, linetype = "dashed", color = "#de2d26", linewidth = 0.7) +
  geom_vline(xintercept = -0.10, linetype = "dashed", color = "#de2d26", linewidth = 0.7) +
  geom_segment(aes(x = 0, xend = SMD, y = Characteristic, yend = Characteristic),
               linewidth = 1.0) +
  geom_point(size = 4) +
  scale_color_manual(values = c("FALSE" = "#52b788", "TRUE" = "#de2d26"),
                     labels = c("|SMD| \u2264 0.10", "|SMD| > 0.10"), name = NULL) +
  labs(x = "Standardized Mean Difference (Treatment \u2212 Control)",
       y = NULL,
       title = "Love Plot: Covariate Balance by Condition",
       subtitle = "Dashed lines = |SMD| = 0.10 threshold  \u00b7  Red = potential imbalance") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(), legend.position = "top")
```

### Omnibus Randomization Test

For a single omnibus test, we use logistic regression predicting treatment from all characteristics. If randomization succeeded, no characteristic should significantly predict assignment, and the model R² should be near zero.

```{r omnibus-test}
#| code-summary: "▶ Omnibus randomization test via logistic regression"

treat_int <- as.integer(treatment) - 0L

omnibus_model <- glm(treat_int ~ ., data = cbind(observed_chars, treat_int),
                     family = binomial)

null_ll <- logLik(glm(treat_int ~ 1, data = cbind(observed_chars, treat_int),
                      family = binomial))
full_ll <- logLik(omnibus_model)
n       <- N_study

r2_nag <- (1 - exp(2 * (null_ll - full_ll) / n)) / (1 - exp(2 * null_ll / n))

lrt_p  <- anova(glm(treat_int ~ 1, data = cbind(observed_chars, treat_int),
                    family = binomial),
                omnibus_model, test = "LRT")$`Pr(>Chi)`[2]

cat(sprintf("Omnibus LRT: p = %.4f\n",  lrt_p))
cat(sprintf("Nagelkerke R\u00b2: %.3f\n", r2_nag))
cat(sprintf("Interpretation: Covariates explain %.1f%% of variance in treatment assignment.\n",
            r2_nag * 100))
cat("Good randomization should produce R\u00b2 \u2248 0 and p >> .05.\n")
```

#### Handling mixed covariate types {.unnumbered}

In practice, your covariates will rarely all be the same type. You may have binary indicators (gender, prior exposure), ordered categories (education level, income bracket), unordered categories (nationality, recruitment channel), and continuous measures (age, prior attitude score). The omnibus logistic regression approach handles all of these — you just need to encode each type correctly before passing them to `glm`:

```{r omnibus-mixed-types, eval=FALSE}
#| code-summary: "▶ Template: omnibus randomization test for mixed covariate types (not run)"

# ── Step 1: Encode each covariate type correctly ──────────────────────────────
# Suppose your covariate data frame is called `covariates` with columns:
#   gender       : binary (0/1 or factor with 2 levels)
#   edu_level    : ordinal (e.g., 1=high school, 2=bachelor, 3=master, 4=PhD)
#   country      : nominal (unordered categories → MUST be factor)
#   age          : continuous
#   prior_att    : continuous (e.g., prior attitude score on 1–7 scale)

covariates$gender    <- factor(covariates$gender)         # binary → factor (1 df)
covariates$edu_level <- as.integer(covariates$edu_level)  # ordinal → numeric (linear trend)
covariates$country   <- factor(covariates$country)        # nominal → factor (K-1 dummies)
# age and prior_att stay as numeric — no conversion needed

# ── Step 2: Omnibus logistic regression ───────────────────────────────────────
# treatment must be a binary 0/1 variable
# glm automatically dummy-codes all factor variables
omnibus_fit  <- glm(treatment ~ ., data = covariates, family = binomial)
null_fit     <- glm(treatment ~ 1, data = covariates, family = binomial)

# ── Step 3: Likelihood-ratio test ─────────────────────────────────────────────
lrt <- anova(null_fit, omnibus_fit, test = "LRT")
cat(sprintf("Omnibus LRT p = %.4f\n", lrt$`Pr(>Chi)`[2]))

# Nagelkerke pseudo-R²
n_obs  <- nrow(covariates)
nag_r2 <- (1 - exp(2 * (logLik(null_fit) - logLik(omnibus_fit)) / n_obs)) /
           (1 - exp(2 * logLik(null_fit) / n_obs))
cat(sprintf("Nagelkerke R² = %.3f\n", nag_r2))

# ── Optional: non-linear ordinal check ────────────────────────────────────────
# If you suspect the ordinal variable has a non-linear relationship with
# treatment assignment, convert it to a factor (dummy-coded) instead:
#   covariates$edu_level <- factor(covariates$edu_level)
# This uses more degrees of freedom but makes no linearity assumption.
```

**Key points:**

- **Binary** covariates: encode as `factor` — produces 1 dummy variable.
- **Ordinal** covariates: encode as `integer` (assumes linear trend across levels). If the relationship might be non-linear, use `factor()` instead — this creates K-1 dummies but uses more degrees of freedom.
- **Nominal** (unordered) covariates: **must** be `factor` — `glm` automatically creates K-1 dummy variables (one per category minus the reference level).
- **Continuous** covariates: pass as-is (numeric). No transformation needed.

The omnibus LRT tests whether **any** covariate — regardless of type — predicts treatment assignment better than chance. If p < .05 or R² > .05, investigate which covariates are driving the imbalance using the individual balance table and Love plot above.

::: {.callout-warning}
## What to look for

| Diagnostic | Good randomization | Potential problem |
|---|---|---|
| Balance table p-values | All p > .05 | Any p < .05 |
| Love plot SMDs | All \|SMD\| < 0.10 | Any \|SMD\| > 0.10 |
| Omnibus LRT p-value | p >> .05 | p < .05 |
| Nagelkerke R² | R² ≈ 0 | R² > 0.05 |

These checks only evaluate the **observable** dimensions you measured. The latent dimensions — all the antecedents of Y you did not think to measure — remain unverified. The exchangeability assumption is always partly an act of faith, and that faith becomes more tenuous the broader and more complex Y is.
:::

---

## When You Can't Check What You Can't See: Diagnosing Latent Imbalance {#sec-latent-imbalance}

The three diagnostics above — balance table, Love plot, omnibus test — are essential, but they share a fundamental limitation:

> **You can only check balance on dimensions you thought to measure.**

If an unmeasured dimension of Y is correlated with treatment assignment, no balance table will catch it. And as the analysis in Section 3.6 showed, the broader your outcome construct, the more such unmeasured dimensions there are likely to be.

What can you do?

### What you can do before data collection

**1. Map the construct before running the study.** Use Module 1's toolkit — confirmatory factor analysis, convergent/discriminant validity checks, latent class analysis — on *pilot data* to identify all the major sub-dimensions of Y. Every dimension you identify is a confounder you can now measure and check balance on.

**2. Blocked (stratified) randomization.** If you know in advance that certain dimensions of Y are likely to be distributed unevenly in your sample, pre-stratify on them. Assign treatment within blocks defined by those dimensions. This guarantees balance on the dimensions you can observe — reducing the degrees of freedom for latent imbalance to exploit.

**3. Collect more observations.** As Section 3.6 showed quantitatively, the probability of achieving approximate orthogonality on all D dimensions grows with N. If your outcome is broad (large D), power calculations based on the treatment effect alone dramatically underestimate the required sample.

### What you can do after data collection

**4. Open-ended text as a nomological net scanner — and an identification strategy.**

As we developed in detail in Section 2.5 (*What's in X? What's in Y?*), open-ended text responses from a small pilot are the most powerful tool for mapping the construct space before collecting main-study data. The nomological net scan reveals the exhaustive list of intermediary variables in the causal chain from X to Y. Every concept the text reveals is a mediator you should measure quantitatively — and a confounder you should check balance on.

**The key insight from Section 2.5 bears repeating here:** without the complete mediator list, the treatment coefficient in a regression is an uninterpretable mixture of all active pathways. With the list, you gain a structural identification strategy: decompose the total treatment effect into its component channels using the measured mediators.

The complete structural modeling demonstration — quantitative mediator scoring, parallel mediation, SEM with latent factors, and DGP recovery comparison — is developed step by step in **Section 2.5** above, where it serves as the practical illustration of how to handle exclusion restriction violations once the nomological net has been mapped.

---

# Part 4: When Exchangeability Breaks Down — Design and Statistical Solutions {#sec-exchangeability-fix}

## Why Exchangeability Fails

Part 3 showed that achieving latent randomization — the property we *actually* need — requires far more than clicking "Randomize." Exchangeability breaks down for two broad reasons:

1. **Biased sampling mechanisms:** The people who end up in your study, or in each condition, are not a random draw from the target population. This includes:
   - **Self-selection:** participants opt into studies, conditions, or platforms that attract systematically different people
   - **Attrition:** differential dropout between conditions (the eco-label condition may lead environment-indifferent participants to give up early)
   - **Exclusion bias:** screening criteria that remove participants who would look different from those retained
   - **Platform effects:** recruiting on Prolific vs. MTurk vs. a university pool changes the effective population

2. **Insufficient coverage of the factorial space:** Even with a clean sampling mechanism, if the design space has 1,024 cells (D = 10 binary dimensions) and you have N = 200, you are severely undersampling the space — and chance alone can produce treatment groups that are unrepresentative of the full distribution of participant types.

Both problems ultimately violate **exchangeability**: the assumption that the labels "Control" and "Eco-label" could be re-shuffled across the observed data without changing the underlying distribution. When they fail, your p-value null distribution — built on exchangeability — is no longer valid.

We now cover solutions at two stages: **before data collection** (experimental design) and **after data collection** (statistical correction).

---

## Experimental Design Solutions: Maximizing Orthogonality Before You Collect Data

### Completely Random Assignment — The Baseline

In completely random assignment (CRA), each participant is independently assigned to a condition with equal probability. This is what most researchers do when they check "randomize" in Qualtrics.

CRA is unbiased in expectation: on average across infinitely many experiments, it achieves balance. But for any single experiment, balance is a matter of chance — and as Part 3 showed, that chance is often disappointingly low when Y is complex.

### Stratified (Blocked) Random Assignment

**The idea:** divide participants into groups (strata or blocks) based on measured covariates that you expect to relate to Y, then randomize *within* each block.

**Concrete example:** In the eco-coffee study, you know from Module 1's analysis that "environmental values" is a strong predictor of WTP (β ≈ 0.8 SD). Before running the study, add a short pre-screening item to measure environmental values. Split participants into high vs. low environmental values strata. Then assign 50% to eco-label and 50% to control *within each stratum*.

**Why it works:** By randomizing within strata, you guarantee that each condition contains equal proportions of high- and low-environmental-values participants. This eliminates one source of latent imbalance — the most important one — by design rather than by chance.

```{r stratified-random, fig.height=5, fig.width=10}
#| code-summary: "▶ Simulate: CRA vs. stratified assignment for 500 experiments"

set.seed(2026)
N_str  <- 200
D_str  <- 3    # three binary covariates to stratify on

n_sims <- 500

# Helper: absolute SMD for a covariate
abs_smd <- function(x, trt) {
  m1 <- mean(x[trt == 1]); m0 <- mean(x[trt == 0])
  s  <- sqrt((var(x[trt == 1]) + var(x[trt == 0])) / 2)
  if (s == 0) return(0)
  abs(m1 - m0) / s
}

# Simulate one run under CRA and one under stratified
one_run <- function() {
  # Participant covariates (binary)
  X1 <- rbinom(N_str, 1, 0.45)  # environmental values
  X2 <- rbinom(N_str, 1, 0.50)  # income
  X3 <- rbinom(N_str, 1, 0.40)  # health focus

  # CRA
  trt_cra  <- sample(rep(0:1, N_str / 2))

  # Stratified: randomize within each of 2^3 = 8 strata
  strata   <- paste(X1, X2, X3, sep = "-")
  trt_strat <- integer(N_str)
  for (s in unique(strata)) {
    idx <- which(strata == s)
    n_s <- length(idx)
    trt_strat[idx] <- sample(c(rep(0, ceiling(n_s / 2)),
                                rep(1, floor(n_s / 2)))[seq_len(n_s)])
  }

  data.frame(
    method = c(rep("CRA", 3L), rep("Stratified", 3L)),
    cov    = rep(c("Env. Values", "Income", "Health Focus"), 2L),
    smd    = c(abs_smd(X1, trt_cra),  abs_smd(X2, trt_cra),  abs_smd(X3, trt_cra),
               abs_smd(X1, trt_strat), abs_smd(X2, trt_strat), abs_smd(X3, trt_strat))
  )
}

set.seed(42)
strat_sims <- bind_rows(replicate(n_sims, one_run(), simplify = FALSE))

strat_summary <- strat_sims |>
  group_by(method, cov) |>
  summarise(mean_smd = mean(smd), p90_smd = quantile(smd, 0.90), .groups = "drop") |>
  mutate(method = factor(method, levels = c("CRA", "Stratified")))

ggplot(strat_summary, aes(x = cov, y = mean_smd, fill = method,
                           ymin = 0, ymax = p90_smd)) +
  geom_col(position = position_dodge(0.7), width = 0.6, alpha = 0.85) +
  geom_errorbar(aes(ymin = mean_smd, ymax = p90_smd),
                position = position_dodge(0.7), width = 0.25, linewidth = 0.8) +
  geom_hline(yintercept = 0.10, linetype = "dashed", color = "gray40", linewidth = 0.8) +
  annotate("text", x = 3.45, y = 0.105, label = "|SMD| = 0.10\nthreshold",
           size = 3.0, color = "gray40", hjust = 0, lineheight = 0.85) +
  scale_fill_manual(values = c("CRA" = "#e63946", "Stratified" = "#52b788"), name = NULL) +
  labs(x = "Covariate", y = "Mean |SMD| (bar) + 90th percentile (error bar)",
       title = "Stratified vs. Completely Random Assignment: Balance on Stratified Covariates",
       subtitle = paste0("N = ", N_str, "  \u00b7  500 simulated experiments  \u00b7  ",
                         "Bars = mean |SMD|  \u00b7  Error bars = 90th percentile worst case")) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(), legend.position = "top")
```

Stratification dramatically reduces both the mean and worst-case imbalance on the stratified covariates. The 90th-percentile bar for CRA often exceeds the 0.10 threshold; the stratified version almost never does.

**The limitation** is the same as in Module 1's measurement section: you can only stratify on what you have *measured*. Unstratified dimensions remain at the mercy of chance — which returns us to the need to map the construct Y first.

### Clustered Random Assignment

**The idea:** sometimes participants come in natural groups (classrooms, teams, branches, households), and the intervention must be assigned at the group level. In this case, you assign entire clusters to conditions.

**Concrete example:** You want to test whether providing eco-label information changes café purchasing behavior. You cannot assign individual customers within a café to different menu conditions simultaneously — contamination would occur. Instead, you assign *cafés* to conditions.

**The tradeoff:** clustering increases statistical efficiency when you have to intervene at the group level, but it also introduces **intraclass correlation (ICC)**. If people within the same café are more similar to each other (they self-select into similar cafés), then your N cafés provide less independent information than N individuals would.

```{r clustered-random, fig.height=4.5, fig.width=9}
#| code-summary: "▶ Illustrate: effective N under clustered assignment for different ICCs"

icc_vals     <- c(0.00, 0.05, 0.10, 0.20, 0.30, 0.50)
cluster_sizes <- c(5, 10, 20, 40)

deff_df <- expand.grid(icc = icc_vals, m = cluster_sizes) |>
  mutate(
    DEFF    = 1 + (m - 1) * icc,        # design effect
    Neff_80 = ceiling(
      power.t.test(delta = 0.40, sd = 1, sig.level = 0.05, power = 0.80,
                   type = "two.sample")$n * DEFF
    ),
    m_label = paste0("Cluster size = ", m)
  )

ggplot(deff_df, aes(x = icc, y = Neff_80,
                    color = factor(m), group = factor(m))) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.8) +
  geom_hline(yintercept = power.t.test(delta = 0.40, sd = 1, sig.level = 0.05,
                                        power = 0.80, type = "two.sample")$n,
             linetype = "dashed", color = "gray40", linewidth = 0.7) +
  annotate("text", x = 0.51, y = power.t.test(delta = 0.40, sd = 1, sig.level = 0.05,
                                                power = 0.80, type = "two.sample")$n + 3,
           label = "Required N\nif individual\nassignment",
           size = 2.8, color = "gray40", hjust = 0, lineheight = 0.85) +
  scale_color_manual(values = c("5" = "#4a90d9", "10" = "#52b788",
                                 "20" = "#f4a261", "40" = "#e63946"),
                     name = "Cluster size (m)") +
  scale_x_continuous(labels = function(x) sprintf("ICC = %.2f", x)) +
  scale_y_continuous(labels = comma) +
  labs(x = "Intraclass correlation (ICC)",
       y = "Required total N (individuals)",
       title = "Clustered Assignment: How ICC Inflates Required Sample Size",
       subtitle = "DEFF = 1 + (m\u22121)\u00d7ICC  \u00b7  Design effect multiplies into required N  \u00b7  Target: d = 0.40, 80% power") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

::: {.callout-note}
## Stratified vs. Clustered — the key contrast

| Feature | Stratified | Clustered |
|---|---|---|
| Unit of randomization | Individual (within strata) | Group / cluster |
| Effect on balance | **Improves** balance on stratified vars | No improvement (can worsen) |
| Effect on required N | Reduces N slightly | **Increases** N via design effect |
| When to use | When measured covariates predict Y strongly | When intervention must be applied at group level |
| Analogy (Module 1) | Pre-specifying subscales before CFA | Multi-level data with non-independent observations |

The core message: stratification and clustering both depart from simple CRA, but in opposite directions for power. Use stratification proactively when you can; account for clustering carefully when you must.
:::

---

## Statistical Solutions: Correcting for Broken Exchangeability After Data Collection

Even with careful design, selection and attrition can still break exchangeability. Several statistical models attempt to recover causal estimates when this happens.

### Heckman Selection Models

::: {.callout-warning}
## The critical caveat — read this before the example

The Heckman model works **only** if you have a valid **instrument for selection** — a variable that (1) predicts whether someone participates, but (2) has **no direct effect** on the outcome Y. Finding such an instrument is the hardest part of applying a Heckman correction. When the instrument is weak or its exclusion restriction is implausible, the Heckman correction can produce estimates that are *more* biased than naive OLS.

**What makes a good vs. bad instrument?**

| | Good instrument | Bad instrument |
|---|---|---|
| **Example** | Random financial incentive to complete the survey ($0–$4 bonus) | Whether the participant's friend also signed up |
| **Predicts participation?** | Yes — higher bonus → more likely to complete | Yes — social influence → more likely to participate |
| **Affects WTP directly?** | No — the cash bonus is for *completing* the survey, not for the product | Yes — friends share consumption preferences, so friend's sign-up correlates with eco-mindedness, which affects WTP |
| **Why it works / fails** | The incentive shifts *who bothers to participate* without changing how they evaluate the eco-label once in the study | The instrument correlates with the unobservable (eco-mindedness) you are trying to control for — it violates the exclusion restriction |

**A further technical requirement:** the instrument should be **continuous** (or at least multi-valued). A binary instrument (e.g., reminder email yes/no) creates only a handful of unique covariate combinations in Stage 1, making the inverse Mills ratio (IMR) nearly a step-function of treatment — which causes severe collinearity in Stage 2 and destabilizes the coefficient estimates.

Always report the first-stage F-statistic (or probit z-score) for the instrument as evidence of its relevance. **Notice the irony:** the statistical fix for a broken exclusion restriction (Heckman) itself requires an exclusion restriction. The problems multiply as Y becomes more complex.
:::

**The problem:** suppose eco-label study participants were recruited via an environmental blog (self-selection). People who click through are systematically more environmentally conscious. The treatment effect you estimate confounds "eco-label response" with "environmentally conscious person response."

**The Heckman correction** (Heckman, 1979) models the selection process explicitly, using a variable called the **inverse Mills ratio (IMR)** to account for the selection into the study:

1. **Stage 1 (Selection equation):** Regress participation (1 = participated, 0 = declined) on all variables that predict selection but do *not* directly affect Y. This requires an **exclusion restriction in the selection model** — a variable that predicts participation but not WTP. In the simulation below, we use a continuous financial incentive ($0–$4 bonus) randomly assigned to potential participants.

2. **Stage 2 (Outcome equation):** Regress Y on X and the IMR from Stage 1. The IMR absorbs the selection-induced correlation between treatment assignment and unobserved factors.

```{r heckman-demo}
#| code-summary: "▶ Simulate Heckman correction for self-selection"

set.seed(42)
N_hk  <- 1000   # pool of potential participants (larger N → more stable Heckman)

# ── Unobserved confounder ─────────────────────────────────────────────────────
# theta = latent eco-mindedness: affects BOTH survey completion AND WTP.
# It is NOT in the data (hence "unobserved"), so naive OLS cannot control for it.
theta <- rnorm(N_hk)

# ── Treatment assignment (random) ─────────────────────────────────────────────
# Eco vs. control randomly assigned to all N_hk potential participants.
eco   <- sample(rep(0:1, N_hk / 2))

# ── Instrument ────────────────────────────────────────────────────────────────
# incentive: a continuous financial incentive ($0–$4 bonus) randomly assigned
# to potential participants.  Higher incentive → more likely to complete the
# survey, but the incentive itself has NO direct effect on WTP for the product.
# This is the EXCLUSION RESTRICTION — a continuous instrument is essential here
# because a binary instrument (e.g., reminder email yes/no) would create only
# 4 unique (eco × reminder) combinations in Stage 1, making the IMR nearly a
# step-function of eco in Stage 2 and causing collinearity that destabilises
# the coefficient estimates.
incentive <- runif(N_hk, 0, 4)   # continuous $0–$4 bonus

# ── Selection into completion ──────────────────────────────────────────────────
# Key mechanism: the eco condition is intrinsically interesting to eco-minded
# (high-theta) people BUT ALSO to some low-theta "curious" people — so the eco
# arm selects in more low-theta completers than the control arm (where only
# truly motivated people bother).  This creates DOWNWARD bias in naive OLS.
# Keeping theta and eco effects moderate ensures the Heckman correction is
# tractable and does not overshoot due to finite-sample IMR estimation error.
sel_latent <- -0.30 +
  0.55 * theta     +    # eco-minded people more likely to complete
  0.65 * eco       +    # eco topic attracts notably more lower-theta completers → bigger OVB
  0.55 * incentive +    # strong continuous instrument for identification
  rnorm(N_hk, 0, 1)
selected <- as.integer(sel_latent > 0)
N_sel    <- sum(selected)

# ── WTP (observed only for completers; theta is the unobserved driver) ─────────
# theta effect on WTP = 0.85: large enough to create detectable downward bias
# in naive OLS, small enough that Heckman correction is stable (doesn't overshoot).
wtp_true <- 5.0 +
  0.50 * eco[selected == 1] +      # TRUE eco-label effect = $0.50
  0.85 * theta[selected == 1] +    # theta drives WTP but is UNOBSERVED in df_hk
  rnorm(N_sel, 0, 1.0)
wtp_true <- pmax(wtp_true, 0)      # floor at 0 (no negative WTP)

# ── Analysis datasets ──────────────────────────────────────────────────────────
# df_hk:   observed data for completers (theta NOT included — it's unobserved)
# df_full: full pool used for Stage 1 probit (includes instrument)
df_hk   <- data.frame(wtp       = wtp_true,
                       eco       = eco[selected == 1],
                       incentive = incentive[selected == 1])
df_full <- data.frame(selected, eco, incentive)   # theta excluded — unobserved

# Naive OLS: ignores selection → eco coefficient is DOWNWARD BIASED because
# eco arm selects in more low-theta participants (lower average WTP)
naive <- lm(wtp ~ eco, data = df_hk)

# Heckman Stage 1: probit of completion on eco + incentive (instrument)
# Note: theta is excluded (unobserved); incentive is the identifying instrument.
# The continuous incentive creates a smooth IMR function → no collinearity.
stage1 <- glm(selected ~ eco + incentive, data = df_full,
              family = binomial(link = "probit"))

# Inverse Mills ratio (IMR) for completers — absorbs the selection-on-theta
xb  <- predict(stage1)[selected == 1]
imr <- dnorm(xb) / pnorm(xb)

# Heckman Stage 2: includes IMR to correct for selective attrition
heckman <- lm(wtp ~ eco + imr, data = cbind(df_hk, imr = imr))

cat(sprintf("Completion rate: %.1f%% overall  |  Eco: %.1f%%  |  Control: %.1f%%\n",
            100 * mean(selected),
            100 * mean(selected[eco == 1]),
            100 * mean(selected[eco == 0])))
cat(sprintf("Mean theta among completers — Eco: %.2f  |  Control: %.2f  (gap creates bias)\n",
            mean(theta[selected == 1 & eco == 1]),
            mean(theta[selected == 1 & eco == 0])))

heck_tab <- data.frame(
  Model = c("Naive OLS (ignores selection)", "Heckman 2-step (corrects attrition)"),
  `Eco-label coefficient` = c(round(coef(naive)["eco"], 3),
                               round(coef(heckman)["eco"], 3)),
  `95% CI (low)` = c(round(confint(naive)["eco", 1], 3),
                      round(confint(heckman)["eco", 1], 3)),
  `95% CI (high)` = c(round(confint(naive)["eco", 2], 3),
                       round(confint(heckman)["eco", 2], 3)),
  `True value` = 0.50,
  check.names = FALSE
)

kable(heck_tab,
      caption = paste0("Heckman correction: differential attrition creates downward bias in naive OLS. ",
                       "Eco arm attracts low-theta completers; control arm retains only high-theta ones. ",
                       "True eco-label effect = $0.50."))
```

Notice how the naive OLS estimate is **downward biased**: the eco condition attracted some low-theta ("curious but not eco-minded") completers who would not have bothered with the control condition. These participants bring down the mean WTP in the eco arm, understating the true eco-label effect. The control arm retained only high-theta (genuinely motivated) completers with high baseline WTP regardless. The Heckman correction uses the IMR to account for this differential attrition and recovers an estimate closer to the true $0.50 value.

### Censored Data Models (Tobit)

In the eco-coffee WTP study, participants could not state WTP below $1 (floor) or above $10 (ceiling). In many studies, the outcome is **censored** — truncated at a boundary because the measurement instrument or the context prevents observing values beyond a threshold.

**Consequence of ignoring censoring:** OLS on a censored outcome is biased and inconsistent. Observations piled at the ceiling (e.g., everyone stating exactly $10.00 WTP) look identical to OLS but represent unobserved heterogeneity (some would have stated $12 or $15 if allowed).

The **Tobit model** explicitly models the censoring threshold and recovers the latent distribution:

$$Y^* = X\beta + \varepsilon, \quad Y = \max(Y_{\text{lower}}, \min(Y^*, Y_{\text{upper}}))$$

```{r tobit-demo}
#| code-summary: "▶ Simulate: OLS vs. Tobit on censored WTP data"

if (!requireNamespace("AER", quietly = TRUE)) install.packages("AER")
library(AER)

set.seed(2026)
N_tb   <- 1200   # large N for stable Tobit estimates (Tobit MLE is less efficient than OLS)
eco_tb <- rep(0:1, each = N_tb / 2)

# ── Latent (true) WTP ─────────────────────────────────────────────────────────
# True eco-label effect = $0.90 on the latent scale.
# Mean = 1.2 (control arm) with sd = 1.8 places ~25% of latent WTPs below $0.
# High censoring creates clear OLS attenuation; N = 1200 makes Tobit converge
# reliably to 0.90 without overshoot.
wtp_lat <- 1.2 + 0.90 * eco_tb + rnorm(N_tb, 0, 1.8)

# Observed WTP: censored at $0 (participants state "I would not pay anything")
wtp_obs      <- pmax(wtp_lat, 0)
pct_censored <- mean(wtp_obs == 0)

cat(sprintf("Floor-censored observations: %d / %d  (%.1f%%)\n",
            sum(wtp_obs == 0), N_tb, 100 * pct_censored))

# OLS on the censored observed values: attenuated toward zero
ols_tb <- lm(wtp_obs ~ eco_tb)

# Tobit: models the latent WTP as normal, recovering the true coefficient
tobit_tb <- tobit(wtp_obs ~ eco_tb, left = 0)

cens_tab <- data.frame(
  Model    = c("OLS (ignores censoring)", "Tobit (corrects for censoring)"),
  `Eco coefficient` = c(round(coef(ols_tb)["eco_tb"], 3),
                         round(coef(tobit_tb)["eco_tb"], 3)),
  `True value` = 0.90,
  `% censored` = paste0(round(pct_censored * 100, 1), "%"),
  check.names = FALSE
)

kable(cens_tab,
      caption = paste0("OLS vs. Tobit on floor-censored WTP (true eco effect = $0.90). ",
                       "~20% floor censoring attenuates OLS; Tobit recovers the latent coefficient."))
```

::: {.callout-note}
## Looking ahead to Module 3

The selection and censored-data models above are a first pass at a much richer literature. In Module 3 we will formalize the causal inference frameworks (potential outcomes / Rubin causal model, DAGs / Pearl's do-calculus) that make precise which assumptions each model requires, and introduce more general solutions:

- **Covariate matching** — pairing treated and control units on observed characteristics to reduce confounding
- **Propensity score methods** (propensity matching, inverse probability weighting, doubly-robust estimators) for observational settings where selection is driven by observed covariates
- **Instrumental variables (IV)** for endogenous regressors — the general solution to the exclusion restriction problem in Part 2
- **Difference-in-differences and panel methods** for settings with longitudinal structure
- **Regression discontinuity** for settings where treatment is assigned by a threshold rule
- **Mediation analysis** — formalizing the decomposition approach we introduced in this module, with careful attention to the causal assumptions required for valid mediation inference

The language we have built here — exchangeability, exclusion restriction, discriminant validity of X, nomological net of Y — will carry through directly into that causal inference framework.
:::

---

# Part 5: Experimental Design Strategies — Using Structure to Buy Efficiency {#sec-designs}

## The Core Tension

Parts 3 and 4 showed a fundamental tension:

- **More dimensions of Y** → larger design space → more N needed for orthogonality
- **More N** → more expensive → temptation to reduce sample → back to poor orthogonality

Section 4 offered two partial design-based solutions (stratification, clustering). Here we go further: rather than trying to *achieve* orthogonality with brute-force N, we ask whether **the experimental design itself** can reduce the size of the design space you need to fill.

## Completely Random, Factorial, and Latin Square Designs

The simplest way to see the tradeoffs is to contrast three design types on a common example.

**Scenario:** You want to test the effect of:

- **Factor A:** Eco-certification badge (present / absent) — 2 levels
- **Factor B:** Price level (low / medium / high) — 3 levels
- **Factor C:** Message framing (individual benefit / social benefit) — 2 levels

```{r design-comparison, fig.height=6, fig.width=11}
#| code-summary: "▶ Visualize: CRD vs. Factorial vs. Latin Square cell structures"

# ── Fully factorial: all A×B×C combinations ──────────────────────────────────
fact_df <- expand.grid(
  EcoBadge  = c("No Badge", "Eco Badge"),
  PriceLevel = c("Low", "Medium", "High"),
  Framing    = c("Individual\nBenefit", "Social\nBenefit")
) |>
  mutate(cell = paste(EcoBadge, PriceLevel, Framing, sep = "\n"),
         design = "Fully Factorial\n(12 cells, 12N total)")

# ── Latin square: Factor C varied across rows and Factor A×B as column×row
# 2×3 Latin Square (accommodate unequal levels by using 3×3 and dropping one row)
ls_grid <- expand.grid(Row = 1:2, Col = 1:3) |>
  mutate(
    EcoBadge   = ifelse(Col %in% c(1, 3), "No Badge", "Eco Badge"),
    PriceLevel = c("Low","Medium","High","High","Low","Medium"),
    Framing    = c("Individual\nBenefit","Social\nBenefit","Individual\nBenefit",
                   "Social\nBenefit","Individual\nBenefit","Social\nBenefit"),
    design     = "Latin Square\n(6 cells, 6N total)"
  )

# ── CRD: just two cells (eco vs. no eco, collapsing B and C) ─────────────────
crd_df <- data.frame(
  EcoBadge   = c("No Badge", "Eco Badge"),
  PriceLevel = "Pooled\n(B,C varied\nbut uncontrolled)",
  Framing    = "Pooled",
  design     = "Simple A/B (CRA)\n(2 cells, N total)",
  Row = 1, Col = 1:2
)

# ── Plot factorial and latin square side by side ──────────────────────────────
p_fact <- ggplot(fact_df, aes(x = PriceLevel, y = Framing, fill = EcoBadge)) +
  geom_tile(color = "white", linewidth = 1.2, width = 0.95, height = 0.95) +
  geom_text(aes(label = paste0("n\u2265", ifelse(EcoBadge == "Eco Badge", "1", "1"))),
            size = 3.5, fontface = "bold", color = "gray20") +
  scale_fill_manual(values = c("No Badge" = "#d6ecff", "Eco Badge" = "#2d6a4f"),
                    name = "Eco Badge") +
  labs(title = "Fully Factorial (12 conditions)",
       subtitle = "Every combination tested  \u00b7  Min N = 12\nTests A, B, C main effects AND interactions",
       x = "Price Level", y = "Message Framing") +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(), legend.position = "top")

p_ls <- ggplot(ls_grid, aes(x = factor(Col), y = factor(Row), fill = EcoBadge)) +
  geom_tile(color = "white", linewidth = 1.5, width = 0.92, height = 0.92) +
  geom_text(aes(label = paste0(EcoBadge, "\n", PriceLevel, "\n", Framing)),
            size = 2.9, lineheight = 0.85, color = "gray10") +
  scale_fill_manual(values = c("No Badge" = "#d6ecff", "Eco Badge" = "#2d6a4f"),
                    name = "Eco Badge") +
  labs(title = "Latin Square (6 conditions)",
       subtitle = "Each factor level appears exactly once per row and column\nMin N = 6  \u00b7  Tests main effects only (no interactions)",
       x = "Column (order slot / price)", y = "Row (participant block)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(), legend.position = "top",
        axis.text = element_text(size = 9))

p_fact + p_ls
```

```{r design-efficiency-table}
#| code-summary: "▶ Table: design comparison on key dimensions"

design_tab <- data.frame(
  Design = c("Simple A/B (CRA)", "Fully Factorial", "Latin Square"),
  `Conditions` = c(2, 12, 6),
  `Min N (1 obs/cell)` = c(2, 12, 6),
  `Main effects testable` = c("A only", "A, B, C", "A, B, C"),
  `Interactions testable` = c("None", "A\u00d7B, A\u00d7C, B\u00d7C, A\u00d7B\u00d7C", "None"),
  `Orthogonality coverage` = c("Only A vs. B,C confounded", "Perfect within cells", "Approximate — by design"),
  check.names = FALSE
)

kable(design_tab,
      caption = "Design comparison: efficiency, testability, and orthogonality coverage")
```

::: {.callout-note}
## Connecting to Section 3: how designs reduce the effective design space

Recall from Section 3.6 that the required N for approximate orthogonality grows with the number of dimensions D and falls as $r_{\max}$ increases. A Latin Square design directly reduces the effective D by **structurally balancing** factors across the design matrix rather than relying on chance:

- A fully factorial 2×3×2 design with N = 48 (4 per cell) achieves exact balance on A, B, and C by construction.
- A Latin Square of the same factors with N = 24 achieves approximate balance by ensuring each factor level appears an equal number of times across the other factors' levels.
- A simple A/B study with N = 48 leaves B and C entirely to chance — back to the orthogonality problem from Section 3.

The Latin Square is a direct structural solution to the Section 3 problem: **it buys you orthogonality through design structure rather than through sample size.**
:::

## Interactive Latin Square Design Builder

Use the code below to generate and visualize your own Latin Square design. Each number in `levels_per_factor` represents one factor and its number of levels. Changing the vector changes the design.

```{r latin-square-builder, fig.height=5, fig.width=8}
#| code-summary: "▶ Latin Square builder — edit levels_per_factor to explore your own design"

# ── Design generator ─────────────────────────────────────────────────────────
generate_latin_design <- function(levels_per_factor) {
  maxL   <- max(levels_per_factor)
  design <- expand.grid(Row = seq_len(maxL), Column = seq_len(maxL))
  for (i in seq_along(levels_per_factor)) {
    L   <- levels_per_factor[i]
    mat <- matrix(NA, nrow = maxL, ncol = maxL)
    for (r in seq_len(maxL)) mat[r, ] <- ((0:(maxL - 1) + (r - 1)) %% L) + 1
    design[[paste0("Factor_", LETTERS[i])]] <- factor(mat[cbind(design$Row, design$Column)])
  }
  invisible(design)
}

# ── Visualizer ───────────────────────────────────────────────────────────────
plot_latin_design <- function(design_df, factor_names = NULL) {
  fac_cols  <- grep("^Factor_", names(design_df), value = TRUE)
  n_factors <- length(fac_cols)
  if (!is.null(factor_names) && length(factor_names) == n_factors)
    names(design_df)[match(fac_cols, names(design_df))] <- factor_names

  fa <- names(design_df)[3]   # first factor = tile fill

  p <- ggplot(design_df, aes(x = Column, y = Row, fill = .data[[fa]])) +
    geom_tile(color = "white", linewidth = 1.1) +
    scale_y_reverse() + coord_equal() +
    scale_fill_brewer(palette = "Set2", name = fa) +
    theme_minimal(base_size = 13) +
    theme(panel.grid = element_blank()) +
    labs(x = "Column (order slot)", y = "Row (participant block)",
         title = "Latin Square Design",
         subtitle = paste0("Factors: ", paste(names(design_df)[-(1:2)], collapse = ", "),
                           "  \u00b7  Total cells: ", nrow(design_df)))

  if (n_factors >= 2) {
    facet_vars    <- names(design_df)[-(1:3)]
    facet_vars_bt <- paste0("`", facet_vars, "`")  # backtick-quote names that may contain spaces
    p <- p + facet_wrap(reformulate(facet_vars_bt), labeller = label_both)
  }
  p
}

# ── Try your own design here ─────────────────────────────────────────────────
# c(2, 3, 2)  = Factor A has 2 levels, Factor B has 3 levels, Factor C has 2 levels
# c(4, 4)     = Two factors, each with 4 levels (perfect 4×4 square)
# c(2, 2, 2)  = Three 2-level factors

my_design <- generate_latin_design(c(2, 3, 2))
plot_latin_design(my_design,
                  factor_names = c("Eco Badge", "Price Level", "Framing"))
```

```{r latin-square-summary}
#| code-summary: "▶ Print the design table"
kable(my_design,
      caption = paste0("Latin Square design table  \u00b7  ",
                       nrow(my_design), " unique cells  \u00b7  ",
                       "Each row = one participant block or sequence"))
```

### Analysis code for Latin Square designs

```{r latin-analysis-code, eval=FALSE}
#| code-summary: "▶ Analysis template: fit Latin Square model"

# Replace Y, Factor_A, Factor_B, Factor_C with your actual variable names.
# Row and Column are nuisance blocking variables — always include them.

# Two-factor model
fit2 <- lm(Y ~ Factor_A + Factor_B + Row + Column, data = your_df)
emmeans(fit2, pairwise ~ Factor_A, adjust = "holm")$contrasts
emmeans(fit2, pairwise ~ Factor_B, adjust = "holm")$contrasts

# Three-factor model
fit3 <- lm(Y ~ Factor_A + Factor_B + Factor_C + Row + Column, data = your_df)
emmeans(fit3, pairwise ~ Factor_A, adjust = "holm")$contrasts
emmeans(fit3, pairwise ~ Factor_B, adjust = "holm")$contrasts
emmeans(fit3, pairwise ~ Factor_C, adjust = "holm")$contrasts

# Note: Latin Square designs cannot test interactions between factors.
# If interactions are theoretically important, use a full factorial design.
```

---

## Metastudies: Pooling Power Across Multiple Factors

The Latin Square is efficient because it uses design structure to study multiple factors simultaneously. A related idea scales this up to the level of a research program: the **metastudy** (DeKay, Rubinchik, Li, & De Boeck, 2022, *Perspectives on Psychological Science*).

A metastudy is a single, pre-registered study that simultaneously tests many experimental factors — what DeKay et al. call *facets* — by crossing them in a balanced factorial design. Each participant sees one randomly sampled combination of facet levels (a *microstudy*); the researcher estimates each facet's main effect by contrasting all participants who received the high level of that facet against all who received the low level, averaged across all other facets.

**Why this provides more power per hypothesis than running separate A/B tests:**

The key insight is about how the fixed participant budget is allocated. In a metastudy, all N participants are exposed to all K factors simultaneously. When estimating the effect of any one factor, the *entire* sample contributes to that estimate — N/2 participants in each "level" of that factor — regardless of how many other factors are also being tested. This is because the factors are orthogonally assigned, so each factor's effect is estimated independently of the others using the full N.

By contrast, if you run K separate A/B tests and divide your budget equally, each test gets only N/K participants. Power for each test depends on N/K, which shrinks as K grows.

**The comparison with fixed total N:** if you have N = 400 participants and want to test K hypotheses:
- **Metastudy:** all 400 test all K factors. Each factor is estimated using all 400 participants → power stays constant.
- **Separate A/B tests:** 400/K participants per factor → as K grows, per-test sample shrinks → power falls.

At K = 1, the two designs are *identical*: a single A/B test with 400 participants. As K grows, they diverge sharply. As DeKay et al. (2022) put it, metastudies "intentionally and completely sacrifice power at the level of the individual microstudies to achieve greater power for testing the effects of interest at the aggregate level" — a counterintuitive but mathematically principled trade-off.

```{r metastudy-power, fig.height=5.5, fig.width=12}
#| code-summary: "▶ Simulate: metastudy maintains power as K grows; separate A/B tests collapse"

set.seed(2026)
K_vec   <- 1:7       # 1 to 7 factors (stop at K = 7)
N_total <- 400       # fixed total budget (same for both designs)
d       <- 0.30      # true effect size per factor
n_sim   <- 1000      # simulations per cell

# ── Metastudy: all N=400 test all K factors simultaneously (orthogonal OLS) ────
# Key: each factor is assigned orthogonally, so estimating Factor 1 uses all N
# participants (N/2 per level), regardless of K. SE ≈ 2*sigma/sqrt(N) = constant.
# At K=1 this is identical to a single A/B test with N participants.
sim_meta_power <- function(N, K, d, n_sim = 500) {
  mean(replicate(n_sim, {
    # Random ±0.5 assignment to each of K orthogonal factors
    X   <- matrix(sample(c(-0.5, 0.5), N * K, replace = TRUE), nrow = N)
    Y   <- X %*% rep(d, K) + rnorm(N, 0, 1)
    fit <- lm(Y ~ X)
    # p-value for Factor 1 (all factors symmetric by design)
    coef(summary(fit))[2, 4] < 0.05
  }))
}

# ── Separate A/B tests: total N = 400 split EQUALLY among K tests ─────────────
# Each of the K factors gets its own study with N/K participants.
# At K=1: N_per_test = 400 → identical to metastudy (K=1 case) → equal power.
# As K grows: N_per_test = 400/K shrinks → power falls.
sim_ab_power <- function(N_per_test, d, n_sim = 500) {
  n_t <- ceiling(N_per_test / 2)
  n_c <- floor(N_per_test / 2)
  mean(replicate(n_sim, {
    obs <- mean(rnorm(n_t,  d / 2, 1)) - mean(rnorm(n_c, -d / 2, 1))
    se  <- sqrt(1/n_t + 1/n_c)
    abs(obs / se) > qnorm(0.975)
  }))
}

meta_powers <- sapply(K_vec, function(K) sim_meta_power(N_total, K, d, n_sim))
ab_powers   <- sapply(K_vec, function(K) sim_ab_power(floor(N_total / K), d, n_sim))

power_df <- data.frame(
  n_factors = rep(K_vec, 2),
  power     = c(meta_powers, ab_powers),
  design    = rep(c("Metastudy (total N = 400; all factors tested simultaneously)",
                    "Separate A/B tests (N = 400 / K per test)"),
                 each = length(K_vec))
)

ggplot(power_df, aes(x = n_factors, y = power,
                     color = design, group = design)) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 3.5) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "gray40") +
  annotate("text", x = 7.05, y = 0.815, label = "80% power",
           color = "gray40", size = 3.2, hjust = 0) +
  scale_color_manual(
    values = c("Metastudy (total N = 400; all factors tested simultaneously)" = "#2d6a4f",
               "Separate A/B tests (N = 400 / K per test)"                   = "#e63946"),
    name = NULL) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
  scale_x_continuous(breaks = K_vec) +
  labs(x = "Number of factors tested (K)",
       y = "Power per factor (d = 0.30)",
       title = "Metastudy Maintains Power as K Grows; Separate A/B Tests Collapse",
       subtitle = paste0(
         "Both designs use the same total N = 400  \u00b7  d = 0.30  \u00b7  At K=1 both designs are identical (one A/B test, N=400)\n",
         "Metastudy: all 400 contribute to each factor\u2019s estimate \u2192 power stays constant  \u00b7  ",
         "Separate A/B: 400/K per test \u2192 power falls")) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(), legend.position = "top",
        legend.text = element_text(size = 9),
        plot.margin = margin(5, 30, 5, 5))
```

**Reading the plot:** at K = 1, both designs are literally the same study — one A/B test with 400 participants — and achieve the same power (~85%). As K grows, the designs diverge. The metastudy *maintains* near-constant high power because all 400 participants contribute to each factor's estimate; the orthogonal design means Factor 1's SE is always ≈ 2/√N regardless of how many other factors are also being tested. Separate A/B tests collapse because the budget is divided: at K = 7, each test gets only 400/7 ≈ 57 participants, pushing power below 20%. The metastudy tests 7× as many hypotheses at the same power level using exactly the same resources.

**The connection to Section 3:** a metastudy also directly reduces the Section 3 problem. By deliberately varying many potential confounders as factors in the design (rather than leaving them to chance), you ensure they are orthogonally distributed across conditions *by construction* — transforming unobserved confounders into *measured experimental factors* and extracting their effects cleanly.

---

## Within-Participant Designs: Same Observations, More Power

### The Intuition

In all the designs above, we compared *different people* in different conditions. A within-participant (repeated-measures) design instead exposes *the same person* to multiple conditions, measuring the outcome each time.

**Why does this increase power?** Consider the source of variance in your data:

$$\text{Total variance in Y} = \underbrace{\text{Between-person variance}}_{\text{individual differences}} + \underbrace{\text{Within-person variance}}_{\text{response to conditions}} + \underbrace{\text{Measurement error}}_{\text{noise}}$$

In a between-participant design, your treatment effect estimate must be separated from *both* between-person variance and measurement error. Between-person variance is typically the largest component — people simply differ from each other a lot, regardless of condition.

In a within-participant design, between-person variance **cancels out** because each person serves as their own control. You are estimating the effect within each person and averaging those within-person effects. The only remaining noise is within-person variance and measurement error — which are much smaller.

```{r within-between-power, fig.height=5.5, fig.width=10}
#| code-summary: "▶ Simulate: between vs. within participant power comparison"

set.seed(2026)

# Parameters
true_effect <- 0.40   # true within-person effect (SD units)
icc_range   <- seq(0.10, 0.90, by = 0.10)  # ICC = between / total variance

power_comparison <- expand.grid(
  icc    = icc_range,
  design = c("Between-participant", "Within-participant")
) |>
  mutate(
    # For fixed total N = 100, n_per_arm = 50 between, N = 50 for within
    N        = 100,
    n_arm    = 50,
    # Between: variance includes between-person component (inflates denominator)
    sd_eff_b = sqrt(1),             # full SD
    # Within: SD of the DIFFERENCE scores = sqrt(2*(1-icc)) * total_sd
    sd_eff_w = sqrt(2 * (1 - icc)),
    power_val = mapply(function(des, sd_e) {
      tryCatch(
        power.t.test(n = 50, delta = true_effect,
                     sd = sd_e, sig.level = 0.05,
                     type = ifelse(des == "Between-participant",
                                   "two.sample", "paired"))$power,
        error = function(e) NA_real_
      )
    }, design, ifelse(design == "Between-participant", sd_eff_b, sd_eff_w))
  )

ggplot(power_comparison, aes(x = icc, y = power_val,
                              color = design, group = design)) +
  geom_line(linewidth = 1.4) +
  geom_point(size = 3.2) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "gray40") +
  annotate("text", x = 0.91, y = 0.82, label = "80% power",
           color = "gray40", size = 3.2) +
  scale_color_manual(values = c("Between-participant" = "#e63946",
                                 "Within-participant"  = "#2d6a4f"),
                     name = NULL) +
  scale_x_continuous(labels = function(x) sprintf("ICC = %.2f\n(%.0f%% between-person)", x, x*100)) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(x = "Intraclass correlation (ICC = proportion of variance between people)",
       y = "Statistical power",
       title = "Within-Participant Designs Have Greater Power — Especially When ICC Is High",
       subtitle = paste0("Fixed N = 100 participants  \u00b7  True effect = d = ", true_effect,
                         "\nWithin-participant design removes between-person variance from the error term")) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(), legend.position = "top")
```

::: {.callout-note}
## Why ICC determines how much you gain

The **intraclass correlation (ICC)** measures what fraction of total variance in Y is due to stable between-person differences. When ICC is high:

- People differ a lot from each other (but are consistent with themselves across conditions)
- Between-participant designs "waste" most of their variance budget on individual differences
- Within-participant designs remove all of this, leaving only the small within-person residual

When ICC is low:
- People are not very consistent — their responses vary randomly from occasion to occasion
- Within-participant designs offer less of an advantage because the between-person variance that cancels out was small to begin with

**Practical benchmark:** for most psychological constructs (attitudes, preferences, perceptions), ICC values between 0.40 and 0.70 are common. In this range, within-participant designs typically require **half to one-third** the sample size of between-participant designs for equivalent power.
:::

### The connection to Section 3: within-participant designs and the design space

Within-participant designs also address the Section 3 problem in a specific and powerful way. Because each person sees *multiple* conditions, the entire design space for that person's dimensions is held constant across conditions — **by construction**. Between-participant designs rely on random assignment to balance the design space across people; within-participant designs remove the need for that balance entirely within a person's data.

The tradeoff is well known: **carryover effects** (condition A changes how the person responds to condition B), **demand characteristics** (participants figure out the hypothesis), and **fatigue or practice effects**. These are most severe when:

- The treatment induces lasting attitude change (hard to reverse)
- The conditions are highly transparent (easy to guess the hypothesis)
- The outcome requires learning or practice (performance tasks)

For short perceptual or preference tasks — including most WTP and product evaluation paradigms — within-participant designs are often feasible and substantially more efficient.

---

::: {.callout-note}
## Module 2 — the complete thread

Every concept in this module is a direct extension of Module 1, applied to the treatment variable and the act of randomization rather than the outcome variable and the act of measurement:

| Module 1 (Measurement of Y) | Module 2 (Manipulation of X and Randomization) |
|---|---|
| Scale items are observables for a latent construct | Treatment assignment is an observable for latent randomization |
| Discriminant validity: does Y measure *only* what it should? | Exclusion restriction: does X activate *only* the intended pathway? |
| Construct breadth determines how many items you need | Construct breadth of Y determines how much N you need for orthogonality |
| Correlated subscales: efficient but can hide distinct pathways | Correlated latent dimensions: can ease orthogonality OR cascade failures |
| Open-ended text reveals the nomological net of Y | Open-ended text identifies unmeasured confounders and latent dimensions |
| Stratified measurement (subscales by domain) | Stratified randomization (block by measured covariates) |
| Multi-level data (participants nested in groups) | Clustered randomization (interventions at group level, ICC inflates required N) |
| Narrow constructs: few items, easy reliability | Narrow Y: small D, easy orthogonality, A/B suffices |
| Broad constructs: many items, hard to achieve discriminant validity | Broad Y: large D, orthogonality requires large N or structural design (Latin Square, metastudy) |

**The common lesson** — across both modules — is that observables are imperfect indicators of latent properties, and the gap between the observable and the latent grows with construct complexity. Researchers who map the latent space carefully (Module 1), design their manipulations and sampling deliberately (Module 2), and use the right statistical corrections when design solutions are insufficient (Module 2, Part 4) give themselves the best chance of making valid causal claims. Module 3 will formalize all of this in the language of causal inference.
:::

---

*This tutorial is updated periodically. Please check for the latest version before your session.*
