---
title: "Part 1: What Is a Causal Effect?"
---

```{r setup, include=FALSE}
options(repos = c(CRAN = "https://cran.rstudio.com/"))

if (!requireNamespace("remotes",  quietly = TRUE)) install.packages("remotes")
if (!requireNamespace("synthdid", quietly = TRUE))
  remotes::install_github("synth-did/synthdid")

pkgs <- c(
  "tidyverse", "broom", "knitr", "patchwork",
  "conflicted",
  "MatchIt", "WeightIt", "cobalt",
  "rdrobust", "rddensity",
  "did",
  "mediation", "robmed",
  "estimatr", "sandwich", "lmtest"
)
for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p, dependencies = TRUE)
  library(p, character.only = TRUE)
}
library(synthdid)

library(conflicted)
conflicts_prefer(dplyr::select, dplyr::filter, dplyr::rename, dplyr::mutate,
                 dplyr::summarise, dplyr::arrange, dplyr::lag)

dir.create("figs", recursive = TRUE, showWarnings = FALSE)
knitr::opts_chunk$set(
  echo = TRUE, fig.width = 9, fig.height = 5,
  fig.path = "figs/", dev = "svg"
)

clr_eco     <- "#2d6a4f"
clr_ctrl    <- "#4a90d9"
clr_treated <- clr_eco

theme_mod3 <- function() {
  theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          strip.background = element_rect(fill = "#f0f4f8", colour = NA),
          legend.position  = "bottom")
}

# ── Lightweight DAG helper ─────────────────────────────────────────────────────
.dag <- function(nodes, solid = NULL, dashed = NULL,
                 unobserved = character(0), title = "") {
  nd <- do.call(rbind, lapply(names(nodes), function(nm)
    data.frame(id=nm, x=nodes[[nm]][1], y=nodes[[nm]][2], stringsAsFactors=FALSE)))
  nd$type <- ifelse(nd$id %in% unobserved, "unobs", "obs")
  segs <- function(pairs, lt) {
    if (is.null(pairs) || length(pairs)==0) return(NULL)
    do.call(rbind, lapply(pairs, function(ab) {
      fr <- nd[nd$id==ab[1],]; to <- nd[nd$id==ab[2],]
      dx <- to$x-fr$x; dy <- to$y-fr$y
      len <- sqrt(dx^2+dy^2)+1e-9; pad <- 0.12
      data.frame(x=fr$x+pad*dx/len, y=fr$y+pad*dy/len,
                 xend=to$x-pad*dx/len, yend=to$y-pad*dy/len,
                 lt=lt, stringsAsFactors=FALSE)
    }))
  }
  edf <- rbind(segs(solid,"solid"), segs(dashed,"dashed"))
  xs <- range(nd$x); ys <- range(nd$y)
  p <- ggplot() + theme_void() + ggtitle(title) +
    coord_cartesian(xlim=c(xs[1]-.3,xs[2]+.3), ylim=c(ys[1]-.3,ys[2]+.3), expand=FALSE) +
    theme(plot.title=element_text(size=10,face="bold",hjust=.5,margin=margin(b=6)),
          plot.background=element_rect(fill="grey98",colour="grey80",linewidth=.4),
          plot.margin=margin(10,16,10,16), legend.position="none")
  if (!is.null(edf) && nrow(edf)>0)
    p <- p +
      geom_segment(data=edf, aes(x=x,y=y,xend=xend,yend=yend,colour=lt,linetype=lt),
                   arrow=arrow(length=unit(.20,"cm"),type="closed"), linewidth=.85) +
      scale_colour_manual(values=c(solid="grey20",dashed="firebrick")) +
      scale_linetype_manual(values=c(solid="solid",dashed="dashed"))
  if (any(nd$type=="obs"))
    p <- p + geom_label(data=nd[nd$type=="obs",], aes(x=x,y=y,label=id),
                        size=4, fontface="bold", fill="white", colour="black",
                        label.size=.9, label.padding=unit(.4,"lines"))
  if (any(nd$type=="unobs"))
    p <- p + geom_label(data=nd[nd$type=="unobs",], aes(x=x,y=y,label=id),
                        size=4, fontface="bold.italic", fill="#fff3e0", colour="#b45309",
                        label.size=1.1, label.padding=unit(.4,"lines"),
                        label.r=unit(.5,"lines"))
  p
}

# ── Mediation extraction helper (handles robmed version differences) ───────────
rob_extract <- function(fit) {
  # point estimate: try $ab, then $indirect, then coef()["ab"]
  ab <- tryCatch(fit$ab,
    error = function(e) tryCatch(fit$indirect,
      error = function(e2) tryCatch(coef(fit)["ab"],
        error = function(e3) NA_real_)))
  # CI: try $ci (matrix or vector), then confint()
  ci <- tryCatch({
    r <- fit$ci
    if (is.matrix(r)) r[1,] else r
  }, error = function(e) tryCatch(confint(fit)["ab",],
    error = function(e2) c(NA_real_, NA_real_)))
  list(est=unname(ab), lo=unname(ci[1]), hi=unname(ci[2]))
}
```

# Part 1: What Do We Mean by "Causal Effect"? {#sec-m3-part1}

<iframe src="https://docs.google.com/presentation/d/e/2PACX-1vQ_Qbgn1mguAIxorFZHqsS0sGJXYUbyYw7yFBsyd37wKXQpQgrpBh5c3wWTxDEtvg/pubembed?start=false&loop=false&delayms=3000" width="100%" height="500px" frameborder="0" allowfullscreen="true" mozallowfullscreen="true" webkitallowfullscreen="true" title="Module 3 Part 1: Causal Effects"></iframe>

## Causal Inference is a Deductive Process {#sec-deductive}

::: {.callout-important}
## What this exercise assumes
The 64 structures shown below represent the complete space of possible causal relationships among three variables under **ideal conditions**: (1) **proper measurement** — X, M, and Y are each measured without discriminant validity failures or systematic error (Module 1), and (2) **no omitted variables** — these three variables are the only players in the system (Module 2). In real research, measurement error and hidden confounders expand this space considerably.
:::

Causal inference is not induction — you cannot *prove* a causal structure from data. It is **deduction by elimination**: you bring every tool in your methodological toolkit to bear and see which structures survive. The exercise below makes this concrete.

The grid shows **all 64 structurally distinct relationships** that can exist among three variables **X**, **M**, and **Y** — arranged as a triangle with X at the lower-left, Y at the lower-right, and M at the top. Each cell shows one possible causal structure. Arrows between any two variables can point forward (A → B), backward (A ← B), bidirectionally (A ↔ B, representing correlated unmeasured causes or feedback), or not exist at all.

Click the buttons below to apply research findings. **Each finding rules out the structures inconsistent with it** — those cells will be crossed out in red. After exhausting your toolkit, notice how many structures remain. That irreducible remainder is why causal inference is hard.

```{ojs}
//| echo: false
// ── Generate all 64 causal structures ─────────────────────────────────────────
// Arrow code per pair: 0=none, 1=A→B (forward), 2=A←B (backward), 3=A↔B (bidir)
// Three pairs: xm (X↔M), my (M↔Y), xy (X↔Y)
dag_states = {
  const arr = [];
  for (let xm = 0; xm < 4; xm++)
    for (let my = 0; my < 4; my++)
      for (let xy = 0; xy < 4; xy++)
        arr.push({ xm, my, xy });
  return arr;
}
```

```{ojs}
//| echo: false
// ── Consistency logic ──────────────────────────────────────────────────────────
// Returns true if state s is consistent with all active tool findings.
//
// Arrow code per pair: 0=none, 1=A→B (forward), 2=A←B (backward), 3=A↔B (bidir)
// Pairs: xm (X,M), my (M,Y), xy (X,Y)
//
// Tool logic (each button represents a POSITIVE finding that rules out
// structures that cannot produce that result):
//
//   reg  — regression Y ~ X + M, both significant → requires direct X-Y link (xy≠0)
//           AND direct M-Y link (my≠0). Does NOT require an X-M link — X and M can
//           be uncorrelated orthogonal predictors.
//   expX — X is experimentally manipulated       → X is exogenous: no arrows INTO X
//           (removes xm=2,3 and xy=2,3)
//   med  — mediation analysis (indirect effect)  → requires X-M link (xm≠0) AND M-Y
//           link (my≠0). Enforces no direct X-Y (xy=0) — pure mediation — unless
//           Reg is also active, in which case xy≠0 is already required by Reg and
//           the combination represents partial mediation (both paths present).
function dagConsistent(s, active) {
  const { xm, my, xy } = s;
  const any = v => v > 0;               // any connection at all

  // Experimental manipulation of X makes it exogenous: no arrows into X.
  // Rules out M→X (xm=2), X↔M (xm=3), Y→X (xy=2), X↔Y (xy=3).
  if (active.includes("expX") && (xm === 2 || xm === 3 || xy === 2 || xy === 3)) return false;

  for (const t of active) {
    if (t === "reg" && !(any(xy) && any(my))) return false; // direct X-Y and M-Y required
    if (t === "med") {
      if (!any(xm) || !any(my)) return false;               // X-M and M-Y links required
      // Pure mediation: no direct X-Y link. But if Regression is also active it
      // already requires xy≠0 (direct effect found), so the combination represents
      // partial mediation — relax xy=0 to keep the button set satisfiable.
      if (!active.includes("reg") && xy !== 0) return false;
    }
  }
  return true;
}
```

```{ojs}
//| echo: false
// ── Tool buttons ───────────────────────────────────────────────────────────────
viewof activeTools = {
  const TOOLS = [
    { id:"reg",  cat:"Regression", label:"Regression Y ~ X + M",  sub:"Both X and M directly predict Y (xy≠0 and my≠0)" },
    { id:"expX", cat:"Experiment", label:"Manipulate X",           sub:"X is exogenous — removes all arrows into X" },
    { id:"med",  cat:"Mediation",  label:"X → M → Y  indirect",   sub:"X–M and M–Y links required; pure mediation removes direct X–Y" },
  ];
  const ICONS = { Regression:"📊", Experiment:"🧪", Mediation:"⛓️" };
  const active = new Set();

  // ── Build the DOM entirely in JS (avoids html-template ID clashes) ──────────
  const wrap = document.createElement("div");
  Object.assign(wrap.style, {
    border:"1px solid #e2e8f0", borderRadius:"10px", padding:"16px 18px",
    background:"#f8fafc", margin:"16px 0"
  });

  // Header row
  const hdr = document.createElement("div");
  Object.assign(hdr.style, { display:"flex", justifyContent:"space-between",
    alignItems:"center", marginBottom:"12px" });
  const title = document.createElement("span");
  title.style.cssText = "font-weight:700;font-size:14px;color:#0f172a;";
  title.innerHTML = `🛠️ Research Tools &nbsp;<span style="font-weight:400;color:#64748b;font-size:12px;">— click a finding to apply it; inconsistent structures are crossed out</span>`;
  const resetBtn = document.createElement("button");
  resetBtn.textContent = "↺ Reset All";
  resetBtn.style.cssText = "padding:5px 12px;border-radius:6px;border:1px solid #94a3b8;background:#fff;cursor:pointer;font-size:12px;color:#475569;";
  hdr.appendChild(title); hdr.appendChild(resetBtn);
  wrap.appendChild(hdr);

  // Button grid — group by category
  const cats = [...new Set(TOOLS.map(t => t.cat))];
  cats.forEach(cat => {
    const row = document.createElement("div");
    row.style.cssText = "display:flex;flex-wrap:wrap;gap:7px;margin-bottom:8px;align-items:center;";
    const catLbl = document.createElement("span");
    catLbl.style.cssText = "font-size:11px;color:#94a3b8;min-width:80px;";
    catLbl.textContent = `${ICONS[cat]} ${cat}`;
    row.appendChild(catLbl);
    TOOLS.filter(t => t.cat === cat).forEach(t => {
      const b = document.createElement("button");
      b.style.cssText = "padding:7px 13px;border-radius:18px;border:2px solid #cbd5e1;background:#fff;cursor:pointer;font-size:12px;color:#334155;transition:all 0.15s;line-height:1.4;text-align:left;";
      b.innerHTML = `<strong>${t.label}</strong><br><span style="font-size:10px;color:#94a3b8;">${t.sub}</span>`;
      b.onclick = () => {
        if (active.has(t.id)) {
          active.delete(t.id);
          b.style.cssText = b.style.cssText.replace(/background:[^;]+/, "background:#fff")
                              .replace(/border-color:[^;]+/, "border-color:#cbd5e1")
                              .replace(/color:[^;]+/, "color:#334155");
          b.style.background = "#fff"; b.style.borderColor = "#cbd5e1"; b.style.color = "#334155";
        } else {
          active.add(t.id);
          b.style.background = "#dc2626"; b.style.borderColor = "#dc2626"; b.style.color = "#fff";
        }
        wrap.value = [...active];
        wrap.dispatchEvent(new Event("input", { bubbles: true }));
      };
      row.appendChild(b);
    });
    wrap.appendChild(row);
  });

  resetBtn.onclick = () => {
    active.clear();
    wrap.querySelectorAll("button:not(:first-of-type)").forEach(b => {
      b.style.background = "#fff"; b.style.borderColor = "#cbd5e1"; b.style.color = "#334155";
    });
    wrap.value = [];
    wrap.dispatchEvent(new Event("input", { bubbles: true }));
  };

  wrap.value = [];
  return wrap;
}
```

```{ojs}
//| echo: false
// ── Status counter ─────────────────────────────────────────────────────────────
dagStatusBar = {
  const n   = dag_states.filter(s => dagConsistent(s, activeTools)).length;
  const pct = Math.round(n / 64 * 100);
  const bar = document.createElement("div");
  Object.assign(bar.style, {
    display:"flex", alignItems:"center", gap:"12px",
    padding:"10px 16px", borderRadius:"8px", margin:"6px 0",
    background: n === 64 ? "#f0fdf4" : n === 1 ? "#f0fdf4" : "#fff7ed",
    border: `1px solid ${n === 64 ? "#bbf7d0" : n === 1 ? "#bbf7d0" : "#fed7aa"}`,
    fontSize:"13px"
  });
  const color = n <= 2 ? "#16a34a" : n <= 12 ? "#d97706" : "#dc2626";
  let msg = n === 64
    ? "Apply tools above to begin eliminating structures."
    : n === 1
    ? "✓ Uniquely identified! (This is rare even with all tools.)"
    : `${64 - n} structures eliminated — <strong style="color:#b45309">${n} remain consistent with your findings</strong>.`;
  bar.innerHTML = `
    <span style="font-size:22px;">🎯</span>
    <span>
      <strong style="color:${color};font-size:18px;">${n}</strong>
      <span style="color:#475569;"> / 64 structures consistent &nbsp;(${pct}%)</span>
      &nbsp;&nbsp;<span style="color:#6b7280;font-size:12px;">${msg}</span>
    </span>`;
  return bar;
}
```

```{ojs}
//| echo: false
// ── Main DAG grid ──────────────────────────────────────────────────────────────
dagGrid = {
  // Cell geometry
  const W = 86, H = 70, GAP = 4;
  // Node positions within each cell (origin = cell top-left)
  const NX = [15, 58], NY = [71, 58], NM = [43, 12];
  const NR = 9, APAD = 10;

  // Layout metrics
  const LPAD = 58, SGAP = 16;
  const SHEAD = 26, CHEAD = 16;
  const secW = 4 * (W + GAP) - GAP;   // 352
  const secH = 4 * (H + GAP) - GAP;   // 292
  const totalW = LPAD + 2 * secW + SGAP + 4;
  const totalH = 2 * (SHEAD + CHEAD + secH) + SGAP + 6;

  const svg = d3.create("svg")
    .attr("width", totalW).attr("height", totalH)
    .style("display", "block").style("overflow", "visible")
    .style("font-family", "system-ui, -apple-system, sans-serif");

  // Arrowhead markers — one normal, one faded
  svg.append("defs").html(`
    <marker id="dah"  viewBox="0 0 8 8" refX="6" refY="4"
            markerWidth="5" markerHeight="5" orient="auto">
      <path d="M0,1 L0,7 L8,4 z" fill="#374151"/>
    </marker>
    <marker id="dahf" viewBox="0 0 8 8" refX="6" refY="4"
            markerWidth="5" markerHeight="5" orient="auto">
      <path d="M0,1 L0,7 L8,4 z" fill="#d1d5db"/>
    </marker>
  `);

  // Draw a single directed arrow from (ax,ay) to (bx,by) with perpendicular offset
  function putArrow(g, ax, ay, bx, by, off, faded) {
    const dx = bx - ax, dy = by - ay;
    const len = Math.sqrt(dx * dx + dy * dy) || 1;
    const ux = dx / len, uy = dy / len;
    const ox = -uy * off, oy = ux * off;
    g.append("line")
      .attr("x1", ax + ux * APAD + ox).attr("y1", ay + uy * APAD + oy)
      .attr("x2", bx - ux * APAD + ox).attr("y2", by - uy * APAD + oy)
      .attr("stroke", faded ? "#d1d5db" : "#374151")
      .attr("stroke-width", faded ? 1.1 : 1.4)
      .attr("marker-end", `url(#${faded ? "dahf" : "dah"})`);
  }

  // Draw one cell at position (cx, cy) in group g
  function drawCell(g, s, ok, cx, cy) {
    const gc = g.append("g").attr("transform", `translate(${cx},${cy})`);
    const f = !ok;

    // Cell background
    gc.append("rect")
      .attr("width", W).attr("height", H).attr("rx", 4)
      .attr("fill",   f ? "rgba(220,38,38,0.07)" : "#ffffff")
      .attr("stroke", f ? "rgba(220,38,38,0.28)" : "#e2e8f0")
      .attr("stroke-width", 0.8);

    // XM arrows
    if (s.xm === 1) putArrow(gc, NX[0],NX[1], NM[0],NM[1], 0, f);
    if (s.xm === 2) putArrow(gc, NM[0],NM[1], NX[0],NX[1], 0, f);
    if (s.xm === 3) {
      putArrow(gc, NX[0],NX[1], NM[0],NM[1], 3.5, f);
      putArrow(gc, NM[0],NM[1], NX[0],NX[1], 3.5, f);
    }
    // MY arrows
    if (s.my === 1) putArrow(gc, NM[0],NM[1], NY[0],NY[1], 0, f);
    if (s.my === 2) putArrow(gc, NY[0],NY[1], NM[0],NM[1], 0, f);
    if (s.my === 3) {
      putArrow(gc, NM[0],NM[1], NY[0],NY[1], 3.5, f);
      putArrow(gc, NY[0],NY[1], NM[0],NM[1], 3.5, f);
    }
    // XY arrows
    if (s.xy === 1) putArrow(gc, NX[0],NX[1], NY[0],NY[1], 0, f);
    if (s.xy === 2) putArrow(gc, NY[0],NY[1], NX[0],NX[1], 0, f);
    if (s.xy === 3) {
      putArrow(gc, NX[0],NX[1], NY[0],NY[1], 4, f);
      putArrow(gc, NY[0],NY[1], NX[0],NX[1], 4, f);
    }

    // Node circles + labels
    const ns = f ? "#d1d5db" : "#374151";
    const nf = f ? "#f9fafb" : "#ffffff";
    const tf = f ? "#9ca3af" : "#111827";
    [[NX,"X"],[NY,"Y"],[NM,"M"]].forEach(([pos, lbl]) => {
      gc.append("circle")
        .attr("cx", pos[0]).attr("cy", pos[1]).attr("r", NR)
        .attr("fill", nf).attr("stroke", ns).attr("stroke-width", 1.5);
      gc.append("text")
        .attr("x", pos[0]).attr("y", pos[1] + 4)
        .attr("text-anchor", "middle").attr("font-size", 10).attr("font-weight", "700")
        .attr("fill", tf).text(lbl);
    });

    // Red X cross for ruled-out structures
    if (f) {
      gc.append("line")
        .attr("x1", 5).attr("y1", 5).attr("x2", W-5).attr("y2", H-5)
        .attr("stroke", "rgba(220,38,38,0.28)").attr("stroke-width", 1.5);
      gc.append("line")
        .attr("x1", W-5).attr("y1", 5).attr("x2", 5).attr("y2", H-5)
        .attr("stroke", "rgba(220,38,38,0.28)").attr("stroke-width", 1.5);
    }
  }

  // Section labels and colours
  const XY_TITLES  = ["X — Y  (no direct link)", "X → Y", "X ← Y", "X ↔ Y"];
  const XY_COLORS  = ["#334155", "#1d4ed8", "#7c3aed", "#0e7490"];
  const XM_LABELS  = ["X—M: —", "X→M", "M→X", "X↔M"];
  const MY_LABELS  = ["—", "M→Y", "Y→M", "M↔Y"];

  // Render 4 sections in a 2×2 layout; each section = one XY relationship type
  [0, 1, 2, 3].forEach((xyVal, si) => {
    const sCol = si % 2;
    const sRow = Math.floor(si / 2);
    const ox = LPAD + sCol * (secW + SGAP);
    const oy = sRow * (SHEAD + CHEAD + secH + SGAP);

    // Section header bar
    svg.append("rect")
      .attr("x", ox).attr("y", oy).attr("width", secW).attr("height", SHEAD - 2)
      .attr("rx", 4).attr("fill", XY_COLORS[xyVal]);
    svg.append("text")
      .attr("x", ox + secW / 2).attr("y", oy + 17)
      .attr("text-anchor", "middle").attr("fill", "white")
      .attr("font-size", 12).attr("font-weight", "700")
      .text(XY_TITLES[xyVal]);

    // Column headers (MY relationship)
    for (let my = 0; my < 4; my++) {
      svg.append("text")
        .attr("x", ox + my * (W + GAP) + W / 2).attr("y", oy + SHEAD + CHEAD - 3)
        .attr("text-anchor", "middle").attr("font-size", 8.5).attr("fill", "#64748b")
        .text(MY_LABELS[my]);
    }

    // Row labels (XM relationship) + cells
    for (let xm = 0; xm < 4; xm++) {
      const cy = oy + SHEAD + CHEAD + xm * (H + GAP);
      // Row label on the left edge of each section column
      svg.append("text")
        .attr("x", ox - 5).attr("y", cy + H / 2 + 4)
        .attr("text-anchor", "end").attr("font-size", 8.5).attr("fill", "#64748b")
        .text(XM_LABELS[xm]);
      for (let my = 0; my < 4; my++) {
        const cx = ox + my * (W + GAP);
        drawCell(svg, { xm, my, xy: xyVal }, dagConsistent({ xm, my, xy: xyVal }, activeTools), cx, cy);
      }
    }
  });

  // Axis label: rows = XM relationship
  svg.append("text")
    .attr("transform", `translate(8, ${totalH / 2}) rotate(-90)`)
    .attr("text-anchor", "middle").attr("font-size", 10).attr("fill", "#94a3b8")
    .text("← XM relationship (rows within each section) →");

  return svg.node();
}
```

### What if There Is an Unobserved Confounder? {#sec-confounded}

The exercise above assumed no hidden variables. Now introduce **U** — an unobserved variable that may have arrows to X, M, and/or Y but is never measured. No research tool can constrain U directly.

Each of the three U-pairs (U–X, U–M, U–Y) can be absent, forward, backward, or bidirectional — 4³ = **64 U configurations** for every one of the 64 X–M–Y structures, giving **4,096 total states**. Rather than draw all 4,096 cells, the grid below preserves the familiar 64 X–M–Y layout. Each cell's **blue shading** encodes how many of the 64 U configurations remain consistent with your findings: deeper blue = U is highly ambiguous for that structure; pale = almost no U configuration survives; crossed out = the X–M–Y structure is ruled out for *every* possible U configuration. The small dashed circle labelled **U** in each cell is a reminder that the confounder is always present but unobserved.

Notice in particular: pressing **Manipulate X** eliminates U→X (since X is now exogenous), but U's connections to M and Y remain entirely free — experimental control of X does not eliminate unmeasured confounding of the M–Y path.

```{ojs}
//| echo: false
// ── Consistency logic for confounded model ─────────────────────────────────
// Six pairs: xm, my, xy (same as before) plus ux (U,X), um (U,M), uy (U,Y)
// Arrow codes: 0=none, 1=A→B, 2=A←B, 3=A↔B
//
//   reg2  — regression Y ~ X + M     → any(xy) AND any(my); U unconstrained
//   expX2 — manipulate X             → X exogenous: xm≠2,3; xy≠2,3; ux≠2,3
//   med2  — mediation X→M→Y          → xm≠0, my≠0, xy=0 (unless reg2 active)
dagConsistent2 = (s, active) => {
  const { xm, my, xy, ux, um, uy } = s;
  const any = v => v > 0;
  if (active.includes("expX2") && (xm===2||xm===3||xy===2||xy===3||ux===2||ux===3)) return false;
  for (const t of active) {
    if (t === "reg2" && !(any(xy) && any(my))) return false;
    if (t === "med2") {
      if (!any(xm) || !any(my)) return false;
      if (!active.includes("reg2") && xy !== 0) return false;
    }
  }
  return true;
}
```

```{ojs}
//| echo: false
// Count how many of the 64 U configurations (ux, um, uy) are consistent
// for a fixed (xm, my, xy) state given the active button set.
countUConfigs = (xmyState, active) => {
  let n = 0;
  for (let ux = 0; ux < 4; ux++)
    for (let um = 0; um < 4; um++)
      for (let uy = 0; uy < 4; uy++)
        if (dagConsistent2({...xmyState, ux, um, uy}, active)) n++;
  return n;
}
```

```{ojs}
//| echo: false
// ── Tool buttons (confounded model) ────────────────────────────────────────
viewof activeTools2 = {
  const TOOLS = [
    { id:"reg2",  cat:"Regression", label:"Regression Y ~ X + M",  sub:"Both X and M directly predict Y (xy≠0 and my≠0)" },
    { id:"expX2", cat:"Experiment", label:"Manipulate X",           sub:"X is exogenous — removes arrows into X, including U→X" },
    { id:"med2",  cat:"Mediation",  label:"X → M → Y  indirect",   sub:"X–M and M–Y links required; pure mediation removes direct X–Y" },
  ];
  const ICONS = { Regression:"📊", Experiment:"🧪", Mediation:"⛓️" };
  const active = new Set();
  const buttons = [];

  const wrap = document.createElement("div");
  Object.assign(wrap.style, {
    border:"1px solid #e2e8f0", borderRadius:"10px", padding:"16px 18px",
    background:"#f8fafc", margin:"16px 0"
  });

  const hdr = document.createElement("div");
  Object.assign(hdr.style, { display:"flex", justifyContent:"space-between",
    alignItems:"center", marginBottom:"12px" });
  const title = document.createElement("span");
  title.style.cssText = "font-weight:700;font-size:14px;color:#0f172a;";
  title.innerHTML = `🛠️ Research Tools &nbsp;<span style="font-weight:400;color:#64748b;font-size:12px;">— U is always present but unobserved; no tool constrains it</span>`;
  const resetBtn = document.createElement("button");
  resetBtn.textContent = "↺ Reset All";
  resetBtn.style.cssText = "padding:5px 12px;border-radius:6px;border:1px solid #94a3b8;background:#fff;cursor:pointer;font-size:12px;color:#475569;";
  hdr.appendChild(title); hdr.appendChild(resetBtn);
  wrap.appendChild(hdr);

  const cats = [...new Set(TOOLS.map(t => t.cat))];
  cats.forEach(cat => {
    const row = document.createElement("div");
    row.style.cssText = "display:flex;flex-wrap:wrap;gap:7px;margin-bottom:8px;align-items:center;";
    const catLbl = document.createElement("span");
    catLbl.style.cssText = "font-size:11px;color:#94a3b8;min-width:80px;";
    catLbl.textContent = `${ICONS[cat]} ${cat}`;
    row.appendChild(catLbl);
    TOOLS.filter(t => t.cat === cat).forEach(t => {
      const b = document.createElement("button");
      b.style.cssText = "padding:7px 13px;border-radius:18px;border:2px solid #cbd5e1;background:#fff;cursor:pointer;font-size:12px;color:#334155;transition:all 0.15s;line-height:1.4;text-align:left;";
      b.innerHTML = `<strong>${t.label}</strong><br><span style="font-size:10px;color:#94a3b8;">${t.sub}</span>`;
      b.onclick = () => {
        if (active.has(t.id)) {
          active.delete(t.id);
          b.style.background = "#fff"; b.style.borderColor = "#cbd5e1"; b.style.color = "#334155";
        } else {
          active.add(t.id);
          b.style.background = "#dc2626"; b.style.borderColor = "#dc2626"; b.style.color = "#fff";
        }
        wrap.value = [...active];
        wrap.dispatchEvent(new Event("input", { bubbles: true }));
      };
      buttons.push(b);
      row.appendChild(b);
    });
    wrap.appendChild(row);
  });

  resetBtn.onclick = () => {
    active.clear();
    buttons.forEach(b => {
      b.style.background = "#fff"; b.style.borderColor = "#cbd5e1"; b.style.color = "#334155";
    });
    wrap.value = [];
    wrap.dispatchEvent(new Event("input", { bubbles: true }));
  };

  wrap.value = [];
  return wrap;
}
```

```{ojs}
//| echo: false
// ── Status counter (out of 4096) ────────────────────────────────────────────
dagStatusBar2 = {
  let total = 0;
  for (let xm = 0; xm < 4; xm++)
    for (let my = 0; my < 4; my++)
      for (let xy = 0; xy < 4; xy++)
        total += countUConfigs({xm, my, xy}, activeTools2);
  const pct = Math.round(total / 4096 * 100);
  const bar = document.createElement("div");
  Object.assign(bar.style, {
    display:"flex", alignItems:"center", gap:"12px",
    padding:"10px 16px", borderRadius:"8px", margin:"6px 0",
    background: total === 4096 ? "#f0fdf4" : total <= 192 ? "#f0fdf4" : "#fff7ed",
    border: `1px solid ${total === 4096 ? "#bbf7d0" : total <= 192 ? "#bbf7d0" : "#fed7aa"}`,
    fontSize:"13px"
  });
  const color = total <= 192 ? "#16a34a" : total <= 1024 ? "#d97706" : "#dc2626";
  const msg = total === 4096
    ? "Apply tools above to begin eliminating structures."
    : `${4096 - total} state combinations eliminated — <strong style="color:#b45309">${total} remain consistent (${pct}%)</strong>.`;
  bar.innerHTML = `
    <span style="font-size:22px;">🎯</span>
    <span>
      <strong style="color:${color};font-size:18px;">${total}</strong>
      <span style="color:#475569;"> / 4,096 states consistent</span>
      &nbsp;&nbsp;<span style="color:#6b7280;font-size:12px;">${msg}</span>
    </span>`;
  return bar;
}
```

```{ojs}
//| echo: false
// ── Confounded DAG grid ─────────────────────────────────────────────────────
// Same 64-cell XMY layout as dagGrid. Cell background shading encodes the
// count of consistent U configurations (0–64): deeper blue = more U ambiguity.
// A small dashed "U" node (top-right of each cell) marks the unobserved variable.
dagGrid2 = {
  const W = 86, H = 70, GAP = 4;
  const NX = [15, 58], NY = [71, 58], NM = [43, 12];
  const NU = [76, 9];          // U node: small dashed circle, top-right corner
  const NR = 9, NRU = 6, APAD = 10;

  const LPAD = 58, SGAP = 16;
  const SHEAD = 26, CHEAD = 16;
  const secW = 4 * (W + GAP) - GAP;
  const secH = 4 * (H + GAP) - GAP;
  const totalW = LPAD + 2 * secW + SGAP + 4;
  const totalH = 2 * (SHEAD + CHEAD + secH) + SGAP + 32; // extra 32 for legend

  const svg = d3.create("svg")
    .attr("width", totalW).attr("height", totalH)
    .style("display", "block").style("overflow", "visible")
    .style("font-family", "system-ui, -apple-system, sans-serif");

  svg.append("defs").html(`
    <marker id="dah2"  viewBox="0 0 8 8" refX="6" refY="4"
            markerWidth="5" markerHeight="5" orient="auto">
      <path d="M0,1 L0,7 L8,4 z" fill="#374151"/>
    </marker>
    <marker id="dahf2" viewBox="0 0 8 8" refX="6" refY="4"
            markerWidth="5" markerHeight="5" orient="auto">
      <path d="M0,1 L0,7 L8,4 z" fill="#d1d5db"/>
    </marker>
  `);

  function putArrow2(g, ax, ay, bx, by, off, faded) {
    const dx = bx-ax, dy = by-ay, len = Math.sqrt(dx*dx+dy*dy)||1;
    const ux = dx/len, uy = dy/len, ox = -uy*off, oy = ux*off;
    g.append("line")
      .attr("x1", ax+ux*APAD+ox).attr("y1", ay+uy*APAD+oy)
      .attr("x2", bx-ux*APAD+ox).attr("y2", by-uy*APAD+oy)
      .attr("stroke", faded ? "#d1d5db" : "#374151")
      .attr("stroke-width", faded ? 1.1 : 1.4)
      .attr("marker-end", `url(#${faded ? "dahf2" : "dah2"})`);
  }

  // Interpolate white → blue-200 (#bfdbfe) based on fraction t ∈ [0,1]
  function uFill(count) {
    if (count === 0) return "rgba(220,38,38,0.07)";
    const t = count / 64;
    return `rgb(${Math.round(255+t*(191-255))},${Math.round(255+t*(219-255))},255)`;
  }
  function uStroke(count) {
    if (count === 0) return "rgba(220,38,38,0.28)";
    const t = count / 64;
    return `rgb(${Math.round(255+t*(147-255))},${Math.round(255+t*(197-255))},253)`;
  }

  function drawCell2(g, s, count, cx, cy) {
    const gc = g.append("g").attr("transform", `translate(${cx},${cy})`);
    const f = count === 0;

    gc.append("rect")
      .attr("width", W).attr("height", H).attr("rx", 4)
      .attr("fill", uFill(count)).attr("stroke", uStroke(count))
      .attr("stroke-width", 0.8);

    // XMY arrows
    if (s.xm===1) putArrow2(gc, NX[0],NX[1], NM[0],NM[1], 0, f);
    if (s.xm===2) putArrow2(gc, NM[0],NM[1], NX[0],NX[1], 0, f);
    if (s.xm===3) { putArrow2(gc,NX[0],NX[1],NM[0],NM[1],3.5,f); putArrow2(gc,NM[0],NM[1],NX[0],NX[1],3.5,f); }
    if (s.my===1) putArrow2(gc, NM[0],NM[1], NY[0],NY[1], 0, f);
    if (s.my===2) putArrow2(gc, NY[0],NY[1], NM[0],NM[1], 0, f);
    if (s.my===3) { putArrow2(gc,NM[0],NM[1],NY[0],NY[1],3.5,f); putArrow2(gc,NY[0],NY[1],NM[0],NM[1],3.5,f); }
    if (s.xy===1) putArrow2(gc, NX[0],NX[1], NY[0],NY[1], 0, f);
    if (s.xy===2) putArrow2(gc, NY[0],NY[1], NX[0],NX[1], 0, f);
    if (s.xy===3) { putArrow2(gc,NX[0],NX[1],NY[0],NY[1],4,f); putArrow2(gc,NY[0],NY[1],NX[0],NX[1],4,f); }

    // XMY node circles
    const ns = f ? "#d1d5db" : "#374151";
    const nf = f ? "#f9fafb" : "#ffffff";
    const tf = f ? "#9ca3af" : "#111827";
    [[NX,"X"],[NY,"Y"],[NM,"M"]].forEach(([pos,lbl]) => {
      gc.append("circle").attr("cx",pos[0]).attr("cy",pos[1]).attr("r",NR)
        .attr("fill",nf).attr("stroke",ns).attr("stroke-width",1.5);
      gc.append("text").attr("x",pos[0]).attr("y",pos[1]+4)
        .attr("text-anchor","middle").attr("font-size",10).attr("font-weight","700")
        .attr("fill",tf).text(lbl);
    });

    // U node — dashed, unobserved
    gc.append("circle").attr("cx",NU[0]).attr("cy",NU[1]).attr("r",NRU)
      .attr("fill", f ? "#f9fafb" : "#f8fafc")
      .attr("stroke", f ? "#d1d5db" : "#94a3b8")
      .attr("stroke-width",1.2).attr("stroke-dasharray","2.5 1.5");
    gc.append("text").attr("x",NU[0]).attr("y",NU[1]+3.5)
      .attr("text-anchor","middle").attr("font-size",7).attr("font-weight","600")
      .attr("fill", f ? "#d1d5db" : "#94a3b8").text("U");

    // U-count badge bottom-right
    if (!f) {
      gc.append("text").attr("x",W-3).attr("y",H-3)
        .attr("text-anchor","end").attr("font-size",7).attr("fill","#64748b")
        .text(`${count}/64`);
    }

    // Red cross for ruled-out
    if (f) {
      gc.append("line").attr("x1",5).attr("y1",5).attr("x2",W-5).attr("y2",H-5)
        .attr("stroke","rgba(220,38,38,0.28)").attr("stroke-width",1.5);
      gc.append("line").attr("x1",W-5).attr("y1",5).attr("x2",5).attr("y2",H-5)
        .attr("stroke","rgba(220,38,38,0.28)").attr("stroke-width",1.5);
    }
  }

  const XY_TITLES = ["X — Y  (no direct link)", "X → Y", "X ← Y", "X ↔ Y"];
  const XY_COLORS = ["#334155","#1d4ed8","#7c3aed","#0e7490"];
  const XM_LABELS = ["X—M: —","X→M","M→X","X↔M"];
  const MY_LABELS = ["—","M→Y","Y→M","M↔Y"];

  [0,1,2,3].forEach((xyVal, si) => {
    const sCol = si % 2, sRow = Math.floor(si/2);
    const ox = LPAD + sCol*(secW+SGAP);
    const oy = sRow*(SHEAD+CHEAD+secH+SGAP);

    svg.append("rect").attr("x",ox).attr("y",oy).attr("width",secW).attr("height",SHEAD-2)
      .attr("rx",4).attr("fill",XY_COLORS[xyVal]);
    svg.append("text").attr("x",ox+secW/2).attr("y",oy+17)
      .attr("text-anchor","middle").attr("fill","white")
      .attr("font-size",12).attr("font-weight","700").text(XY_TITLES[xyVal]);

    for (let my = 0; my < 4; my++) {
      svg.append("text")
        .attr("x", ox+my*(W+GAP)+W/2).attr("y", oy+SHEAD+CHEAD-3)
        .attr("text-anchor","middle").attr("font-size",8.5).attr("fill","#64748b")
        .text(MY_LABELS[my]);
    }
    for (let xm = 0; xm < 4; xm++) {
      const cy = oy+SHEAD+CHEAD+xm*(H+GAP);
      svg.append("text").attr("x",ox-5).attr("y",cy+H/2+4)
        .attr("text-anchor","end").attr("font-size",8.5).attr("fill","#64748b")
        .text(XM_LABELS[xm]);
      for (let my = 0; my < 4; my++) {
        const cx = ox+my*(W+GAP);
        drawCell2(svg, {xm, my, xy:xyVal}, countUConfigs({xm,my,xy:xyVal}, activeTools2), cx, cy);
      }
    }
  });

  // Color legend
  const legY = totalH - 20;
  const legX = LPAD;
  svg.append("text").attr("x",legX).attr("y",legY)
    .attr("font-size",8).attr("fill","#64748b").text("U configs consistent:");
  for (let i = 0; i <= 8; i++) {
    const t = i/8;
    svg.append("rect")
      .attr("x", legX+130+i*16).attr("y",legY-10).attr("width",15).attr("height",10).attr("rx",2)
      .attr("fill",`rgb(${Math.round(255+t*(191-255))},${Math.round(255+t*(219-255))},255)`);
  }
  svg.append("text").attr("x",legX+128).attr("y",legY)
    .attr("text-anchor","end").attr("font-size",7.5).attr("fill","#94a3b8").text("0");
  svg.append("text").attr("x",legX+130+9*16+2).attr("y",legY)
    .attr("font-size",7.5).attr("fill","#94a3b8").text("64");

  svg.append("text")
    .attr("transform",`translate(8,${(totalH-32)/2}) rotate(-90)`)
    .attr("text-anchor","middle").attr("font-size",10).attr("fill","#94a3b8")
    .text("← XM relationship (rows within each section) →");

  return svg.node();
}
```

---

## The Fundamental Problem of Causal Inference {#sec-fundamental}

Every participant in the AlterEco eco-label study had **two potential outcomes**:

- $Y_i(0)$: their WTP if they did *not* see the eco-label
- $Y_i(1)$: their WTP if they *did* see the eco-label

Their **individual treatment effect** is $\tau_i = Y_i(1) - Y_i(0)$.

The problem? **We can only observe one of these for any individual.** This is called the **Fundamental Problem of Causal Inference** (Holland, 1986). Everything that follows is a strategy for getting around it.

```{r dgs-potential-outcomes}
#| code-summary: "▶ Build the data-generating system"
set.seed(2024)
N <- 2000

env        <- rnorm(N, 0, 1)
price_sens <- rnorm(N, 0, 1)
income     <- rnorm(N, 0, 1)

Y0 <- 5.50 + 0.80*env - 0.50*price_sens + 0.30*income + rnorm(N, 0, 1.10)
Y0 <- pmax(1, pmin(10, Y0))

ITE <- 0.60 + 0.80*env - 0.55*price_sens + rnorm(N, 0, 0.40)
Y1  <- pmax(1, pmin(10, Y0 + ITE))

ATE_true <- mean(Y1 - Y0)
cat(sprintf("True ATE  = %.3f\nTrue ITE SD = %.3f\n", ATE_true, sd(Y1 - Y0)))
```

```{r fundamental-problem-table, echo=FALSE}
#| code-summary: "▶ Show the missing-data structure"
set.seed(42)
demo_idx <- c(which(env > 1.2)[1:2], which(abs(env) < 0.15)[1], which(env < -1.2)[1:2])
assigned <- c("Control","Treatment","Control","Treatment","Control")
tibble(
  Person             = c("Sophie (high concern)","Rens (moderate+)","Lotte (neutral)","Daan (low concern)","Kim (price-sensitive)"),
  `Env concern`      = round(env[demo_idx], 1),
  `Y(0) No label`   = round(Y0[demo_idx], 2),
  `Y(1) With label` = round(Y1[demo_idx], 2),
  `True ITE`        = round(Y1[demo_idx]-Y0[demo_idx], 2),
  `Assigned to`     = assigned,
  `What we observe` = c(
    sprintf("Y(0)=$%.2f  Y(1)=???", Y0[demo_idx[1]]),
    sprintf("Y(0)=???   Y(1)=$%.2f", Y1[demo_idx[2]]),
    sprintf("Y(0)=$%.2f  Y(1)=???", Y0[demo_idx[3]]),
    sprintf("Y(0)=???   Y(1)=$%.2f", Y1[demo_idx[4]]),
    sprintf("Y(0)=$%.2f  Y(1)=???", Y0[demo_idx[5]])
  )
) |> knitr::kable(caption="The fundamental problem: exactly ONE potential outcome per person is observed.")
```

## Individual Treatment Effects (ITE) {#sec-ite}

```{r ite-distribution, fig.height=4.5, fig.width=9}
#| code-summary: "▶ Plot the distribution of individual treatment effects"
tibble(ITE=ITE,
       env_grp=cut(env, c(-Inf,-0.5,0.5,Inf), labels=c("Low","Moderate","High"))) |>
  ggplot(aes(x=ITE, fill=env_grp)) +
  geom_histogram(bins=50, alpha=0.75, position="identity") +
  geom_vline(xintercept=ATE_true, linetype="dashed", linewidth=1.1, colour="grey20") +
  annotate("text", x=ATE_true+0.06, y=82,
           label=sprintf("ATE = $%.2f", ATE_true), hjust=0, size=3.5, fontface="bold") +
  scale_fill_manual(values=c("High"=clr_eco,"Moderate"="#74b9a0","Low"=clr_ctrl)) +
  labs(x="Individual Treatment Effect  Y(1)−Y(0)  ($)", y="Count", fill=NULL,
       title="The eco-label effect is wildly different across individuals",
       subtitle="In real research you never see this distribution — only the ATE is estimable") +
  theme_mod3()
```

## Average Treatment Effect (ATE) {#sec-ate}

$$\text{ATE} = E[\tau_i] = E[Y_i(1) - Y_i(0)]$$

In a **randomised experiment**, random assignment makes the treated and control groups exchangeable, so a simple difference in group means is unbiased for the ATE.

```{r ate-rct}
treat_rct <- rbinom(N, 1, 0.5)
Y_obs_rct <- ifelse(treat_rct==1, Y1, Y0)
df_rct    <- tibble(id=1:N, treat=treat_rct, Y_obs=Y_obs_rct, Y0, Y1, ITE, env, price_sens, income)

ATE_est <- mean(Y_obs_rct[treat_rct==1]) - mean(Y_obs_rct[treat_rct==0])
lm(Y_obs ~ treat, data=df_rct) |> tidy() |>
  filter(term=="treat") |>
  transmute(Estimand="ATE", `True value`=round(ATE_true,3), Estimate=round(estimate,3),
            SE=round(std.error,3),
            `95% CI`=sprintf("[%.3f, %.3f]", estimate-1.96*std.error, estimate+1.96*std.error)) |>
  knitr::kable(caption="ATE estimate from randomised eco-label experiment")
```

## Conditional Average Treatment Effect (CATE) {#sec-cate}

```{r cate-by-subgroup, fig.height=4.5, fig.width=9}
df_rct |>
  mutate(env_grp=cut(env, c(-Inf,-1,0,1,Inf), labels=c("Very Low","Low","High","Very High"))) |>
  group_by(env_grp) |>
  summarise(CATE_true=mean(Y1-Y0), CATE_est=mean(Y_obs[treat==1])-mean(Y_obs[treat==0]),
            n=n(), .groups="drop") |>
  pivot_longer(starts_with("CATE"), names_to="type", values_to="val") |>
  ggplot(aes(x=env_grp, y=val, fill=type)) +
  geom_col(position="dodge", alpha=0.85, width=0.65) +
  geom_hline(yintercept=ATE_true, linetype="dashed", colour="grey30") +
  annotate("text", x=0.65, y=ATE_true+0.05, label="Overall ATE", size=3.5) +
  scale_fill_manual(values=c(CATE_true=clr_eco, CATE_est=clr_ctrl),
                    labels=c("True CATE","Estimated CATE")) +
  labs(x="Environmental Concern", y="Eco-label effect ($)", fill=NULL,
       title="CATE: the label works far better for environmentally engaged consumers",
       subtitle="Reporting only the ATE hides this story completely") +
  theme_mod3()
```

## Compliance, ITT, LATE, ATT, and ATC {#sec-late}

```{r compliance-dgs}
set.seed(2024)
Z <- treat_rct
p_comply_if_Z1 <- plogis(0.8 + 0.6*env)
p_always_taker <- plogis(-2.5 + 0.5*env)
D_if_Z1 <- rbinom(N, 1, p_comply_if_Z1)
D_if_Z0 <- rbinom(N, 1, p_always_taker)
D       <- ifelse(Z==1, D_if_Z1, D_if_Z0)
compliance_type <- case_when(
  D_if_Z1==1 & D_if_Z0==0 ~ "Complier",
  D_if_Z1==1 & D_if_Z0==1 ~ "Always-taker",
  D_if_Z1==0 & D_if_Z0==0 ~ "Never-taker",
  D_if_Z1==0 & D_if_Z0==1 ~ "Defier"
)
knitr::kable(table(compliance_type), col.names=c("Compliance type","Count"),
             caption="True compliance breakdown (observable only in simulation)")
```

```{r itt-late-att-atc}
Y_obs_iv <- ifelse(D==1, Y1, Y0)
df_iv    <- tibble(Z, D, Y_obs_iv, Y0, Y1, ITE, compliance_type, env, price_sens)

ITT       <- mean(Y_obs_iv[Z==1]) - mean(Y_obs_iv[Z==0])
FS        <- mean(D[Z==1]) - mean(D[Z==0])
LATE_est  <- ITT / FS
LATE_true <- mean(ITE[compliance_type=="Complier"])
ATT_est   <- mean(ITE[D==1])
ATC_est   <- mean(ITE[D==0])

tibble(
  Estimand    = c("ATE","ITT","LATE (Wald)","ATT","ATC"),
  Description = c(
    "Effect averaged over all 2,000 consumers",
    "Effect of *assigning* the label regardless of compliance",
    "Effect for consumers who complied *because* they were assigned",
    "Effect for consumers who *actually saw* the label",
    "Effect for those who *did not see* it"
  ),
  `True value` = round(c(ATE_true, ITT, LATE_true, ATT_est, ATC_est), 3),
  Estimated    = round(c(ATE_true, ITT, LATE_est,  ATT_est, ATC_est), 3)
) |> knitr::kable(caption="Five ways to summarise 'the effect' of the eco-label")
```

```{r estimands-plot, fig.height=4, fig.width=8, echo=FALSE}
tibble(
  Estimand=c("ATE\n(all consumers)","ITT\n(assigned, any)","LATE\n(compliers only)",
             "ATT\n(label seen)","ATC\n(label not seen)"),
  Value=c(ATE_true, ITT, LATE_est, ATT_est, ATC_est),
  Type=c("Population","Assignment","Compliers","Treated","Control")
) |>
  mutate(Estimand=fct_reorder(Estimand, Value)) |>
  ggplot(aes(y=Estimand, x=Value, fill=Type)) +
  geom_col(width=0.6, alpha=0.85) +
  geom_vline(xintercept=ATE_true, linetype="dashed", colour="grey30") +
  scale_fill_brewer(palette="Set2") +
  labs(x="Eco-label effect on WTP ($)", y=NULL, fill=NULL,
       title="Five estimands — same experiment, different questions",
       subtitle="ATT > ATE > ATC: consumers who seek eco-labels respond more strongly to them") +
  theme_mod3()
```

::: {.callout-warning}
## Connection to Module 2
Module 2's permutation test estimated the **ITT** — comparing all participants assigned to the eco-label condition against all assigned to control, regardless of whether they actually processed the label. In a well-controlled lab study, ITT ≈ ATE. In field settings — where consumers scroll past labels or ignore shelf information — the distinction matters greatly.
:::

---
