---
title: "Part 1: What Is a Causal Effect?"
---

```{r setup, include=FALSE}
options(repos = c(CRAN = "https://cran.rstudio.com/"))

if (!requireNamespace("remotes",  quietly = TRUE)) install.packages("remotes")
if (!requireNamespace("synthdid", quietly = TRUE))
  remotes::install_github("synth-did/synthdid")

pkgs <- c(
  "tidyverse", "broom", "knitr", "patchwork",
  "conflicted",
  "MatchIt", "WeightIt", "cobalt",
  "rdrobust", "rddensity",
  "did",
  "mediation", "robmed",
  "estimatr", "sandwich", "lmtest"
)
for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p, dependencies = TRUE)
  library(p, character.only = TRUE)
}
library(synthdid)

library(conflicted)
conflicts_prefer(dplyr::select, dplyr::filter, dplyr::rename, dplyr::mutate,
                 dplyr::summarise, dplyr::arrange, dplyr::lag)

dir.create("figs", recursive = TRUE, showWarnings = FALSE)
knitr::opts_chunk$set(
  echo = TRUE, fig.width = 9, fig.height = 5,
  fig.path = "figs/", dev = "svg"
)

clr_eco     <- "#2d6a4f"
clr_ctrl    <- "#4a90d9"
clr_treated <- clr_eco

theme_mod3 <- function() {
  theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          strip.background = element_rect(fill = "#f0f4f8", colour = NA),
          legend.position  = "bottom")
}

# ── Lightweight DAG helper ─────────────────────────────────────────────────────
.dag <- function(nodes, solid = NULL, dashed = NULL,
                 unobserved = character(0), title = "") {
  nd <- do.call(rbind, lapply(names(nodes), function(nm)
    data.frame(id=nm, x=nodes[[nm]][1], y=nodes[[nm]][2], stringsAsFactors=FALSE)))
  nd$type <- ifelse(nd$id %in% unobserved, "unobs", "obs")
  segs <- function(pairs, lt) {
    if (is.null(pairs) || length(pairs)==0) return(NULL)
    do.call(rbind, lapply(pairs, function(ab) {
      fr <- nd[nd$id==ab[1],]; to <- nd[nd$id==ab[2],]
      dx <- to$x-fr$x; dy <- to$y-fr$y
      len <- sqrt(dx^2+dy^2)+1e-9; pad <- 0.12
      data.frame(x=fr$x+pad*dx/len, y=fr$y+pad*dy/len,
                 xend=to$x-pad*dx/len, yend=to$y-pad*dy/len,
                 lt=lt, stringsAsFactors=FALSE)
    }))
  }
  edf <- rbind(segs(solid,"solid"), segs(dashed,"dashed"))
  xs <- range(nd$x); ys <- range(nd$y)
  p <- ggplot() + theme_void() + ggtitle(title) +
    coord_cartesian(xlim=c(xs[1]-.3,xs[2]+.3), ylim=c(ys[1]-.3,ys[2]+.3), expand=FALSE) +
    theme(plot.title=element_text(size=10,face="bold",hjust=.5,margin=margin(b=6)),
          plot.background=element_rect(fill="grey98",colour="grey80",linewidth=.4),
          plot.margin=margin(10,16,10,16), legend.position="none")
  if (!is.null(edf) && nrow(edf)>0)
    p <- p +
      geom_segment(data=edf, aes(x=x,y=y,xend=xend,yend=yend,colour=lt,linetype=lt),
                   arrow=arrow(length=unit(.20,"cm"),type="closed"), linewidth=.85) +
      scale_colour_manual(values=c(solid="grey20",dashed="firebrick")) +
      scale_linetype_manual(values=c(solid="solid",dashed="dashed"))
  if (any(nd$type=="obs"))
    p <- p + geom_label(data=nd[nd$type=="obs",], aes(x=x,y=y,label=id),
                        size=4, fontface="bold", fill="white", colour="black",
                        label.size=.9, label.padding=unit(.4,"lines"))
  if (any(nd$type=="unobs"))
    p <- p + geom_label(data=nd[nd$type=="unobs",], aes(x=x,y=y,label=id),
                        size=4, fontface="bold.italic", fill="#fff3e0", colour="#b45309",
                        label.size=1.1, label.padding=unit(.4,"lines"),
                        label.r=unit(.5,"lines"))
  p
}

# ── Mediation extraction helper (handles robmed version differences) ───────────
rob_extract <- function(fit) {
  # point estimate: try $ab, then $indirect, then coef()["ab"]
  ab <- tryCatch(fit$ab,
    error = function(e) tryCatch(fit$indirect,
      error = function(e2) tryCatch(coef(fit)["ab"],
        error = function(e3) NA_real_)))
  # CI: try $ci (matrix or vector), then confint()
  ci <- tryCatch({
    r <- fit$ci
    if (is.matrix(r)) r[1,] else r
  }, error = function(e) tryCatch(confint(fit)["ab",],
    error = function(e2) c(NA_real_, NA_real_)))
  list(est=unname(ab), lo=unname(ci[1]), hi=unname(ci[2]))
}
```

# Part 1: What Do We Mean by "Causal Effect"? {#sec-m3-part1}

<iframe src="https://docs.google.com/presentation/d/e/2PACX-1vQ_Qbgn1mguAIxorFZHqsS0sGJXYUbyYw7yFBsyd37wKXQpQgrpBh5c3wWTxDEtvg/pubembed?start=false&loop=false&delayms=3000" width="100%" height="500px" frameborder="0" allowfullscreen="true" mozallowfullscreen="true" webkitallowfullscreen="true" title="Module 3 Part 1: Causal Effects"></iframe>

## Causal Inference is a Deductive Process {#sec-deductive}

::: {.callout-important}
## What this exercise assumes
The 64 structures shown below represent the complete space of possible causal relationships among three variables under **ideal conditions**: (1) **proper measurement** — X, M, and Y are each measured without discriminant validity failures or systematic error (Module 1), and (2) **no omitted variables** — these three variables are the only players in the system (Module 2). In real research, measurement error and hidden confounders expand this space considerably.
:::

Causal inference is not induction — you cannot *prove* a causal structure from data. It is **deduction by elimination**: you bring every tool in your methodological toolkit to bear and see which structures survive. The exercise below makes this concrete.

The grid shows **all 64 structurally distinct relationships** that can exist among three variables **X**, **M**, and **Y** — arranged as a triangle with X at the lower-left, Y at the lower-right, and M at the top. Each cell shows one possible causal structure. Arrows between any two variables can point forward (A → B), backward (A ← B), bidirectionally (A ↔ B, representing correlated unmeasured causes or feedback), or not exist at all.

Click the buttons below to apply research findings. **Each finding rules out the structures inconsistent with it** — those cells will be crossed out in red. After exhausting your toolkit, notice how many structures remain. That irreducible remainder is why causal inference is hard.

```{ojs}
//| echo: false
// ── Generate all 64 causal structures ─────────────────────────────────────────
// Arrow code per pair: 0=none, 1=A→B (forward), 2=A←B (backward), 3=A↔B (bidir)
// Three pairs: xm (X↔M), my (M↔Y), xy (X↔Y)
dag_states = {
  const arr = [];
  for (let xm = 0; xm < 4; xm++)
    for (let my = 0; my < 4; my++)
      for (let xy = 0; xy < 4; xy++)
        arr.push({ xm, my, xy });
  return arr;
}
```

```{ojs}
//| echo: false
// ── Consistency logic ──────────────────────────────────────────────────────────
// Returns true if state s is consistent with all active tool findings.
//
// Arrow code per pair: 0=none, 1=A→B (forward), 2=A←B (backward), 3=A↔B (bidir)
// Pairs: xm (X,M), my (M,Y), xy (X,Y)
//
// Tool logic (each button represents a POSITIVE finding that rules out
// structures that cannot produce that result):
//
//   reg  — regression Y ~ X + M runs      → requires X-M association (xm≠0) AND M-Y association (my≠0)
//           Does NOT constrain xy, and does NOT identify the direction of associations:
//           each could be causal, reverse-causal, or arise from a collider or shared cause.
//   expX — X experimentally manipulated   → X exogenous: xm≠2,3; xy≠2,3
//   med  — indirect-pattern finding       → xm≠0; my≠0; xy=0
//           Extends reg by requiring no direct X–Y association conditional on M.
//           Rules out direct X→Y, but X→M←Y (Y causes M) and X→M↔Y still survive.
function dagConsistent(s, active) {
  const { xm, my, xy } = s;
  const any = v => v > 0;

  // X exogenous when manipulated: no arrows into X
  if (active.includes("expX") && (xm === 2 || xm === 3 || xy === 2 || xy === 3)) return false;

  for (const t of active) {
    if (t === "reg" && !(any(xm) && any(my))) return false; // X-M and M-Y links required
    if (t === "med" && !(any(xm) && any(my) && xy === 0))  return false; // pure mediation
  }
  return true;
}
```

```{ojs}
//| echo: false
// ── Tool buttons ───────────────────────────────────────────────────────────────
viewof activeTools = {
  const TOOLS = [
    { id:"reg",  cat:"Regression", label:"Regression Y ~ X + M",             sub:"X–M and M–Y associations detected; direction of each association is unconstrained" },
    { id:"expX", cat:"Experiment", label:"Manipulate X",                       sub:"X is exogenous — removes all arrows into X" },
    { id:"med",  cat:"Mediation",  label:"No direct X–Y, given X–M and M–Y",  sub:"X–Y association vanishes conditional on M; consistent with mediation, reverse causation, or reciprocal M–Y" },
  ];
  const ICONS = { Regression:"📊", Experiment:"🧪", Mediation:"⛓️" };
  const active = new Set();

  // ── Build the DOM entirely in JS (avoids html-template ID clashes) ──────────
  const wrap = document.createElement("div");
  Object.assign(wrap.style, {
    border:"1px solid #e2e8f0", borderRadius:"10px", padding:"16px 18px",
    background:"#f8fafc", margin:"16px 0"
  });

  // Header row
  const hdr = document.createElement("div");
  Object.assign(hdr.style, { display:"flex", justifyContent:"space-between",
    alignItems:"center", marginBottom:"12px" });
  const title = document.createElement("span");
  title.style.cssText = "font-weight:700;font-size:14px;color:#0f172a;";
  title.innerHTML = `🛠️ Research Tools &nbsp;<span style="font-weight:400;color:#64748b;font-size:12px;">— click a finding to apply it; inconsistent structures are crossed out</span>`;
  const resetBtn = document.createElement("button");
  resetBtn.textContent = "↺ Reset All";
  resetBtn.style.cssText = "padding:5px 12px;border-radius:6px;border:1px solid #94a3b8;background:#fff;cursor:pointer;font-size:12px;color:#475569;";
  hdr.appendChild(title); hdr.appendChild(resetBtn);
  wrap.appendChild(hdr);

  // Button grid — group by category
  const cats = [...new Set(TOOLS.map(t => t.cat))];
  cats.forEach(cat => {
    const row = document.createElement("div");
    row.style.cssText = "display:flex;flex-wrap:wrap;gap:7px;margin-bottom:8px;align-items:center;";
    const catLbl = document.createElement("span");
    catLbl.style.cssText = "font-size:11px;color:#94a3b8;min-width:80px;";
    catLbl.textContent = `${ICONS[cat]} ${cat}`;
    row.appendChild(catLbl);
    TOOLS.filter(t => t.cat === cat).forEach(t => {
      const b = document.createElement("button");
      b.style.cssText = "padding:7px 13px;border-radius:18px;border:2px solid #cbd5e1;background:#fff;cursor:pointer;font-size:12px;color:#334155;transition:all 0.15s;line-height:1.4;text-align:left;";
      b.innerHTML = `<strong>${t.label}</strong><br><span style="font-size:10px;color:#94a3b8;">${t.sub}</span>`;
      b.onclick = () => {
        if (active.has(t.id)) {
          active.delete(t.id);
          b.style.cssText = b.style.cssText.replace(/background:[^;]+/, "background:#fff")
                              .replace(/border-color:[^;]+/, "border-color:#cbd5e1")
                              .replace(/color:[^;]+/, "color:#334155");
          b.style.background = "#fff"; b.style.borderColor = "#cbd5e1"; b.style.color = "#334155";
        } else {
          active.add(t.id);
          b.style.background = "#dc2626"; b.style.borderColor = "#dc2626"; b.style.color = "#fff";
        }
        wrap.value = [...active];
        wrap.dispatchEvent(new Event("input", { bubbles: true }));
      };
      row.appendChild(b);
    });
    wrap.appendChild(row);
  });

  resetBtn.onclick = () => {
    active.clear();
    wrap.querySelectorAll("button:not(:first-of-type)").forEach(b => {
      b.style.background = "#fff"; b.style.borderColor = "#cbd5e1"; b.style.color = "#334155";
    });
    wrap.value = [];
    wrap.dispatchEvent(new Event("input", { bubbles: true }));
  };

  wrap.value = [];
  return wrap;
}
```

```{ojs}
//| echo: false
// ── Status counter ─────────────────────────────────────────────────────────────
dagStatusBar = {
  const n   = dag_states.filter(s => dagConsistent(s, activeTools)).length;
  const bar = document.createElement("div");
  Object.assign(bar.style, {
    display:"flex", alignItems:"center", gap:"12px",
    padding:"10px 16px", borderRadius:"8px", margin:"6px 0",
    background: n === 64 ? "#f0fdf4" : n === 1 ? "#f0fdf4" : "#fff7ed",
    border: `1px solid ${n === 64 ? "#bbf7d0" : n === 1 ? "#bbf7d0" : "#fed7aa"}`,
    fontSize:"13px"
  });
  const color = n <= 2 ? "#16a34a" : n <= 12 ? "#d97706" : "#dc2626";
  let msg = n === 64
    ? "Apply tools above to begin eliminating structures."
    : n === 1
    ? "✓ Uniquely identified! (This is rare even with all tools.)"
    : `${64 - n} structures eliminated — <strong style="color:#b45309">${n} remain consistent with your findings</strong>.`;
  bar.innerHTML = `
    <span style="font-size:22px;">🎯</span>
    <span>
      <strong style="color:${color};font-size:18px;">${n}</strong>
      <span style="color:#475569;"> / 64 structures consistent</span>
      &nbsp;&nbsp;<span style="color:#6b7280;font-size:12px;">${msg}</span>
    </span>`;
  return bar;
}
```

```{ojs}
//| echo: false
// ── Main DAG grid ──────────────────────────────────────────────────────────────
dagGrid = {
  // Cell geometry
  const W = 86, H = 70, GAP = 4;
  // Node positions within each cell (origin = cell top-left)
  const NX = [15, 58], NY = [71, 58], NM = [43, 12];
  const NR = 9, APAD = 10;

  // Layout metrics
  const LPAD = 58, SGAP = 16;
  const SHEAD = 26, CHEAD = 16;
  const secW = 4 * (W + GAP) - GAP;   // 352
  const secH = 4 * (H + GAP) - GAP;   // 292
  const totalW = LPAD + 2 * secW + SGAP + 4;
  const totalH = 2 * (SHEAD + CHEAD + secH) + SGAP + 6;

  const svg = d3.create("svg")
    .attr("width", totalW).attr("height", totalH)
    .style("display", "block").style("overflow", "visible")
    .style("font-family", "system-ui, -apple-system, sans-serif");

  // Arrowhead markers — end (normal/faded) + start (auto-start-reverse for ↔)
  svg.append("defs").html(`
    <marker id="dah"   viewBox="0 0 8 8" refX="6" refY="4"
            markerWidth="5" markerHeight="5" orient="auto">
      <path d="M0,1 L0,7 L8,4 z" fill="#374151"/>
    </marker>
    <marker id="dahf"  viewBox="0 0 8 8" refX="6" refY="4"
            markerWidth="5" markerHeight="5" orient="auto">
      <path d="M0,1 L0,7 L8,4 z" fill="#d1d5db"/>
    </marker>
    <marker id="dah-s"  viewBox="0 0 8 8" refX="6" refY="4"
            markerWidth="5" markerHeight="5" orient="auto-start-reverse">
      <path d="M0,1 L0,7 L8,4 z" fill="#374151"/>
    </marker>
    <marker id="dahf-s" viewBox="0 0 8 8" refX="6" refY="4"
            markerWidth="5" markerHeight="5" orient="auto-start-reverse">
      <path d="M0,1 L0,7 L8,4 z" fill="#d1d5db"/>
    </marker>
  `);

  // Draw a single directed arrow from (ax,ay) to (bx,by)
  function putArrow(g, ax, ay, bx, by, off, faded) {
    const dx = bx - ax, dy = by - ay;
    const len = Math.sqrt(dx * dx + dy * dy) || 1;
    const ux = dx / len, uy = dy / len;
    const ox = -uy * off, oy = ux * off;
    g.append("line")
      .attr("x1", ax + ux * APAD + ox).attr("y1", ay + uy * APAD + oy)
      .attr("x2", bx - ux * APAD + ox).attr("y2", by - uy * APAD + oy)
      .attr("stroke", faded ? "#d1d5db" : "#374151")
      .attr("stroke-width", faded ? 1.1 : 1.4)
      .attr("marker-end", `url(#${faded ? "dahf" : "dah"})`);
  }
  // Draw a bidirectional arrow — one line, arrowheads at both ends
  function putBiArrow(g, ax, ay, bx, by, faded) {
    const dx = bx - ax, dy = by - ay;
    const len = Math.sqrt(dx * dx + dy * dy) || 1;
    const ux = dx / len, uy = dy / len;
    g.append("line")
      .attr("x1", ax + ux * APAD).attr("y1", ay + uy * APAD)
      .attr("x2", bx - ux * APAD).attr("y2", by - uy * APAD)
      .attr("stroke", faded ? "#d1d5db" : "#374151")
      .attr("stroke-width", faded ? 1.1 : 1.4)
      .attr("marker-start", `url(#${faded ? "dahf-s" : "dah-s"})`)
      .attr("marker-end",   `url(#${faded ? "dahf"   : "dah"})`);
  }

  // Draw one cell at position (cx, cy) in group g
  function drawCell(g, s, ok, cx, cy) {
    const gc = g.append("g").attr("transform", `translate(${cx},${cy})`);
    const f = !ok;

    // Cell background
    gc.append("rect")
      .attr("width", W).attr("height", H).attr("rx", 4)
      .attr("fill",   f ? "rgba(220,38,38,0.07)" : "#ffffff")
      .attr("stroke", f ? "rgba(220,38,38,0.28)" : "#e2e8f0")
      .attr("stroke-width", 0.8);

    // XM arrows
    if (s.xm === 1) putArrow(gc, NX[0],NX[1], NM[0],NM[1], 0, f);
    if (s.xm === 2) putArrow(gc, NM[0],NM[1], NX[0],NX[1], 0, f);
    if (s.xm === 3) putBiArrow(gc, NX[0],NX[1], NM[0],NM[1], f);
    // MY arrows
    if (s.my === 1) putArrow(gc, NM[0],NM[1], NY[0],NY[1], 0, f);
    if (s.my === 2) putArrow(gc, NY[0],NY[1], NM[0],NM[1], 0, f);
    if (s.my === 3) putBiArrow(gc, NM[0],NM[1], NY[0],NY[1], f);
    // XY arrows
    if (s.xy === 1) putArrow(gc, NX[0],NX[1], NY[0],NY[1], 0, f);
    if (s.xy === 2) putArrow(gc, NY[0],NY[1], NX[0],NX[1], 0, f);
    if (s.xy === 3) putBiArrow(gc, NX[0],NX[1], NY[0],NY[1], f);

    // Node circles + labels
    const ns = f ? "#d1d5db" : "#374151";
    const nf = f ? "#f9fafb" : "#ffffff";
    const tf = f ? "#9ca3af" : "#111827";
    [[NX,"X"],[NY,"Y"],[NM,"M"]].forEach(([pos, lbl]) => {
      gc.append("circle")
        .attr("cx", pos[0]).attr("cy", pos[1]).attr("r", NR)
        .attr("fill", nf).attr("stroke", ns).attr("stroke-width", 1.5);
      gc.append("text")
        .attr("x", pos[0]).attr("y", pos[1] + 4)
        .attr("text-anchor", "middle").attr("font-size", 10).attr("font-weight", "700")
        .attr("fill", tf).text(lbl);
    });

    // Red X cross for ruled-out structures
    if (f) {
      gc.append("line")
        .attr("x1", 5).attr("y1", 5).attr("x2", W-5).attr("y2", H-5)
        .attr("stroke", "rgba(220,38,38,0.28)").attr("stroke-width", 1.5);
      gc.append("line")
        .attr("x1", W-5).attr("y1", 5).attr("x2", 5).attr("y2", H-5)
        .attr("stroke", "rgba(220,38,38,0.28)").attr("stroke-width", 1.5);
    }
  }

  // Section labels and colours
  const XY_TITLES  = ["X — Y  (no direct link)", "X → Y", "X ← Y", "X ↔ Y"];
  const XY_COLORS  = ["#334155", "#1d4ed8", "#7c3aed", "#0e7490"];
  const XM_LABELS  = ["X—M: —", "X→M", "M→X", "X↔M"];
  const MY_LABELS  = ["—", "M→Y", "Y→M", "M↔Y"];

  // Render 4 sections in a 2×2 layout; each section = one XY relationship type
  [0, 1, 2, 3].forEach((xyVal, si) => {
    const sCol = si % 2;
    const sRow = Math.floor(si / 2);
    const ox = LPAD + sCol * (secW + SGAP);
    const oy = sRow * (SHEAD + CHEAD + secH + SGAP);

    // Section header bar
    svg.append("rect")
      .attr("x", ox).attr("y", oy).attr("width", secW).attr("height", SHEAD - 2)
      .attr("rx", 4).attr("fill", XY_COLORS[xyVal]);
    svg.append("text")
      .attr("x", ox + secW / 2).attr("y", oy + 17)
      .attr("text-anchor", "middle").attr("fill", "white")
      .attr("font-size", 12).attr("font-weight", "700")
      .text(XY_TITLES[xyVal]);

    // Column headers (MY relationship)
    for (let my = 0; my < 4; my++) {
      svg.append("text")
        .attr("x", ox + my * (W + GAP) + W / 2).attr("y", oy + SHEAD + CHEAD - 3)
        .attr("text-anchor", "middle").attr("font-size", 8.5).attr("fill", "#64748b")
        .text(MY_LABELS[my]);
    }

    // Row labels (XM relationship) + cells
    for (let xm = 0; xm < 4; xm++) {
      const cy = oy + SHEAD + CHEAD + xm * (H + GAP);
      // Row label on the left edge of each section column
      svg.append("text")
        .attr("x", ox - 5).attr("y", cy + H / 2 + 4)
        .attr("text-anchor", "end").attr("font-size", 8.5).attr("fill", "#64748b")
        .text(XM_LABELS[xm]);
      for (let my = 0; my < 4; my++) {
        const cx = ox + my * (W + GAP);
        drawCell(svg, { xm, my, xy: xyVal }, dagConsistent({ xm, my, xy: xyVal }, activeTools), cx, cy);
      }
    }
  });

  // Axis label: rows = XM relationship
  svg.append("text")
    .attr("transform", `translate(8, ${totalH / 2}) rotate(-90)`)
    .attr("text-anchor", "middle").attr("font-size", 10).attr("fill", "#94a3b8")
    .text("← XM relationship (rows within each section) →");

  return svg.node();
}
```

::: {.callout-important}
## What pressing all three buttons shows

After applying all three tools together — regression, experimental manipulation of X, and the indirect-pattern finding — **three causal structures survive**, not one:

| Structure | Interpretation |
|---|---|
| X → M → Y | Classic mediation: X causes M, M causes Y |
| X → M ← Y | Y causes M (reverse causation on the M–Y link; M is a collider) |
| X → M ↔ Y | M and Y are reciprocally linked or share an unmeasured common cause |

All three are consistent with every piece of evidence you can collect in this system without additional assumptions. Causal inference is deduction by elimination — each tool removes some structures — but **the elimination rarely converges to one uniquely identified story**. Ruling out the two alternatives requires additional design choices (e.g., time-ordering evidence that Y cannot precede M, an instrument for M, or a manipulation of M) that go beyond regression and a single experiment.
:::

### What if There Is an Unobserved Confounder? {#sec-confounded}

The exercise above assumed no hidden variables. Now introduce **U** — an unobserved variable that may have arrows to X, M, and/or Y but is never measured. No research tool can constrain U directly.

Each of the three U-pairs (U–X, U–M, U–Y) can be absent, forward, backward, or bidirectional — 4³ = **64 U configurations** for every one of the 64 X–M–Y structures, giving **4,096 total states**. Rather than draw all 4,096 cells, the grid below preserves the familiar 64 X–M–Y layout. Each cell's **blue shading** encodes how many of the 64 U configurations remain consistent with your findings: deeper blue = U is highly ambiguous for that structure; pale = almost no U configuration survives; crossed out = the X–M–Y structure is ruled out for *every* possible U configuration. The small dashed circle labelled **U** in each cell is a reminder that the confounder is always present but unobserved.

Notice in particular: pressing **Manipulate X** eliminates U→X (since X is now exogenous), but U's connections to M and Y remain entirely free — experimental control of X does not eliminate unmeasured confounding of the M–Y path.

```{ojs}
//| echo: false
// ── Consistency logic for confounded model ─────────────────────────────────
// Six pairs: xm, my, xy (same as before) plus ux (U,X), um (U,M), uy (U,Y)
// Arrow codes: 0=none, 1=A→B, 2=A←B, 3=A↔B
//
//   reg2  — regression Y ~ X + M   → xm≠0; my≠0; xy unconstrained
//   expX2 — manipulate X           → X exogenous: xm≠2,3; xy≠2,3; ux≠2,3
//   med2  — mediation X→M→Y (pure) → xm≠0; my≠0; xy=0
//           Extends reg2 by requiring no direct X→Y. reg2+med2 = med2 alone.
dagConsistent2 = (s, active) => {
  const { xm, my, xy, ux, um, uy } = s;
  const any = v => v > 0;
  if (active.includes("expX2") && (xm===2||xm===3||xy===2||xy===3||ux===2||ux===3)) return false;
  for (const t of active) {
    if (t === "reg2" && !(any(xm) && any(my))) return false;
    if (t === "med2" && !(any(xm) && any(my) && xy === 0)) return false;
  }
  return true;
}
```

```{ojs}
//| echo: false
// Count how many of the 64 U configurations (ux, um, uy) are consistent
// for a fixed (xm, my, xy) state given the active button set.
countUConfigs = (xmyState, active) => {
  let n = 0;
  for (let ux = 0; ux < 4; ux++)
    for (let um = 0; um < 4; um++)
      for (let uy = 0; uy < 4; uy++)
        if (dagConsistent2({...xmyState, ux, um, uy}, active)) n++;
  return n;
}
```

```{ojs}
//| echo: false
// ── Tool buttons (confounded model) ────────────────────────────────────────
viewof activeTools2 = {
  const TOOLS = [
    { id:"reg2",  cat:"Regression", label:"Regression Y ~ X + M",             sub:"X–M and M–Y associations detected; direction of each association is unconstrained" },
    { id:"expX2", cat:"Experiment", label:"Manipulate X",                       sub:"X is exogenous — removes arrows into X, including U→X" },
    { id:"med2",  cat:"Mediation",  label:"No direct X–Y, given X–M and M–Y",  sub:"X–Y association vanishes conditional on M; consistent with mediation, reverse causation, or reciprocal M–Y" },
  ];
  const ICONS = { Regression:"📊", Experiment:"🧪", Mediation:"⛓️" };
  const active = new Set();
  const buttons = [];

  const wrap = document.createElement("div");
  Object.assign(wrap.style, {
    border:"1px solid #e2e8f0", borderRadius:"10px", padding:"16px 18px",
    background:"#f8fafc", margin:"16px 0"
  });

  const hdr = document.createElement("div");
  Object.assign(hdr.style, { display:"flex", justifyContent:"space-between",
    alignItems:"center", marginBottom:"12px" });
  const title = document.createElement("span");
  title.style.cssText = "font-weight:700;font-size:14px;color:#0f172a;";
  title.innerHTML = `🛠️ Research Tools &nbsp;<span style="font-weight:400;color:#64748b;font-size:12px;">— U is always present but unobserved; no tool constrains it</span>`;
  const resetBtn = document.createElement("button");
  resetBtn.textContent = "↺ Reset All";
  resetBtn.style.cssText = "padding:5px 12px;border-radius:6px;border:1px solid #94a3b8;background:#fff;cursor:pointer;font-size:12px;color:#475569;";
  hdr.appendChild(title); hdr.appendChild(resetBtn);
  wrap.appendChild(hdr);

  const cats = [...new Set(TOOLS.map(t => t.cat))];
  cats.forEach(cat => {
    const row = document.createElement("div");
    row.style.cssText = "display:flex;flex-wrap:wrap;gap:7px;margin-bottom:8px;align-items:center;";
    const catLbl = document.createElement("span");
    catLbl.style.cssText = "font-size:11px;color:#94a3b8;min-width:80px;";
    catLbl.textContent = `${ICONS[cat]} ${cat}`;
    row.appendChild(catLbl);
    TOOLS.filter(t => t.cat === cat).forEach(t => {
      const b = document.createElement("button");
      b.style.cssText = "padding:7px 13px;border-radius:18px;border:2px solid #cbd5e1;background:#fff;cursor:pointer;font-size:12px;color:#334155;transition:all 0.15s;line-height:1.4;text-align:left;";
      b.innerHTML = `<strong>${t.label}</strong><br><span style="font-size:10px;color:#94a3b8;">${t.sub}</span>`;
      b.onclick = () => {
        if (active.has(t.id)) {
          active.delete(t.id);
          b.style.background = "#fff"; b.style.borderColor = "#cbd5e1"; b.style.color = "#334155";
        } else {
          active.add(t.id);
          b.style.background = "#dc2626"; b.style.borderColor = "#dc2626"; b.style.color = "#fff";
        }
        wrap.value = [...active];
        wrap.dispatchEvent(new Event("input", { bubbles: true }));
      };
      buttons.push(b);
      row.appendChild(b);
    });
    wrap.appendChild(row);
  });

  resetBtn.onclick = () => {
    active.clear();
    buttons.forEach(b => {
      b.style.background = "#fff"; b.style.borderColor = "#cbd5e1"; b.style.color = "#334155";
    });
    wrap.value = [];
    wrap.dispatchEvent(new Event("input", { bubbles: true }));
  };

  wrap.value = [];
  return wrap;
}
```

```{ojs}
//| echo: false
// ── Status counter (out of 4096) ────────────────────────────────────────────
dagStatusBar2 = {
  let total = 0;
  for (let xm = 0; xm < 4; xm++)
    for (let my = 0; my < 4; my++)
      for (let xy = 0; xy < 4; xy++)
        total += countUConfigs({xm, my, xy}, activeTools2);
  const bar = document.createElement("div");
  Object.assign(bar.style, {
    display:"flex", alignItems:"center", gap:"12px",
    padding:"10px 16px", borderRadius:"8px", margin:"6px 0",
    background: total === 4096 ? "#f0fdf4" : total <= 192 ? "#f0fdf4" : "#fff7ed",
    border: `1px solid ${total === 4096 ? "#bbf7d0" : total <= 192 ? "#bbf7d0" : "#fed7aa"}`,
    fontSize:"13px"
  });
  const color = total <= 192 ? "#16a34a" : total <= 1024 ? "#d97706" : "#dc2626";
  const msg = total === 4096
    ? "Apply tools above to begin eliminating structures."
    : `${4096 - total} state combinations eliminated — <strong style="color:#b45309">${total} remain consistent</strong>.`;
  bar.innerHTML = `
    <span style="font-size:22px;">🎯</span>
    <span>
      <strong style="color:${color};font-size:18px;">${total}</strong>
      <span style="color:#475569;"> / 4,096 states consistent</span>
      &nbsp;&nbsp;<span style="color:#6b7280;font-size:12px;">${msg}</span>
    </span>`;
  return bar;
}
```

```{ojs}
//| echo: false
// ── Confounded DAG grid ─────────────────────────────────────────────────────
// Same 64-cell XMY layout as dagGrid. Cell background shading encodes the
// count of consistent U configurations (0–64): deeper blue = more U ambiguity.
// A small dashed "U" node (top-right of each cell) marks the unobserved variable.
dagGrid2 = {
  const W = 86, H = 70, GAP = 4;
  const NX = [15, 58], NY = [71, 58], NM = [43, 12];
  const NU = [76, 9];          // U node: small dashed circle, top-right corner
  const NR = 9, NRU = 6, APAD = 10;

  const LPAD = 58, SGAP = 16;
  const SHEAD = 26, CHEAD = 16;
  const secW = 4 * (W + GAP) - GAP;
  const secH = 4 * (H + GAP) - GAP;
  const totalW = LPAD + 2 * secW + SGAP + 4;
  const totalH = 2 * (SHEAD + CHEAD + secH) + SGAP + 32; // extra 32 for legend

  const svg = d3.create("svg")
    .attr("width", totalW).attr("height", totalH)
    .style("display", "block").style("overflow", "visible")
    .style("font-family", "system-ui, -apple-system, sans-serif");

  svg.append("defs").html(`
    <marker id="dah2"   viewBox="0 0 8 8" refX="6" refY="4"
            markerWidth="5" markerHeight="5" orient="auto">
      <path d="M0,1 L0,7 L8,4 z" fill="#374151"/>
    </marker>
    <marker id="dahf2"  viewBox="0 0 8 8" refX="6" refY="4"
            markerWidth="5" markerHeight="5" orient="auto">
      <path d="M0,1 L0,7 L8,4 z" fill="#d1d5db"/>
    </marker>
    <marker id="dah2-s"  viewBox="0 0 8 8" refX="6" refY="4"
            markerWidth="5" markerHeight="5" orient="auto-start-reverse">
      <path d="M0,1 L0,7 L8,4 z" fill="#374151"/>
    </marker>
    <marker id="dahf2-s" viewBox="0 0 8 8" refX="6" refY="4"
            markerWidth="5" markerHeight="5" orient="auto-start-reverse">
      <path d="M0,1 L0,7 L8,4 z" fill="#d1d5db"/>
    </marker>
  `);

  function putArrow2(g, ax, ay, bx, by, off, faded) {
    const dx = bx-ax, dy = by-ay, len = Math.sqrt(dx*dx+dy*dy)||1;
    const ux = dx/len, uy = dy/len, ox = -uy*off, oy = ux*off;
    g.append("line")
      .attr("x1", ax+ux*APAD+ox).attr("y1", ay+uy*APAD+oy)
      .attr("x2", bx-ux*APAD+ox).attr("y2", by-uy*APAD+oy)
      .attr("stroke", faded ? "#d1d5db" : "#374151")
      .attr("stroke-width", faded ? 1.1 : 1.4)
      .attr("marker-end", `url(#${faded ? "dahf2" : "dah2"})`);
  }
  // Bidirectional: one line, arrowheads at both ends
  function putBiArrow2(g, ax, ay, bx, by, faded) {
    const dx = bx-ax, dy = by-ay, len = Math.sqrt(dx*dx+dy*dy)||1;
    const ux = dx/len, uy = dy/len;
    g.append("line")
      .attr("x1", ax+ux*APAD).attr("y1", ay+uy*APAD)
      .attr("x2", bx-ux*APAD).attr("y2", by-uy*APAD)
      .attr("stroke", faded ? "#d1d5db" : "#374151")
      .attr("stroke-width", faded ? 1.1 : 1.4)
      .attr("marker-start", `url(#${faded ? "dahf2-s" : "dah2-s"})`)
      .attr("marker-end",   `url(#${faded ? "dahf2"   : "dah2"})`);
  }

  // Interpolate white → blue-200 (#bfdbfe) based on fraction t ∈ [0,1]
  function uFill(count) {
    if (count === 0) return "rgba(220,38,38,0.07)";
    const t = count / 64;
    return `rgb(${Math.round(255+t*(191-255))},${Math.round(255+t*(219-255))},255)`;
  }
  function uStroke(count) {
    if (count === 0) return "rgba(220,38,38,0.28)";
    const t = count / 64;
    return `rgb(${Math.round(255+t*(147-255))},${Math.round(255+t*(197-255))},253)`;
  }

  function drawCell2(g, s, count, cx, cy) {
    const gc = g.append("g").attr("transform", `translate(${cx},${cy})`);
    const f = count === 0;

    gc.append("rect")
      .attr("width", W).attr("height", H).attr("rx", 4)
      .attr("fill", uFill(count)).attr("stroke", uStroke(count))
      .attr("stroke-width", 0.8);

    // XMY arrows
    if (s.xm===1) putArrow2(gc, NX[0],NX[1], NM[0],NM[1], 0, f);
    if (s.xm===2) putArrow2(gc, NM[0],NM[1], NX[0],NX[1], 0, f);
    if (s.xm===3) putBiArrow2(gc, NX[0],NX[1], NM[0],NM[1], f);
    if (s.my===1) putArrow2(gc, NM[0],NM[1], NY[0],NY[1], 0, f);
    if (s.my===2) putArrow2(gc, NY[0],NY[1], NM[0],NM[1], 0, f);
    if (s.my===3) putBiArrow2(gc, NM[0],NM[1], NY[0],NY[1], f);
    if (s.xy===1) putArrow2(gc, NX[0],NX[1], NY[0],NY[1], 0, f);
    if (s.xy===2) putArrow2(gc, NY[0],NY[1], NX[0],NX[1], 0, f);
    if (s.xy===3) putBiArrow2(gc, NX[0],NX[1], NY[0],NY[1], f);

    // XMY node circles
    const ns = f ? "#d1d5db" : "#374151";
    const nf = f ? "#f9fafb" : "#ffffff";
    const tf = f ? "#9ca3af" : "#111827";
    [[NX,"X"],[NY,"Y"],[NM,"M"]].forEach(([pos,lbl]) => {
      gc.append("circle").attr("cx",pos[0]).attr("cy",pos[1]).attr("r",NR)
        .attr("fill",nf).attr("stroke",ns).attr("stroke-width",1.5);
      gc.append("text").attr("x",pos[0]).attr("y",pos[1]+4)
        .attr("text-anchor","middle").attr("font-size",10).attr("font-weight","700")
        .attr("fill",tf).text(lbl);
    });

    // U node — dashed, unobserved
    gc.append("circle").attr("cx",NU[0]).attr("cy",NU[1]).attr("r",NRU)
      .attr("fill", f ? "#f9fafb" : "#f8fafc")
      .attr("stroke", f ? "#d1d5db" : "#94a3b8")
      .attr("stroke-width",1.2).attr("stroke-dasharray","2.5 1.5");
    gc.append("text").attr("x",NU[0]).attr("y",NU[1]+3.5)
      .attr("text-anchor","middle").attr("font-size",7).attr("font-weight","600")
      .attr("fill", f ? "#d1d5db" : "#94a3b8").text("U");

    // U-count badge top-left
    if (!f) {
      gc.append("text").attr("x",3).attr("y",10)
        .attr("text-anchor","start").attr("font-size",7).attr("fill","#64748b")
        .text(`${count}/64`);
    }

    // Red cross for ruled-out
    if (f) {
      gc.append("line").attr("x1",5).attr("y1",5).attr("x2",W-5).attr("y2",H-5)
        .attr("stroke","rgba(220,38,38,0.28)").attr("stroke-width",1.5);
      gc.append("line").attr("x1",W-5).attr("y1",5).attr("x2",5).attr("y2",H-5)
        .attr("stroke","rgba(220,38,38,0.28)").attr("stroke-width",1.5);
    }
  }

  const XY_TITLES = ["X — Y  (no direct link)", "X → Y", "X ← Y", "X ↔ Y"];
  const XY_COLORS = ["#334155","#1d4ed8","#7c3aed","#0e7490"];
  const XM_LABELS = ["X—M: —","X→M","M→X","X↔M"];
  const MY_LABELS = ["—","M→Y","Y→M","M↔Y"];

  [0,1,2,3].forEach((xyVal, si) => {
    const sCol = si % 2, sRow = Math.floor(si/2);
    const ox = LPAD + sCol*(secW+SGAP);
    const oy = sRow*(SHEAD+CHEAD+secH+SGAP);

    svg.append("rect").attr("x",ox).attr("y",oy).attr("width",secW).attr("height",SHEAD-2)
      .attr("rx",4).attr("fill",XY_COLORS[xyVal]);
    svg.append("text").attr("x",ox+secW/2).attr("y",oy+17)
      .attr("text-anchor","middle").attr("fill","white")
      .attr("font-size",12).attr("font-weight","700").text(XY_TITLES[xyVal]);

    for (let my = 0; my < 4; my++) {
      svg.append("text")
        .attr("x", ox+my*(W+GAP)+W/2).attr("y", oy+SHEAD+CHEAD-3)
        .attr("text-anchor","middle").attr("font-size",8.5).attr("fill","#64748b")
        .text(MY_LABELS[my]);
    }
    for (let xm = 0; xm < 4; xm++) {
      const cy = oy+SHEAD+CHEAD+xm*(H+GAP);
      svg.append("text").attr("x",ox-5).attr("y",cy+H/2+4)
        .attr("text-anchor","end").attr("font-size",8.5).attr("fill","#64748b")
        .text(XM_LABELS[xm]);
      for (let my = 0; my < 4; my++) {
        const cx = ox+my*(W+GAP);
        drawCell2(svg, {xm, my, xy:xyVal}, countUConfigs({xm,my,xy:xyVal}, activeTools2), cx, cy);
      }
    }
  });

  // Color legend
  const legY = totalH - 20;
  const legX = LPAD;
  svg.append("text").attr("x",legX).attr("y",legY)
    .attr("font-size",8).attr("fill","#64748b").text("U configs consistent:");
  for (let i = 0; i <= 8; i++) {
    const t = i/8;
    svg.append("rect")
      .attr("x", legX+130+i*16).attr("y",legY-10).attr("width",15).attr("height",10).attr("rx",2)
      .attr("fill",`rgb(${Math.round(255+t*(191-255))},${Math.round(255+t*(219-255))},255)`);
  }
  svg.append("text").attr("x",legX+128).attr("y",legY)
    .attr("text-anchor","end").attr("font-size",7.5).attr("fill","#94a3b8").text("0");
  svg.append("text").attr("x",legX+130+9*16+2).attr("y",legY)
    .attr("font-size",7.5).attr("fill","#94a3b8").text("64");

  svg.append("text")
    .attr("transform",`translate(8,${(totalH-32)/2}) rotate(-90)`)
    .attr("text-anchor","middle").attr("font-size",10).attr("fill","#94a3b8")
    .text("← XM relationship (rows within each section) →");

  return svg.node();
}
```

---

## The Fundamental Problem of Causal Inference {#sec-fundamental}

Every participant in the AlterEco eco-label study had **two potential outcomes**:

- $Y_i(0)$: their WTP if they did *not* see the eco-label
- $Y_i(1)$: their WTP if they *did* see the eco-label

Their **individual treatment effect** is $\tau_i = Y_i(1) - Y_i(0)$.

The problem? **We can only observe one of these for any individual.** This is called the **Fundamental Problem of Causal Inference** (Holland, 1986). Everything that follows is a strategy for getting around it.

```{r dgs-potential-outcomes}
#| code-summary: "▶ Build the data-generating system"
set.seed(2024)
N <- 2000

env        <- rnorm(N, 0, 1)
price_sens <- rnorm(N, 0, 1)
income     <- rnorm(N, 0, 1)

Y0 <- 5.50 + 0.80*env - 0.50*price_sens + 0.30*income + rnorm(N, 0, 1.10)
Y0 <- pmax(1, pmin(10, Y0))

ITE <- 0.60 + 0.80*env - 0.55*price_sens + rnorm(N, 0, 0.40)
Y1  <- pmax(1, pmin(10, Y0 + ITE))

ATE_true <- mean(Y1 - Y0)
cat(sprintf("True ATE  = %.3f\nTrue ITE SD = %.3f\n", ATE_true, sd(Y1 - Y0)))
```

```{r fundamental-problem-table, echo=FALSE}
#| code-summary: "▶ Show the missing-data structure"
set.seed(42)
demo_idx <- c(which(env > 1.2)[1:2], which(abs(env) < 0.15)[1], which(env < -1.2)[1:2])
assigned <- c("Control","Treatment","Control","Treatment","Control")
tibble(
  Person             = c("Sophie (high concern)","Rens (moderate+)","Lotte (neutral)","Daan (low concern)","Kim (price-sensitive)"),
  `Env concern`      = round(env[demo_idx], 1),
  `Y(0) No label`   = round(Y0[demo_idx], 2),
  `Y(1) With label` = round(Y1[demo_idx], 2),
  `True ITE`        = round(Y1[demo_idx]-Y0[demo_idx], 2),
  `Assigned to`     = assigned,
  `What we observe` = c(
    sprintf("Y(0)=$%.2f  Y(1)=???", Y0[demo_idx[1]]),
    sprintf("Y(0)=???   Y(1)=$%.2f", Y1[demo_idx[2]]),
    sprintf("Y(0)=$%.2f  Y(1)=???", Y0[demo_idx[3]]),
    sprintf("Y(0)=???   Y(1)=$%.2f", Y1[demo_idx[4]]),
    sprintf("Y(0)=$%.2f  Y(1)=???", Y0[demo_idx[5]])
  )
) |> knitr::kable(caption="The fundamental problem: exactly ONE potential outcome per person is observed.")
```

### What is a counterfactual?

A **counterfactual** is the answer to a "what if" question about the road not taken. For Sophie, who saw the eco-label and paid \$8.20, the counterfactual is: *what would she have paid if she had not seen it?* That number — say, \$6.80 — never happened in our world, but it is what the word "caused" points at. The eco-label *caused* a \$1.40 increase in Sophie's WTP precisely because \$8.20 (observed) minus \$6.80 (counterfactual) equals \$1.40.

This might sound abstract, but counterfactuals are built into everyday causal language:

- *"Taking the earlier train saved me 20 minutes"* — the counterfactual is arriving 20 minutes later had you not taken it.
- *"The aspirin cured my headache"* — the counterfactual is the headache persisting without aspirin.
- *"The eco-label increased her WTP by \$1.40"* — the counterfactual is what she would have paid without it.

The table above makes the problem concrete: we assigned Sophie to the control condition and observed her pay \$6.80, so her counterfactual *Y(1)* — what she would have paid with the label — is invisible. We assigned Daan to the treatment and observed *Y(1)*, so his *Y(0)* is invisible. Both missing cells are counterfactuals. The **fundamental problem of causal inference** is that every individual appears in exactly one condition, so every individual's causal effect requires reasoning about a world that did not occur.

::: {.callout-tip}
## Counterfactuals in your own research

Before running any analysis, try writing out the counterfactual claim your paper makes:

> *"If participant* $i$ *had been assigned to condition A instead of condition B, their outcome would have been \_\_\_ rather than \_\_\_."*

If you cannot fill in the blank — or if filling it in requires assuming things you have not measured — your causal claim is more fragile than it looks. This exercise is the fastest way to spot whether your design actually supports the inference you intend to draw.
:::

## Individual Treatment Effects (ITE) {#sec-ite}

The **individual treatment effect** for person $i$ is:

$$\tau_i = Y_i(1) - Y_i(0)$$

This is the *only* truly assumption-free definition of causality: it is the effect *on this person*, *in this moment*, caused by *this treatment*. Did the eco-label change **this consumer's** willingness to pay? That is the ITE.

The ITE is simultaneously the most interpretable estimand and the one we can *never* directly observe — because every person appears in only one condition at one point in time. The histogram below shows the true ITE distribution that exists in reality but is entirely invisible to any researcher. Notice how wide it is: some consumers gain over $2 in WTP from seeing the label; others gain nothing; a few are actually put off by it. This heterogeneity is the central fact that all other estimands must grapple with.

```{r ite-distribution, fig.height=4.5, fig.width=9}
#| code-summary: "▶ Plot the distribution of individual treatment effects"
tibble(ITE=ITE,
       env_grp=cut(env, c(-Inf,-0.5,0.5,Inf), labels=c("Low","Moderate","High"))) |>
  ggplot(aes(x=ITE, fill=env_grp)) +
  geom_histogram(bins=50, alpha=0.75, position="identity") +
  geom_vline(xintercept=ATE_true, linetype="dashed", linewidth=1.1, colour="grey20") +
  annotate("text", x=ATE_true+0.06, y=82,
           label=sprintf("ATE = $%.2f", ATE_true), hjust=0, size=3.5, fontface="bold") +
  scale_fill_manual(values=c("High"=clr_eco,"Moderate"="#74b9a0","Low"=clr_ctrl)) +
  labs(x="Individual Treatment Effect  Y(1)−Y(0)  ($)", y="Count", fill=NULL,
       title="The eco-label effect is wildly different across individuals",
       subtitle="In real research you never see this distribution — only the ATE is estimable") +
  theme_mod3()
```

## Average Treatment Effect (ATE) {#sec-ate}

$$\text{ATE} = E[\tau_i] = E[Y_i(1) - Y_i(0)]$$

The ATE is the expectation of the ITE across the entire target population — the answer to "what would happen on average if this treatment were applied to everyone?"

**The assumption needed to use ATE as a proxy for ITE:** **Treatment effect homogeneity** — the belief that every unit responds the same way ($\tau_i = \tau$ for all $i$). If the eco-label added exactly the same WTP boost to every consumer, the ATE and every ITE would be identical, and knowing the ATE would fully characterise all individual responses.

In practice, homogeneity is almost never plausible. The ITE histogram above makes this plain: effects range from negative to strongly positive. **This means the ATE answers a different question than the ITE.** The ATE is the right estimand for population-level policy decisions — "should supermarkets mandate eco-labels on all products?" — because its answer applies in aggregate even when individual effects vary wildly. But it is a poor guide to predicting any individual consumer's response, because it averages over enormous heterogeneity.

**Why the ATE is still the gold standard for experiments:** In a randomised trial, random assignment makes the treated and control groups exchangeable in expectation, so the observed mean difference is an unbiased estimate of the ATE *without* assuming homogeneity. The experiment recovers the ATE honestly; it is the *subsequent inference* from ATE to individual effect that requires the homogeneity assumption.

That said, the exchangeability assumption is more fragile than it appears — and more fragile than is often acknowledged. Random assignment guarantees balance *in expectation* across infinite replications, but any single study can produce unlucky covariate imbalance. More importantly, the behavioural conditions for exchangeability extend beyond the draw of assignment: if participants in the treatment condition behave differently because they *know* they are treated — through demand characteristics, Hawthorne effects, or awareness of experimental purpose — or if the sample differs systematically from the target population, then the observed mean difference estimates something other than the population ATE. Module 2 explored these threats in detail: the design features that protect exchangeability (pre-registration, cover stories, unobtrusive measures, stimulus sampling, representative recruitment) are precisely what make an ATE estimate credible rather than merely randomised.

```{r ate-rct}
treat_rct <- rbinom(N, 1, 0.5)
Y_obs_rct <- ifelse(treat_rct==1, Y1, Y0)
df_rct    <- tibble(id=1:N, treat=treat_rct, Y_obs=Y_obs_rct, Y0, Y1, ITE, env, price_sens, income)

ATE_est <- mean(Y_obs_rct[treat_rct==1]) - mean(Y_obs_rct[treat_rct==0])
lm(Y_obs ~ treat, data=df_rct) |> tidy() |>
  filter(term=="treat") |>
  transmute(Estimand="ATE", `True value`=round(ATE_true,3), Estimate=round(estimate,3),
            SE=round(std.error,3),
            `95% CI`=sprintf("[%.3f, %.3f]", estimate-1.96*std.error, estimate+1.96*std.error)) |>
  knitr::kable(caption="ATE estimate from randomised eco-label experiment")
```

## Conditional Average Treatment Effect (CATE) {#sec-cate}

$$\text{CATE}(x) = E[\tau_i \mid X_i = x]$$

The CATE partitions the population by pre-treatment characteristics $X$ and estimates a separate average treatment effect within each subgroup. The eco-label may add \$1.40 WTP for consumers with high environmental concern but only \$0.15 for low-concern consumers — the ATE buries this story.

**What CATE relaxes:** The global homogeneity assumption. CATE allows different *types* of consumers to respond differently; it only requires homogeneity *within* each subgroup (everyone with the same covariate profile $x$ is assumed to respond identically to their group's average).

**What CATE still assumes — and why the gap to ITE remains:**

1. **You have measured the right moderators — and measured them well.** You can only condition on variables you have collected. If the strongest driver of heterogeneity is something unmeasured — say, a consumer's specific shopping occasion or the brand's existing equity — your CATEs are averages over that unobserved heterogeneity and may not correspond to any real subgroup's true effect. But even when you *have* collected the right moderator, measurement quality matters enormously: a construct that suffers from discriminant validity failures, cross-group non-invariance, or systematic response bias (see Module 1) produces distorted CATE estimates even if the causal heterogeneity is real. A CATE model built on "environmental concern" items that actually blend concern with general pro-social identity will attribute the wrong heterogeneity to the wrong mechanism. The entire CATE machinery is predicated on the measurement infrastructure being sound.

2. **Correct functional form when extrapolating.** Estimating CATE for a continuous moderator (e.g., environmental concern score) requires assuming a relationship between the moderator and the treatment effect — linearity, correct interactions, no omitted non-linearities. Even a well-estimated linear CATE model will give wrong predictions for individuals whose true effect relationship is non-linear.

3. **Within-cell homogeneity.** Even if you carve out fine-grained cells, predicting an *individual's* response from their cell's CATE still assumes every person in that cell responds identically to the cell average. As cells get finer this becomes a stronger claim about fewer people, eventually reducing to the same problem as the ITE itself.

The plot below shows CATE estimates versus the (unobservable) true CATEs across four environmental concern bins. The estimates track the truth well here — but each bar still summarises a distribution of individual effects within the bin.

```{r cate-by-subgroup, fig.height=4.5, fig.width=9}
df_rct |>
  mutate(env_grp=cut(env, c(-Inf,-1,0,1,Inf), labels=c("Very Low","Low","High","Very High"))) |>
  group_by(env_grp) |>
  summarise(CATE_true=mean(Y1-Y0), CATE_est=mean(Y_obs[treat==1])-mean(Y_obs[treat==0]),
            n=n(), .groups="drop") |>
  pivot_longer(starts_with("CATE"), names_to="type", values_to="val") |>
  ggplot(aes(x=env_grp, y=val, fill=type)) +
  geom_col(position="dodge", alpha=0.85, width=0.65) +
  geom_hline(yintercept=ATE_true, linetype="dashed", colour="grey30") +
  annotate("text", x=0.65, y=ATE_true+0.05, label="Overall ATE", size=3.5) +
  scale_fill_manual(values=c(CATE_true=clr_eco, CATE_est=clr_ctrl),
                    labels=c("True CATE","Estimated CATE")) +
  labs(x="Environmental Concern", y="Eco-label effect ($)", fill=NULL,
       title="CATE: the label works far better for environmentally engaged consumers",
       subtitle="Reporting only the ATE hides this story completely") +
  theme_mod3()
```

## Compliance, ITT, LATE, ATT, and ATC {#sec-late}

When the treatment of interest cannot be directly assigned — only *encouraged* — the population stratifies into four **compliance types** based on how each unit would respond to assignment under both conditions:

| Type | $D_i(Z{=}0)$ | $D_i(Z{=}1)$ | Description |
|---|---|---|---|
| **Complier** | 0 | 1 | Takes treatment only when assigned |
| **Always-taker** | 1 | 1 | Takes treatment regardless |
| **Never-taker** | 0 | 0 | Never takes treatment |
| **Defier** | 1 | 0 | Takes treatment only when *not* assigned |

In the eco-label context: a store may be assigned to display eco-labels ($Z=1$), but individual consumers may or may not actually look at them ($D=1$). Consumers who would read *any* label they see are always-takers; those who read the label only because the store displays it prominently are compliers; those who never engage with shelf labels are never-takers.

Each estimand below targets a *different slice* of the ITE distribution — and each translation back to the ITE requires different assumptions:

**Intent-to-treat (ITT):** The effect of *assignment* to the treatment condition regardless of actual receipt. In an imperfect-compliance world, ITT underestimates the ATE by a factor equal to the compliance rate, because some assigned units are not actually treated. ITT is the most conservative and assumption-light estimand here, but it answers the wrong question if you care about the treatment's biological or psychological mechanism rather than the logistical reality of assignment.

**Local Average Treatment Effect (LATE):** The ITT scaled by the first-stage compliance rate (the Wald estimator). This recovers the average ITE *for compliers only* — those who take the treatment because they were assigned to, not because they always would have. The LATE is the right estimand when you care about the policy lever of assignment, and it rests on two additional assumptions beyond randomisation:

- **Monotonicity:** No defiers. Every unit's probability of taking treatment is weakly *higher* under assignment ($D_i(1) \geq D_i(0)$ for all $i$). This rules out consumers who deliberately avoid eco-labels when stores promote them.
- **Exclusion restriction:** Assignment $Z$ affects the outcome *only through* take-up $D$ — not through any direct path. If displaying eco-labels changes store atmosphere and affects WTP independently of whether any consumer reads them, the exclusion restriction fails. This is a structural version of the manipulation validity problem from Module 2: just as a laboratory manipulation can inadvertently activate demand characteristics or alter the psychological context beyond the intended treatment, an instrument can influence outcomes through channels the researcher neither intended nor anticipated. The same design rigour that Module 2 prescribed for protecting internal validity — careful operationalisation, manipulation checks, ruling out alternative accounts of the treatment — applies equally to establishing that an instrument is truly excludable.

Even with these assumptions satisfied, LATE still describes an *average* over compliers' ITEs, requiring within-complier homogeneity to predict any individual complier's response.

**ATT and ATC:** The average treatment effect for units that *did* (ATT) or *did not* (ATC) actually receive the treatment. Because treatment receipt may be self-selected — consumers who seek out eco-labels are likely those who care most about sustainability and therefore respond most strongly — ATT $\neq$ ATE $\neq$ ATC. Translating from ATT to ITE for a specific treated unit requires assuming that unit is representative of the treated group, which is a homogeneity claim within the treated subpopulation.

::: {.callout-note}
## The chain of assumptions back to ITE

Every estimand is a different answer to a different version of the causal question. The further you move from the population-wide ATE toward more targeted estimands, the more specific the claim — but also the more assumptions required to bridge back to the individual level:

| Estimand | Population | Additional assumption(s) to reach ITE |
|---|---|---|
| **ATE** | All units | Treatment effect homogeneity |
| **CATE** | Subgroup $X = x$ | Correct moderator specification; within-cell homogeneity |
| **LATE** | Compliers | Monotonicity; exclusion restriction; within-complier homogeneity |
| **ATT** | Treated units | Representativeness of the treated for the target individual |
| **ATC** | Control units | Representativeness of controls for the target individual |

None of these estimands *is* the ITE. All of them are averages over a group of individuals. The ITE is the only true unit-level causal effect, and it is the only one that requires no assumptions about other units — but it is also the only one we can never observe. Applied causal inference is the art of choosing the estimand whose assumptions are most defensible given your design and question, while being explicit that a gap always remains between your estimate and the individual-level effect that treatment ultimately acts upon.
:::

```{r compliance-dgs}
set.seed(2024)
Z <- treat_rct
p_comply_if_Z1 <- plogis(0.8 + 0.6*env)
p_always_taker <- plogis(-2.5 + 0.5*env)
D_if_Z1 <- rbinom(N, 1, p_comply_if_Z1)
D_if_Z0 <- rbinom(N, 1, p_always_taker)
# Enforce monotonicity: no defiers (D_i(1) >= D_i(0) for all i)
D_if_Z0 <- pmin(D_if_Z0, D_if_Z1)
D       <- ifelse(Z==1, D_if_Z1, D_if_Z0)
compliance_type <- case_when(
  D_if_Z1==1 & D_if_Z0==0 ~ "Complier",
  D_if_Z1==1 & D_if_Z0==1 ~ "Always-taker",
  D_if_Z1==0 & D_if_Z0==0 ~ "Never-taker",
  D_if_Z1==0 & D_if_Z0==1 ~ "Defier"
)
knitr::kable(table(compliance_type), col.names=c("Compliance type","Count"),
             caption="True compliance breakdown (observable only in simulation)")
```

```{r itt-late-att-atc}
Y_obs_iv <- ifelse(D==1, Y1, Y0)
df_iv    <- tibble(Z, D, Y_obs_iv, Y0, Y1, ITE, compliance_type, env, price_sens)

ITT       <- mean(Y_obs_iv[Z==1]) - mean(Y_obs_iv[Z==0])
FS        <- mean(D[Z==1]) - mean(D[Z==0])
LATE_est  <- ITT / FS
LATE_true <- mean(ITE[compliance_type=="Complier"])
ATT_est   <- mean(ITE[D==1])
ATC_est   <- mean(ITE[D==0])

tibble(
  Estimand    = c("ATE","ITT","LATE (Wald)","ATT","ATC"),
  Description = c(
    "Effect averaged over all 2,000 consumers",
    "Effect of *assigning* the label regardless of compliance",
    "Effect for consumers who complied *because* they were assigned",
    "Effect for consumers who *actually saw* the label",
    "Effect for those who *did not see* it"
  ),
  `True value` = round(c(ATE_true, ITT, LATE_true, ATT_est, ATC_est), 3),
  Estimated    = round(c(ATE_true, ITT, LATE_est,  ATT_est, ATC_est), 3)
) |> knitr::kable(caption="Five ways to summarise 'the effect' of the eco-label")
```

```{r estimands-plot, fig.height=4, fig.width=8, echo=FALSE}
tibble(
  Estimand=c("ATE\n(all consumers)","ITT\n(assigned, any)","LATE\n(compliers only)",
             "ATT\n(label seen)","ATC\n(label not seen)"),
  Value=c(ATE_true, ITT, LATE_est, ATT_est, ATC_est),
  Type=c("Population","Assignment","Compliers","Treated","Control")
) |>
  mutate(Estimand=fct_reorder(Estimand, Value)) |>
  ggplot(aes(y=Estimand, x=Value, fill=Type)) +
  geom_col(width=0.6, alpha=0.85) +
  geom_vline(xintercept=ATE_true, linetype="dashed", colour="grey30") +
  scale_fill_brewer(palette="Set2") +
  labs(x="Eco-label effect on WTP ($)", y=NULL, fill=NULL,
       title="Five estimands — same experiment, different questions",
       subtitle="ATT > ATE > ATC: consumers who seek eco-labels respond more strongly to them") +
  theme_mod3()
```

---

## Causal Graphs and the Logic of Identification {#sec-dags}

Potential outcomes give you a precise language for *what* you want to estimate. But they are largely silent on *how* to estimate it — which variables to control for, which to leave alone, and why. **Directed acyclic graphs (DAGs)**, developed by Judea Pearl and colleagues, fill that gap. They provide a visual language for encoding your causal assumptions and a set of rules for reading off, directly from the picture, which adjustment strategies identify causal effects and which produce bias.

### A pipe-wall analogy

Before the formalism, an analogy that keeps the concepts concrete.

Imagine you are standing in front of a sealed wall. On the left side is an input valve, $X$. On the right side is an output spout, $Y$. You can see the openings and gauges on the outside, but the plumbing inside the wall is hidden.

A **DAG is your proposed blueprint of the pipes behind the wall.** Every arrow is a claim about how water can flow. Every missing arrow is equally a claim: it says there is no direct pipe between those two openings.

The causal question is: *"If I turn the $X$ valve myself, how much does the $Y$ spout change?"* That is different from asking: *"When the $X$ gauge reads high, is the $Y$ gauge also high?"* The second is observational — you are watching natural variation. The first is interventional — you are doing something. The pipe analogy will make each path structure, and the distinction between them, easy to see.

### What is a DAG?

A DAG has two ingredients:

- **Nodes** — variables (observed or unobserved)
- **Directed edges (arrows)** — each arrow $A \rightarrow B$ asserts that $A$ directly causes $B$, holding all other direct causes of $B$ constant

The "acyclic" part means no variable can cause itself through any chain of arrows — time must flow forward. You can represent feedback by treating each time-point as a separate node ($X_t \rightarrow Y_t \rightarrow X_{t+1}$), but within a cross-sectional DAG, cycles are forbidden.

Every arrow you draw is an assumption. Every missing arrow is also an assumption — it claims the two variables are not directly linked. A DAG made honest is a complete inventory of what you believe about the causal structure of your system.

### Three fundamental path structures

Three building blocks cover nearly every situation you will encounter:

```{r dag-three-structures, fig.height=3.2, fig.width=10, echo=FALSE}
#| code-summary: "▶ Show the three fundamental DAG structures"

# 1. Chain (mediator)
p_chain <- .dag(
  nodes   = list(X=c(0,0), M=c(1,0), Y=c(2,0)),
  solid   = list(c("X","M"), c("M","Y")),
  title   = "Chain  X → M → Y\n(mediator)"
)

# 2. Fork (common cause / confounder)
p_fork <- .dag(
  nodes   = list(U=c(1,1), X=c(0,0), Y=c(2,0)),
  solid   = list(c("U","X"), c("U","Y")),
  unobserved = "U",
  title   = "Fork  X ← U → Y\n(confounder)"
)

# 3. Collider
p_collider <- .dag(
  nodes   = list(X=c(0,0), M=c(1,0), Y=c(2,0)),
  solid   = list(c("X","M"), c("Y","M")),
  title   = "Collider  X → M ← Y\n(collider on M)"
)

p_chain | p_fork | p_collider
```

**Chain (mediator):** $X \rightarrow M \rightarrow Y$. $X$ causes $M$, $M$ causes $Y$. In this simple structure — where there is no direct $X \rightarrow Y$ path and no confounding — the entire association between $X$ and $Y$ runs through $M$. If you block $M$ by controlling for it, you sever the only route and the $X$–$Y$ association disappears. In the pipe analogy: the pipe from $X$ to $Y$ passes through a visible middle gauge $M$. Clamping $M$ shut stops the flow — which is useful if you are asking whether anything bypasses $M$, but wrong if you just want to know the total effect of opening $X$.

**Fork (confounder):** $X \leftarrow U \rightarrow Y$. A common cause $U$ produces an association between $X$ and $Y$ even if $X$ does not cause $Y$ at all. Controlling for $U$ blocks this spurious path. In the pipe analogy: there is a hidden water source $U$ feeding both the $X$ side and the $Y$ side. The $X$ and $Y$ gauges move together not because $X$ feeds $Y$ but because both are fed from the same upstream source.

**Collider:** $X \rightarrow M \leftarrow Y$. Both $X$ and $Y$ cause $M$, but the arrows flow *into* $M$, so the path is blocked — $X$ and $Y$ are not associated in the general population. Conditioning on $M$ (controlling for it, stratifying on it, or selecting a sample where $M$ is fixed) *opens* a spurious association that did not previously exist. In the pipe analogy: $M$ is a warning light triggered by either $X$ pressure or $Y$ pressure. In the general system $X$ and $Y$ are unrelated. But if you look only at cases where the warning light is on, you create a false connection — low $X$ pressure implies $Y$ must have been high enough to trigger it, and vice versa.

::: {.callout-important}
## The key rule

A path transmits association between two variables **unless** it contains a collider *or* you condition on a non-collider node on the path. Conditioning on a collider **opens** a path that was closed.

This has an immediate practical implication: **not every control variable improves your estimate.** If you control for a collider, you introduce bias. Choosing covariates by "putting in everything correlated with Y" is precisely the wrong strategy.
:::

### Collider bias: a concrete example

Suppose you study whether conscientiousness ($X$) causes job performance ($Y$). You restrict your sample to employees who were *hired* — conditioning on $M = \text{hired}$. The structure is:

$$\text{Conscientiousness} \rightarrow \text{Hired} \leftarrow \text{Prior performance signals}$$

and prior performance signals also predict later job performance ($Y$). Hiring is a collider: both conscientiousness and prior performance independently raise the probability of being hired. In the full applicant pool, conscientiousness and prior performance signals are unrelated. But inside the hired group, they are negatively correlated: a candidate with low conscientiousness who was still hired must have had strong prior performance to compensate, and vice versa. Conditioning on the collider opens the path Conscientiousness $\leftrightarrow$ Prior signals $\rightarrow$ Performance, creating a spurious *negative* association between conscientiousness and performance within your sample. This is why studies restricted to elite samples — top MBA programmes, Fortune 500 firms, clinical trial completers — routinely find attenuated or reversed relationships compared to population-level estimates.

### Intervention vs. observation: the do-operator

Observational data tell you about the world as it is — how variables co-occur. Causal questions ask about the world as it *would be* if you intervened. Pearl's **do-operator** makes this distinction precise.

- $P(Y \mid X = x)$ — the probability of $Y$ given that you *observe* $X = x$ (observational)
- $P(Y \mid \mathit{do}(X = x))$ — the probability of $Y$ given that you *set* $X = x$ by intervention (interventional)

These are equal only when $X$ has no upstream common cause with $Y$. Suppose environmental values ($U$) cause both label exposure ($X$) and WTP ($Y$). Then $P(Y \mid X = \text{high})$ partly reflects the kind of person who ends up in high-$X$ situations — one with high environmental values, and therefore high WTP regardless. $P(Y \mid \mathit{do}(X = \text{high}))$ removes that selection: you have forced $X$ to be high, so $U$ no longer determines who gets which value of $X$, and the observed $Y$ difference is purely causal.

In the pipe analogy: observing high $X$ pressure may just tell you the hidden upstream source $U$ is running strong. *Setting* $X$ yourself — attaching your own controlled pump — cuts $X$ free from whatever was feeding it naturally. That is why randomisation identifies causal effects: it is a physical implementation of $\mathit{do}(X)$.

Random assignment is the cleanest implementation of $\mathit{do}(X = x)$: by forcing $X$ to the value you choose, you sever all incoming arrows to $X$. The resulting $P(Y \mid \mathit{do}(X))$ is identified without further adjustment. When you cannot randomise, identification requires finding an *adjustment set* that mimics what randomisation would have done.

### The backdoor criterion {#sec-backdoor}

A **backdoor path** from $X$ to $Y$ is any path that starts by going *backwards* along an incoming arrow into $X$. Backdoor paths carry confounding association — they are routes through which $X$ and $Y$ move together for reasons that have nothing to do with $X$ causing $Y$.

In the pipe analogy: a backdoor path is a hidden pipe that enters $X$ from behind (upstream contamination) and then reaches $Y$. The water appears to flow from $X$ to $Y$ but is really entering through a hidden joint behind the $X$ valve. To estimate the true effect of turning the $X$ valve, you need to close every backdoor route — without closing pipes that are downstream consequences of $X$.

The **backdoor criterion** says: a set of variables $Z$ is sufficient to identify the causal effect of $X$ on $Y$ if $Z$ (1) blocks all backdoor paths from $X$ to $Y$, and (2) does not include any descendant of $X$ — no post-treatment variables or mediators, since those lie downstream and blocking them would remove genuine causal flow.

```{r dag-backdoor, fig.height=3.5, fig.width=6, echo=FALSE}
#| code-summary: "▶ Show backdoor confounding"

# Eco-label example: U = environmental values affects both store choice (X) and WTP (Y)
.dag(
  nodes      = list(U=c(1,1.4), X=c(0,0), Y=c(2,0)),
  solid      = list(c("X","Y")),
  dashed     = list(c("U","X"), c("U","Y")),
  unobserved = "U",
  title      = "Backdoor: U confounds X → Y"
)
```

In the eco-label study, suppose consumers who chose to shop at eco-conscious stores ($X$ = saw label) differ systematically from other shoppers in their underlying environmental values ($U$). $U$ causes both store choice (and thus label exposure) and WTP ($Y$). The path $X \leftarrow U \rightarrow Y$ is a backdoor path. If we could measure and control for $U$, we would block this path and identify the causal effect. Randomising store assignment — so $X$ is no longer caused by $U$ — severs the backdoor path entirely.

When a valid backdoor adjustment set exists, the causal effect is:

$$P(Y \mid \mathit{do}(X)) = \sum_z P(Y \mid X, Z=z)\, P(Z=z)$$

This is just a weighted average of the $X \rightarrow Y$ association within strata of $Z$ — familiar as regression adjustment or stratification.

### Controlling for mediators: when to and when not to {#sec-mediator-control}

The mediator structure ($X \rightarrow M \rightarrow Y$) creates a choice: do you control for $M$ or not? The answer depends entirely on your research question — not on convention or habit.

**If you want the total effect of $X$ on $Y$**, do not control for $M$. Controlling for $M$ blocks the indirect path $X \rightarrow M \rightarrow Y$ and returns only whatever direct $X \rightarrow Y$ path exists outside $M$. In the pipe analogy: if you clamp the middle gauge shut, you are no longer measuring the total flow caused by opening $X$ — you have blocked the route through which most of $X$'s effect travels.

**If you want the direct effect of $X$ on $Y$ holding $M$ fixed**, you may control for $M$ — but only if the $M \rightarrow Y$ relationship is itself unconfounded. If there is an unmeasured variable that affects both $M$ and $Y$, then conditioning on $M$ no longer identifies the direct effect: the mediator–outcome confounding is left unblocked, and you may introduce additional bias rather than remove it. The requirement is not just that $M$ is measured, but that you can credibly claim the $M$–$Y$ relationship is clean.

In the eco-label context: suppose the label works by (a) directly signalling product quality and (b) activating environmental guilt ($M$), which then raises WTP. If you want to know "does the label raise WTP overall?", do not control for guilt — you would delete the mechanism. If you want to know "does the label raise WTP *beyond* the guilt pathway?", you may control for guilt — but only after establishing that no unmeasured variable (say, general pro-social identity) jointly drives both guilt and WTP.

### The frontdoor criterion {#sec-frontdoor}

Sometimes the backdoor path cannot be blocked — $U$ is unobserved and there is no valid instrument. But if a mediator $M$ satisfies three conditions, the causal effect is still identifiable through the **frontdoor criterion**:

1. $M$ is causally between $X$ and $Y$ (all causal effect of $X$ on $Y$ flows through $M$)
2. There is no unblocked backdoor path from $X$ to $M$
3. All backdoor paths from $M$ to $Y$ are blocked by $X$

```{r dag-frontdoor, fig.height=3.5, fig.width=6, echo=FALSE}
#| code-summary: "▶ Show frontdoor structure"

.dag(
  nodes      = list(U=c(1,1.4), X=c(0,0), M=c(1,0), Y=c(2,0)),
  solid      = list(c("X","M"), c("M","Y")),
  dashed     = list(c("U","X"), c("U","Y")),
  unobserved = "U",
  title      = "Frontdoor: M identifies X → Y despite unobserved U"
)
```

The classic example: smoking ($X$) causes lung cancer ($Y$) through tar deposits ($M$). Genetic predisposition ($U$) confounds the $X$–$Y$ relationship — heavier smokers may also carry genetic risk factors that independently raise cancer risk. But the $X \rightarrow M$ path is clean: whether someone smokes determines tar accumulation, and there is no direct route for $U$ to affect tar independently of smoking. And the backdoor paths from $M$ to $Y$ all pass through $X$, which is observed and can be adjusted for. So we can (1) estimate the effect of smoking on tar using variation in smoking status, then (2) estimate the effect of tar on cancer adjusting for smoking status to close the backdoor, and finally (3) combine these two stages to recover the total causal effect of smoking on cancer — even though $U$ is never measured.

In the pipe analogy: the back of the wall is inaccessible (you cannot measure $U$), but all the water from $X$ to $Y$ must pass through a visible meter $M$. You can read the $X \rightarrow M$ flow directly (the $X$–$M$ segment is clean), and you can read the $M \rightarrow Y$ flow after accounting for the $X$ value (which blocks the only remaining backdoor into $M$). Combining both readings reconstructs the total flow from $X$ to $Y$ without ever seeing the hidden upstream source.

In management research, the frontdoor is relevant when you have a clear causal chain — policy $\rightarrow$ behaviour $\rightarrow$ outcome — and you can measure the behaviour even though the policy's take-up is confounded by unobservables. It is less commonly exploited than instrumental variables but rests on weaker assumptions when the full mediating structure is genuinely present and measurable.

::: {.callout-note}
## DAGs in practice

You do not need to run Pearl's full do-calculus algebra to benefit from DAGs. Drawing the DAG before your analysis forces three productive habits:

1. **Enumerate your assumptions** — every absent arrow is a claim that two variables are not directly linked.
2. **Classify your covariates** — before adding a control variable, ask: is it a confounder (should adjust), a mediator (adjust only for direct effect), or a collider (never adjust)?
3. **Identify the backdoor paths** — these are the precise channels through which confounding flows, and blocking them is the definition of identification.

Modules 3.2–3.5 will apply this framework to regression discontinuity, difference-in-differences, and mediation analysis, where the DAG determines whether each design's identifying assumption is plausible.
:::

---

## Researcher Checklist: Causal Effects and Estimands {#sec-checklist-part1}

::: {.callout-note}
## Key questions before choosing a method

- [ ] **Which estimand answers your research question — ATE, ATT, or LATE?** These are not interchangeable. The ATT estimates the effect for the self-selected treated group; the ATE estimates it for everyone; the LATE estimates it for units near a threshold. Choose based on the substantive question, not on which is easiest to identify.
- [ ] **Have you drawn the causal DAG?** The DAG determines which variables are confounders (must be controlled), colliders (must not be controlled), and mediators (control or not depending on whether you want the total or direct effect). Choosing covariates without a DAG risks collider bias.
- [ ] **What is your source of as-good-as-random variation?** Every causal claim requires a mechanism that makes treatment assignment independent of potential outcomes. Identify yours explicitly — randomization, a threshold, a policy rollout, a matched comparison — and state the assumption it requires.
- [ ] **What evidence rules out reverse causality and confounding?** Causal inference is deduction by elimination. Each methodological tool (time ordering, randomization, IV, matching) removes some alternative structures. Which ones remain?
:::
