---
title: "Part 1: What Is a Causal Effect?"
---

```{r setup, include=FALSE}
options(repos = c(CRAN = "https://cran.rstudio.com/"))

if (!requireNamespace("remotes",  quietly = TRUE)) install.packages("remotes")
if (!requireNamespace("synthdid", quietly = TRUE))
  remotes::install_github("synth-did/synthdid")

pkgs <- c(
  "tidyverse", "broom", "knitr", "patchwork",
  "conflicted",
  "MatchIt", "WeightIt", "cobalt",
  "rdrobust", "rddensity",
  "did",
  "mediation", "robmed",
  "estimatr", "sandwich", "lmtest"
)
for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p, dependencies = TRUE)
  library(p, character.only = TRUE)
}
library(synthdid)

library(conflicted)
conflicts_prefer(dplyr::select, dplyr::filter, dplyr::rename, dplyr::mutate,
                 dplyr::summarise, dplyr::arrange, dplyr::lag)

dir.create("figs", recursive = TRUE, showWarnings = FALSE)
knitr::opts_chunk$set(
  echo = TRUE, fig.width = 9, fig.height = 5,
  fig.path = "figs/", dev = "svg"
)

clr_eco     <- "#2d6a4f"
clr_ctrl    <- "#4a90d9"
clr_treated <- clr_eco

theme_mod3 <- function() {
  theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          strip.background = element_rect(fill = "#f0f4f8", colour = NA),
          legend.position  = "bottom")
}

# ── Lightweight DAG helper ─────────────────────────────────────────────────────
.dag <- function(nodes, solid = NULL, dashed = NULL,
                 unobserved = character(0), title = "") {
  nd <- do.call(rbind, lapply(names(nodes), function(nm)
    data.frame(id=nm, x=nodes[[nm]][1], y=nodes[[nm]][2], stringsAsFactors=FALSE)))
  nd$type <- ifelse(nd$id %in% unobserved, "unobs", "obs")
  segs <- function(pairs, lt) {
    if (is.null(pairs) || length(pairs)==0) return(NULL)
    do.call(rbind, lapply(pairs, function(ab) {
      fr <- nd[nd$id==ab[1],]; to <- nd[nd$id==ab[2],]
      dx <- to$x-fr$x; dy <- to$y-fr$y
      len <- sqrt(dx^2+dy^2)+1e-9; pad <- 0.12
      data.frame(x=fr$x+pad*dx/len, y=fr$y+pad*dy/len,
                 xend=to$x-pad*dx/len, yend=to$y-pad*dy/len,
                 lt=lt, stringsAsFactors=FALSE)
    }))
  }
  edf <- rbind(segs(solid,"solid"), segs(dashed,"dashed"))
  xs <- range(nd$x); ys <- range(nd$y)
  p <- ggplot() + theme_void() + ggtitle(title) +
    coord_cartesian(xlim=c(xs[1]-.3,xs[2]+.3), ylim=c(ys[1]-.3,ys[2]+.3), expand=FALSE) +
    theme(plot.title=element_text(size=10,face="bold",hjust=.5,margin=margin(b=6)),
          plot.background=element_rect(fill="grey98",colour="grey80",linewidth=.4),
          plot.margin=margin(10,16,10,16), legend.position="none")
  if (!is.null(edf) && nrow(edf)>0)
    p <- p +
      geom_segment(data=edf, aes(x=x,y=y,xend=xend,yend=yend,colour=lt,linetype=lt),
                   arrow=arrow(length=unit(.20,"cm"),type="closed"), linewidth=.85) +
      scale_colour_manual(values=c(solid="grey20",dashed="firebrick")) +
      scale_linetype_manual(values=c(solid="solid",dashed="dashed"))
  if (any(nd$type=="obs"))
    p <- p + geom_label(data=nd[nd$type=="obs",], aes(x=x,y=y,label=id),
                        size=4, fontface="bold", fill="white", colour="black",
                        label.size=.9, label.padding=unit(.4,"lines"))
  if (any(nd$type=="unobs"))
    p <- p + geom_label(data=nd[nd$type=="unobs",], aes(x=x,y=y,label=id),
                        size=4, fontface="bold.italic", fill="#fff3e0", colour="#b45309",
                        label.size=1.1, label.padding=unit(.4,"lines"),
                        label.r=unit(.5,"lines"))
  p
}

# ── Mediation extraction helper (handles robmed version differences) ───────────
rob_extract <- function(fit) {
  # point estimate: try $ab, then $indirect, then coef()["ab"]
  ab <- tryCatch(fit$ab,
    error = function(e) tryCatch(fit$indirect,
      error = function(e2) tryCatch(coef(fit)["ab"],
        error = function(e3) NA_real_)))
  # CI: try $ci (matrix or vector), then confint()
  ci <- tryCatch({
    r <- fit$ci
    if (is.matrix(r)) r[1,] else r
  }, error = function(e) tryCatch(confint(fit)["ab",],
    error = function(e2) c(NA_real_, NA_real_)))
  list(est=unname(ab), lo=unname(ci[1]), hi=unname(ci[2]))
}
```

# Part 1: What Do We Mean by "Causal Effect"? {#sec-m3-part1}

<iframe src="https://liveeur-my.sharepoint.com/personal/58542dsc_eur_nl/_layouts/15/Doc.aspx?sourcedoc={ac5b8424-7776-44d2-9bde-03770c0128d9}&action=embedview&wdAr=1.7777777777777777" width="100%" height="500px" frameborder="0" allowfullscreen title="Module 3 Part 1: Causal Effects"></iframe>

## The Fundamental Problem of Causal Inference {#sec-fundamental}

Every participant in the AlterEco eco-label study had **two potential outcomes**:

- $Y_i(0)$: their WTP if they did *not* see the eco-label
- $Y_i(1)$: their WTP if they *did* see the eco-label

Their **individual treatment effect** is $\tau_i = Y_i(1) - Y_i(0)$.

The problem? **We can only observe one of these for any individual.** This is called the **Fundamental Problem of Causal Inference** (Holland, 1986). Everything that follows is a strategy for getting around it.

```{r dgs-potential-outcomes}
#| code-summary: "▶ Build the data-generating system"
set.seed(2024)
N <- 2000

env        <- rnorm(N, 0, 1)
price_sens <- rnorm(N, 0, 1)
income     <- rnorm(N, 0, 1)

Y0 <- 5.50 + 0.80*env - 0.50*price_sens + 0.30*income + rnorm(N, 0, 1.10)
Y0 <- pmax(1, pmin(10, Y0))

ITE <- 0.60 + 0.80*env - 0.55*price_sens + rnorm(N, 0, 0.40)
Y1  <- pmax(1, pmin(10, Y0 + ITE))

ATE_true <- mean(Y1 - Y0)
cat(sprintf("True ATE  = %.3f\nTrue ITE SD = %.3f\n", ATE_true, sd(Y1 - Y0)))
```

```{r fundamental-problem-table, echo=FALSE}
#| code-summary: "▶ Show the missing-data structure"
set.seed(42)
demo_idx <- c(which(env > 1.2)[1:2], which(abs(env) < 0.15)[1], which(env < -1.2)[1:2])
assigned <- c("Control","Treatment","Control","Treatment","Control")
tibble(
  Person             = c("Sophie (high concern)","Rens (moderate+)","Lotte (neutral)","Daan (low concern)","Kim (price-sensitive)"),
  `Env concern`      = round(env[demo_idx], 1),
  `Y(0) No label`   = round(Y0[demo_idx], 2),
  `Y(1) With label` = round(Y1[demo_idx], 2),
  `True ITE`        = round(Y1[demo_idx]-Y0[demo_idx], 2),
  `Assigned to`     = assigned,
  `What we observe` = c(
    sprintf("Y(0)=$%.2f  Y(1)=???", Y0[demo_idx[1]]),
    sprintf("Y(0)=???   Y(1)=$%.2f", Y1[demo_idx[2]]),
    sprintf("Y(0)=$%.2f  Y(1)=???", Y0[demo_idx[3]]),
    sprintf("Y(0)=???   Y(1)=$%.2f", Y1[demo_idx[4]]),
    sprintf("Y(0)=$%.2f  Y(1)=???", Y0[demo_idx[5]])
  )
) |> knitr::kable(caption="The fundamental problem: exactly ONE potential outcome per person is observed.")
```

## Individual Treatment Effects (ITE) {#sec-ite}

```{r ite-distribution, fig.height=4.5, fig.width=9}
#| code-summary: "▶ Plot the distribution of individual treatment effects"
tibble(ITE=ITE,
       env_grp=cut(env, c(-Inf,-0.5,0.5,Inf), labels=c("Low","Moderate","High"))) |>
  ggplot(aes(x=ITE, fill=env_grp)) +
  geom_histogram(bins=50, alpha=0.75, position="identity") +
  geom_vline(xintercept=ATE_true, linetype="dashed", linewidth=1.1, colour="grey20") +
  annotate("text", x=ATE_true+0.06, y=82,
           label=sprintf("ATE = $%.2f", ATE_true), hjust=0, size=3.5, fontface="bold") +
  scale_fill_manual(values=c("High"=clr_eco,"Moderate"="#74b9a0","Low"=clr_ctrl)) +
  labs(x="Individual Treatment Effect  Y(1)−Y(0)  ($)", y="Count", fill=NULL,
       title="The eco-label effect is wildly different across individuals",
       subtitle="In real research you never see this distribution — only the ATE is estimable") +
  theme_mod3()
```

## Average Treatment Effect (ATE) {#sec-ate}

$$\text{ATE} = E[\tau_i] = E[Y_i(1) - Y_i(0)]$$

In a **randomised experiment**, random assignment makes the treated and control groups exchangeable, so a simple difference in group means is unbiased for the ATE.

```{r ate-rct}
treat_rct <- rbinom(N, 1, 0.5)
Y_obs_rct <- ifelse(treat_rct==1, Y1, Y0)
df_rct    <- tibble(id=1:N, treat=treat_rct, Y_obs=Y_obs_rct, Y0, Y1, ITE, env, price_sens, income)

ATE_est <- mean(Y_obs_rct[treat_rct==1]) - mean(Y_obs_rct[treat_rct==0])
lm(Y_obs ~ treat, data=df_rct) |> tidy() |>
  filter(term=="treat") |>
  transmute(Estimand="ATE", `True value`=round(ATE_true,3), Estimate=round(estimate,3),
            SE=round(std.error,3),
            `95% CI`=sprintf("[%.3f, %.3f]", estimate-1.96*std.error, estimate+1.96*std.error)) |>
  knitr::kable(caption="ATE estimate from randomised eco-label experiment")
```

## Conditional Average Treatment Effect (CATE) {#sec-cate}

```{r cate-by-subgroup, fig.height=4.5, fig.width=9}
df_rct |>
  mutate(env_grp=cut(env, c(-Inf,-1,0,1,Inf), labels=c("Very Low","Low","High","Very High"))) |>
  group_by(env_grp) |>
  summarise(CATE_true=mean(Y1-Y0), CATE_est=mean(Y_obs[treat==1])-mean(Y_obs[treat==0]),
            n=n(), .groups="drop") |>
  pivot_longer(starts_with("CATE"), names_to="type", values_to="val") |>
  ggplot(aes(x=env_grp, y=val, fill=type)) +
  geom_col(position="dodge", alpha=0.85, width=0.65) +
  geom_hline(yintercept=ATE_true, linetype="dashed", colour="grey30") +
  annotate("text", x=0.65, y=ATE_true+0.05, label="Overall ATE", size=3.5) +
  scale_fill_manual(values=c(CATE_true=clr_eco, CATE_est=clr_ctrl),
                    labels=c("True CATE","Estimated CATE")) +
  labs(x="Environmental Concern", y="Eco-label effect ($)", fill=NULL,
       title="CATE: the label works far better for environmentally engaged consumers",
       subtitle="Reporting only the ATE hides this story completely") +
  theme_mod3()
```

## Compliance, ITT, LATE, ATT, and ATC {#sec-late}

```{r compliance-dgs}
set.seed(2024)
Z <- treat_rct
p_comply_if_Z1 <- plogis(0.8 + 0.6*env)
p_always_taker <- plogis(-2.5 + 0.5*env)
D_if_Z1 <- rbinom(N, 1, p_comply_if_Z1)
D_if_Z0 <- rbinom(N, 1, p_always_taker)
D       <- ifelse(Z==1, D_if_Z1, D_if_Z0)
compliance_type <- case_when(
  D_if_Z1==1 & D_if_Z0==0 ~ "Complier",
  D_if_Z1==1 & D_if_Z0==1 ~ "Always-taker",
  D_if_Z1==0 & D_if_Z0==0 ~ "Never-taker",
  D_if_Z1==0 & D_if_Z0==1 ~ "Defier"
)
knitr::kable(table(compliance_type), col.names=c("Compliance type","Count"),
             caption="True compliance breakdown (observable only in simulation)")
```

```{r itt-late-att-atc}
Y_obs_iv <- ifelse(D==1, Y1, Y0)
df_iv    <- tibble(Z, D, Y_obs_iv, Y0, Y1, ITE, compliance_type, env, price_sens)

ITT       <- mean(Y_obs_iv[Z==1]) - mean(Y_obs_iv[Z==0])
FS        <- mean(D[Z==1]) - mean(D[Z==0])
LATE_est  <- ITT / FS
LATE_true <- mean(ITE[compliance_type=="Complier"])
ATT_est   <- mean(ITE[D==1])
ATC_est   <- mean(ITE[D==0])

tibble(
  Estimand    = c("ATE","ITT","LATE (Wald)","ATT","ATC"),
  Description = c(
    "Effect averaged over all 2,000 consumers",
    "Effect of *assigning* the label regardless of compliance",
    "Effect for consumers who complied *because* they were assigned",
    "Effect for consumers who *actually saw* the label",
    "Effect for those who *did not see* it"
  ),
  `True value` = round(c(ATE_true, ITT, LATE_true, ATT_est, ATC_est), 3),
  Estimated    = round(c(ATE_true, ITT, LATE_est,  ATT_est, ATC_est), 3)
) |> knitr::kable(caption="Five ways to summarise 'the effect' of the eco-label")
```

```{r estimands-plot, fig.height=4, fig.width=8, echo=FALSE}
tibble(
  Estimand=c("ATE\n(all consumers)","ITT\n(assigned, any)","LATE\n(compliers only)",
             "ATT\n(label seen)","ATC\n(label not seen)"),
  Value=c(ATE_true, ITT, LATE_est, ATT_est, ATC_est),
  Type=c("Population","Assignment","Compliers","Treated","Control")
) |>
  mutate(Estimand=fct_reorder(Estimand, Value)) |>
  ggplot(aes(y=Estimand, x=Value, fill=Type)) +
  geom_col(width=0.6, alpha=0.85) +
  geom_vline(xintercept=ATE_true, linetype="dashed", colour="grey30") +
  scale_fill_brewer(palette="Set2") +
  labs(x="Eco-label effect on WTP ($)", y=NULL, fill=NULL,
       title="Five estimands — same experiment, different questions",
       subtitle="ATT > ATE > ATC: consumers who seek eco-labels respond more strongly to them") +
  theme_mod3()
```

::: {.callout-warning}
## Connection to Module 2
Module 2's permutation test estimated the **ITT** — comparing all participants assigned to the eco-label condition against all assigned to control, regardless of whether they actually processed the label. In a well-controlled lab study, ITT ≈ ATE. In field settings — where consumers scroll past labels or ignore shelf information — the distinction matters greatly.
:::

---
