---
title: "Part 2: Selection on Observables"
---

```{r setup, include=FALSE}
options(repos = c(CRAN = "https://cran.rstudio.com/"))

if (!requireNamespace("remotes",  quietly = TRUE)) install.packages("remotes")
if (!requireNamespace("synthdid", quietly = TRUE))
  remotes::install_github("synth-did/synthdid")

pkgs <- c(
  "tidyverse", "broom", "knitr", "patchwork",
  "conflicted",
  "MatchIt", "WeightIt", "cobalt",
  "rdrobust", "rddensity",
  "did",
  "mediation", "robmed",
  "estimatr", "sandwich", "lmtest"
)
for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p, dependencies = TRUE)
  library(p, character.only = TRUE)
}
library(synthdid)

library(conflicted)
conflicts_prefer(dplyr::select, dplyr::filter, dplyr::rename, dplyr::mutate,
                 dplyr::summarise, dplyr::arrange, dplyr::lag)

dir.create("figs", recursive = TRUE, showWarnings = FALSE)
knitr::opts_chunk$set(
  echo = TRUE, fig.width = 9, fig.height = 5,
  fig.path = "figs/", dev = "svg"
)

clr_eco     <- "#2d6a4f"
clr_ctrl    <- "#4a90d9"
clr_treated <- clr_eco

theme_mod3 <- function() {
  theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          strip.background = element_rect(fill = "#f0f4f8", colour = NA),
          legend.position  = "bottom")
}

# ── Lightweight DAG helper ─────────────────────────────────────────────────────
.dag <- function(nodes, solid = NULL, dashed = NULL,
                 unobserved = character(0), title = "") {
  nd <- do.call(rbind, lapply(names(nodes), function(nm)
    data.frame(id=nm, x=nodes[[nm]][1], y=nodes[[nm]][2], stringsAsFactors=FALSE)))
  nd$type <- ifelse(nd$id %in% unobserved, "unobs", "obs")
  segs <- function(pairs, lt) {
    if (is.null(pairs) || length(pairs)==0) return(NULL)
    do.call(rbind, lapply(pairs, function(ab) {
      fr <- nd[nd$id==ab[1],]; to <- nd[nd$id==ab[2],]
      dx <- to$x-fr$x; dy <- to$y-fr$y
      len <- sqrt(dx^2+dy^2)+1e-9; pad <- 0.12
      data.frame(x=fr$x+pad*dx/len, y=fr$y+pad*dy/len,
                 xend=to$x-pad*dx/len, yend=to$y-pad*dy/len,
                 lt=lt, stringsAsFactors=FALSE)
    }))
  }
  edf <- rbind(segs(solid,"solid"), segs(dashed,"dashed"))
  xs <- range(nd$x); ys <- range(nd$y)
  p <- ggplot() + theme_void() + ggtitle(title) +
    coord_cartesian(xlim=c(xs[1]-.3,xs[2]+.3), ylim=c(ys[1]-.3,ys[2]+.3), expand=FALSE) +
    theme(plot.title=element_text(size=10,face="bold",hjust=.5,margin=margin(b=6)),
          plot.background=element_rect(fill="grey98",colour="grey80",linewidth=.4),
          plot.margin=margin(10,16,10,16), legend.position="none")
  if (!is.null(edf) && nrow(edf)>0)
    p <- p +
      geom_segment(data=edf, aes(x=x,y=y,xend=xend,yend=yend,colour=lt,linetype=lt),
                   arrow=arrow(length=unit(.20,"cm"),type="closed"), linewidth=.85) +
      scale_colour_manual(values=c(solid="grey20",dashed="firebrick")) +
      scale_linetype_manual(values=c(solid="solid",dashed="dashed"))
  if (any(nd$type=="obs"))
    p <- p + geom_label(data=nd[nd$type=="obs",], aes(x=x,y=y,label=id),
                        size=4, fontface="bold", fill="white", colour="black",
                        label.size=.9, label.padding=unit(.4,"lines"))
  if (any(nd$type=="unobs"))
    p <- p + geom_label(data=nd[nd$type=="unobs",], aes(x=x,y=y,label=id),
                        size=4, fontface="bold.italic", fill="#fff3e0", colour="#b45309",
                        label.size=1.1, label.padding=unit(.4,"lines"),
                        label.r=unit(.5,"lines"))
  p
}

# ── Mediation extraction helper (handles robmed version differences) ───────────
rob_extract <- function(fit) {
  # point estimate: try $ab, then $indirect, then coef()["ab"]
  ab <- tryCatch(fit$ab,
    error = function(e) tryCatch(fit$indirect,
      error = function(e2) tryCatch(coef(fit)["ab"],
        error = function(e3) NA_real_)))
  # CI: try $ci (matrix or vector), then confint()
  ci <- tryCatch({
    r <- fit$ci
    if (is.matrix(r)) r[1,] else r
  }, error = function(e) tryCatch(confint(fit)["ab",],
    error = function(e2) c(NA_real_, NA_real_)))
  list(est=unname(ab), lo=unname(ci[1]), hi=unname(ci[2]))
}
```

::: {.callout-tip}
## How to use the code in this section

The code blocks throughout this section contain inline annotations explaining **what data structures go in**, **what the key function arguments control**, and **what to look for in the output**. Click any "▶ Code" triangle to expand a block and read the annotations alongside the code. Treat them as a guided checklist: wherever you see a comment starting with `# ──`, it marks a decision point you would need to revisit with your own data.
:::

<iframe src="https://docs.google.com/presentation/d/e/2PACX-1vQDZkkwB4SGjnilPnrjkLIiEc48HRNlQ26hvFYLUFfMUftwN-xCXndPGlFDdG_DIw/pubembed?start=false&loop=false&delayms=3000" width="100%" height="500px" frameborder="0" allowfullscreen="true" mozallowfullscreen="true" webkitallowfullscreen="true" title="Module 3 Part 2: Selection on Observables"></iframe>

## Heckman Selection Models {#sec-heckman}

::: {.callout-warning}
## The critical caveat — read this before the example

The Heckman model works **only** if you have a valid **instrument for selection** — a variable that (1) predicts whether someone participates, but (2) has **no direct effect** on the outcome Y. Finding such an instrument is the hardest part of applying a Heckman correction. When the instrument is weak or its exclusion restriction is implausible, the Heckman correction can produce estimates that are *more* biased than naive OLS.

**What makes a good vs. bad instrument?**

| | Good instrument | Bad instrument |
|---|---|---|
| **Example** | Random financial incentive to complete the survey ($0–$4 bonus) | Whether the participant's friend also signed up |
| **Predicts participation?** | Yes — higher bonus → more likely to complete | Yes — social influence → more likely to participate |
| **Affects WTP directly?** | No — the cash bonus is for *completing* the survey, not for the product | Yes — friends share consumption preferences, so friend's sign-up correlates with eco-mindedness, which affects WTP |
| **Why it works / fails** | The incentive shifts *who bothers to participate* without changing how they evaluate the eco-label once in the study | The instrument correlates with the unobservable (eco-mindedness) you are trying to control for — it violates the exclusion restriction |

**A further technical requirement:** the instrument should be **continuous** (or at least multi-valued). A binary instrument (e.g., reminder email yes/no) creates only a handful of unique covariate combinations in Stage 1, making the inverse Mills ratio (IMR) nearly a step-function of treatment — which causes severe collinearity in Stage 2 and destabilizes the coefficient estimates.

Always report the first-stage F-statistic (or probit z-score) for the instrument as evidence of its relevance. **Notice the irony:** the statistical fix for a broken exclusion restriction (Heckman) itself requires an exclusion restriction. The problems multiply as Y becomes more complex.
:::

**The problem:** suppose eco-label study participants were recruited via an environmental blog (self-selection). People who click through are systematically more environmentally conscious. The treatment effect you estimate confounds "eco-label response" with "environmentally conscious person response."

**The Heckman correction** (Heckman, 1979) models the selection process explicitly, using a variable called the **inverse Mills ratio (IMR)** to account for the selection into the study:

1. **Stage 1 (Selection equation):** Regress participation (1 = participated, 0 = declined) on all variables that predict selection but do *not* directly affect Y. This requires an **exclusion restriction in the selection model** — a variable that predicts participation but not WTP. In the simulation below, we use a continuous financial incentive ($0–$4 bonus) randomly assigned to potential participants.

2. **Stage 2 (Outcome equation):** Regress Y on X and the IMR from Stage 1. The IMR absorbs the selection-induced correlation between treatment assignment and unobserved factors.

```{r heckman-demo}
#| code-summary: "▶ Simulate Heckman correction for self-selection"

set.seed(42)
N_hk  <- 1000   # pool of potential participants (larger N → more stable Heckman)

# ── Unobserved confounder ─────────────────────────────────────────────────────
# theta = latent eco-mindedness: affects BOTH survey completion AND WTP.
# It is NOT in the data (hence "unobserved"), so naive OLS cannot control for it.
theta <- rnorm(N_hk)

# ── Treatment assignment (random) ─────────────────────────────────────────────
# Eco vs. control randomly assigned to all N_hk potential participants.
eco   <- sample(rep(0:1, N_hk / 2))

# ── Instrument ────────────────────────────────────────────────────────────────
# incentive: a continuous financial incentive ($0–$4 bonus) randomly assigned
# to potential participants.  Higher incentive → more likely to complete the
# survey, but the incentive itself has NO direct effect on WTP for the product.
# This is the EXCLUSION RESTRICTION — a continuous instrument is essential here
# because a binary instrument (e.g., reminder email yes/no) would create only
# 4 unique (eco × reminder) combinations in Stage 1, making the IMR nearly a
# step-function of eco in Stage 2 and causing collinearity that destabilises
# the coefficient estimates.
incentive <- runif(N_hk, 0, 4)   # continuous $0–$4 bonus

# ── Selection into completion ──────────────────────────────────────────────────
# Key mechanism: the eco condition is intrinsically interesting to eco-minded
# (high-theta) people BUT ALSO to some low-theta "curious" people — so the eco
# arm selects in more low-theta completers than the control arm (where only
# truly motivated people bother).  This creates DOWNWARD bias in naive OLS.
# Keeping theta and eco effects moderate ensures the Heckman correction is
# tractable and does not overshoot due to finite-sample IMR estimation error.
sel_latent <- -0.30 +
  0.55 * theta     +    # eco-minded people more likely to complete
  0.65 * eco       +    # eco topic attracts notably more lower-theta completers → bigger OVB
  0.55 * incentive +    # strong continuous instrument for identification
  rnorm(N_hk, 0, 1)
selected <- as.integer(sel_latent > 0)
N_sel    <- sum(selected)

# ── WTP (observed only for completers; theta is the unobserved driver) ─────────
# theta effect on WTP = 0.85: large enough to create detectable downward bias
# in naive OLS, small enough that Heckman correction is stable (doesn't overshoot).
wtp_true <- 5.0 +
  0.50 * eco[selected == 1] +      # TRUE eco-label effect = $0.50
  0.85 * theta[selected == 1] +    # theta drives WTP but is UNOBSERVED in df_hk
  rnorm(N_sel, 0, 1.0)
wtp_true <- pmax(wtp_true, 0)      # floor at 0 (no negative WTP)

# ── Analysis datasets ──────────────────────────────────────────────────────────
# df_hk:   observed data for completers (theta NOT included — it's unobserved)
# df_full: full pool used for Stage 1 probit (includes instrument)
df_hk   <- data.frame(wtp       = wtp_true,
                       eco       = eco[selected == 1],
                       incentive = incentive[selected == 1])
df_full <- data.frame(selected, eco, incentive)   # theta excluded — unobserved

# Naive OLS: ignores selection → eco coefficient is DOWNWARD BIASED because
# eco arm selects in more low-theta participants (lower average WTP)
naive <- lm(wtp ~ eco, data = df_hk)

# Heckman Stage 1: probit of completion on eco + incentive (instrument)
# Note: theta is excluded (unobserved); incentive is the identifying instrument.
# The continuous incentive creates a smooth IMR function → no collinearity.
stage1 <- glm(selected ~ eco + incentive, data = df_full,
              family = binomial(link = "probit"))

# Inverse Mills ratio (IMR) for completers — absorbs the selection-on-theta
xb  <- predict(stage1)[selected == 1]
imr <- dnorm(xb) / pnorm(xb)

# Heckman Stage 2: includes IMR to correct for selective attrition
heckman <- lm(wtp ~ eco + imr, data = cbind(df_hk, imr = imr))

cat(sprintf("Completion rate: %.1f%% overall  |  Eco: %.1f%%  |  Control: %.1f%%\n",
            100 * mean(selected),
            100 * mean(selected[eco == 1]),
            100 * mean(selected[eco == 0])))
cat(sprintf("Mean theta among completers — Eco: %.2f  |  Control: %.2f  (gap creates bias)\n",
            mean(theta[selected == 1 & eco == 1]),
            mean(theta[selected == 1 & eco == 0])))

heck_tab <- data.frame(
  Model = c("Naive OLS (ignores selection)", "Heckman 2-step (corrects attrition)"),
  `Eco-label coefficient` = c(round(coef(naive)["eco"], 3),
                               round(coef(heckman)["eco"], 3)),
  `95% CI (low)` = c(round(confint(naive)["eco", 1], 3),
                      round(confint(heckman)["eco", 1], 3)),
  `95% CI (high)` = c(round(confint(naive)["eco", 2], 3),
                       round(confint(heckman)["eco", 2], 3)),
  `True value` = 0.50,
  check.names = FALSE
)

kable(heck_tab,
      caption = paste0("Heckman correction: differential attrition creates downward bias in naive OLS. ",
                       "Eco arm attracts low-theta completers; control arm retains only high-theta ones. ",
                       "True eco-label effect = $0.50."))
```

Notice how the naive OLS estimate is **downward biased**: the eco condition attracted some low-theta ("curious but not eco-minded") completers who would not have bothered with the control condition. These participants bring down the mean WTP in the eco arm, understating the true eco-label effect. The control arm retained only high-theta (genuinely motivated) completers with high baseline WTP regardless. The Heckman correction uses the IMR to account for this differential attrition and recovers an estimate closer to the true $0.50 value.

## When Instruments Fail: The Bad Instrument Case {#sec-bad-instrument}

The exclusion restriction — that the instrument affects the outcome only through selection — is the most critical and least verifiable assumption in a Heckman model. It cannot be tested from the data alone. The example below shows what happens when this assumption is violated.

```{r bad-instrument-demo}
#| code-summary: "▶ Simulate: Heckman with a bad (invalid) instrument"

# All dataset objects from the heckman-demo chunk above are already defined:
# N_hk, theta, eco, selected, df_hk, df_full, naive, heckman

# ── Bad instrument: whether the participant's friend also signed up ─────────────
# friend_signed_up is correlated with theta (eco-mindedness) because
# environmentally conscious people tend to have eco-conscious friends.
# This violates the exclusion restriction: friend_signed_up predicts selection
# (social influence), but it also correlates with theta, which directly affects
# WTP. The instrument is therefore NOT excludable from the outcome equation.
friend_signed_up <- rbinom(N_hk, 1, plogis(0.8 * theta))

# ── Stage 1 probit with the bad instrument ────────────────────────────────────
stage1_bad <- glm(selected ~ eco + friend_signed_up,
                  data = data.frame(selected, eco, friend_signed_up),
                  family = binomial(link = "probit"))

# ── IMR from the bad Stage 1 ──────────────────────────────────────────────────
# Because friend_signed_up correlates with theta, the IMR it generates is itself
# confounded: it absorbs some of theta's variance but also carries its own
# theta-contaminated signal into Stage 2, biasing the eco coefficient.
xb_bad  <- predict(stage1_bad)[selected == 1]
imr_bad <- dnorm(xb_bad) / pnorm(xb_bad)

# ── Stage 2 with bad IMR ──────────────────────────────────────────────────────
heckman_bad <- lm(wtp ~ eco + imr_bad, data = cbind(df_hk, imr_bad = imr_bad))

# ── Comparison table ──────────────────────────────────────────────────────────
bad_tab <- data.frame(
  Model = c(
    "Naive OLS (ignores selection)",
    "Heckman — good instrument (incentive)",
    "Heckman — bad instrument (friend signed up)"
  ),
  `Eco-label coefficient` = c(
    round(coef(naive)["eco"],        3),
    round(coef(heckman)["eco"],      3),
    round(coef(heckman_bad)["eco"],  3)
  ),
  `True value` = 0.50,
  check.names = FALSE
)

kable(bad_tab,
      caption = paste0("Bad instrument demo: friend_signed_up correlates with theta (eco-mindedness), ",
                       "violating the exclusion restriction. The IMR it generates is confounded, ",
                       "potentially producing an estimate worse than naive OLS. True eco effect = $0.50."))
```

A bad instrument can produce estimates that are *worse* than naive OLS — the Heckman correction amplifies bias when the exclusion restriction fails rather than correcting it. The instrument must genuinely satisfy two conditions: it must predict selection (relevance), and it must have no path to the outcome except through selection (exclusion). The second condition is untestable, which is why instrument choice demands substantive, not statistical, justification.

## Censored Data Models (Tobit) {#sec-tobit}

In the eco-coffee WTP study, participants could not state WTP below $1 (floor) or above $10 (ceiling). In many studies, the outcome is **censored** — truncated at a boundary because the measurement instrument or the context prevents observing values beyond a threshold.

**Consequence of ignoring censoring:** OLS on a censored outcome is biased and inconsistent. Observations piled at the ceiling (e.g., everyone stating exactly $10.00 WTP) look identical to OLS but represent unobserved heterogeneity (some would have stated $12 or $15 if allowed).

The **Tobit model** explicitly models the censoring threshold and recovers the latent distribution:

$$Y^* = X\beta + \varepsilon, \quad Y = \max(Y_{\text{lower}}, \min(Y^*, Y_{\text{upper}}))$$

```{r tobit-demo}
#| code-summary: "▶ Simulate: OLS vs. Tobit on censored WTP data"

if (!requireNamespace("AER", quietly = TRUE)) install.packages("AER")
library(AER)

set.seed(2026)
N_tb   <- 1200   # large N for stable Tobit estimates (Tobit MLE is less efficient than OLS)
eco_tb <- rep(0:1, each = N_tb / 2)

# ── Latent (true) WTP ─────────────────────────────────────────────────────────
# True eco-label effect = $0.90 on the latent scale.
# Mean = 1.2 (control arm) with sd = 1.8 places ~25% of latent WTPs below $0.
# High censoring creates clear OLS attenuation; N = 1200 makes Tobit converge
# reliably to 0.90 without overshoot.
wtp_lat <- 1.2 + 0.90 * eco_tb + rnorm(N_tb, 0, 1.8)

# Observed WTP: censored at $0 (participants state "I would not pay anything")
wtp_obs      <- pmax(wtp_lat, 0)
pct_censored <- mean(wtp_obs == 0)

cat(sprintf("Floor-censored observations: %d / %d  (%.1f%%)\n",
            sum(wtp_obs == 0), N_tb, 100 * pct_censored))

# OLS on the censored observed values: attenuated toward zero
ols_tb <- lm(wtp_obs ~ eco_tb)

# Tobit: models the latent WTP as normal, recovering the true coefficient
tobit_tb <- tobit(wtp_obs ~ eco_tb, left = 0)

cens_tab <- data.frame(
  Model    = c("OLS (ignores censoring)", "Tobit (corrects for censoring)"),
  `Eco coefficient` = c(round(coef(ols_tb)["eco_tb"], 3),
                         round(coef(tobit_tb)["eco_tb"], 3)),
  `True value` = 0.90,
  `% censored` = paste0(round(pct_censored * 100, 1), "%"),
  check.names = FALSE
)

kable(cens_tab,
      caption = paste0("OLS vs. Tobit on floor-censored WTP (true eco effect = $0.90). ",
                       "~20% floor censoring attenuates OLS; Tobit recovers the latent coefficient."))
```

## The Problem: Consumers Choose to Buy Eco-Labelled Products {#sec-obs-problem}

Outside the lab, consumers freely *choose* whether to pick up the eco-labelled bag of coffee. This creates **selection bias**. The DGP below adds two realistic complications beyond Part 1:

1. **Non-linear confounding**: environmental concern has a *quadratic* effect on baseline WTP and *interacts* with income. A consumer with extremely high env-concern sees a doubly large WTP boost. Simple linear regression will miss this.
2. **Interaction in the propensity score**: the probability of buying eco-labelled coffee depends on the *joint* level of env-concern and income, not just their individual levels.

These features mean that naive and simple-linear corrections will have remaining bias, while flexible and doubly-robust methods will do better.

```{r obs-dgs}
#| code-summary: "▶ Simulate observational data with non-linear confounding"
set.seed(2024)
N <- 2000
# ── YOUR DATA: replace N with nrow(your_df); the columns env, price_sens,
#    income, and age map to your pre-treatment covariates (standardise continuous
#    ones with scale() so distance-based methods work correctly)

env        <- rnorm(N, 0, 1)
price_sens <- rnorm(N, 0, 1)
income     <- rnorm(N, 0, 1)
age        <- round(rnorm(N, 38, 12))

# Non-linear outcome model: env has quadratic effect + env:income interaction
# Linear regression misses these → remaining bias after covariate adjustment
Y0 <- pmax(1, pmin(10,
  5.20 + 0.80*env - 0.50*price_sens + 0.30*income +
  0.25*env^2 + 0.18*env*income +
  rnorm(N, 0, 0.85)
))

ITE <- 0.55 + 0.65*env - 0.38*price_sens + rnorm(N, 0, 0.35)
Y1  <- pmax(1, pmin(10, Y0 + ITE))

# True PS includes env:income interaction
# Standard logit (no interaction) will slightly misspecify the PS
ps_true   <- plogis(-0.40 + 0.95*env + 0.42*income - 0.32*price_sens + 0.14*env*income)
treat_obs <- rbinom(N, 1, ps_true)
Y_obs     <- ifelse(treat_obs==1, Y1, Y0)

# Pre-compute nonlinear features for use in flexible models
env_sq      <- env^2
env_income  <- env * income

df_obs <- tibble(id=1:N, treat=treat_obs, Y_obs, Y0, Y1, ITE,
                 env, price_sens, income, age, env_sq, env_income, ps_true)
# ── YOUR DATA: df_obs should be your real data frame with columns:
#    treat (0/1 treatment indicator), Y_obs (observed outcome),
#    and all pre-treatment covariates. Drop Y0, Y1, ITE, ps_true —
#    those are unobservable in real data; they exist here only for
#    ground-truth evaluation.

ATE_true_obs <- mean(Y1 - Y0)
ATT_true_obs <- mean(ITE[treat_obs==1])
cat(sprintf("True ATE = %.3f  |  True ATT = %.3f  |  Treatment rate = %.1f%%\n",
            ATE_true_obs, ATT_true_obs, 100*mean(treat_obs)))
# ── CHECK: treatment rate far from 50% (< 10% or > 90%) signals severe overlap
#    problems; near-zero or near-1 PS values will produce extreme IPW weights.
```

```{r naive-bias, fig.height=4.5, fig.width=9}
#| code-summary: "▶ Show why naive comparison overestimates the eco-label effect"
naive_diff <- mean(Y_obs[treat_obs==1]) - mean(Y_obs[treat_obs==0])
cat(sprintf("True ATE             = %.3f\nNaive diff-in-means  = %.3f\nBias                 = +%.3f\n",
            ATE_true_obs, naive_diff, naive_diff - ATE_true_obs))

bind_rows(
  tibble(group="Eco-label buyers",     env=env[treat_obs==1]),
  tibble(group="Non-eco-label buyers", env=env[treat_obs==0])
) |>
  ggplot(aes(x=env, fill=group)) +
  geom_density(alpha=0.55) +
  scale_fill_manual(values=c("Eco-label buyers"=clr_eco,"Non-eco-label buyers"=clr_ctrl)) +
  labs(x="Environmental Concern (standardised)", y="Density",
       title="Eco-label buyers have higher environmental concern AND higher baseline WTP",
       subtitle="Positive selection on a cause of the outcome inflates the naive comparison",
       fill=NULL) +
  theme_mod3()
```

## Regression Adjustment {#sec-reg-adj}

**What it does:** Controls for confounders in the outcome regression, comparing eco-label buyers and non-buyers who are *similar on measured covariates*.

**Intuition:** Among consumers with the same env-concern, income, and price-sensitivity, any remaining WTP difference is more plausibly causal. But if the true relationship between covariates and WTP is *non-linear*, a linear regression will not fully remove the confounding.

```{r regression-adjustment}
#| code-summary: "▶ Linear vs. flexible regression adjustment"
# ── YOUR DATA: replace Y_obs with your outcome variable and treat with your
#    treatment indicator; add your own covariates in place of env, price_sens,
#    income. Flexible adjustment (m_flex) adds squared and interaction terms —
#    include these whenever you suspect non-linear covariate effects.
m_naive   <- lm(Y_obs ~ treat, data=df_obs)
m_linear  <- lm(Y_obs ~ treat + env + price_sens + income, data=df_obs)
m_flex    <- lm(Y_obs ~ treat + env + price_sens + income + env_sq + env_income, data=df_obs)
# ── KEY ARGS: the formula after ~ lists covariates to adjust for; env_sq and
#    env_income are pre-computed non-linear terms — create analogous columns in
#    your data frame if you suspect curvilinear or interactive confounding.

bind_rows(
  tidy(m_naive)  |> mutate(model="1. Unadjusted"),
  tidy(m_linear) |> mutate(model="2. Linear adjustment"),
  tidy(m_flex)   |> mutate(model="3. Flexible adjustment\n(+ env² + env×income)")
) |>
  filter(term=="treat") |>
  transmute(Model=model, Estimate=round(estimate,3), SE=round(std.error,3),
            `95% CI`=sprintf("[%.3f, %.3f]",estimate-1.96*std.error,estimate+1.96*std.error)) |>
  add_row(Model="True ATE", Estimate=round(ATE_true_obs,3), SE=NA, `95% CI`="—") |>
  knitr::kable(caption="Flexible regression (with non-linear terms) gets closer to the true ATE than linear")
```

::: {.callout-tip}
The flexible model including `env²` and `env × income` tracks the true ATE noticeably better than the linear adjustment. The lesson: **the form of your covariate adjustment matters**. If you don't know the true functional form (you never do in practice), doubly robust estimation provides insurance.
:::

## Inverse Probability Weighting (IPW) {#sec-ipw}

**What it does:** Reweights each observation by the inverse of its probability of being in its actual group, creating a pseudo-population where eco-label buyers and non-buyers look like random draws from the same population.

**Concrete weights:**
- An eco-conscious, high-income consumer (PS = 0.85) who buys eco-labelled coffee: weight = $1/0.85 = 1.18$ — not surprising, downweighted.
- A price-sensitive, low-income consumer (PS = 0.08) who buys eco-labelled coffee: weight = $1/0.08 = 12.5$ — very surprising, heavily upweighted.

**Entropy balancing** (via `WeightIt`) takes this further: instead of estimating a parametric PS model, it directly finds weights that exactly balance the *moments* (means, variances, and cross-products) of the covariate distribution between groups. No model to misspecify.

```{r ipw}
#| code-summary: "▶ Standard IPW vs. entropy-balanced IPW"
# Standard IPW: logistic regression on linear terms (misses interaction)
# ── YOUR DATA: replace treat ~ env + price_sens + income with your treatment
#    variable and pre-treatment covariates. Use the same covariate set throughout
#    all methods so comparisons are meaningful.
ps_logit <- glm(treat ~ env + price_sens + income, data=df_obs, family=binomial)
ps_hat   <- fitted(ps_logit)

df_obs <- df_obs |>
  mutate(ps_hat  = ps_hat,
         ipw_wt  = ifelse(treat==1, 1/ps_hat, 1/(1-ps_hat)),
         # Stabilised weights (reduces variance from extreme PS values)
         ipw_stab= ifelse(treat==1, mean(treat)/ps_hat, mean(1-treat)/(1-ps_hat)))

ATE_ipw_raw   <- with(df_obs,
  weighted.mean(Y_obs[treat==1], ipw_wt[treat==1]) -
  weighted.mean(Y_obs[treat==0], ipw_wt[treat==0]))
ATE_ipw_stab  <- lm(Y_obs ~ treat, data=df_obs, weights=ipw_stab) |>
  tidy() |> filter(term=="treat") |> pull(estimate)
# ── CHECK: inspect summary(df_obs$ipw_wt) — any weights > 20–30 signal near-
#    zero or near-one PS values (positivity violation); use stabilised weights
#    or entropy balancing instead to reduce variance inflation.

# Entropy balancing: directly balance means + variances + covariances
# Includes quadratic and interaction terms for thorough moment matching
# ── KEY ARGS: method="ebal" finds minimum-variance weights that exactly match
#    covariate moments; estimand="ATE" targets the population ATE (use "ATT" if
#    you only care about the effect among treated units).
wb_ebal <- weightit(treat ~ env + price_sens + income + env_sq + env_income,
                    data=df_obs, method="ebal", estimand="ATE")
ATE_ebal <- lm(Y_obs ~ treat, data=df_obs, weights=wb_ebal$weights) |>
  tidy() |> filter(term=="treat") |> pull(estimate)
# ── CHECK: run summary(wb_ebal) to see effective sample size (ESS) after
#    weighting — large weight loss (ESS much smaller than N) means poor overlap.

cat(sprintf(
  "True ATE              = %.3f\nIPW (standard logit)  = %.3f\nIPW (stabilised)      = %.3f\nEntropy balancing     = %.3f\n",
  ATE_true_obs, ATE_ipw_raw, ATE_ipw_stab, ATE_ebal
))
```

```{r ps-overlap, fig.height=4.5, fig.width=9}
#| code-summary: "▶ Propensity score overlap and entropy-balanced weight distribution"
p_overlap <- df_obs |>
  mutate(group=ifelse(treat==1,"Eco-label buyers","Non-buyers")) |>
  ggplot(aes(x=ps_hat, fill=group)) +
  geom_histogram(bins=40, alpha=0.65, position="identity") +
  geom_vline(xintercept=c(0.05,0.95), linetype="dashed", colour="firebrick") +
  scale_fill_manual(values=c("Eco-label buyers"=clr_eco,"Non-buyers"=clr_ctrl)) +
  labs(x="Estimated PS", y="Count", title="PS overlap", fill=NULL,
       subtitle="Red lines = extreme weights zone") +
  theme_mod3()

p_wts <- tibble(wt=wb_ebal$weights, group=ifelse(treat_obs==1,"Treated","Control")) |>
  ggplot(aes(x=wt, fill=group)) +
  geom_histogram(bins=40, alpha=0.7, position="identity") +
  scale_fill_manual(values=c("Treated"=clr_eco,"Control"=clr_ctrl)) +
  labs(x="Entropy-balanced weight", y="Count",
       title="Entropy-balanced weights",
       subtitle="Well-behaved: no extreme spikes", fill=NULL) +
  theme_mod3()

p_overlap + p_wts
```

::: {.callout-warning}
## IPW can be unstable with extreme propensity scores

Unstabilised IPW weights become enormous near PS = 0 or 1. Entropy balancing avoids this by finding the *minimum-variance* weights that achieve balance — it never requires extreme weights to correct for outlier PS values.
:::

## Covariate Matching (Mahalanobis) {#sec-matching}

**What it does:** For each eco-label buyer, finds the most similar non-buyer based on raw covariate values (Mahalanobis distance accounts for covariate correlations). The matched sample mimics a randomised experiment — but only for the *treated* group (estimates ATT, not ATE).

```{r covariate-matching}
#| code-summary: "▶ 1:1 Mahalanobis nearest-neighbour matching"
# ── YOUR DATA: replace df_obs with your data frame; treat ~ env + price_sens +
#    income should list your pre-treatment covariates (continuous or binary).
#    Mahalanobis distance accounts for correlations between covariates —
#    useful when covariates are on very different scales.
# ── KEY ARGS: method="nearest" does greedy nearest-neighbour matching;
#    distance="mahalanobis" can be changed to "gower" (handles mixed types) or
#    replaced by a propensity-score distance (see psm-comparison chunk);
#    ratio=1 means 1 control matched per treated unit — increase (e.g., ratio=2)
#    to gain precision at the cost of match quality.
m_match_maha <- matchit(treat ~ env + price_sens + income, data=df_obs,
                        method="nearest", distance="mahalanobis", ratio=1)
df_matched_maha <- match.data(m_match_maha)
ATT_match_maha  <- lm(Y_obs ~ treat, data=df_matched_maha, weights=weights) |>
  tidy() |> filter(term=="treat") |> pull(estimate)
# ── CHECK: run summary(m_match_maha) and inspect SMD (Std. Mean Difference)
#    for each covariate — SMD < 0.1 means good balance; large SMD means the
#    matched sample still has systematic differences on that covariate.

cat(sprintf("True ATT (treated group)              = %.3f\nMatching (Mahalanobis, 1:1 NN) est.   = %.3f\n",
            ATT_true_obs, ATT_match_maha))
```

## Propensity Score Matching (PSM) {#sec-psm}

**What it does:** Condenses all covariates into a single scalar — the propensity score — and matches on that. The catch: **the PS model must be specified correctly**. Different models give different matched samples and often quite different estimates.

```{r psm-comparison, fig.height=5.5, fig.width=10}
#| code-summary: "▶ Compare five PS model specifications"
# ── YOUR DATA: replace treat ~ env + price_sens + income with your treatment
#    column and pre-treatment covariates. Run multiple matchit() calls varying
#    the PS formula (try linear, polynomial, interaction terms) to test how
#    sensitive your ATT estimate is to PS model choice — if estimates shift a
#    lot, the result is fragile.
# ── KEY ARGS: distance="logit" or "probit" estimates PS via logistic/probit
#    regression; caliper= trims matches where PS difference > threshold SD
#    (caliper=0.1 is a common rule-of-thumb); std.caliper=TRUE means the
#    caliper is in units of the PS's standard deviation.
# Misspecified models (common practice — miss the env:income interaction)
m_ps1 <- matchit(treat ~ env + price_sens + income,
                 data=df_obs, method="nearest", distance="logit")
m_ps2 <- matchit(treat ~ env * price_sens + income,
                 data=df_obs, method="nearest", distance="probit")
m_ps3 <- matchit(treat ~ poly(env,2) + poly(price_sens,2) + income,
                 data=df_obs, method="nearest", distance="logit")
m_ps4 <- matchit(treat ~ env + price_sens + income,
                 data=df_obs, method="nearest", distance="logit",
                 caliper=0.1, std.caliper=TRUE)
# Correctly specified: matches the true DGP (includes env^2 and env:income)
m_ps5 <- matchit(treat ~ env + price_sens + income + env_sq + env_income,
                 data=df_obs, method="nearest", distance="logit",
                 caliper=0.15, std.caliper=TRUE)

results_psm <- map_dfr(
  list("1. Logit (linear)"              = m_ps1,
       "2. Probit + env×price"          = m_ps2,
       "3. Logit + poly(env,2)"         = m_ps3,
       "4. Logit + caliper (0.1 SD)"    = m_ps4,
       "5. Logit + env²+env×income\n+ caliper (correct PS)" = m_ps5),
  function(m) {
    d   <- match.data(m)
    est <- lm(Y_obs ~ treat, data=d, weights=weights) |>
      tidy() |> filter(term=="treat")
    tibble(estimate=est$estimate, se=est$std.error,
           n_matched=nrow(d[d$treat==1,]))
  }, .id="PS model"
)

results_psm |>
  mutate(lo=estimate-1.96*se, hi=estimate+1.96*se,
         `N retained`=n_matched,
         correct = str_detect(`PS model`, "correct")) |>
  ggplot(aes(y=`PS model`, x=estimate, xmin=lo, xmax=hi,
             colour=correct, shape=correct)) +
  geom_vline(xintercept=ATT_true_obs, linetype="dashed", colour=clr_eco, linewidth=1) +
  geom_vline(xintercept=ATE_true_obs, linetype="dotted", colour=clr_ctrl, linewidth=1) +
  geom_pointrange(size=0.8) +
  geom_text(aes(label=sprintf("N=%d", `N retained`), x=hi+0.03), hjust=0, size=3) +
  annotate("text", x=ATT_true_obs+0.01, y=5.45,
           label="True ATT", hjust=0, size=3.2, colour=clr_eco) +
  annotate("text", x=ATE_true_obs+0.01, y=5.1,
           label="True ATE", hjust=0, size=3.2, colour=clr_ctrl) +
  scale_colour_manual(values=c("FALSE"="grey50","TRUE"="#b45309"),
                      labels=c("Misspecified PS","Correctly specified PS")) +
  scale_shape_manual(values=c("FALSE"=16,"TRUE"=18),
                     labels=c("Misspecified PS","Correctly specified PS")) +
  labs(x="ATT estimate — eco-label effect ($)", y=NULL, colour=NULL, shape=NULL,
       title="PSM estimates converge on the true ATT only when the PS model is correctly specified",
       subtitle="The correctly specified model (orange) recovers the true ATT most accurately") +
  theme_mod3()
```

```{r covariate-balance, fig.height=5, fig.width=8}
#| code-summary: "▶ Love plot: balance before and after correctly-specified PSM"
love.plot(m_ps5, thresholds=c(m=0.1), stars="std",
          title="Covariate balance: correctly specified PSM (env² + env×income + caliper)",
          colors=c(clr_ctrl, clr_eco))
```

## Doubly Robust Estimation {#sec-doubly-robust}

**What it does:** Combines an outcome model *and* a propensity score model. The estimate is consistent if **either** model is correctly specified — it only needs one to be right.

Here both the outcome model and PS model include the non-linear terms (`env²`, `env × income`), so both are correctly specified. The DR estimate should perform best.

```{r doubly-robust}
#| code-summary: "▶ AIPW with correctly-specified outcome and PS models"
# ── YOUR DATA: replace Y_obs, treat, and the covariate list with your own
#    variable names. Fit mu1_model on treated rows only, mu0_model on control
#    rows only — these are separate outcome regressions for each arm.
# ── KEY ARGS: include all covariates you believe affect the outcome in the
#    outcome models (mu1/mu0) AND all covariates that predict treatment in the
#    PS model. Adding non-linear terms (env_sq, env_income) reduces bias when
#    the true relationship is non-linear. The AIPW formula is consistent if
#    EITHER the outcome model OR the PS model is correctly specified.
# Outcome models (correctly specified: include env^2 and env:income)
mu1_model <- lm(Y_obs ~ env + price_sens + income + env_sq + env_income,
                data=df_obs[df_obs$treat==1,])
mu0_model <- lm(Y_obs ~ env + price_sens + income + env_sq + env_income,
                data=df_obs[df_obs$treat==0,])
mu1_hat   <- predict(mu1_model, newdata=df_obs)
mu0_hat   <- predict(mu0_model, newdata=df_obs)

# PS model (correctly specified)
ps_hat_dr <- fitted(glm(treat ~ env + price_sens + income + env_sq + env_income,
                        data=df_obs, family=binomial))

# AIPW estimator
Y_v <- df_obs$Y_obs; D_v <- df_obs$treat
tau_dr <- (mu1_hat - mu0_hat) +
          D_v*(Y_v - mu1_hat)/ps_hat_dr -
          (1-D_v)*(Y_v - mu0_hat)/(1-ps_hat_dr)

ATE_dr <- mean(tau_dr)
SE_dr  <- sd(tau_dr)/sqrt(N)
# ── CHECK: compare ATE_dr to IPW and regression adjustment estimates —
#    if DR diverges sharply from both, a model may be badly misspecified.
#    For inference with real data use a sandwich SE or bootstrap rather than
#    the naive sd(tau_dr)/sqrt(N) used here for illustration.

cat(sprintf(
  "True ATE              = %.3f\nNaive estimate        = %.3f\nLinear reg. adj.      = %.3f\nFlexible reg. adj.    = %.3f\nIPW (stabilised)      = %.3f\nEntropy balancing     = %.3f\nDR/AIPW (correct)     = %.3f  (SE = %.3f)\n",
  ATE_true_obs, naive_diff,
  coef(m_linear)["treat"], coef(m_flex)["treat"],
  ATE_ipw_stab, ATE_ebal,
  ATE_dr, SE_dr
))
```

## How Well Do All Section 2 Methods Recover the True ATE? {#sec-sec2-comparison}

```{r section2-summary-plot, fig.height=5.5, fig.width=10, echo=FALSE}
#| code-summary: "▶ Comprehensive comparison of all observational estimators"
tibble(
  Method   = c(
    "Naive (no adjustment)",
    "Linear regression\nadjustment",
    "Flexible regression\n(env² + env×income)",
    "IPW — stabilised\n(logit, linear PS)",
    "Entropy balancing\n(moment-matching IPW)",
    "Matching — Mahalanobis\n1:1 NN",
    "PSM — misspecified\n(logit linear)",
    "PSM — correctly\nspecified + caliper",
    "Doubly Robust\nAIPW (correct models)"
  ),
  Estimate = c(
    naive_diff,
    coef(m_linear)["treat"],
    coef(m_flex)["treat"],
    ATE_ipw_stab,
    ATE_ebal,
    ATT_match_maha,
    results_psm$estimate[1],
    results_psm$estimate[5],
    ATE_dr
  ),
  Target  = c("ATE","ATE","ATE","ATE","ATE","ATT","ATT","ATT","ATE"),
  Quality = c("Naive","Basic","Flexible","IPW","Advanced","Matching","PSM","PSM","Advanced")
) |>
  mutate(Method  = fct_inorder(Method),
         Bias    = Estimate - ATE_true_obs) |>
  ggplot(aes(y=Method, x=Estimate, colour=Quality, shape=Target)) +
  geom_vline(xintercept=ATE_true_obs, linetype="dashed", colour=clr_eco,  linewidth=1.1) +
  geom_vline(xintercept=ATT_true_obs, linetype="dotted", colour=clr_ctrl, linewidth=1.1) +
  geom_point(size=4) +
  annotate("text", x=ATE_true_obs+0.01, y=9.45,
           label="True ATE", hjust=0, size=3.3, colour=clr_eco, fontface="bold") +
  annotate("text", x=ATT_true_obs+0.01, y=9.0,
           label="True ATT", hjust=0, size=3.3, colour=clr_ctrl) +
  scale_colour_manual(
    values=c(Naive="#c0392b", Basic="#e67e22", Flexible="#f1c40f",
             IPW="#3498db", Matching="#9b59b6", PSM="#16a085", Advanced="#2d6a4f")) +
  scale_shape_manual(values=c(ATE=16, ATT=18)) +
  labs(x="Estimate ($)", y=NULL, colour="Method class", shape="Target estimand",
       title="Nine estimators compared against the known true ATE and ATT",
       subtitle="Non-linear DGP: methods that model non-linearity or balance moments perform best") +
  theme_mod3()
```

::: {.callout-tip}
## What the gradient shows

| Method class | Why it performs that way |
|---|---|
| **Naive** | No correction — pure selection bias |
| **Linear regression** | Removes linear confounding; missing `env²` and `env×income` leaves residual bias |
| **Flexible regression** | Includes non-linear terms → less bias |
| **IPW (stabilised)** | PS model also misspecified (linear logit) → some remaining bias |
| **Entropy balancing** | Directly balances covariate *moments* — robust to PS misspecification |
| **Matching/PSM** | Targets ATT (higher than ATE because treated have stronger responses) |
| **Doubly Robust** | Both outcome model AND PS model correctly specified → lowest bias |

The **doubly robust** and **entropy balancing** approaches consistently outperform simpler methods when the true confounding is non-linear. The correctly specified PSM recovers the ATT accurately; the misspecified PSM does less well.
:::

## Synthetic Controls {#sec-synth}

::: {.callout-warning}
## Synthetic controls: validity depends on an unverifiable assumption

A synthetic control constructs a weighted combination of untreated units to mimic the counterfactual for the treated unit. The approach is transparent and often produces compelling-looking pre-period fits — but there is no way to verify that the synthetic world it generates is a sufficiently representative stand-in for the real counterfactual post-treatment. Synthetic controls assume that the same weighting that matched pre-treatment trends would have continued to match post-treatment trends in the absence of treatment. This is an untestable assumption, and deviations — due to unobserved shocks, compositional changes, or structural breaks — will bias the estimated treatment effect without any diagnostic flag.

Use synthetic controls when you have a single treated unit, a long pre-treatment window, and a credible pool of potential donor units. But always report sensitivity analyses varying the donor pool and be explicit about what would have to be true for the synthetic counterfactual to be valid.
:::

What if you only have **one treated unit** — say, AlterEco's flagship Rotterdam store introduced eco-labels?

```{r synthetic-control, fig.height=5, fig.width=9}
#| code-summary: "▶ Synthetic control for Rotterdam flagship store"
set.seed(2024)
n_stores <- 21; n_pre <- 20; n_post <- 10; treat_store <- 1
store_fe <- rnorm(n_stores, 0, 0.8)

panel_df <- expand.grid(store=1:n_stores, period=1:(n_pre+n_post)) |>
  as_tibble() |>
  mutate(treated=(store==treat_store & period>n_pre),
         WTP=5.5 + store_fe[store] + 0.02*(period-n_pre) +
             treated*0.80 + rnorm(n(), 0, 0.40))

pre_means <- panel_df |> filter(period<=n_pre) |>
  group_by(store) |> summarise(pre_mean=mean(WTP), .groups="drop")
treated_pre_mean <- pre_means$pre_mean[pre_means$store==treat_store]

control_stores <- pre_means |> filter(store!=treat_store) |>
  mutate(dist=abs(pre_mean-treated_pre_mean), w_raw=1/(dist+0.01), w=w_raw/sum(w_raw))

synth_ts <- panel_df |> filter(store!=treat_store) |>
  left_join(control_stores[,c("store","w")], by="store") |>
  group_by(period) |> summarise(WTP_synth=weighted.mean(WTP,w), .groups="drop") |>
  mutate(series="Synthetic Rotterdam")

treated_ts <- panel_df |> filter(store==treat_store) |>
  dplyr::select(period, WTP) |> rename(WTP_synth=WTP) |>
  mutate(series="Rotterdam (actual)")

bind_rows(synth_ts, treated_ts) |>
  ggplot(aes(x=period, y=WTP_synth, colour=series, linetype=series)) +
  geom_line(linewidth=1) +
  geom_vline(xintercept=n_pre+0.5, linetype="dashed", colour="grey50") +
  annotate("text", x=n_pre+1, y=5.0, label="Eco-label\nintroduced", hjust=0, size=3.5, colour="grey40") +
  scale_colour_manual(values=c("Rotterdam (actual)"=clr_eco,"Synthetic Rotterdam"=clr_ctrl)) +
  scale_linetype_manual(values=c("Rotterdam (actual)"="solid","Synthetic Rotterdam"="dashed")) +
  labs(x="Period", y="Mean WTP ($)", colour=NULL, linetype=NULL,
       title="Synthetic Control: Rotterdam vs. its synthetic twin built from control stores",
       subtitle="Good pre-period fit validates the synthetic match; gap after dashed line = causal effect") +
  theme_mod3()
```

---
