---
title: "Part 5: Mediation — Does the Eco-Label Work Through Perceived Sustainability?"
---

```{r setup, include=FALSE}
options(repos = c(CRAN = "https://cran.rstudio.com/"))

if (!requireNamespace("remotes",  quietly = TRUE)) install.packages("remotes")
if (!requireNamespace("synthdid", quietly = TRUE))
  remotes::install_github("synth-did/synthdid")

pkgs <- c(
  "tidyverse", "broom", "knitr", "patchwork",
  "conflicted",
  "MatchIt", "WeightIt", "cobalt",
  "rdrobust", "rddensity",
  "did",
  "mediation", "robmed",
  "estimatr", "sandwich", "lmtest"
)
for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p, dependencies = TRUE)
  library(p, character.only = TRUE)
}
library(synthdid)

library(conflicted)
conflicts_prefer(dplyr::select, dplyr::filter, dplyr::rename, dplyr::mutate,
                 dplyr::summarise, dplyr::arrange, dplyr::lag)

dir.create("figs", recursive = TRUE, showWarnings = FALSE)
knitr::opts_chunk$set(
  echo = TRUE, fig.width = 9, fig.height = 5,
  fig.path = "figs/", dev = "svg"
)

clr_eco     <- "#2d6a4f"
clr_ctrl    <- "#4a90d9"
clr_treated <- clr_eco

theme_mod3 <- function() {
  theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          strip.background = element_rect(fill = "#f0f4f8", colour = NA),
          legend.position  = "bottom")
}

# ── Lightweight DAG helper ─────────────────────────────────────────────────────
.dag <- function(nodes, solid = NULL, dashed = NULL,
                 unobserved = character(0), title = "") {
  nd <- do.call(rbind, lapply(names(nodes), function(nm)
    data.frame(id=nm, x=nodes[[nm]][1], y=nodes[[nm]][2], stringsAsFactors=FALSE)))
  nd$type <- ifelse(nd$id %in% unobserved, "unobs", "obs")
  segs <- function(pairs, lt) {
    if (is.null(pairs) || length(pairs)==0) return(NULL)
    do.call(rbind, lapply(pairs, function(ab) {
      fr <- nd[nd$id==ab[1],]; to <- nd[nd$id==ab[2],]
      dx <- to$x-fr$x; dy <- to$y-fr$y
      len <- sqrt(dx^2+dy^2)+1e-9; pad <- 0.12
      data.frame(x=fr$x+pad*dx/len, y=fr$y+pad*dy/len,
                 xend=to$x-pad*dx/len, yend=to$y-pad*dy/len,
                 lt=lt, stringsAsFactors=FALSE)
    }))
  }
  edf <- rbind(segs(solid,"solid"), segs(dashed,"dashed"))
  xs <- range(nd$x); ys <- range(nd$y)
  p <- ggplot() + theme_void() + ggtitle(title) +
    coord_cartesian(xlim=c(xs[1]-.3,xs[2]+.3), ylim=c(ys[1]-.3,ys[2]+.3), expand=FALSE) +
    theme(plot.title=element_text(size=10,face="bold",hjust=.5,margin=margin(b=6)),
          plot.background=element_rect(fill="grey98",colour="grey80",linewidth=.4),
          plot.margin=margin(10,16,10,16), legend.position="none")
  if (!is.null(edf) && nrow(edf)>0)
    p <- p +
      geom_segment(data=edf, aes(x=x,y=y,xend=xend,yend=yend,colour=lt,linetype=lt),
                   arrow=arrow(length=unit(.20,"cm"),type="closed"), linewidth=.85) +
      scale_colour_manual(values=c(solid="grey20",dashed="firebrick")) +
      scale_linetype_manual(values=c(solid="solid",dashed="dashed"))
  if (any(nd$type=="obs"))
    p <- p + geom_label(data=nd[nd$type=="obs",], aes(x=x,y=y,label=id),
                        size=4, fontface="bold", fill="white", colour="black",
                        label.size=.9, label.padding=unit(.4,"lines"))
  if (any(nd$type=="unobs"))
    p <- p + geom_label(data=nd[nd$type=="unobs",], aes(x=x,y=y,label=id),
                        size=4, fontface="bold.italic", fill="#fff3e0", colour="#b45309",
                        label.size=1.1, label.padding=unit(.4,"lines"),
                        label.r=unit(.5,"lines"))
  p
}

# ── Mediation extraction helper (handles robmed version differences) ───────────
rob_extract <- function(fit) {
  # point estimate: try $ab, then $indirect, then coef()["ab"]
  ab <- tryCatch(fit$ab,
    error = function(e) tryCatch(fit$indirect,
      error = function(e2) tryCatch(coef(fit)["ab"],
        error = function(e3) NA_real_)))
  # CI: try $ci (matrix or vector), then confint()
  ci <- tryCatch({
    r <- fit$ci
    if (is.matrix(r)) r[1,] else r
  }, error = function(e) tryCatch(confint(fit)["ab",],
    error = function(e2) c(NA_real_, NA_real_)))
  list(est=unname(ab), lo=unname(ci[1]), hi=unname(ci[2]))
}
```

::: {.callout-tip}
## How to use the code in this section

The code blocks throughout this section contain inline annotations explaining **what data structures go in**, **what the key function arguments control**, and **what to look for in the output**. Click any "▶ Code" triangle to expand a block and read the annotations alongside the code. Treat them as a guided checklist: wherever you see a comment starting with `# ──`, it marks a decision point you would need to revisit with your own data.
:::

Mediation analysis asks whether the eco-label affects WTP **directly** or **through a mechanism** — here, *perceived sustainability* (M). Three estimators are compared across three scenarios that each challenge a different method.

| Scenario | Key feature | Who wins |
|---|---|---|
| **1: Clean normal data** | No violations — a baseline | All methods perform well |
| **2: Large outliers** | Correlated high-leverage points inflate OLS b-path | **ROBMED** wins |
| **3: M–Y confounded** | Unobserved U causes both M and Y; true ACME = 0 | **Imai sensitivity** catches it |

::: {.callout-note}
## The mediation arithmetic — and where it can break

Baron–Kenny mediation decomposes X's total effect on Y into two paths:

$$\underbrace{X \xrightarrow{\;a\;} M \xrightarrow{\;b\;} Y}_{\text{indirect (ACME) = } a \times b} \quad + \quad \underbrace{X \xrightarrow{\;c'\;} Y}_{\text{direct (ADE)}}$$

The **ACME** (Average Causal Mediation Effect) is the product $a \times b$. It has two failure modes that the three scenarios below are specifically designed to reveal:

1. **The *b*-path can be estimated on the wrong data** — if outliers inflate the apparent slope between M and Y, $\hat{b}$ is too large, and so is $\hat{a} \times \hat{b}$. ROBMED addresses this.
2. **The *b*-path can be estimated on the wrong model** — if an unobserved variable $U^*$ drives both M and Y, the OLS $\hat{b}$ picks up the $U^*$ signal as if it were causal M→Y. No regression method alone can fix this; Imai's sensitivity analysis quantifies how severe the problem would need to be to explain away the result.

**A Module 1 failure mode:** There is a third, often overlooked failure mode — **measurement error in M**. If *perceived sustainability* is measured as a single noisy item rather than a well-validated multi-item scale (Module 1 — reliability, CFA, convergent validity), the $\hat{a}$ path is attenuated (like any regression coefficient on a mismeasured predictor) and the $\hat{b}$ path is biased as well, because M carries both signal and noise into the Y regression. The practical implication: investing in high-quality mediator measurement at the design stage (multi-item constructs, established scales, pilot reliability checks) is not just good practice — it is a precondition for trustworthy mediation estimates. A large ACME from a single-item mediator may be entirely an artifact of differential measurement error.

**A Module 2 connection:** Mediation analysis is typically conducted on experimental data where X is randomised (as in Scenarios 1–3 here). This randomisation satisfies the *a*-path assumption (no confounding of X→M). But randomisation does *not* satisfy the *b*-path assumption — M is not randomised, so the M→Y step faces all the observational identification challenges of Part 2 of this module. The sensitivity parameter $\rho^*$ below is a direct quantification of how much M–Y confounding would be needed to nullify the ACME, just as power analysis in Module 2 quantified how much sampling error could explain an apparent effect.
:::

## Causal Structures {#sec-med-dags}

```{r med-dag-scenarios, fig.height=3.2, fig.width=10, echo=FALSE}
#| code-summary: "▶ DAGs: true mediation (Scenarios 1 & 2) vs. confounded (Scenario 3)"
dA <- .dag(nodes = list(X=c(0,.5), M=c(.5,.9), Y=c(1,.5)),
           solid  = list(c("X","M"), c("M","Y"), c("X","Y")),
           title  = "Scenarios 1 & 2: True mediation\nX → M → Y  (+ direct path X → Y)")
dB <- .dag(nodes     = list(X=c(0,.5), M=c(.5,.9), Y=c(1,.5), U=c(.5,.1)),
           solid     = list(c("X","M"), c("X","Y"), c("U","M"), c("U","Y")),
           dashed    = list(c("M","Y")),
           unobserved= "U",
           title     = "Scenario 3: No causal M→Y\nU* confounds M and Y  (M→Y is spurious)")
dA | dB
```

## Setup: Three DGPs {#sec-med-dgp}

```{r med-dgp-setup}
#| code-summary: "▶ Three mediation DGPs: clean normal data, large outliers, M–Y confounded"
set.seed(2025)
N_med <- 350
X     <- rbinom(N_med, 1, 0.5)   # eco-label (1 = exposed to label)
# ── YOUR DATA: X is your binary treatment (0/1); replace rbinom() with your
#    actual treatment column. N_med should equal nrow(your_df).
#    Each df_sN below maps to your real data frame with columns:
#    eco_label (or your treatment name), perc_sust (your mediator), WTP (outcome).

# True parameters (Scenarios 1 & 2)
a_true    <- 0.80   # eco_label → perc_sust  (strong a-path)
b_true    <- 0.75   # perc_sust → WTP        (strong b-path)
c_prime   <- 0.30   # eco_label → WTP (direct path)
ACME_true <- a_true * b_true   # 0.600  (large → needs big ρ to nullify → ROBUST)
# ── YOUR DATA: in real data you do NOT know a_true or b_true; these are set
#    here only for ground-truth evaluation. With real data, run the
#    run_mediation_triple() function directly on your data frame.

# ── Scenario 1: Clean normal data ────────────────────────────────────────────
# No violations; both OLS and ROBMED should recover ACME ≈ 0.275
M_s1   <- a_true * X + rnorm(N_med, 0, 0.60)
Y_s1   <- c_prime * X + b_true * M_s1 + rnorm(N_med, 0, 0.60)
df_s1  <- data.frame(eco_label = X, perc_sust = M_s1, WTP = Y_s1)

# ── Scenario 2: Large correlated outliers (12% contamination) ───────────────
# Correlated shocks to M and Y residuals in the same direction: outlier obs
# suggest a much steeper M→Y slope → OLS b-path inflated → OLS ACME >> 0.275
# ROBMED MM-estimator downweights these high-leverage points → stays near truth
out_n   <- ceiling(N_med * 0.12)
out_idx <- sample(N_med, out_n)
out_sgn <- rep(1, out_n)
out_mag <- runif(out_n, 6, 10)   # 10–17× the baseline SD: strong leverage
eps_M2  <- rnorm(N_med, 0, 0.60)
eps_Y2  <- rnorm(N_med, 0, 0.60)
eps_M2[out_idx] <- eps_M2[out_idx] + out_sgn * out_mag          # M outlier
eps_Y2[out_idx] <- eps_Y2[out_idx] + out_sgn * out_mag * 0.90   # correlated Y outlier
M_s2   <- a_true * X + eps_M2
Y_s2   <- c_prime * X + b_true * M_s2 + eps_Y2
df_s2  <- data.frame(eco_label = X, perc_sust = M_s2, WTP = Y_s2)
# ── CHECK (Scenario 2): plot perc_sust vs WTP and colour outlier points (out_idx)
#    red — if a small cluster of points visually drives the OLS slope, ROBMED
#    is important for your data; inspect with plot(df_s2$perc_sust, df_s2$WTP).

# ── Scenario 3: M–Y confounded via unobserved U* (true ACME = 0) ─────────────
# U* drives both perc_sust and WTP; there is NO causal M → Y path
# OLS and ROBMED both find spurious ACME (Type I error) because they cannot
# separate causal M→Y from U-induced M–Y correlation.
# U* coefficients are deliberately modest (0.28, 0.32) so the induced residual
# correlation ρ ≈ 0.22 — Imai's medsens() correctly identifies this low ρ* as
# FRAGILE, in sharp contrast to Scenario 1's ρ* ≈ 0.80.
U      <- rnorm(N_med, 0, 1)
M_s3   <- a_true * X + 0.28 * U + rnorm(N_med, 0, 0.55)
Y_s3   <- c_prime * X + 0.32 * U + rnorm(N_med, 0, 0.55)   # no M_s3 term!
df_s3  <- data.frame(eco_label = X, perc_sust = M_s3, WTP = Y_s3)
ACME_true_s3 <- 0   # no causal M → Y
# ── YOUR DATA: Scenario 3 represents the most dangerous real-world failure —
#    an unmeasured variable (e.g., prior eco-engagement, health consciousness)
#    drives both your mediator and outcome. Always run medsens() on your real
#    data and report the critical ρ* alongside the ACME estimate.

cat(sprintf("N = %d  |  Treatment rate = %.0f%%\n", N_med, 100 * mean(X)))
cat(sprintf("Scenarios 1 & 2: true ACME = %.3f  (a=%.2f, b=%.2f, c'=%.2f)\n",
            ACME_true, a_true, b_true, c_prime))
cat(sprintf("Scenario 3:      true ACME = %.3f  (U* confounds both M and Y)\n",
            ACME_true_s3))
```

## Estimating All Three Methods {#sec-med-run}

```{r run-all-mediation, cache=FALSE}
#| code-summary: "▶ Run OLS delta method, ROBMED, and Imai on all three scenarios"
# ── YOUR DATA: to apply this to your own data, call run_mediation_triple()
#    with a data frame that has columns named eco_label (treatment), perc_sust
#    (mediator), and WTP (outcome). Rename your columns to match, or edit the
#    column references inside the function to match your variable names.
run_mediation_triple <- function(df, R = 500) {
  m_m <- lm(perc_sust ~ eco_label, data = df)
  m_y <- lm(WTP ~ eco_label + perc_sust, data = df)
  # ── KEY ARGS: m_m is the a-path model (treatment → mediator);
  #    m_y is the b-path / outcome model (treatment + mediator → outcome).
  #    Both must share the same data frame. Add covariates to both models
  #    if confounders of the M→Y relationship are available (e.g., age, income).

  # 1. OLS: a × b with delta-method CI
  a_hat   <- coef(m_m)["eco_label"]
  b_hat   <- coef(m_y)["perc_sust"]
  ols_est <- a_hat * b_hat
  se_ab   <- sqrt(b_hat^2 * vcov(m_m)["eco_label","eco_label"] +
                  a_hat^2 * vcov(m_y)["perc_sust","perc_sust"])
  ols_lo  <- ols_est - 1.96 * se_ab
  ols_hi  <- ols_est + 1.96 * se_ab

  # 2. ROBMED: MM-robust regression (downweights high-leverage outliers)
  # ── KEY ARGS: WTP ~ m(perc_sust) + eco_label specifies outcome ~ m(mediator) +
  #    treatment; the m() wrapper identifies the mediator for robmed.
  #    method="regression" uses regression-based mediation; robust=TRUE activates
  #    the MM-estimator. R= controls bootstrap replications — 500 for exploration,
  #    2000+ for final publication figures.
  set.seed(99)
  med_rob  <- test_mediation(WTP ~ m(perc_sust) + eco_label,
                             data = df, method = "regression", robust = TRUE, R = R)
  rob      <- rob_extract(med_rob)

  # 3. Imai et al.: quasi-Bayesian (used for medsens() sensitivity analysis)
  # ── KEY ARGS: treat= names the treatment column; mediator= names the mediator;
  #    boot=FALSE uses quasi-Bayesian simulation (faster); sims= controls the
  #    number of draws — increase to 1000 for publication.
  set.seed(99)
  med_imai <- mediate(m_m, m_y, treat = "eco_label", mediator = "perc_sust",
                      boot = FALSE, sims = R)

  list(a_hat    = a_hat,          b_hat    = b_hat,
       ols_est  = ols_est,        ols_lo   = ols_lo,       ols_hi  = ols_hi,
       rob_est  = rob$est,        rob_lo   = rob$lo,       rob_hi  = rob$hi,
       imai_est = med_imai$d0,    imai_lo  = med_imai$d0.ci[1],
       imai_hi  = med_imai$d0.ci[2],
       med_imai = med_imai,       med_rob  = med_rob)
}

cat("Running three mediation scenarios (OLS + ROBMED + Imai, 500 sims each)...\n")
res_s1 <- run_mediation_triple(df_s1)
res_s2 <- run_mediation_triple(df_s2)
res_s3 <- run_mediation_triple(df_s3)
# ── CHECK: compare ols_est vs. rob_est — if they differ substantially (> 0.05
#    in standardised units), outliers are influential; prefer ROBMED.
#    Always follow up with medsens() (see sensitivity chunk) to report ρ*.
cat(sprintf("Scenario 1 — OLS: %.3f  ROBMED: %.3f  (truth = %.3f)\n",
            res_s1$ols_est, res_s1$rob_est, ACME_true))
cat(sprintf("Scenario 2 — OLS: %.3f  ROBMED: %.3f  (truth = %.3f)\n",
            res_s2$ols_est, res_s2$rob_est, ACME_true))
cat(sprintf("Scenario 3 — OLS: %.3f  ROBMED: %.3f  (truth = %.3f [no M→Y path])\n",
            res_s3$ols_est, res_s3$rob_est, ACME_true_s3))
```

## Scenario 1: Clean Normal Data — All Methods Should Work {#sec-med-clean}

With textbook normal errors and no confounding, OLS, ROBMED, and Imai all estimate ACME accurately. This is the baseline case each method was designed for.

```{r med-scenario1, echo=FALSE}
#| code-summary: "▶ Scenario 1 results: clean normal data"
tibble(
  Method   = c("OLS delta method", "ROBMED (MM-estimator)", "Imai (quasi-Bayes)"),
  Paths    = c(sprintf("a=%.3f, b=%.3f", res_s1$a_hat, res_s1$b_hat),
               "-- (robust MM)",
               "-- (uses OLS)"),
  Estimate = round(c(res_s1$ols_est, res_s1$rob_est, res_s1$imai_est), 3),
  CI95     = sprintf("[%.3f, %.3f]",
                     c(res_s1$ols_lo, res_s1$rob_lo, res_s1$imai_lo),
                     c(res_s1$ols_hi, res_s1$rob_hi, res_s1$imai_hi)),
  Bias     = sprintf("%+.3f",
                     c(res_s1$ols_est, res_s1$rob_est, res_s1$imai_est) - ACME_true),
  Verdict  = c("On target", "On target", "On target")
) |>
  knitr::kable(
    col.names = c("Method", "a & b paths", "Estimate", "95% CI", "Bias", "Verdict"),
    caption   = paste0("Scenario 1 \u2014 Clean normal data | true ACME = ",
                       round(ACME_true, 3), " | N = ", N_med),
    align     = c("l","l","r","c","r","c")
  )
```

## Scenario 2: Large Outliers — ROBMED's Advantage {#sec-med-outliers}

Twelve percent of participants show large, **correlated** deviations in both perceived sustainability (M) and WTP (Y) in the same direction. These high-leverage observations suggest a much steeper M→Y slope than the true 0.75, inflating the OLS b-path and thus ACME. The MM-estimator in ROBMED downweights points that deviate strongly from the main data cloud.

::: {.callout-note}
## Why correlated outliers inflate the *b*-path specifically

The key is *direction*: when participants who score unusually high on M also score unusually high on Y (same-sign shocks), they create data points that pull the OLS regression line steeply upward. To OLS, these are just informative extreme observations — it can't distinguish "genuinely high M causes genuinely high Y" from "an external shock hit both M and Y simultaneously."

**ROBMED's MM-estimator** assigns lower weight to observations whose residuals are large relative to the bulk of the data. The outlier participants — who are far from the M–Y regression cloud — receive close to zero weight, leaving the slope estimate dominated by the non-contaminated 88%.

**Why Imai doesn't help here**: `mediate()` calls the same `lm()` models internally. When those models are biased, the quasi-Bayesian sampling averages over biased posteriors. ROBMED is the right tool; Imai is not.
:::


```{r med-scenario2, echo=FALSE}
#| code-summary: "▶ Scenario 2 results: large outliers inflate OLS but not ROBMED"
tibble(
  Method   = c("OLS delta method", "ROBMED (MM-estimator)", "Imai (quasi-Bayes)"),
  `b-path` = sprintf("%.3f (true = %.2f)",
                     c(res_s2$b_hat, res_s2$rob_est / res_s2$a_hat, res_s2$b_hat),
                     rep(b_true, 3)),
  Estimate = round(c(res_s2$ols_est, res_s2$rob_est, res_s2$imai_est), 3),
  `95% CI` = sprintf("[%.3f, %.3f]",
                     c(res_s2$ols_lo, res_s2$rob_lo, res_s2$imai_lo),
                     c(res_s2$ols_hi, res_s2$rob_hi, res_s2$imai_hi)),
  Bias     = sprintf("%+.3f",
                     c(res_s2$ols_est, res_s2$rob_est, res_s2$imai_est) - ACME_true),
  Verdict  = c("\u2716 Inflated \u2014 outliers pull b-path up",
               "\u2713 Robust \u2014 outliers down-weighted",
               "\u2716 Inflated \u2014 uses OLS internally")
) |>
  knitr::kable(
    caption = sprintf(
      "Scenario 2 \u2014 12%% correlated outlier contamination | true ACME = %.3f | N = %d",
      ACME_true, N_med),
    align = c("l","l","r","c","r","l")
  )
```

```{r med-outlier-scatter, fig.height=3.5, fig.width=10, echo=FALSE}
#| code-summary: "▶ Scatter: clean vs contaminated data showing outlier leverage on OLS slope"
p_s1_sc <- ggplot(df_s1, aes(x = perc_sust, y = WTP)) +
  geom_point(alpha = 0.35, size = 1.5, colour = clr_ctrl) +
  geom_smooth(method = "lm", se = TRUE, colour = "firebrick", linewidth = 1.2) +
  labs(x = "Perceived sustainability (M)", y = "WTP ($)",
       title = "Scenario 1: Clean data",
       subtitle = sprintf("OLS b = %.3f  (true = %.2f)", res_s1$b_hat, b_true)) +
  theme_mod3()

df_s2_vis <- df_s2 |> mutate(outlier = seq_len(n()) %in% out_idx)
p_s2_sc <- ggplot(df_s2_vis, aes(x = perc_sust, y = WTP, colour = outlier)) +
  geom_point(alpha = 0.45, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, colour = "firebrick", linewidth = 1.2) +
  scale_colour_manual(values = c("FALSE" = "grey55", "TRUE" = "#e74c3c"),
                      labels = c("FALSE" = "Clean obs", "TRUE" = "Outlier (12%)"),
                      name = NULL) +
  labs(x = "Perceived sustainability (M)", y = "WTP ($)",
       title = "Scenario 2: Outliers (red) inflate the OLS slope",
       subtitle = sprintf("OLS b = %.3f  ROBMED b ≈ %.3f  (true = %.2f)",
                          res_s2$b_hat, res_s2$rob_est / res_s2$a_hat, b_true)) +
  theme_mod3()

p_s1_sc | p_s2_sc
```

## Scenario 3: M–Y Confounded — Regression-Based Methods Fail, Imai Diagnoses {#sec-med-confounded}

U* drives both perceived sustainability and WTP. **There is no causal M→Y path.** The true ACME = 0. Yet both OLS and ROBMED find a large, significant "mediation effect" because they cannot distinguish causal M→Y from U-induced M–Y correlation. This is a structural identification failure, not an estimation problem — no regression method alone can fix it. But Imai's sensitivity analysis reveals that the result is fragile.

```{r med-scenario3, echo=FALSE}
#| code-summary: "▶ Scenario 3: both OLS and ROBMED produce a Type I error"
tibble(
  Method   = c("OLS delta method", "ROBMED (MM-estimator)", "Imai (quasi-Bayes)"),
  `b-path (spurious)` = sprintf("%.3f (true = 0.00)", c(res_s3$b_hat, NA_real_, res_s3$b_hat)) |>
                        (\(x) { x[2] <- sprintf("%.3f (true = 0.00)", res_s3$rob_est / res_s3$a_hat); x })(),
  Estimate = round(c(res_s3$ols_est, res_s3$rob_est, res_s3$imai_est), 3),
  `95% CI` = sprintf("[%.3f, %.3f]",
                     c(res_s3$ols_lo, res_s3$rob_lo, res_s3$imai_lo),
                     c(res_s3$ols_hi, res_s3$rob_hi, res_s3$imai_hi)),
  Bias     = sprintf("%+.3f",
                     c(res_s3$ols_est, res_s3$rob_est, res_s3$imai_est) - ACME_true_s3),
  Verdict  = c("\u2716 Type I error \u2014 spurious mediation",
               "\u2716 Type I error \u2014 U* not outlier-based",
               "\u2716 Type I error \u2014 see \u03c1* below")
) |>
  knitr::kable(
    caption = sprintf(
      "Scenario 3 \u2014 M\u2013Y confounded via U* | true ACME = %.3f (no M\u2192Y path) | N = %d",
      ACME_true_s3, N_med),
    align = c("l","l","r","c","r","l")
  )
```

## Where Imai's Framework Shines: Sensitivity Analysis {#sec-med-sensitivity}

Imai et al.'s `medsens()` asks: *how large would the correlation between M-equation and Y-equation residuals (ρ) need to be to drive ACME to zero?* This **critical ρ\*** measures fragility:

- **Large ρ***: only substantial unmeasured confounding could explain away the result → **robust**
- **Small ρ***: even modest unmeasured confounding suffices → **fragile**

```{r med-sensitivity-compare, fig.height=4.5, fig.width=10}
#| code-summary: "▶ Sensitivity analysis: clean DGP (robust) vs. confounded DGP (fragile)"
# ── YOUR DATA: replace res_s1$med_imai and res_s3$med_imai with the med_imai
#    object returned by run_mediation_triple() on your own data frame.
#    Run medsens() on every mediation model you report — it costs little time
#    and is now expected by reviewers in top journals.
# ── KEY ARGS: rho.by = 0.05 evaluates sensitivity at ρ steps of 0.05 over
#    [-1, 1]; effect.type = "indirect" targets the ACME specifically;
#    sims = 500 controls bootstrap draws — use 1000+ for final results.
set.seed(99)
sens_s1 <- medsens(res_s1$med_imai, rho.by = 0.05, effect.type = "indirect", sims = 500)
set.seed(99)
sens_s3 <- medsens(res_s3$med_imai, rho.by = 0.05, effect.type = "indirect", sims = 500)
# ── CHECK: look at crit_rho_s1 and crit_rho_s3 printed below —
#    ρ* > 0.4 is generally considered robust (substantial confounding needed);
#    ρ* < 0.2 is fragile (modest confounding would nullify the result).
#    Report ρ* in your manuscript alongside the ACME point estimate and CI.

safe_crit_rho <- function(s) {
  tryCatch({
    idx_neg <- which(s$d0 <= 0)
    if (length(idx_neg) == 0) NA_real_
    else round(min(abs(s$rho[idx_neg]), na.rm = TRUE), 2)
  }, error = function(e) NA_real_)
}
crit_rho_s1 <- safe_crit_rho(sens_s1)
crit_rho_s3 <- safe_crit_rho(sens_s3)

fmt_rho <- function(x) if (is.na(x)) "> 1.0" else sprintf("%.2f", x)
cat(sprintf("Scenario 1 (true ACME = %.3f): critical ρ* ≈ %s  → ROBUST finding\n",
            ACME_true, fmt_rho(crit_rho_s1)))
cat(sprintf("Scenario 3 (true ACME = 0):     critical ρ* ≈ %s  → FRAGILE finding\n",
            fmt_rho(crit_rho_s3)))
cat("\nSmaller ρ* = less confounding is needed to explain away the result.\n")

# Build tidy sensitivity curves using only the guaranteed $rho and $d0 fields
lbl_s1 <- sprintf("Scenario 1: Clean data (true ACME = %.3f)", ACME_true)
lbl_s3 <- "Scenario 3: M\u2013Y confounded (true ACME = 0)"

sens_df <- bind_rows(
  data.frame(rho      = sens_s1$rho,
             ACME     = sens_s1$d0,
             Scenario = lbl_s1,
             stringsAsFactors = FALSE),
  data.frame(rho      = sens_s3$rho,
             ACME     = sens_s3$d0,
             Scenario = lbl_s3,
             stringsAsFactors = FALSE)
)

# y-annotation positions scale with the ACME range
y_robust  <-  ACME_true * 0.30   # ~30% of way up from zero
y_fragile <- -ACME_true * 0.30   # ~30% of way down from zero

ggplot(sens_df, aes(x = rho, y = ACME, colour = Scenario)) +
  geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.9) +
  geom_line(linewidth = 1.4) +
  { if (!is.na(crit_rho_s1))
      list(
        geom_vline(xintercept =  crit_rho_s1, linetype = "dashed",
                   colour = clr_eco, linewidth = 0.9),
        geom_vline(xintercept = -crit_rho_s1, linetype = "dashed",
                   colour = clr_eco, linewidth = 0.9),
        annotate("text", x = crit_rho_s1 + 0.03, y = y_robust,
                 label = sprintf("\u03c1*\u2248%s\n(ROBUST)", fmt_rho(crit_rho_s1)),
                 hjust = 0, size = 3.0, colour = clr_eco, fontface = "bold")
      ) } +
  { if (!is.na(crit_rho_s3))
      list(
        geom_vline(xintercept =  crit_rho_s3, linetype = "dashed",
                   colour = clr_ctrl, linewidth = 0.9),
        geom_vline(xintercept = -crit_rho_s3, linetype = "dashed",
                   colour = clr_ctrl, linewidth = 0.9),
        annotate("text", x = crit_rho_s3 + 0.03, y = y_fragile,
                 label = sprintf("\u03c1*\u2248%s\n(FRAGILE)", fmt_rho(crit_rho_s3)),
                 hjust = 0, size = 3.0, colour = clr_ctrl, fontface = "bold")
      ) } +
  scale_colour_manual(
    values = setNames(c(clr_eco, clr_ctrl), c(lbl_s1, lbl_s3))) +
  labs(x       = expression(paste("Sensitivity parameter ", rho,
                                  "  (correlation between M- and Y-equation residuals)")),
       y       = "ACME (indirect effect)",
       colour  = NULL,
       title   = "Imai sensitivity: how much confounding would nullify the ACME?",
       subtitle = paste0("\u03c1* = critical value at which ACME crosses zero",
                         "  \u2014  smaller \u03c1* = more fragile result")) +
  theme_mod3() +
  theme(legend.position = "top")
```

::: {.callout-tip}
## Reading the sensitivity plot

Both Scenario 1 and Scenario 3 produce a *statistically significant* ACME from OLS — they look equally convincing at face value. The sensitivity analysis reveals very different levels of trust:

- **Scenario 1 (clean)**: ACME reaches zero only at $\rho^* \approx$ `r fmt_rho(crit_rho_s1)`. Even substantial unmeasured M–Y confounding cannot explain away the result. The finding is **robust**.
- **Scenario 3 (confounded)**: ACME crosses zero at $\rho^* \approx$ `r fmt_rho(crit_rho_s3)`. Only modest confounding is enough to nullify the result. The "significant" ACME is **fragile** — correctly so, since the true ACME is 0.

**The Imai advantage is not a better point estimate** (it uses the same linear models as OLS, so it inherits the same bias in Scenario 3). The advantage is the *sensitivity analysis itself* — an honest diagnostic that Baron–Kenny completely omits. A finding that reports both the ROBMED estimate and a large ρ* from `medsens()` is far more credible than one that does neither.
:::

## Comprehensive Bias Comparison Across All Scenarios {#sec-med-comparison}

```{r mediation-bias-comparison, fig.height=6.5, fig.width=9, echo=FALSE}
#| code-summary: "▶ Forest plot: OLS vs ROBMED vs Imai across all three scenarios"
comp_df <- tibble(
  Scenario = rep(c(
    sprintf("Scenario 1: Clean normal data\n(true ACME = %.3f)", ACME_true),
    sprintf("Scenario 2: Large outliers\n(true ACME = %.3f)", ACME_true),
    sprintf("Scenario 3: M\u2013Y confounded\n(true ACME = %.3f)", ACME_true_s3)
  ), each = 3),
  Method   = rep(c("OLS delta method", "ROBMED", "Imai (quasi-Bayes)"), 3),
  Truth    = rep(c(ACME_true, ACME_true, ACME_true_s3), each = 3),
  Estimate = c(res_s1$ols_est, res_s1$rob_est, res_s1$imai_est,
               res_s2$ols_est, res_s2$rob_est, res_s2$imai_est,
               res_s3$ols_est, res_s3$rob_est, res_s3$imai_est),
  CI_lo    = c(res_s1$ols_lo,  res_s1$rob_lo,  res_s1$imai_lo,
               res_s2$ols_lo,  res_s2$rob_lo,  res_s2$imai_lo,
               res_s3$ols_lo,  res_s3$rob_lo,  res_s3$imai_lo),
  CI_hi    = c(res_s1$ols_hi,  res_s1$rob_hi,  res_s1$imai_hi,
               res_s2$ols_hi,  res_s2$rob_hi,  res_s2$imai_hi,
               res_s3$ols_hi,  res_s3$rob_hi,  res_s3$imai_hi)
) |>
  mutate(
    Bias         = Estimate - Truth,
    Scenario     = fct_inorder(Scenario),
    Method       = factor(Method, levels = c("Imai (quasi-Bayes)", "ROBMED", "OLS delta method")),
    covers_truth = (CI_lo <= Truth & CI_hi >= Truth)
  )

method_colours <- c(
  "OLS delta method"  = clr_ctrl,
  "ROBMED"            = clr_eco,
  "Imai (quasi-Bayes)" = "#8e44ad"
)

ggplot(comp_df, aes(y = Method, x = Estimate, xmin = CI_lo, xmax = CI_hi,
                    colour = Method, shape = covers_truth)) +
  geom_vline(aes(xintercept = Truth), linetype = "dashed",
             colour = "grey30", linewidth = 1) +
  geom_pointrange(size = 0.7, linewidth = 1) +
  geom_text(aes(label = sprintf("bias=%+.3f", Bias), x = CI_hi + 0.012),
            hjust = 0, size = 2.8, colour = "grey20") +
  facet_wrap(~Scenario, ncol = 1, scales = "free_x") +
  scale_colour_manual(values = method_colours) +
  scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 4),
                     labels = c("TRUE" = "CI covers truth", "FALSE" = "CI misses truth"),
                     name   = NULL) +
  labs(x = "Estimated ACME (indirect effect)", y = NULL,
       colour = "Estimator",
       title  = "Bias comparison: OLS vs. ROBMED vs. Imai across three mediation scenarios",
       subtitle = paste0("Dashed line = true ACME  |  \u00d7 = CI misses truth  |  labels show signed bias\n",
                         "Scenario 3 note: all point estimates are spurious \u2014 ",
                         "Imai adds value via sensitivity analysis, not point estimates")) +
  theme_mod3() +
  theme(legend.position = "top",
        legend.box = "horizontal",
        strip.text = element_text(face = "bold", size = 9),
        legend.key.size = unit(0.7, "cm"))
```

::: {.callout-warning}
## Method × scenario decision guide

| Scenario | OLS | ROBMED | Imai `medsens()` |
|---|---|---|---|
| **Clean normal data** | ✔ Unbiased | ✔ Unbiased | ✔ Large ρ* confirms robustness |
| **Large outliers** | ✖ Biased upward | ✔ Downweights outliers | ✖ Inherits OLS bias |
| **M–Y confounded** | ✖ Type I error | ✖ Type I error | ✔ Small ρ* flags fragility |

**Key lesson**: ROBMED addresses *estimation* failures (non-normality, outliers). Imai's sensitivity analysis addresses *identification* failures (unmeasured confounding). They solve different problems and should be used together.

**In practice**: report ROBMED as your primary ACME estimate (robustness to outliers) *and* `medsens()` to quantify fragility to unmeasured confounding. A finding that survives both tests — ROBMED agrees with OLS, and ρ* is large — is the most credible mediation result you can report.
:::

## Why Confounding Causes Type I Errors: The 2×2 Design {#sec-med-2x2}

The three-scenario analysis above treats confounding as a special case. The 2×2 design below makes the mechanism explicit by crossing two independent dimensions:

- **Rows**: true mediation exists vs. does not exist (b = 0)
- **Columns**: M–Y confounded vs. unconfounded

The critical cell is **top-right** — no true mediation, but confounding is active. Both OLS and ROBMED report a large, significant indirect effect despite b = 0. This is a Type I error caused entirely by the unmeasured confounder U*, not by any true mediation.

```{r med-2x2-setup, echo=FALSE}
#| code-summary: "▶ 2×2 simulation setup"

# ── 2×2 parameters ────────────────────────────────────────────────────────────
p2_mx   <- 0.50   # X → M  (a path)
p2_my   <- 0.55   # M → Y  (b path, used in "true mediation" rows)
p2_xy   <- 0.30   # X → Y  (direct c' path)
p2_um   <- 0.55   # U → M  (confounding into mediator)
p2_uy   <- 0.60   # U → Y  (confounding into outcome)
p2_xu   <- 0.30   # X → U
N2      <- 350
R2      <- 300

set.seed(2024)

run_2x2 <- function(label, confound_m, confound_y, b_val) {
  p_um_act <- if (confound_m) p2_um else 0
  p_uy_act <- if (confound_y) p2_uy else 0
  w_a <- c(p2_mx, p_um_act, 0.30); w_a <- w_a / sum(w_a)
  w_c <- c(p2_xy, b_val,  p_uy_act, 0.30); w_c <- w_c / sum(w_c)
  w_u <- c(p2_xu, 0.30);  w_u <- w_u / sum(w_u)
  x   <- rbinom(N2, 1, 0.5)
  u   <- w_u[1]*x + w_u[2]*rnorm(N2)
  m   <- w_a[1]*x + w_a[2]*u + w_a[3]*rnorm(N2)
  y   <- w_c[1]*x + w_c[2]*m + w_c[3]*u + w_c[4]*rnorm(N2)
  df  <- data.frame(x = x, m = m, y = y)
  mm  <- lm(m ~ x, data = df)
  my  <- lm(y ~ x + m, data = df)
  med <- mediate(mm, my, treat = "x", mediator = "m",
                 boot = TRUE, sims = R2)
  true_indirect <- w_a[1] * w_c[2]
  data.frame(
    label    = label,
    confound = confound_m | confound_y,
    b_true   = b_val,
    true_ind = true_indirect,
    est      = med$d0,
    lo       = med$d0.ci[1],
    hi       = med$d0.ci[2],
    covers   = (med$d0.ci[1] <= true_indirect & med$d0.ci[2] >= true_indirect),
    stringsAsFactors = FALSE
  )
}

scenarios_2x2 <- list(
  list(label = "No true mediation\nUnconfounded",   cm = FALSE, co = FALSE, b = 0),
  list(label = "No true mediation\nConfounded (U*)", cm = TRUE,  co = TRUE,  b = 0),
  list(label = "True mediation\nUnconfounded",       cm = FALSE, co = FALSE, b = p2_my),
  list(label = "True mediation\nConfounded (U*)",    cm = TRUE,  co = TRUE,  b = p2_my)
)

res_2x2 <- do.call(rbind, lapply(scenarios_2x2, function(s) {
  run_2x2(s$label, s$cm, s$co, s$b)
}))
```

```{r med-2x2-dags, fig.height=5, fig.width=11, echo=FALSE}
#| code-summary: "▶ DAGs: the four 2×2 scenarios"

# ── DAG drawing helper (adapted from Mediation-Simulation.Rmd) ────────────────
draw_med_dag <- function(confound_m, confound_y, title = "", show_b = TRUE) {
  nodes <- data.frame(
    label = c("X", "M", "Y", "U*"),
    x     = c(0,   2,   4,   2),
    y     = c(0,   0,   0,   2)
  )
  b_etype <- if (show_b) "path" else "null_b"
  b_col   <- if (show_b) "grey30" else "grey70"
  b_lbl   <- if (show_b) "b" else "b\u00a0=\u00a00"

  edges <- rbind(
    data.frame(x=0, y=0, xend=2, yend=0, etype="path"),
    data.frame(x=2, y=0, xend=4, yend=0, etype=b_etype),
    data.frame(x=0, y=0, xend=2, yend=2, etype="neutral"),
    if (confound_m) data.frame(x=2, y=2, xend=2, yend=0.1, etype="confound") else NULL,
    if (confound_y) data.frame(x=2, y=2, xend=4, yend=0.1, etype="confound") else NULL
  )
  direct <- data.frame(x=0, y=0, xend=4, yend=0)

  ggplot() +
    geom_curve(data = direct, aes(x=x, y=y, xend=xend, yend=yend),
               curvature = -0.35, colour = "#2166ac", linewidth = 0.75,
               arrow = arrow(length = unit(0.16, "cm"), type = "open")) +
    annotate("text", x = 2, y = -0.85, colour = "#2166ac", size = 2.8,
             label = "c\u2019 (direct)", fontface = "italic") +
    geom_segment(data = edges,
                 aes(x=x, y=y, xend=xend, yend=yend,
                     colour=etype, linetype=etype),
                 linewidth = 0.85,
                 arrow = arrow(length = unit(0.16, "cm"), type = "open")) +
    annotate("text", x = 1,   y = 0.25, label = "a", size = 3.1,
             colour = "grey30", fontface = "italic") +
    annotate("text", x = 3,   y = 0.25, label = b_lbl, size = 3.1,
             colour = b_col, fontface = "italic") +
    geom_point(data = nodes, aes(x=x, y=y),
               shape = 21, size = 18, fill = "white", colour = "grey25", stroke = 1.3) +
    geom_text(data = nodes, aes(x=x, y=y, label=label), size = 4, fontface = "bold") +
    scale_colour_manual(guide = "none",
      values = c("path"="grey25", "null_b"="grey75",
                 "neutral"="grey60", "confound"="#d6604d")) +
    scale_linetype_manual(guide = "none",
      values = c("path"="solid", "null_b"="dotted",
                 "neutral"="solid", "confound"="dashed")) +
    coord_cartesian(xlim = c(-0.6, 4.6), ylim = c(-1.1, 2.5)) +
    labs(title = title) +
    theme_void(base_size = 11) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 9.5,
                                    margin = margin(b = 5)))
}

dag_list_2x2 <- list(
  draw_med_dag(FALSE, FALSE, "No true mediation \u00b7 Unconfounded\n\u2705 Correct null",  show_b = FALSE),
  draw_med_dag(TRUE,  TRUE,  "No true mediation \u00b7 Confounded\n\U0001F6A8 FALSE POSITIVE", show_b = FALSE),
  draw_med_dag(FALSE, FALSE, "True mediation \u00b7 Unconfounded\n\u2705 Unbiased",          show_b = TRUE),
  draw_med_dag(TRUE,  TRUE,  "True mediation \u00b7 Confounded\n\u26a0\ufe0f Biased upward",  show_b = TRUE)
)

(dag_list_2x2[[1]] | dag_list_2x2[[2]]) / (dag_list_2x2[[3]] | dag_list_2x2[[4]])
```

```{r med-2x2-plot, fig.height=5.5, fig.width=9, echo=FALSE}
#| code-summary: "▶ True vs. estimated ACME across all four cells"

cell_colours <- c(
  "No true mediation\nUnconfounded"    = "#4dac26",
  "No true mediation\nConfounded (U*)" = "#ca0020",
  "True mediation\nUnconfounded"       = "#0571b0",
  "True mediation\nConfounded (U*)"    = "#f4a582"
)

res_2x2$label <- factor(res_2x2$label, levels = names(cell_colours))

ggplot(res_2x2, aes(x = est, xmin = lo, xmax = hi,
                    y = "Indirect effect",
                    colour = label, shape = covers)) +
  geom_vline(aes(xintercept = true_ind), linetype = "dashed",
             colour = "grey30", linewidth = 0.9) +
  geom_pointrange(size = 0.9, linewidth = 1.2, position = position_dodge(width = 0.5)) +
  geom_point(aes(x = true_ind), shape = 17, size = 3.5,
             colour = "#333", position = position_dodge(width = 0.5)) +
  facet_wrap(~ label, ncol = 2, scales = "free_x") +
  scale_colour_manual(values = cell_colours, guide = "none") +
  scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 4),
                     labels = c("TRUE" = "CI covers truth", "FALSE" = "CI misses truth"),
                     name = NULL) +
  labs(x = "Estimated ACME (indirect effect)",
       y = NULL,
       title = "True vs. Estimated ACME \u2014 2\u00d72 Design",
       subtitle = paste0(
         "\u25b2 Black triangle = true DGP value  |  \u25cf Circle + bar = estimate \u00b1 95% CI\n",
         "\U0001F6A8 Top-right panel: b = 0 in truth, yet the model reports significant mediation \u2014 ",
         "a Type I error driven entirely by U*"
       )) +
  theme_mod3() +
  theme(
    strip.text         = element_text(face = "bold", size = 10, colour = "white",
                                      margin = margin(4, 4, 4, 4)),
    strip.background   = element_rect(fill = "grey30", colour = NA),
    panel.spacing      = unit(1.2, "lines"),
    legend.position    = "top",
    panel.grid.major.y = element_blank()
  )
```

::: {.callout-important}
## The Type I error mechanism in one sentence

When both U→M and U→Y are active, M and Y share a hidden common cause. The mediation model mistakes their spurious correlation for a real b path — and the bootstrap CI confidently narrows around a biased estimate, making the false positive look decisive.

**The fix is not a better estimator.** Imai's `medsens()` reveals how fragile the finding is; ROBMED guards against estimation errors from outliers. But neither can eliminate the Type I error if U* is real and unmeasured. The only solutions are design-based: measure U*, block the back-door path by design, or use an instrument for M.
:::

---

## Summary of Methods {#sec-summary-table}

```{r summary-table, echo=FALSE}
tibble(
  Method = c(
    "Randomised Experiment",
    "Linear Regression Adjustment",
    "Flexible Regression Adjustment",
    "IPW (stabilised)",
    "Entropy Balancing (WeightIt)",
    "Covariate Matching (Mahalanobis)",
    "Propensity Score Matching",
    "Doubly Robust (AIPW)",
    "Synthetic Control",
    "Regression Discontinuity",
    "Difference-in-Differences (TWFE)",
    "Synthetic DiD",
    "Mediation (Baron–Kenny OLS)",
    "Mediation (Imai et al.)",
    "ROBMED"
  ),
  `Key identifying assumption` = c(
    "Random assignment; SUTVA",
    "All confounders measured; linear functional form",
    "All confounders measured; correct non-linear specification",
    "All confounders measured; positivity (overlap)",
    "All confounders measured; moment balance sufficient",
    "All confounders measured; sufficient overlap",
    "All confounders measured; PS model correctly specified",
    "Either PS model OR outcome model is correctly specified",
    "Good pre-treatment fit; no interference",
    "Continuity at cutoff; no manipulation; no other discontinuities",
    "Parallel trends (in absence of treatment)",
    "Parallel trends after reweighting control units",
    "Sequential ignorability; no M–Y confounders",
    "Sequential ignorability; sensitivity analysis quantifies fragility",
    "Sequential ignorability; robust to outliers and non-normality"
  ),
  `Target estimand` = c(
    "ATE","ATE","ATE","ATE","ATE","ATT","ATT","ATE",
    "ATT (one treated unit)","LATE (near threshold)","ATT","ATT",
    "ACME","ACME","ACME"
  ),
  `When to use` = c(
    "Can randomise",
    "Observational; confounders measured; linear DGP",
    "Observational; confounders measured; non-linear DGP",
    "Observational; want ATE; good PS overlap",
    "Observational; want ATE; many covariates to balance",
    "Observational; want ATT; sufficient control units",
    "Observational; many covariates; report PS model sensitivity",
    "Observational; insurance against one misspecified model",
    "Few treated units; many pre-treatment periods",
    "Sharp threshold in a running variable",
    "Panel data; policy change in subset of units",
    "Panel data; parallel trends may not hold exactly",
    "Mediation hypothesis; well-controlled experiment",
    "Mediation with explicit assumptions + sensitivity analysis",
    "Mediation; non-normal WTP data or suspected outliers"
  )
) |>
  knitr::kable(caption="All methods in Module 3: assumptions, estimands, and use cases")
```

---

## Session Info {.unnumbered}

```{r session-info}
sessionInfo()
```
