---
title: "Part 3: Regression Discontinuity Design"
---

```{r setup, include=FALSE}
options(repos = c(CRAN = "https://cran.rstudio.com/"))

if (!requireNamespace("remotes",  quietly = TRUE)) install.packages("remotes")
if (!requireNamespace("synthdid", quietly = TRUE))
  remotes::install_github("synth-did/synthdid")

pkgs <- c(
  "tidyverse", "broom", "knitr", "patchwork",
  "conflicted",
  "MatchIt", "WeightIt", "cobalt",
  "rdrobust", "rddensity",
  "did",
  "mediation", "robmed",
  "estimatr", "sandwich", "lmtest"
)
for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p, dependencies = TRUE)
  library(p, character.only = TRUE)
}
library(synthdid)

library(conflicted)
conflicts_prefer(dplyr::select, dplyr::filter, dplyr::rename, dplyr::mutate,
                 dplyr::summarise, dplyr::arrange, dplyr::lag)

dir.create("figs", recursive = TRUE, showWarnings = FALSE)
knitr::opts_chunk$set(
  echo = TRUE, fig.width = 9, fig.height = 5,
  fig.path = "figs/", dev = "svg"
)

clr_eco     <- "#2d6a4f"
clr_ctrl    <- "#4a90d9"
clr_treated <- clr_eco

theme_mod3 <- function() {
  theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          strip.background = element_rect(fill = "#f0f4f8", colour = NA),
          legend.position  = "bottom")
}

# ── Lightweight DAG helper ─────────────────────────────────────────────────────
.dag <- function(nodes, solid = NULL, dashed = NULL,
                 unobserved = character(0), title = "") {
  nd <- do.call(rbind, lapply(names(nodes), function(nm)
    data.frame(id=nm, x=nodes[[nm]][1], y=nodes[[nm]][2], stringsAsFactors=FALSE)))
  nd$type <- ifelse(nd$id %in% unobserved, "unobs", "obs")
  segs <- function(pairs, lt) {
    if (is.null(pairs) || length(pairs)==0) return(NULL)
    do.call(rbind, lapply(pairs, function(ab) {
      fr <- nd[nd$id==ab[1],]; to <- nd[nd$id==ab[2],]
      dx <- to$x-fr$x; dy <- to$y-fr$y
      len <- sqrt(dx^2+dy^2)+1e-9; pad <- 0.12
      data.frame(x=fr$x+pad*dx/len, y=fr$y+pad*dy/len,
                 xend=to$x-pad*dx/len, yend=to$y-pad*dy/len,
                 lt=lt, stringsAsFactors=FALSE)
    }))
  }
  edf <- rbind(segs(solid,"solid"), segs(dashed,"dashed"))
  xs <- range(nd$x); ys <- range(nd$y)
  p <- ggplot() + theme_void() + ggtitle(title) +
    coord_cartesian(xlim=c(xs[1]-.3,xs[2]+.3), ylim=c(ys[1]-.3,ys[2]+.3), expand=FALSE) +
    theme(plot.title=element_text(size=10,face="bold",hjust=.5,margin=margin(b=6)),
          plot.background=element_rect(fill="grey98",colour="grey80",linewidth=.4),
          plot.margin=margin(10,16,10,16), legend.position="none")
  if (!is.null(edf) && nrow(edf)>0)
    p <- p +
      geom_segment(data=edf, aes(x=x,y=y,xend=xend,yend=yend,colour=lt,linetype=lt),
                   arrow=arrow(length=unit(.20,"cm"),type="closed"), linewidth=.85) +
      scale_colour_manual(values=c(solid="grey20",dashed="firebrick")) +
      scale_linetype_manual(values=c(solid="solid",dashed="dashed"))
  if (any(nd$type=="obs"))
    p <- p + geom_label(data=nd[nd$type=="obs",], aes(x=x,y=y,label=id),
                        size=4, fontface="bold", fill="white", colour="black",
                        label.size=.9, label.padding=unit(.4,"lines"))
  if (any(nd$type=="unobs"))
    p <- p + geom_label(data=nd[nd$type=="unobs",], aes(x=x,y=y,label=id),
                        size=4, fontface="bold.italic", fill="#fff3e0", colour="#b45309",
                        label.size=1.1, label.padding=unit(.4,"lines"),
                        label.r=unit(.5,"lines"))
  p
}

# ── Mediation extraction helper (handles robmed version differences) ───────────
rob_extract <- function(fit) {
  # point estimate: try $ab, then $indirect, then coef()["ab"]
  ab <- tryCatch(fit$ab,
    error = function(e) tryCatch(fit$indirect,
      error = function(e2) tryCatch(coef(fit)["ab"],
        error = function(e3) NA_real_)))
  # CI: try $ci (matrix or vector), then confint()
  ci <- tryCatch({
    r <- fit$ci
    if (is.matrix(r)) r[1,] else r
  }, error = function(e) tryCatch(confint(fit)["ab",],
    error = function(e2) c(NA_real_, NA_real_)))
  list(est=unname(ab), lo=unname(ci[1]), hi=unname(ci[2]))
}
```

::: {.callout-tip}
## How to use the code in this section

The code blocks throughout this section contain inline annotations explaining **what data structures go in**, **what the key function arguments control**, and **what to look for in the output**. Click any "▶ Code" triangle to expand a block and read the annotations alongside the code. Treat them as a guided checklist: wherever you see a comment starting with `# ──`, it marks a decision point you would need to revisit with your own data.
:::

## The Core Idea {#sec-rdd-idea}

AlterEco's retailer awards a **"Certified Sustainable"** badge to any coffee product scoring **70 or above** on a third-party environmental audit (scored 0–100). Products at 69 miss the badge; products at 71 get it. Because the audit score has measurement noise, which side of the cutoff a product lands on is *nearly random* near 70. We exploit this near-randomness.

::: {.callout-note}
## Why the jump at 70 is causal — the "as-good-as-random" intuition

Imagine two products with audit scores of 69 and 71. They have almost identical green credentials — similar supply chains, similar materials — but the auditor's ruler is imprecise. Those two points of difference are largely measurement noise. Yet one gets the badge and the other doesn't. **Near the cutoff, badge assignment is as good as a coin flip.**

This local near-randomness is what makes the jump causal. It is equivalent to a mini natural experiment: products just above and just below 70 are exchangeable in every way except for badge receipt. The jump in WTP at that threshold is therefore attributable *entirely* to the badge.

**The cost of this elegance**: the estimate is a *local* average treatment effect (LATE). It tells you the causal effect of receiving the badge for brands hovering around a 70-point score — not for the top-scoring 90s or the low-scoring 40s, which may respond to the badge very differently. This LATE is conceptually identical to the LATE from Part 1 of this module — both arise because the causal estimate is anchored to a specific sub-population (compliers near the threshold) rather than the full population.

**A Module 2 connection:** The "as-good-as-random near the cutoff" logic is local randomisation — a naturally occurring version of the randomised experiment studied in Module 2. The same assumption that makes experiments valid (exchangeability of treated and control) holds here, but only locally. In Module 2, randomisation made exchangeability a design guarantee; here, it is an *empirical claim* that must be verified through the diagnostics below.

**A Module 1 connection:** The running variable itself is a measurement. If audit scores are measured with substantial error — auditors apply different standards, or scores reflect lobbying as much as sustainability — then the near-threshold units may *not* be as similar as assumed. Measurement error in the running variable biases the RDD estimate toward zero (attenuation at the boundary) and inflates the apparent smoothness of pre-cutoff trends. This is the causal-inference analogue of the reliability problem in Module 1: a noisy ruler undermines the precision of every inference drawn from it.

**A practical check to run first**: If brands can game the audit score to land just above 70, the "as-good-as-random" assumption breaks down. The density test below checks for suspicious bunching right at the cutoff.
:::

## Simulating an RDD {#sec-rdd-sim}

```{r rdd-dgp, fig.height=4.5, fig.width=9}
#| code-summary: "▶ Simulate 1,200 products near the 70-point certification threshold"
set.seed(2024)
N_rdd <- 1200; cutoff <- 70; true_jump <- 0.80
# ── YOUR DATA: replace audit_score with your running variable (the continuous
#    score that determines treatment assignment), cutoff with the actual policy
#    threshold, and WTP_rdd with your outcome variable.
#    df_rdd must contain: the running variable, a 0/1 treatment indicator
#    (= 1 if running variable >= cutoff), the outcome, and running = running - cutoff.
audit_score <- runif(N_rdd, 30, 100)
treat_rdd   <- as.integer(audit_score >= cutoff)

WTP_rdd <- 4.0 + 0.02*(audit_score-cutoff) - 0.0003*(audit_score-cutoff)^2 +
           true_jump*treat_rdd + rnorm(N_rdd, 0, 0.60)
WTP_rdd <- pmax(1, pmin(10, WTP_rdd))
df_rdd  <- tibble(audit_score, treat_rdd, WTP_rdd, running=audit_score-cutoff)
# ── CHECK: plot outcome vs. running variable (as below) — a visible jump at
#    the cutoff, with smooth trends on each side, is the visual signature of
#    a valid RDD. Irregular or absent jumps warrant investigating the DGP.

df_rdd |>
  ggplot(aes(x=audit_score, y=WTP_rdd, colour=factor(treat_rdd))) +
  geom_point(alpha=0.30, size=0.9) +
  geom_vline(xintercept=cutoff, linewidth=1, colour="grey30") +
  geom_smooth(method="lm", formula=y~poly(x,2), se=TRUE, alpha=0.2) +
  scale_colour_manual(values=c("0"=clr_ctrl,"1"=clr_eco),
                      labels=c("Below 70 (no badge)","Above 70 (badge)")) +
  annotate("text", x=71, y=9.2, label="Cutoff = 70", hjust=0, size=3.5, fontface="bold") +
  annotate("segment", x=70, xend=70, y=3.85, yend=4.85, colour="firebrick",
           arrow=arrow(ends="both", length=unit(.2,"cm")), linewidth=0.8) +
  annotate("text", x=71, y=4.35,
           label=sprintf("~$%.2f jump\n(badge effect)", true_jump),
           hjust=0, size=3, colour="firebrick") +
  labs(x="Audit Score (running variable)", y="Consumer WTP ($)", colour=NULL,
       title="Regression Discontinuity: the jump at 70 estimates the badge's causal effect") +
  theme_mod3()
```

## Key Assumptions and How to Test Them {#sec-rdd-assumptions}

### Assumption 1: No Manipulation

```{r rdd-density-test, fig.height=4, fig.width=7}
# ── YOUR DATA: replace df_rdd$audit_score with your running variable column
#    and c=cutoff with your policy threshold value.
# ── KEY ARGS: rddensity(X, c) tests whether the density of the running variable
#    is continuous at the cutoff; a significant p-value (< .05) means units are
#    sorting — manipulating the running variable to land just above (or below)
#    the threshold, which invalidates the "as-good-as-random" RDD assumption.
dens_test <- rddensity(X=df_rdd$audit_score, c=cutoff)
summary(dens_test)
# ── CHECK: look at the p-value in the summary output; p > .05 is required for
#    the no-manipulation assumption; also inspect the density plot for visible
#    bunching or a gap immediately around the cutoff.
invisible(rdplotdensity(dens_test, X=df_rdd$audit_score,
              title="Density test: checking for bunching at the 70-point cutoff",
              xlabel="Audit score", ylabel="Density"))
```

A significant p-value signals that brands are gaming audit scores to land just above 70. With our simulated uniform running variable, the density is continuous.

### Assumption 2: Covariate Continuity

```{r rdd-covariate-balance}
env_rdd <- rnorm(N_rdd, 0 + 0.01*(audit_score-cutoff)/10, 1)
rdd_env <- rdrobust(y=env_rdd, x=df_rdd$audit_score, c=cutoff)
cat("Covariate balance — env. concern should NOT jump at the threshold:\n")
summary(rdd_env)
```

The covariate continuity test is the RDD's randomisation check — conceptually identical to the balance tables you ran in Module 2 before analysing experimental data. In an experiment, balance across pre-treatment covariates is a consequence of random assignment; here it must be demonstrated empirically as evidence that the threshold is not systematically sorting units on unobserved characteristics.

## Estimating the RDD Effect {#sec-rdd-estimate}

```{r rdd-estimate, fig.height=4.5, fig.width=8}
# ── YOUR DATA: replace y= with your outcome variable, x= with your running
#    variable, and c= with your cutoff value. Both y and x must be numeric vectors
#    of the same length with no missing values at the cutoff.
# ── KEY ARGS: rdrobust() selects bandwidth automatically (MSE-optimal rule);
#    you can override with h= to specify a fixed bandwidth, or kernel= to change
#    from the default triangular kernel (options: "epanechnikov", "uniform").
#    The default uses local linear regression on each side; p=2 switches to
#    local quadratic (more flexible, higher variance).
rdd_est <- rdrobust(y=df_rdd$WTP_rdd, x=df_rdd$audit_score, c=cutoff)
summary(rdd_est)
# ── CHECK: in the summary output, look at "Conventional" coef for the point
#    estimate and "Robust" CI for the bias-corrected confidence interval —
#    use the Robust CI for inference. Also note the automatically chosen
#    bandwidth (h) and effective sample size on each side of the cutoff.
rdplot(y=df_rdd$WTP_rdd, x=df_rdd$audit_score, c=cutoff,
       title="RDD estimate: causal effect of sustainability badge on WTP",
       x.label="Audit Score", y.label="WTP ($)")
```

```{r rdd-bandwidth-sensitivity, fig.height=4, fig.width=8}
# ── YOUR DATA: adjust the bandwidth grid c(5,8,10,15,20,25,30) to be
#    meaningful in your running variable's units — e.g., if your score runs
#    0–1000, sensible bandwidths might be c(25,50,75,100,150,200).
#    Keep the grid symmetric around the MSE-optimal h chosen by rdrobust() above.
# ── KEY ARGS: h=bw fixes the bandwidth; est$coef["Conventional",1] is the
#    point estimate; est$se["Robust",1] is the bias-corrected robust SE.
bw_results <- map_dfr(c(5,8,10,15,20,25,30), function(bw) {
  est <- rdrobust(y=df_rdd$WTP_rdd, x=df_rdd$audit_score, c=cutoff, h=bw)
  tibble(bandwidth=bw, estimate=est$coef["Conventional",1], se=est$se["Robust",1])
})
# ── CHECK: estimates should be roughly flat across bandwidths — large swings
#    indicate sensitivity to bandwidth choice (a red flag). If estimates collapse
#    at very narrow bandwidths, the effective N is too small for precision.

bw_results |>
  mutate(lo=estimate-1.96*se, hi=estimate+1.96*se) |>
  ggplot(aes(x=bandwidth, y=estimate, ymin=lo, ymax=hi)) +
  geom_hline(yintercept=true_jump, linetype="dashed", colour="grey40", linewidth=1) +
  geom_ribbon(alpha=0.15, fill=clr_eco) +
  geom_line(colour=clr_eco, linewidth=1) + geom_point(colour=clr_eco, size=2.5) +
  annotate("text", x=5.5, y=true_jump+0.05, label=sprintf("True = $%.2f", true_jump),
           hjust=0, size=3.5) +
  labs(x="Bandwidth (audit score units)", y="Estimated badge effect ($)",
       title="RDD estimates stable across bandwidths h = 5 to 30",
       subtitle="Instability would indicate sensitivity to bandwidth — a red flag") +
  theme_mod3()
```

::: {.callout-note}
## The bandwidth trade-off: bias vs. variance

The bandwidth $h$ controls which products are included in the RDD comparison — only those with audit scores in $[70-h,\ 70+h]$. This creates a fundamental tension:

| Narrow $h$ | Wide $h$ |
|---|---|
| Products are very similar to each other near the cutoff ✔ | More observations → tighter CI ✔ |
| Few observations → wide CI ✖ | Products far from 70 differ systematically from each other ✖ |

**`rdrobust` selects the bandwidth automatically** using a mean-squared-error optimal rule — it finds the $h$ that best balances bias and variance. The bandwidth-sensitivity plot above is a sanity check: stable estimates across a wide range of $h$ suggest the result is not an artefact of a specific choice.

When estimates collapse or swing wildly as $h$ changes, it usually means the polynomial fit on one side of the cutoff is extrapolating too aggressively. That's a red flag that warrants narrowing $h$ or using a local linear fit.
:::

## Type I Error from Running Variable Manipulation {#sec-rdd-manipulation-mc}

The density test and covariate continuity checks are the right diagnostics — but how much Type I error accumulates before those tests catch the problem? Here we simulate a concrete manipulation scenario and track false-positive rates as gaming becomes more common.

**The mechanism.** Two independent variables matter: (1) the *audit score* — operational compliance across energy use, waste, and supply chain — and (2) latent *brand eco-reputation* Z, an unobserved dimension reflecting genuine brand prestige and loyal eco-customer base. A brand can have strong reputation yet score just below 70: a complex supply chain, an unfamiliar auditor, or a borderline factory can drag the score down through measurement noise. Z does two things: it **motivates gaming** (high-reputation brands near the threshold believe they deserve the badge and have marketing resources to appeal, re-audit, or make cosmetic improvements) and it **independently drives WTP** (eco-conscious customers already follow high-reputation brands regardless of the badge). **The true badge effect is zero.**

The bias accumulates via *compositional sorting*. Most brands in the sample are clustered around the cutoff (audit scores ~ N(65, 12) — many borderline brands, fewer extreme ones). After gaming, just above 70 is flooded with high-Z departures from just below; just below 70 is left depleted of those same high-Z brands. The RDD compares these compositionally distinct groups and interprets the Z-driven WTP gap as a badge effect.

```{r rdd-manipulation-mc, fig.height=5, fig.width=10}
#| code-summary: "▶ Monte Carlo: Type I error and density-test power vs. gaming rate (200 sims)"
set.seed(2025)
N_SIM_RDD   <- 200
N_RDD_MC    <- 500
CUTOFF_MC   <- 70
GAME_WINDOW <- 10   # brands within 10 pts below cutoff are eligible to game

rdd_mc_one <- function(p_game) {
  mat <- replicate(N_SIM_RDD, {
    # Z: latent brand eco-reputation — the UNOBSERVED confounder.
    # Z is independent of operational audit score but drives both gaming and WTP.
    Z <- rnorm(N_RDD_MC, 0, 1)

    # Audit scores clustered near cutoff: most brands are borderline.
    # A brand's score reflects operational compliance, NOT brand reputation (Z).
    audit <- pmin(100, pmax(30,
                            rnorm(N_RDD_MC, mean = 65, sd = 12)))

    # High-reputation brands (Z > 0) just below the threshold game the audit.
    # They move to just above 70 — bringing their high-Z customer base with them.
    near_below <- audit >= (CUTOFF_MC - GAME_WINDOW) & audit < CUTOFF_MC
    game_mask  <- near_below & (Z > 0) & (runif(N_RDD_MC) < p_game)
    audit[game_mask] <- pmin(100,
                             audit[game_mask] + runif(sum(game_mask), 1, 8))

    # WTP is driven by brand eco-reputation Z alone — badge has NO causal effect.
    # Beta_Z = 1.5: a 1-SD reputation advantage is worth $1.50 in WTP.
    # Low idiosyncratic noise (SD = $0.30) so the Z signal dominates.
    # Post-gaming: RDD sees a spurious WTP jump at 70 driven by Z-sorting, not badge.
    WTP_mc <- 5.5 + 1.5 * Z + rnorm(N_RDD_MC, 0, 0.30)
    WTP_mc <- pmax(1, pmin(10, WTP_mc))

    rdd_p <- tryCatch({
      est <- rdrobust(y = WTP_mc, x = audit, c = CUTOFF_MC)
      2 * pnorm(-abs(est$coef[1] / est$se[1]))
    }, error = function(e) NA_real_)

    dens_p <- tryCatch({
      rd <- rddensity(X = audit, c = CUTOFF_MC)
      p  <- rd$test$p_jk
      if (is.null(p) || !is.finite(p)) 1.0 else as.numeric(p)[1]
    }, error = function(e) 1.0)

    c(rdd_p, dens_p)
  })

  data.frame(p_game    = p_game,
             rdd_pval  = mat[1, ],
             dens_pval = mat[2, ])
}

game_levels <- c(0, 0.10, 0.20, 0.30, 0.50, 0.70)
sim_rdd_mc  <- map_dfr(game_levels, rdd_mc_one)
cat(sprintf("RDD simulation complete: %d conditions x %d sims = %d total.\n",
            length(game_levels), N_SIM_RDD, nrow(sim_rdd_mc)))
```

```{r rdd-mc-plot, fig.height=5, fig.width=10, echo=FALSE}
rdd_mc_summary <- sim_rdd_mc |>
  group_by(p_game) |>
  summarise(
    `RDD Type I error\n(false badge effect)` = mean(rdd_pval  < 0.05, na.rm = TRUE),
    `Density test power\n(catches manipulation)` = mean(dens_pval < 0.05, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_longer(-p_game, names_to = "Metric", values_to = "Rate")

ggplot(rdd_mc_summary, aes(x = p_game * 100, y = Rate, colour = Metric, shape = Metric)) +
  geom_hline(yintercept = 0.05, linetype = "dotted", colour = "grey40", linewidth = 0.9) +
  annotate("text", x = 2, y = 0.075, label = "Nominal 5%", size = 3.2,
           colour = "grey40", hjust = 0) +
  geom_line(linewidth = 1.2) + geom_point(size = 3.5) +
  scale_colour_manual(values = c(
    "RDD Type I error\n(false badge effect)"       = "firebrick",
    "Density test power\n(catches manipulation)" = "#2563a8"
  )) +
  scale_shape_manual(values = c(
    "RDD Type I error\n(false badge effect)"       = 17,
    "Density test power\n(catches manipulation)" = 16
  )) +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(
    x      = "Probability that an eligible firm games the audit score",
    y      = "Rate",
    colour = NULL, shape = NULL,
    title  = "RDD Type I error from manipulation (true badge effect = $0)",
    subtitle = sprintf(
      "High-credential brands (high Z) game into just-above 70; low-Z brands remain below — spurious WTP gap grows with gaming rate  |  %d sims per condition",
      N_SIM_RDD)
  ) +
  theme_mod3() +
  theme(legend.key.width = unit(2, "cm"))
```

```{r rdd-mc-table, echo=FALSE}
sim_rdd_mc |>
  group_by(p_game) |>
  summarise(
    `RDD Type I error`   = sprintf("%.0f%%", 100 * mean(rdd_pval  < 0.05, na.rm = TRUE)),
    `Density test power` = sprintf("%.0f%%", 100 * mean(dens_pval < 0.05, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  transmute(
    `Gaming rate` = sprintf("%.0f%%", p_game * 100),
    `RDD Type I error`, `Density test power`
  ) |>
  knitr::kable(
    caption = sprintf(
      "RDD simulation: %d sims per condition, N = %d products, true badge effect = $0. Gaming firms are high-credential brands near the threshold.",
      N_SIM_RDD, N_RDD_MC)
  )
```

::: {.callout-warning}
## The key pattern: RDD Type I error races toward 100% while the density test lags

As gaming becomes common, the density of audit scores just above 70 swells (gamers pile in from below) and the density just below depletes. The RDD treats this Z-sorted composition as a real treatment effect — **Type I error climbs steeply toward 100%**. The density test (McCrary/rddensity) eventually picks up the bunching, but it has substantially less power than the spurious RDD finding itself: a researcher who checks `rddensity`, sees a borderline p-value, and proceeds will still produce highly confident false positives.

**Why does Z go undetected?** The audit score is observable; Z (brand reputation) is not. Standard covariate balance checks only test *observed* pre-treatment variables. If no survey measure of brand reputation was collected, the sorting on Z is completely invisible — yet it inflates Type I error just as severely.

**What to do:** (1) Collect measures of *why* firms are near the threshold — if you can measure Z, include it as a covariate and check its continuity. (2) Run a donut RDD excluding observations within a narrow band of the cutoff, where gaming is most concentrated. (3) If gaming is plausible from domain knowledge, treat any RDD near a known incentivized threshold with extra skepticism regardless of what the density test says.
:::

---

## Researcher Checklist: Regression Discontinuity Design {#sec-checklist-part3-m3}

::: {.callout-note}
## Key questions about your discontinuity

- [ ] **Can units manipulate their position on the running variable?** If firms or individuals can precisely control their score to land just above the threshold, the local quasi-randomness assumption is violated. Run a density test (`rddensity`) to check for systematic bunching at the cutoff.
- [ ] **Are pre-treatment covariates smooth through the threshold?** Run RDD regressions with pre-determined characteristics as the outcome. Any discontinuity in pre-treatment variables signals selection, not randomness.
- [ ] **Is the estimate stable across bandwidth choices?** If the point estimate changes dramatically as you narrow the bandwidth from the MSE-optimal value, the result depends on extrapolating far from the cutoff — not on the local quasi-experiment.
- [ ] **Do you understand that your estimate is a LATE?** RDD identifies the effect only for units near the threshold — those on the margin of qualifying. This may or may not generalize to the broader population of treated units.
:::
