---
title: "Module 3 Code Tutorial"
subtitle: "Treatment Effects, Natural Experiments, and Causal Mechanisms"
---

```{r setup, include=FALSE}
options(repos = c(CRAN = "https://cran.rstudio.com/"))

if (!requireNamespace("remotes",  quietly = TRUE)) install.packages("remotes")
if (!requireNamespace("synthdid", quietly = TRUE))
  remotes::install_github("synth-did/synthdid")

pkgs <- c(
  "tidyverse", "broom", "knitr", "patchwork",
  "conflicted",
  "MatchIt", "WeightIt", "cobalt",
  "rdrobust", "rddensity",
  "did",
  "mediation", "robmed",
  "estimatr", "sandwich", "lmtest"
)
for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p, dependencies = TRUE)
  library(p, character.only = TRUE)
}
library(synthdid)

library(conflicted)
conflicts_prefer(dplyr::select, dplyr::filter, dplyr::rename, dplyr::mutate,
                 dplyr::summarise, dplyr::arrange, dplyr::lag)

dir.create("figs", recursive = TRUE, showWarnings = FALSE)
knitr::opts_chunk$set(
  echo = TRUE, fig.width = 9, fig.height = 5,
  fig.path = "figs/", dev = "svg"
)

clr_eco     <- "#2d6a4f"
clr_ctrl    <- "#4a90d9"
clr_treated <- clr_eco

theme_mod3 <- function() {
  theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          strip.background = element_rect(fill = "#f0f4f8", colour = NA),
          legend.position  = "bottom")
}

# ── Lightweight DAG helper ─────────────────────────────────────────────────────
.dag <- function(nodes, solid = NULL, dashed = NULL,
                 unobserved = character(0), title = "") {
  nd <- do.call(rbind, lapply(names(nodes), function(nm)
    data.frame(id=nm, x=nodes[[nm]][1], y=nodes[[nm]][2], stringsAsFactors=FALSE)))
  nd$type <- ifelse(nd$id %in% unobserved, "unobs", "obs")
  segs <- function(pairs, lt) {
    if (is.null(pairs) || length(pairs)==0) return(NULL)
    do.call(rbind, lapply(pairs, function(ab) {
      fr <- nd[nd$id==ab[1],]; to <- nd[nd$id==ab[2],]
      dx <- to$x-fr$x; dy <- to$y-fr$y
      len <- sqrt(dx^2+dy^2)+1e-9; pad <- 0.12
      data.frame(x=fr$x+pad*dx/len, y=fr$y+pad*dy/len,
                 xend=to$x-pad*dx/len, yend=to$y-pad*dy/len,
                 lt=lt, stringsAsFactors=FALSE)
    }))
  }
  edf <- rbind(segs(solid,"solid"), segs(dashed,"dashed"))
  xs <- range(nd$x); ys <- range(nd$y)
  p <- ggplot() + theme_void() + ggtitle(title) +
    coord_cartesian(xlim=c(xs[1]-.3,xs[2]+.3), ylim=c(ys[1]-.3,ys[2]+.3), expand=FALSE) +
    theme(plot.title=element_text(size=10,face="bold",hjust=.5,margin=margin(b=6)),
          plot.background=element_rect(fill="grey98",colour="grey80",linewidth=.4),
          plot.margin=margin(10,16,10,16), legend.position="none")
  if (!is.null(edf) && nrow(edf)>0)
    p <- p +
      geom_segment(data=edf, aes(x=x,y=y,xend=xend,yend=yend,colour=lt,linetype=lt),
                   arrow=arrow(length=unit(.20,"cm"),type="closed"), linewidth=.85) +
      scale_colour_manual(values=c(solid="grey20",dashed="firebrick")) +
      scale_linetype_manual(values=c(solid="solid",dashed="dashed"))
  if (any(nd$type=="obs"))
    p <- p + geom_label(data=nd[nd$type=="obs",], aes(x=x,y=y,label=id),
                        size=4, fontface="bold", fill="white", colour="black",
                        label.size=.9, label.padding=unit(.4,"lines"))
  if (any(nd$type=="unobs"))
    p <- p + geom_label(data=nd[nd$type=="unobs",], aes(x=x,y=y,label=id),
                        size=4, fontface="bold.italic", fill="#fff3e0", colour="#b45309",
                        label.size=1.1, label.padding=unit(.4,"lines"),
                        label.r=unit(.5,"lines"))
  p
}

# ── Mediation extraction helper (handles robmed version differences) ───────────
rob_extract <- function(fit) {
  # point estimate: try $ab, then $indirect, then coef()["ab"]
  ab <- tryCatch(fit$ab,
    error = function(e) tryCatch(fit$indirect,
      error = function(e2) tryCatch(coef(fit)["ab"],
        error = function(e3) NA_real_)))
  # CI: try $ci (matrix or vector), then confint()
  ci <- tryCatch({
    r <- fit$ci
    if (is.matrix(r)) r[1,] else r
  }, error = function(e) tryCatch(confint(fit)["ab",],
    error = function(e2) c(NA_real_, NA_real_)))
  list(est=unname(ab), lo=unname(ci[1]), hi=unname(ci[2]))
}
```

```{=html}
<style>
#global-output-toggle {
  position: fixed; bottom: 0; left: 0; right: 0; z-index: 9999;
  display: flex; flex-wrap: wrap; gap: 10px;
  justify-content: center; align-items: center;
  padding: 11px 24px;
  background: #1e3a5f; border-top: 3px solid #4a90d9;
  box-shadow: 0 -3px 16px rgba(0,0,0,0.35);
}
#global-output-toggle span {
  color: #c8d8ed; font-size: 0.85em; font-weight: 700;
  letter-spacing: 0.6px; text-transform: uppercase; margin-right: 6px;
}
body { padding-bottom: 70px !important; }
.gtbtn {
  padding: 9px 24px; color: #ffffff; border-radius: 7px;
  font-size: 1em; font-weight: 700; cursor: pointer;
  letter-spacing: 0.2px; transition: background 0.15s, border-color 0.15s;
  white-space: nowrap; border: 2px solid transparent;
}
.gtbtn.code-btn   { background: #2563a8; border-color: #5b9bd5; }
.gtbtn.code-btn:hover { background: #1a4a80; border-color: #90bfe8; }
.gtbtn.output-btn { background: #2d6a4f; border-color: #52b788; }
.gtbtn.output-btn:hover { background: #1b4332; border-color: #74c69d; }
details.out-fold { margin: 6px 0 14px 0; }
summary.out-fold-btn {
  display: inline-flex; align-items: center; gap: 7px;
  padding: 8px 20px; background: #eef4fb;
  border: 2px solid #4a90d9; border-radius: 7px; cursor: pointer;
  font-size: 0.94em; font-weight: 700; color: #1a3a6b;
  user-select: none; list-style: none; transition: background 0.12s;
}
summary.out-fold-btn::-webkit-details-marker { display: none; }
summary.out-fold-btn:hover { background: #cfe2f7; }
details[open].out-fold > summary.out-fold-btn { background: #d6ecff; }
summary.out-fold-btn .fold-icon { font-size: 1em; }
details[open] .fold-icon-closed { display: none; }
details:not([open]) .fold-icon-open  { display: none; }
</style>
<script>
document.addEventListener('DOMContentLoaded', function () {
  var outputSels = ['.cell-output-stdout','.cell-output-stderr','.cell-output-display'];
  document.querySelectorAll(outputSels.join(',')).forEach(function (el) {
    if (el.closest('.callout') || el.closest('details.out-fold')) return;
    var hasImg = !!el.querySelector('img, svg, canvas');
    var hasTbl = !!el.querySelector('table');
    var label  = hasImg ? '📊  Show plot' : hasTbl ? '📋  Show table' : '📄  Show output';
    var det = document.createElement('details'); det.className = 'out-fold';
    var sum = document.createElement('summary'); sum.className = 'out-fold-btn';
    sum.innerHTML = '<span class="fold-icon fold-icon-closed">▶</span><span class="fold-icon fold-icon-open">▼</span>' + label;
    el.parentNode.insertBefore(det, el); det.appendChild(sum); det.appendChild(el);
  });
  var bar = document.createElement('div'); bar.id = 'global-output-toggle';
  bar.innerHTML = '<span>Quick controls:</span><button id="code-toggle-btn" class="gtbtn code-btn" onclick="toggleCode()">\u25bc Expand All Code</button><button id="output-toggle-btn" class="gtbtn output-btn" onclick="toggleOutput()">\u25bc Expand All Output</button>';
  document.body.appendChild(bar);
});
var codeOpen = false, outputOpen = false;
function toggleCode() {
  codeOpen = !codeOpen;
  document.querySelectorAll('details').forEach(function(d) {
    if (!d.classList.contains('out-fold') && d.querySelector('.sourceCode')) d.open = codeOpen;
  });
  var btn = document.getElementById('code-toggle-btn');
  if (btn) btn.textContent = codeOpen ? '\u25b2 Collapse All Code' : '\u25bc Expand All Code';
}
function toggleOutput() {
  outputOpen = !outputOpen;
  document.querySelectorAll('details.out-fold').forEach(function(d) { d.open = outputOpen; });
  var btn = document.getElementById('output-toggle-btn');
  if (btn) btn.textContent = outputOpen ? '\u25b2 Collapse All Output' : '\u25bc Expand All Output';
}
</script>
```

## What This Tutorial Is About {.unnumbered}

There is a single thread running through all three modules:

> **Your observable may not accurately reflect the latent quantity you want to study.**

In **Module 1** that problem lived in your outcome variable — your scale picked up variance from constructs you never intended to measure. In **Module 2** it appeared in your treatment variable and your significance test — a manipulated treatment can inject multiple signals simultaneously, and a p-value only means what it means when randomisation is actually achieved.

Both modules used the same running example: **AlterEco Coffee's eco-label experiment**, where participants were randomly assigned to see (or not see) an eco-label and reported their willingness to pay (WTP, \$1–\$10).

This module asks a harder set of questions — what if we cannot run that clean experiment?

| Question | Tool |
|---|---|
| *What does "the effect" mean for an individual?* | Potential outcomes / treatment effect taxonomy |
| *What if consumers freely choose eco-labelled products?* | Selection on observables |
| *What if a certification rule creates a natural threshold?* | Regression discontinuity |
| *What if eco-labelling was rolled out by policy over time?* | Difference-in-differences |
| *Does the eco-label work **because** of perceived sustainability?* | Causal mediation analysis |

Every section uses simulated data with **known true effects** so you can see exactly when each method recovers the truth — and when it does not.

::: {.callout-note}
## Learning Objectives
- Distinguish ATE, ATT, ATC, LATE, ITT, and CATE — and know which one answers your question
- Apply regression adjustment, IPW, covariate matching, propensity-score matching, entropy balancing, and doubly-robust estimation
- Conduct an RDD analysis and test its assumptions
- Implement DiD, an event study, and synthetic DiD
- Diagnose when standard mediation gives wrong answers (three structural failure modes)
- Use `robmed` and `mediation` packages; interpret sensitivity analyses
- Compare bias across methods using a known data-generating system
:::

---

# Part 1: What Do We Mean by "Causal Effect"? {#sec-m3-part1}

<iframe src="https://liveeur-my.sharepoint.com/personal/58542dsc_eur_nl/_layouts/15/Doc.aspx?sourcedoc={ac5b8424-7776-44d2-9bde-03770c0128d9}&action=embedview&wdAr=1.7777777777777777" width="100%" height="500px" frameborder="0" allowfullscreen title="Module 3 Part 1: Causal Effects"></iframe>

## The Fundamental Problem of Causal Inference {#sec-fundamental}

Every participant in the AlterEco eco-label study had **two potential outcomes**:

- $Y_i(0)$: their WTP if they did *not* see the eco-label
- $Y_i(1)$: their WTP if they *did* see the eco-label

Their **individual treatment effect** is $\tau_i = Y_i(1) - Y_i(0)$.

The problem? **We can only observe one of these for any individual.** This is called the **Fundamental Problem of Causal Inference** (Holland, 1986). Everything that follows is a strategy for getting around it.

```{r dgs-potential-outcomes}
#| code-summary: "▶ Build the data-generating system"
set.seed(2024)
N <- 2000

env        <- rnorm(N, 0, 1)
price_sens <- rnorm(N, 0, 1)
income     <- rnorm(N, 0, 1)

Y0 <- 5.50 + 0.80*env - 0.50*price_sens + 0.30*income + rnorm(N, 0, 1.10)
Y0 <- pmax(1, pmin(10, Y0))

ITE <- 0.60 + 0.80*env - 0.55*price_sens + rnorm(N, 0, 0.40)
Y1  <- pmax(1, pmin(10, Y0 + ITE))

ATE_true <- mean(Y1 - Y0)
cat(sprintf("True ATE  = %.3f\nTrue ITE SD = %.3f\n", ATE_true, sd(Y1 - Y0)))
```

```{r fundamental-problem-table, echo=FALSE}
#| code-summary: "▶ Show the missing-data structure"
set.seed(42)
demo_idx <- c(which(env > 1.2)[1:2], which(abs(env) < 0.15)[1], which(env < -1.2)[1:2])
assigned <- c("Control","Treatment","Control","Treatment","Control")
tibble(
  Person             = c("Sophie (high concern)","Rens (moderate+)","Lotte (neutral)","Daan (low concern)","Kim (price-sensitive)"),
  `Env concern`      = round(env[demo_idx], 1),
  `Y(0) No label`   = round(Y0[demo_idx], 2),
  `Y(1) With label` = round(Y1[demo_idx], 2),
  `True ITE`        = round(Y1[demo_idx]-Y0[demo_idx], 2),
  `Assigned to`     = assigned,
  `What we observe` = c(
    sprintf("Y(0)=$%.2f  Y(1)=???", Y0[demo_idx[1]]),
    sprintf("Y(0)=???   Y(1)=$%.2f", Y1[demo_idx[2]]),
    sprintf("Y(0)=$%.2f  Y(1)=???", Y0[demo_idx[3]]),
    sprintf("Y(0)=???   Y(1)=$%.2f", Y1[demo_idx[4]]),
    sprintf("Y(0)=$%.2f  Y(1)=???", Y0[demo_idx[5]])
  )
) |> knitr::kable(caption="The fundamental problem: exactly ONE potential outcome per person is observed.")
```

## Individual Treatment Effects (ITE) {#sec-ite}

```{r ite-distribution, fig.height=4.5, fig.width=9}
#| code-summary: "▶ Plot the distribution of individual treatment effects"
tibble(ITE=ITE,
       env_grp=cut(env, c(-Inf,-0.5,0.5,Inf), labels=c("Low","Moderate","High"))) |>
  ggplot(aes(x=ITE, fill=env_grp)) +
  geom_histogram(bins=50, alpha=0.75, position="identity") +
  geom_vline(xintercept=ATE_true, linetype="dashed", linewidth=1.1, colour="grey20") +
  annotate("text", x=ATE_true+0.06, y=82,
           label=sprintf("ATE = $%.2f", ATE_true), hjust=0, size=3.5, fontface="bold") +
  scale_fill_manual(values=c("High"=clr_eco,"Moderate"="#74b9a0","Low"=clr_ctrl)) +
  labs(x="Individual Treatment Effect  Y(1)−Y(0)  ($)", y="Count", fill=NULL,
       title="The eco-label effect is wildly different across individuals",
       subtitle="In real research you never see this distribution — only the ATE is estimable") +
  theme_mod3()
```

## Average Treatment Effect (ATE) {#sec-ate}

$$\text{ATE} = E[\tau_i] = E[Y_i(1) - Y_i(0)]$$

In a **randomised experiment**, random assignment makes the treated and control groups exchangeable, so a simple difference in group means is unbiased for the ATE.

```{r ate-rct}
treat_rct <- rbinom(N, 1, 0.5)
Y_obs_rct <- ifelse(treat_rct==1, Y1, Y0)
df_rct    <- tibble(id=1:N, treat=treat_rct, Y_obs=Y_obs_rct, Y0, Y1, ITE, env, price_sens, income)

ATE_est <- mean(Y_obs_rct[treat_rct==1]) - mean(Y_obs_rct[treat_rct==0])
lm(Y_obs ~ treat, data=df_rct) |> tidy() |>
  filter(term=="treat") |>
  transmute(Estimand="ATE", `True value`=round(ATE_true,3), Estimate=round(estimate,3),
            SE=round(std.error,3),
            `95% CI`=sprintf("[%.3f, %.3f]", estimate-1.96*std.error, estimate+1.96*std.error)) |>
  knitr::kable(caption="ATE estimate from randomised eco-label experiment")
```

## Conditional Average Treatment Effect (CATE) {#sec-cate}

```{r cate-by-subgroup, fig.height=4.5, fig.width=9}
df_rct |>
  mutate(env_grp=cut(env, c(-Inf,-1,0,1,Inf), labels=c("Very Low","Low","High","Very High"))) |>
  group_by(env_grp) |>
  summarise(CATE_true=mean(Y1-Y0), CATE_est=mean(Y_obs[treat==1])-mean(Y_obs[treat==0]),
            n=n(), .groups="drop") |>
  pivot_longer(starts_with("CATE"), names_to="type", values_to="val") |>
  ggplot(aes(x=env_grp, y=val, fill=type)) +
  geom_col(position="dodge", alpha=0.85, width=0.65) +
  geom_hline(yintercept=ATE_true, linetype="dashed", colour="grey30") +
  annotate("text", x=0.65, y=ATE_true+0.05, label="Overall ATE", size=3.5) +
  scale_fill_manual(values=c(CATE_true=clr_eco, CATE_est=clr_ctrl),
                    labels=c("True CATE","Estimated CATE")) +
  labs(x="Environmental Concern", y="Eco-label effect ($)", fill=NULL,
       title="CATE: the label works far better for environmentally engaged consumers",
       subtitle="Reporting only the ATE hides this story completely") +
  theme_mod3()
```

## Compliance, ITT, LATE, ATT, and ATC {#sec-late}

```{r compliance-dgs}
set.seed(2024)
Z <- treat_rct
p_comply_if_Z1 <- plogis(0.8 + 0.6*env)
p_always_taker <- plogis(-2.5 + 0.5*env)
D_if_Z1 <- rbinom(N, 1, p_comply_if_Z1)
D_if_Z0 <- rbinom(N, 1, p_always_taker)
D       <- ifelse(Z==1, D_if_Z1, D_if_Z0)
compliance_type <- case_when(
  D_if_Z1==1 & D_if_Z0==0 ~ "Complier",
  D_if_Z1==1 & D_if_Z0==1 ~ "Always-taker",
  D_if_Z1==0 & D_if_Z0==0 ~ "Never-taker",
  D_if_Z1==0 & D_if_Z0==1 ~ "Defier"
)
knitr::kable(table(compliance_type), col.names=c("Compliance type","Count"),
             caption="True compliance breakdown (observable only in simulation)")
```

```{r itt-late-att-atc}
Y_obs_iv <- ifelse(D==1, Y1, Y0)
df_iv    <- tibble(Z, D, Y_obs_iv, Y0, Y1, ITE, compliance_type, env, price_sens)

ITT       <- mean(Y_obs_iv[Z==1]) - mean(Y_obs_iv[Z==0])
FS        <- mean(D[Z==1]) - mean(D[Z==0])
LATE_est  <- ITT / FS
LATE_true <- mean(ITE[compliance_type=="Complier"])
ATT_est   <- mean(ITE[D==1])
ATC_est   <- mean(ITE[D==0])

tibble(
  Estimand    = c("ATE","ITT","LATE (Wald)","ATT","ATC"),
  Description = c(
    "Effect averaged over all 2,000 consumers",
    "Effect of *assigning* the label regardless of compliance",
    "Effect for consumers who complied *because* they were assigned",
    "Effect for consumers who *actually saw* the label",
    "Effect for those who *did not see* it"
  ),
  `True value` = round(c(ATE_true, ITT, LATE_true, ATT_est, ATC_est), 3),
  Estimated    = round(c(ATE_true, ITT, LATE_est,  ATT_est, ATC_est), 3)
) |> knitr::kable(caption="Five ways to summarise 'the effect' of the eco-label")
```

```{r estimands-plot, fig.height=4, fig.width=8, echo=FALSE}
tibble(
  Estimand=c("ATE\n(all consumers)","ITT\n(assigned, any)","LATE\n(compliers only)",
             "ATT\n(label seen)","ATC\n(label not seen)"),
  Value=c(ATE_true, ITT, LATE_est, ATT_est, ATC_est),
  Type=c("Population","Assignment","Compliers","Treated","Control")
) |>
  mutate(Estimand=fct_reorder(Estimand, Value)) |>
  ggplot(aes(y=Estimand, x=Value, fill=Type)) +
  geom_col(width=0.6, alpha=0.85) +
  geom_vline(xintercept=ATE_true, linetype="dashed", colour="grey30") +
  scale_fill_brewer(palette="Set2") +
  labs(x="Eco-label effect on WTP ($)", y=NULL, fill=NULL,
       title="Five estimands — same experiment, different questions",
       subtitle="ATT > ATE > ATC: consumers who seek eco-labels respond more strongly to them") +
  theme_mod3()
```

::: {.callout-warning}
## Connection to Module 2
Module 2's permutation test estimated the **ITT** — comparing all participants assigned to the eco-label condition against all assigned to control, regardless of whether they actually processed the label. In a well-controlled lab study, ITT ≈ ATE. In field settings — where consumers scroll past labels or ignore shelf information — the distinction matters greatly.
:::

---

# Part 2: Selection on Observables {#sec-m3-part2}

<iframe src="https://liveeur-my.sharepoint.com/personal/58542dsc_eur_nl/_layouts/15/Doc.aspx?sourcedoc={c4e4767b-beb7-4702-82bf-5717967f663e}&action=embedview&wdAr=1.7777777777777777" width="100%" height="500px" frameborder="0" allowfullscreen title="Module 3 Part 2: Selection on Observables"></iframe>

## The Problem: Consumers Choose to Buy Eco-Labelled Products {#sec-obs-problem}

Outside the lab, consumers freely *choose* whether to pick up the eco-labelled bag of coffee. This creates **selection bias**. The DGP below adds two realistic complications beyond Part 1:

1. **Non-linear confounding**: environmental concern has a *quadratic* effect on baseline WTP and *interacts* with income. A consumer with extremely high env-concern sees a doubly large WTP boost. Simple linear regression will miss this.
2. **Interaction in the propensity score**: the probability of buying eco-labelled coffee depends on the *joint* level of env-concern and income, not just their individual levels.

These features mean that naive and simple-linear corrections will have remaining bias, while flexible and doubly-robust methods will do better.

```{r obs-dgs}
#| code-summary: "▶ Simulate observational data with non-linear confounding"
set.seed(2024)
N <- 2000

env        <- rnorm(N, 0, 1)
price_sens <- rnorm(N, 0, 1)
income     <- rnorm(N, 0, 1)
age        <- round(rnorm(N, 38, 12))

# Non-linear outcome model: env has quadratic effect + env:income interaction
# Linear regression misses these → remaining bias after covariate adjustment
Y0 <- pmax(1, pmin(10,
  5.20 + 0.80*env - 0.50*price_sens + 0.30*income +
  0.25*env^2 + 0.18*env*income +
  rnorm(N, 0, 0.85)
))

ITE <- 0.55 + 0.65*env - 0.38*price_sens + rnorm(N, 0, 0.35)
Y1  <- pmax(1, pmin(10, Y0 + ITE))

# True PS includes env:income interaction
# Standard logit (no interaction) will slightly misspecify the PS
ps_true   <- plogis(-0.40 + 0.95*env + 0.42*income - 0.32*price_sens + 0.14*env*income)
treat_obs <- rbinom(N, 1, ps_true)
Y_obs     <- ifelse(treat_obs==1, Y1, Y0)

# Pre-compute nonlinear features for use in flexible models
env_sq      <- env^2
env_income  <- env * income

df_obs <- tibble(id=1:N, treat=treat_obs, Y_obs, Y0, Y1, ITE,
                 env, price_sens, income, age, env_sq, env_income, ps_true)

ATE_true_obs <- mean(Y1 - Y0)
ATT_true_obs <- mean(ITE[treat_obs==1])
cat(sprintf("True ATE = %.3f  |  True ATT = %.3f  |  Treatment rate = %.1f%%\n",
            ATE_true_obs, ATT_true_obs, 100*mean(treat_obs)))
```

```{r naive-bias, fig.height=4.5, fig.width=9}
#| code-summary: "▶ Show why naive comparison overestimates the eco-label effect"
naive_diff <- mean(Y_obs[treat_obs==1]) - mean(Y_obs[treat_obs==0])
cat(sprintf("True ATE             = %.3f\nNaive diff-in-means  = %.3f\nBias                 = +%.3f\n",
            ATE_true_obs, naive_diff, naive_diff - ATE_true_obs))

bind_rows(
  tibble(group="Eco-label buyers",     env=env[treat_obs==1]),
  tibble(group="Non-eco-label buyers", env=env[treat_obs==0])
) |>
  ggplot(aes(x=env, fill=group)) +
  geom_density(alpha=0.55) +
  scale_fill_manual(values=c("Eco-label buyers"=clr_eco,"Non-eco-label buyers"=clr_ctrl)) +
  labs(x="Environmental Concern (standardised)", y="Density",
       title="Eco-label buyers have higher environmental concern AND higher baseline WTP",
       subtitle="Positive selection on a cause of the outcome inflates the naive comparison",
       fill=NULL) +
  theme_mod3()
```

## Regression Adjustment {#sec-reg-adj}

**What it does:** Controls for confounders in the outcome regression, comparing eco-label buyers and non-buyers who are *similar on measured covariates*.

**Intuition:** Among consumers with the same env-concern, income, and price-sensitivity, any remaining WTP difference is more plausibly causal. But if the true relationship between covariates and WTP is *non-linear*, a linear regression will not fully remove the confounding.

```{r regression-adjustment}
#| code-summary: "▶ Linear vs. flexible regression adjustment"
m_naive   <- lm(Y_obs ~ treat, data=df_obs)
m_linear  <- lm(Y_obs ~ treat + env + price_sens + income, data=df_obs)
m_flex    <- lm(Y_obs ~ treat + env + price_sens + income + env_sq + env_income, data=df_obs)

bind_rows(
  tidy(m_naive)  |> mutate(model="1. Unadjusted"),
  tidy(m_linear) |> mutate(model="2. Linear adjustment"),
  tidy(m_flex)   |> mutate(model="3. Flexible adjustment\n(+ env² + env×income)")
) |>
  filter(term=="treat") |>
  transmute(Model=model, Estimate=round(estimate,3), SE=round(std.error,3),
            `95% CI`=sprintf("[%.3f, %.3f]",estimate-1.96*std.error,estimate+1.96*std.error)) |>
  add_row(Model="True ATE", Estimate=round(ATE_true_obs,3), SE=NA, `95% CI`="—") |>
  knitr::kable(caption="Flexible regression (with non-linear terms) gets closer to the true ATE than linear")
```

::: {.callout-tip}
The flexible model including `env²` and `env × income` tracks the true ATE noticeably better than the linear adjustment. The lesson: **the form of your covariate adjustment matters**. If you don't know the true functional form (you never do in practice), doubly robust estimation provides insurance.
:::

## Inverse Probability Weighting (IPW) {#sec-ipw}

**What it does:** Reweights each observation by the inverse of its probability of being in its actual group, creating a pseudo-population where eco-label buyers and non-buyers look like random draws from the same population.

**Concrete weights:**
- An eco-conscious, high-income consumer (PS = 0.85) who buys eco-labelled coffee: weight = $1/0.85 = 1.18$ — not surprising, downweighted.
- A price-sensitive, low-income consumer (PS = 0.08) who buys eco-labelled coffee: weight = $1/0.08 = 12.5$ — very surprising, heavily upweighted.

**Entropy balancing** (via `WeightIt`) takes this further: instead of estimating a parametric PS model, it directly finds weights that exactly balance the *moments* (means, variances, and cross-products) of the covariate distribution between groups. No model to misspecify.

```{r ipw}
#| code-summary: "▶ Standard IPW vs. entropy-balanced IPW"
# Standard IPW: logistic regression on linear terms (misses interaction)
ps_logit <- glm(treat ~ env + price_sens + income, data=df_obs, family=binomial)
ps_hat   <- fitted(ps_logit)

df_obs <- df_obs |>
  mutate(ps_hat  = ps_hat,
         ipw_wt  = ifelse(treat==1, 1/ps_hat, 1/(1-ps_hat)),
         # Stabilised weights (reduces variance from extreme PS values)
         ipw_stab= ifelse(treat==1, mean(treat)/ps_hat, mean(1-treat)/(1-ps_hat)))

ATE_ipw_raw   <- with(df_obs,
  weighted.mean(Y_obs[treat==1], ipw_wt[treat==1]) -
  weighted.mean(Y_obs[treat==0], ipw_wt[treat==0]))
ATE_ipw_stab  <- lm(Y_obs ~ treat, data=df_obs, weights=ipw_stab) |>
  tidy() |> filter(term=="treat") |> pull(estimate)

# Entropy balancing: directly balance means + variances + covariances
# Includes quadratic and interaction terms for thorough moment matching
wb_ebal <- weightit(treat ~ env + price_sens + income + env_sq + env_income,
                    data=df_obs, method="ebal", estimand="ATE")
ATE_ebal <- lm(Y_obs ~ treat, data=df_obs, weights=wb_ebal$weights) |>
  tidy() |> filter(term=="treat") |> pull(estimate)

cat(sprintf(
  "True ATE              = %.3f\nIPW (standard logit)  = %.3f\nIPW (stabilised)      = %.3f\nEntropy balancing     = %.3f\n",
  ATE_true_obs, ATE_ipw_raw, ATE_ipw_stab, ATE_ebal
))
```

```{r ps-overlap, fig.height=4.5, fig.width=9}
#| code-summary: "▶ Propensity score overlap and entropy-balanced weight distribution"
p_overlap <- df_obs |>
  mutate(group=ifelse(treat==1,"Eco-label buyers","Non-buyers")) |>
  ggplot(aes(x=ps_hat, fill=group)) +
  geom_histogram(bins=40, alpha=0.65, position="identity") +
  geom_vline(xintercept=c(0.05,0.95), linetype="dashed", colour="firebrick") +
  scale_fill_manual(values=c("Eco-label buyers"=clr_eco,"Non-buyers"=clr_ctrl)) +
  labs(x="Estimated PS", y="Count", title="PS overlap", fill=NULL,
       subtitle="Red lines = extreme weights zone") +
  theme_mod3()

p_wts <- tibble(wt=wb_ebal$weights, group=ifelse(treat_obs==1,"Treated","Control")) |>
  ggplot(aes(x=wt, fill=group)) +
  geom_histogram(bins=40, alpha=0.7, position="identity") +
  scale_fill_manual(values=c("Treated"=clr_eco,"Control"=clr_ctrl)) +
  labs(x="Entropy-balanced weight", y="Count",
       title="Entropy-balanced weights",
       subtitle="Well-behaved: no extreme spikes", fill=NULL) +
  theme_mod3()

p_overlap + p_wts
```

::: {.callout-warning}
## IPW can be unstable with extreme propensity scores

Unstabilised IPW weights become enormous near PS = 0 or 1. Entropy balancing avoids this by finding the *minimum-variance* weights that achieve balance — it never requires extreme weights to correct for outlier PS values.
:::

## Covariate Matching (Mahalanobis) {#sec-matching}

**What it does:** For each eco-label buyer, finds the most similar non-buyer based on raw covariate values (Mahalanobis distance accounts for covariate correlations). The matched sample mimics a randomised experiment — but only for the *treated* group (estimates ATT, not ATE).

```{r covariate-matching}
#| code-summary: "▶ 1:1 Mahalanobis nearest-neighbour matching"
m_match_maha <- matchit(treat ~ env + price_sens + income, data=df_obs,
                        method="nearest", distance="mahalanobis", ratio=1)
df_matched_maha <- match.data(m_match_maha)
ATT_match_maha  <- lm(Y_obs ~ treat, data=df_matched_maha, weights=weights) |>
  tidy() |> filter(term=="treat") |> pull(estimate)

cat(sprintf("True ATT (treated group)              = %.3f\nMatching (Mahalanobis, 1:1 NN) est.   = %.3f\n",
            ATT_true_obs, ATT_match_maha))
```

## Propensity Score Matching (PSM) {#sec-psm}

**What it does:** Condenses all covariates into a single scalar — the propensity score — and matches on that. The catch: **the PS model must be specified correctly**. Different models give different matched samples and often quite different estimates.

```{r psm-comparison, fig.height=5.5, fig.width=10}
#| code-summary: "▶ Compare five PS model specifications"
# Misspecified models (common practice — miss the env:income interaction)
m_ps1 <- matchit(treat ~ env + price_sens + income,
                 data=df_obs, method="nearest", distance="logit")
m_ps2 <- matchit(treat ~ env * price_sens + income,
                 data=df_obs, method="nearest", distance="probit")
m_ps3 <- matchit(treat ~ poly(env,2) + poly(price_sens,2) + income,
                 data=df_obs, method="nearest", distance="logit")
m_ps4 <- matchit(treat ~ env + price_sens + income,
                 data=df_obs, method="nearest", distance="logit",
                 caliper=0.1, std.caliper=TRUE)
# Correctly specified: matches the true DGP (includes env^2 and env:income)
m_ps5 <- matchit(treat ~ env + price_sens + income + env_sq + env_income,
                 data=df_obs, method="nearest", distance="logit",
                 caliper=0.15, std.caliper=TRUE)

results_psm <- map_dfr(
  list("1. Logit (linear)"              = m_ps1,
       "2. Probit + env×price"          = m_ps2,
       "3. Logit + poly(env,2)"         = m_ps3,
       "4. Logit + caliper (0.1 SD)"    = m_ps4,
       "5. Logit + env²+env×income\n+ caliper (correct PS)" = m_ps5),
  function(m) {
    d   <- match.data(m)
    est <- lm(Y_obs ~ treat, data=d, weights=weights) |>
      tidy() |> filter(term=="treat")
    tibble(estimate=est$estimate, se=est$std.error,
           n_matched=nrow(d[d$treat==1,]))
  }, .id="PS model"
)

results_psm |>
  mutate(lo=estimate-1.96*se, hi=estimate+1.96*se,
         `N retained`=n_matched,
         correct = str_detect(`PS model`, "correct")) |>
  ggplot(aes(y=`PS model`, x=estimate, xmin=lo, xmax=hi,
             colour=correct, shape=correct)) +
  geom_vline(xintercept=ATT_true_obs, linetype="dashed", colour=clr_eco, linewidth=1) +
  geom_vline(xintercept=ATE_true_obs, linetype="dotted", colour=clr_ctrl, linewidth=1) +
  geom_pointrange(size=0.8) +
  geom_text(aes(label=sprintf("N=%d", `N retained`), x=hi+0.03), hjust=0, size=3) +
  annotate("text", x=ATT_true_obs+0.01, y=5.45,
           label="True ATT", hjust=0, size=3.2, colour=clr_eco) +
  annotate("text", x=ATE_true_obs+0.01, y=5.1,
           label="True ATE", hjust=0, size=3.2, colour=clr_ctrl) +
  scale_colour_manual(values=c("FALSE"="grey50","TRUE"="#b45309"),
                      labels=c("Misspecified PS","Correctly specified PS")) +
  scale_shape_manual(values=c("FALSE"=16,"TRUE"=18),
                     labels=c("Misspecified PS","Correctly specified PS")) +
  labs(x="ATT estimate — eco-label effect ($)", y=NULL, colour=NULL, shape=NULL,
       title="PSM estimates converge on the true ATT only when the PS model is correctly specified",
       subtitle="The correctly specified model (orange) recovers the true ATT most accurately") +
  theme_mod3()
```

```{r covariate-balance, fig.height=5, fig.width=8}
#| code-summary: "▶ Love plot: balance before and after correctly-specified PSM"
love.plot(m_ps5, thresholds=c(m=0.1), stars="std",
          title="Covariate balance: correctly specified PSM (env² + env×income + caliper)",
          colors=c(clr_ctrl, clr_eco))
```

## Doubly Robust Estimation {#sec-doubly-robust}

**What it does:** Combines an outcome model *and* a propensity score model. The estimate is consistent if **either** model is correctly specified — it only needs one to be right.

Here both the outcome model and PS model include the non-linear terms (`env²`, `env × income`), so both are correctly specified. The DR estimate should perform best.

```{r doubly-robust}
#| code-summary: "▶ AIPW with correctly-specified outcome and PS models"
# Outcome models (correctly specified: include env^2 and env:income)
mu1_model <- lm(Y_obs ~ env + price_sens + income + env_sq + env_income,
                data=df_obs[df_obs$treat==1,])
mu0_model <- lm(Y_obs ~ env + price_sens + income + env_sq + env_income,
                data=df_obs[df_obs$treat==0,])
mu1_hat   <- predict(mu1_model, newdata=df_obs)
mu0_hat   <- predict(mu0_model, newdata=df_obs)

# PS model (correctly specified)
ps_hat_dr <- fitted(glm(treat ~ env + price_sens + income + env_sq + env_income,
                        data=df_obs, family=binomial))

# AIPW estimator
Y_v <- df_obs$Y_obs; D_v <- df_obs$treat
tau_dr <- (mu1_hat - mu0_hat) +
          D_v*(Y_v - mu1_hat)/ps_hat_dr -
          (1-D_v)*(Y_v - mu0_hat)/(1-ps_hat_dr)

ATE_dr <- mean(tau_dr)
SE_dr  <- sd(tau_dr)/sqrt(N)

cat(sprintf(
  "True ATE              = %.3f\nNaive estimate        = %.3f\nLinear reg. adj.      = %.3f\nFlexible reg. adj.    = %.3f\nIPW (stabilised)      = %.3f\nEntropy balancing     = %.3f\nDR/AIPW (correct)     = %.3f  (SE = %.3f)\n",
  ATE_true_obs, naive_diff,
  coef(m_linear)["treat"], coef(m_flex)["treat"],
  ATE_ipw_stab, ATE_ebal,
  ATE_dr, SE_dr
))
```

## How Well Do All Section 2 Methods Recover the True ATE? {#sec-sec2-comparison}

```{r section2-summary-plot, fig.height=5.5, fig.width=10, echo=FALSE}
#| code-summary: "▶ Comprehensive comparison of all observational estimators"
tibble(
  Method   = c(
    "Naive (no adjustment)",
    "Linear regression\nadjustment",
    "Flexible regression\n(env² + env×income)",
    "IPW — stabilised\n(logit, linear PS)",
    "Entropy balancing\n(moment-matching IPW)",
    "Matching — Mahalanobis\n1:1 NN",
    "PSM — misspecified\n(logit linear)",
    "PSM — correctly\nspecified + caliper",
    "Doubly Robust\nAIPW (correct models)"
  ),
  Estimate = c(
    naive_diff,
    coef(m_linear)["treat"],
    coef(m_flex)["treat"],
    ATE_ipw_stab,
    ATE_ebal,
    ATT_match_maha,
    results_psm$estimate[1],
    results_psm$estimate[5],
    ATE_dr
  ),
  Target  = c("ATE","ATE","ATE","ATE","ATE","ATT","ATT","ATT","ATE"),
  Quality = c("Naive","Basic","Flexible","IPW","Advanced","Matching","PSM","PSM","Advanced")
) |>
  mutate(Method  = fct_inorder(Method),
         Bias    = Estimate - ATE_true_obs) |>
  ggplot(aes(y=Method, x=Estimate, colour=Quality, shape=Target)) +
  geom_vline(xintercept=ATE_true_obs, linetype="dashed", colour=clr_eco,  linewidth=1.1) +
  geom_vline(xintercept=ATT_true_obs, linetype="dotted", colour=clr_ctrl, linewidth=1.1) +
  geom_point(size=4) +
  annotate("text", x=ATE_true_obs+0.01, y=9.45,
           label="True ATE", hjust=0, size=3.3, colour=clr_eco, fontface="bold") +
  annotate("text", x=ATT_true_obs+0.01, y=9.0,
           label="True ATT", hjust=0, size=3.3, colour=clr_ctrl) +
  scale_colour_manual(
    values=c(Naive="#c0392b", Basic="#e67e22", Flexible="#f1c40f",
             IPW="#3498db", Matching="#9b59b6", PSM="#16a085", Advanced="#2d6a4f")) +
  scale_shape_manual(values=c(ATE=16, ATT=18)) +
  labs(x="Estimate ($)", y=NULL, colour="Method class", shape="Target estimand",
       title="Nine estimators compared against the known true ATE and ATT",
       subtitle="Non-linear DGP: methods that model non-linearity or balance moments perform best") +
  theme_mod3()
```

::: {.callout-tip}
## What the gradient shows

| Method class | Why it performs that way |
|---|---|
| **Naive** | No correction — pure selection bias |
| **Linear regression** | Removes linear confounding; missing `env²` and `env×income` leaves residual bias |
| **Flexible regression** | Includes non-linear terms → less bias |
| **IPW (stabilised)** | PS model also misspecified (linear logit) → some remaining bias |
| **Entropy balancing** | Directly balances covariate *moments* — robust to PS misspecification |
| **Matching/PSM** | Targets ATT (higher than ATE because treated have stronger responses) |
| **Doubly Robust** | Both outcome model AND PS model correctly specified → lowest bias |

The **doubly robust** and **entropy balancing** approaches consistently outperform simpler methods when the true confounding is non-linear. The correctly specified PSM recovers the ATT accurately; the misspecified PSM does less well.
:::

## Synthetic Controls {#sec-synth}

What if you only have **one treated unit** — say, AlterEco's flagship Rotterdam store introduced eco-labels?

```{r synthetic-control, fig.height=5, fig.width=9}
#| code-summary: "▶ Synthetic control for Rotterdam flagship store"
set.seed(2024)
n_stores <- 21; n_pre <- 20; n_post <- 10; treat_store <- 1
store_fe <- rnorm(n_stores, 0, 0.8)

panel_df <- expand.grid(store=1:n_stores, period=1:(n_pre+n_post)) |>
  as_tibble() |>
  mutate(treated=(store==treat_store & period>n_pre),
         WTP=5.5 + store_fe[store] + 0.02*(period-n_pre) +
             treated*0.80 + rnorm(n(), 0, 0.40))

pre_means <- panel_df |> filter(period<=n_pre) |>
  group_by(store) |> summarise(pre_mean=mean(WTP), .groups="drop")
treated_pre_mean <- pre_means$pre_mean[pre_means$store==treat_store]

control_stores <- pre_means |> filter(store!=treat_store) |>
  mutate(dist=abs(pre_mean-treated_pre_mean), w_raw=1/(dist+0.01), w=w_raw/sum(w_raw))

synth_ts <- panel_df |> filter(store!=treat_store) |>
  left_join(control_stores[,c("store","w")], by="store") |>
  group_by(period) |> summarise(WTP_synth=weighted.mean(WTP,w), .groups="drop") |>
  mutate(series="Synthetic Rotterdam")

treated_ts <- panel_df |> filter(store==treat_store) |>
  dplyr::select(period, WTP) |> rename(WTP_synth=WTP) |>
  mutate(series="Rotterdam (actual)")

bind_rows(synth_ts, treated_ts) |>
  ggplot(aes(x=period, y=WTP_synth, colour=series, linetype=series)) +
  geom_line(linewidth=1) +
  geom_vline(xintercept=n_pre+0.5, linetype="dashed", colour="grey50") +
  annotate("text", x=n_pre+1, y=5.0, label="Eco-label\nintroduced", hjust=0, size=3.5, colour="grey40") +
  scale_colour_manual(values=c("Rotterdam (actual)"=clr_eco,"Synthetic Rotterdam"=clr_ctrl)) +
  scale_linetype_manual(values=c("Rotterdam (actual)"="solid","Synthetic Rotterdam"="dashed")) +
  labs(x="Period", y="Mean WTP ($)", colour=NULL, linetype=NULL,
       title="Synthetic Control: Rotterdam vs. its synthetic twin built from control stores",
       subtitle="Good pre-period fit validates the synthetic match; gap after dashed line = causal effect") +
  theme_mod3()
```

---

# Part 3: Regression Discontinuity Design {#sec-m3-part3}

## The Core Idea {#sec-rdd-idea}

AlterEco's retailer awards a **"Certified Sustainable"** badge to any coffee product scoring **70 or above** on a third-party environmental audit (scored 0–100). Products at 69 miss the badge; products at 71 get it. Because the audit score has measurement noise, which side of the cutoff a product lands on is *nearly random* near 70. We exploit this near-randomness.

::: {.callout-note}
## Why the jump at 70 is causal — the "as-good-as-random" intuition

Imagine two products with audit scores of 69 and 71. They have almost identical green credentials — similar supply chains, similar materials — but the auditor's ruler is imprecise. Those two points of difference are largely measurement noise. Yet one gets the badge and the other doesn't. **Near the cutoff, badge assignment is as good as a coin flip.**

This local near-randomness is what makes the jump causal. It is equivalent to a mini natural experiment: products just above and just below 70 are exchangeable in every way except for badge receipt. The jump in WTP at that threshold is therefore attributable *entirely* to the badge.

**The cost of this elegance**: the estimate is a *local* average treatment effect (LATE). It tells you the causal effect of receiving the badge for brands hovering around a 70-point score — not for the top-scoring 90s or the low-scoring 40s, which may respond to the badge very differently.

**A practical check to run first**: If brands can game the audit score to land just above 70, the "as-good-as-random" assumption breaks down. The density test below checks for suspicious bunching right at the cutoff.
:::

## Simulating an RDD {#sec-rdd-sim}

```{r rdd-dgp, fig.height=4.5, fig.width=9}
#| code-summary: "▶ Simulate 1,200 products near the 70-point certification threshold"
set.seed(2024)
N_rdd <- 1200; cutoff <- 70; true_jump <- 0.80
audit_score <- runif(N_rdd, 30, 100)
treat_rdd   <- as.integer(audit_score >= cutoff)

WTP_rdd <- 4.0 + 0.02*(audit_score-cutoff) - 0.0003*(audit_score-cutoff)^2 +
           true_jump*treat_rdd + rnorm(N_rdd, 0, 0.60)
WTP_rdd <- pmax(1, pmin(10, WTP_rdd))
df_rdd  <- tibble(audit_score, treat_rdd, WTP_rdd, running=audit_score-cutoff)

df_rdd |>
  ggplot(aes(x=audit_score, y=WTP_rdd, colour=factor(treat_rdd))) +
  geom_point(alpha=0.30, size=0.9) +
  geom_vline(xintercept=cutoff, linewidth=1, colour="grey30") +
  geom_smooth(method="lm", formula=y~poly(x,2), se=TRUE, alpha=0.2) +
  scale_colour_manual(values=c("0"=clr_ctrl,"1"=clr_eco),
                      labels=c("Below 70 (no badge)","Above 70 (badge)")) +
  annotate("text", x=71, y=9.2, label="Cutoff = 70", hjust=0, size=3.5, fontface="bold") +
  annotate("segment", x=70, xend=70, y=3.85, yend=4.85, colour="firebrick",
           arrow=arrow(ends="both", length=unit(.2,"cm")), linewidth=0.8) +
  annotate("text", x=71, y=4.35,
           label=sprintf("~$%.2f jump\n(badge effect)", true_jump),
           hjust=0, size=3, colour="firebrick") +
  labs(x="Audit Score (running variable)", y="Consumer WTP ($)", colour=NULL,
       title="Regression Discontinuity: the jump at 70 estimates the badge's causal effect") +
  theme_mod3()
```

## Key Assumptions and How to Test Them {#sec-rdd-assumptions}

### Assumption 1: No Manipulation

```{r rdd-density-test, fig.height=4, fig.width=7}
dens_test <- rddensity(X=df_rdd$audit_score, c=cutoff)
summary(dens_test)
rdplotdensity(dens_test, X=df_rdd$audit_score,
              title="Density test: checking for bunching at the 70-point cutoff",
              xlabel="Audit score", ylabel="Density")
```

A significant p-value signals that brands are gaming audit scores to land just above 70. With our simulated uniform running variable, the density is continuous.

### Assumption 2: Covariate Continuity

```{r rdd-covariate-balance}
env_rdd <- rnorm(N_rdd, 0 + 0.01*(audit_score-cutoff)/10, 1)
rdd_env <- rdrobust(y=env_rdd, x=df_rdd$audit_score, c=cutoff)
cat("Covariate balance — env. concern should NOT jump at the threshold:\n")
summary(rdd_env)
```

## Estimating the RDD Effect {#sec-rdd-estimate}

```{r rdd-estimate, fig.height=4.5, fig.width=8}
rdd_est <- rdrobust(y=df_rdd$WTP_rdd, x=df_rdd$audit_score, c=cutoff)
summary(rdd_est)
rdplot(y=df_rdd$WTP_rdd, x=df_rdd$audit_score, c=cutoff,
       title="RDD estimate: causal effect of sustainability badge on WTP",
       x.label="Audit Score", y.label="WTP ($)")
```

```{r rdd-bandwidth-sensitivity, fig.height=4, fig.width=8}
bw_results <- map_dfr(c(5,8,10,15,20,25,30), function(bw) {
  est <- rdrobust(y=df_rdd$WTP_rdd, x=df_rdd$audit_score, c=cutoff, h=bw)
  tibble(bandwidth=bw, estimate=est$coef["Conventional",1], se=est$se["Robust",1])
})

bw_results |>
  mutate(lo=estimate-1.96*se, hi=estimate+1.96*se) |>
  ggplot(aes(x=bandwidth, y=estimate, ymin=lo, ymax=hi)) +
  geom_hline(yintercept=true_jump, linetype="dashed", colour="grey40", linewidth=1) +
  geom_ribbon(alpha=0.15, fill=clr_eco) +
  geom_line(colour=clr_eco, linewidth=1) + geom_point(colour=clr_eco, size=2.5) +
  annotate("text", x=5.5, y=true_jump+0.05, label=sprintf("True = $%.2f", true_jump),
           hjust=0, size=3.5) +
  labs(x="Bandwidth (audit score units)", y="Estimated badge effect ($)",
       title="RDD estimates stable across bandwidths h = 5 to 30",
       subtitle="Instability would indicate sensitivity to bandwidth — a red flag") +
  theme_mod3()
```

::: {.callout-note}
## The bandwidth trade-off: bias vs. variance

The bandwidth $h$ controls which products are included in the RDD comparison — only those with audit scores in $[70-h,\ 70+h]$. This creates a fundamental tension:

| Narrow $h$ | Wide $h$ |
|---|---|
| Products are very similar to each other near the cutoff ✔ | More observations → tighter CI ✔ |
| Few observations → wide CI ✖ | Products far from 70 differ systematically from each other ✖ |

**`rdrobust` selects the bandwidth automatically** using a mean-squared-error optimal rule — it finds the $h$ that best balances bias and variance. The bandwidth-sensitivity plot above is a sanity check: stable estimates across a wide range of $h$ suggest the result is not an artefact of a specific choice.

When estimates collapse or swing wildly as $h$ changes, it usually means the polynomial fit on one side of the cutoff is extrapolating too aggressively. That's a red flag that warrants narrowing $h$ or using a local linear fit.
:::

---

# Part 4: Difference-in-Differences {#sec-m3-part4}

## The Setup {#sec-did-setup}

The Dutch government introduces a **mandatory eco-labelling law** in **2022**. Belgian supermarkets are not subject to it. We observe mean WTP in 15 Dutch stores (treated) and 20 Belgian stores (control).

## Simulating Panel Data {#sec-did-sim}

```{r did-dgp, fig.height=4.5, fig.width=9}
set.seed(2024)
n_dutch <- 15; n_belgian <- 20; n_stores_did <- n_dutch+n_belgian
years <- 2018:2024; T_treat <- 2022

panel_did <- expand.grid(store=1:n_stores_did, year=years) |>
  as_tibble() |>
  mutate(
    country      = ifelse(store<=n_dutch,"Netherlands","Belgium"),
    treated      = (country=="Netherlands"),
    post         = (year>=T_treat),
    did_indicator= treated & post,
    store_fe     = rnorm(n_stores_did, 0, 0.5)[store],
    time_trend   = 0.10*(year-2018),
    country_fe   = ifelse(country=="Netherlands", 0.30, 0),
    treat_effect = did_indicator * 0.70,
    WTP_did      = 5.20 + store_fe + time_trend + country_fe + treat_effect + rnorm(n(), 0, 0.35)
  )

panel_did |>
  group_by(country, year) |>
  summarise(mean_WTP=mean(WTP_did), .groups="drop") |>
  ggplot(aes(x=year, y=mean_WTP, colour=country, group=country)) +
  geom_vline(xintercept=T_treat-0.5, linetype="dashed", colour="grey50") +
  geom_line(linewidth=1.2) + geom_point(size=2.5) +
  annotate("text", x=T_treat+0.1, y=5.1,
           label="Dutch eco-label\nlaw takes effect", hjust=0, size=3.2) +
  scale_colour_manual(values=c("Netherlands"=clr_eco,"Belgium"=clr_ctrl)) +
  labs(x="Year", y="Mean WTP ($)", colour=NULL,
       title="DiD: parallel pre-trends are the key identifying assumption",
       subtitle="Dutch and Belgian stores track each other in slope before 2022; Dutch stores jump after") +
  theme_mod3()
```

::: {.callout-note}
## What "parallel trends" actually means — and why Belgium is the key

DiD hinges on one unanswerable question: *what would Dutch stores have done after 2022 if the eco-label law had never passed?* We can't observe this counterfactual. **The parallel trends assumption supplies the answer**: Belgian stores tell us.

More precisely, the assumption states that **absent treatment, the Dutch–Belgian WTP gap would have remained constant** — both series drifting up or down by identical amounts each year. Under this assumption:

$$\underbrace{\Delta Y_{\text{Dutch}}}_{\text{observed}} - \underbrace{\Delta Y_{\text{Belgian}}}_{\text{counterfactual proxy}} = \delta_{\text{DiD}}$$

**Reading the plot above:** The two lines run nearly parallel from 2018–2021 (same slope, constant gap). After 2022 the Dutch line separates from the Belgian. DiD attributes that separation — roughly \$0.70 — to the eco-label law, *after* subtracting whatever movement Belgium showed over the same period (which would have happened anyway).

**What makes this assumption fail?** Anything that makes Dutch and Belgian stores diverge *for reasons unrelated to the law* — e.g., Dutch shoppers becoming greener faster, or a Dutch-only economic shock. The parallel trends violation section below shows exactly this scenario.
:::

## Two-Way Fixed Effects (TWFE) {#sec-did-twfe}

$$Y_{it} = \alpha_i + \lambda_t + \delta \cdot D_{it} + \varepsilon_{it}$$

```{r did-twfe}
did_twfe <- lm_robust(WTP_did ~ did_indicator + factor(store) + factor(year),
                      data=panel_did, clusters=store)
did_coef <- tidy(did_twfe) |> filter(term=="did_indicatorTRUE")
cat(sprintf(
  "True DiD effect  = 0.70\nEstimated DiD    = %.3f  (SE = %.3f, p = %.4f)\n95%% CI: [%.3f, %.3f]\n",
  did_coef$estimate, did_coef$std.error, did_coef$p.value,
  did_coef$conf.low, did_coef$conf.high
))
```

::: {.callout-note}
## What the two sets of fixed effects absorb

$$Y_{it} = \underbrace{\alpha_i}_{\text{store FE}} + \underbrace{\lambda_t}_{\text{year FE}} + \delta \cdot D_{it} + \varepsilon_{it}$$

Each type of fixed effect removes a specific source of confounding:

| Term | What it controls for | Example here |
|---|---|---|
| $\alpha_i$ — store fixed effects | Time-invariant differences between stores | Store size, neighbourhood demographics, pre-existing customer base |
| $\lambda_t$ — year fixed effects | Year-level shocks that hit *all* stores equally | Euro-area inflation, global supply chain costs, macro consumer sentiment |
| $\delta$ — the DiD estimator | How much *more* the treated group changed *after* treatment vs. the control group | Causal effect of the Dutch eco-label law |

**The critical residual $\varepsilon_{it}$**: whatever the fixed effects can't absorb — differential trends (treated and control groups moving at different rates), time-varying store-specific shocks — ends up here. This is exactly where parallel-trends violations hide. Adding pre-treatment event-study coefficients is a way to interrogate whether $\varepsilon_{it}$ contains systematic pre-treatment divergence.
:::

## Testing Parallel Trends: Event Study {#sec-did-event}

```{r did-event-study, fig.height=4.5, fig.width=9}
panel_did <- panel_did |>
  mutate(year_rel=year-T_treat,
         year_rel_fac=relevel(factor(year_rel), ref="-1"))

event_formula <- as.formula("WTP_did ~ treated:year_rel_fac + factor(store) + factor(year)")
did_event     <- lm_robust(event_formula, data=panel_did, clusters=store)

event_tbl <- tidy(did_event) |>
  filter(str_detect(term,"year_rel_fac")) |>
  mutate(yr=as.integer(str_extract(term,"-?\\d+$")),
         period=if_else(yr<0,"Pre-treatment","Post-treatment")) |>
  add_row(yr=-1L, estimate=0, conf.low=0, conf.high=0, period="Reference (−1)")

event_tbl |>
  ggplot(aes(x=yr, y=estimate, ymin=conf.low, ymax=conf.high, colour=period)) +
  geom_hline(yintercept=0, linetype="dashed", colour="grey50") +
  geom_vline(xintercept=-0.5, linetype="dotted", colour="grey60") +
  geom_pointrange(size=0.7) +
  scale_colour_manual(values=c("Pre-treatment"=clr_ctrl,"Post-treatment"=clr_eco,
                               "Reference (−1)"="grey40")) +
  labs(x="Years relative to Dutch eco-label law (0 = 2022)", y="DiD coefficient ($)",
       colour=NULL,
       title="Event study: pre-treatment coefficients near zero — parallel trends supported",
       subtitle="Post-treatment estimates rise consistently — evidence of a sustained law effect") +
  theme_mod3()
```

::: {.callout-note}
## How to read an event study plot

Each point is the DiD coefficient for that relative year — how much the Dutch–Belgian gap in that year differed from the gap in the reference year (here: year $-1$, the year just before treatment).

**Pre-treatment years (left of the dotted line)**: If parallel trends holds, these should be near zero with overlapping CIs. A systematic upward or downward trend here is evidence of a *pre-existing differential trend* — the most direct threat to identification.

**Post-treatment years (right of the dotted line)**: These estimate the treatment effect over time. A sustained positive trajectory (as here) suggests a genuine, growing effect of the law rather than a one-off jump.

**What to worry about**: (1) Any pre-period estimate far from zero and statistically significant; (2) CIs so wide they can't rule out large violations — this happens when you have very few pre-periods (see the next section).
:::

## Synthetic Difference-in-Differences {#sec-synthdid}

```{r synthdid, fig.height=4, fig.width=8}
sdid_wide <- panel_did |>
  dplyr::select(store, year, WTP_did, treated) |>
  pivot_wider(names_from=year, values_from=WTP_did) |>
  arrange(treated)

N0 <- sum(!sdid_wide$treated); T0 <- sum(years < T_treat)
Y_matrix <- sdid_wide |> dplyr::select(-store,-treated) |> as.matrix()

sdid_est <- synthdid_estimate(Y_matrix, N0, T0)
sc_est   <- sc_estimate(Y_matrix, N0, T0)
did_est  <- did_estimate(Y_matrix, N0, T0)

tibble(
  Method       = c("Standard DiD","Synthetic Control","Synthetic DiD"),
  Estimate     = round(c(as.numeric(did_est),as.numeric(sc_est),as.numeric(sdid_est)), 3),
  `True effect`= 0.70,
  Bias         = round(c(as.numeric(did_est),as.numeric(sc_est),as.numeric(sdid_est))-0.70, 3)
) |> knitr::kable(caption="All three estimators vs. the true effect of 0.70")
```

```{r synthdid-plot, fig.height=5, fig.width=8}
plot(sdid_est, se.method="placebo")
```

## When Parallel Trends Fails — and Pre-Testing Cannot Always Save You {#sec-pt-failure}

Parallel trends is **not directly testable** — you never observe what Dutch stores would have done without the eco-label law. What you *can* test is whether pre-treatment trajectories were parallel. The event study above is exactly this test. But two practical problems cripple it in most real datasets:

1. **Too few pre-treatment periods.** Most DiD studies in marketing and management observe units for only 2–4 years before treatment. With so few periods, the event study has very low statistical power to detect violations.
2. **You collect whatever data exists.** Pre-treatment archives are often limited — the number of pre-periods is determined by data availability, not statistical power considerations.

::: {.callout-important}
## What the pre-trend test does — and does not — establish

The event study tests whether *observed pre-treatment outcomes* moved in parallel. It does **not** test:

- Whether unmeasured confounders were evolving differently for treated and control units
- Whether the differential dynamic would have continued into the post-period absent treatment
- Whether a violation is simply too small to detect given the available pre-periods

Passing the pre-trend test is necessary but far from sufficient for causal identification.
:::

### A Realistic Parallel Trends Violation {#sec-pt-violation-viz}

Suppose Dutch stores were *already* attracting more environmentally conscious shoppers whose WTP grew \$0.15/year faster than Belgian shoppers — not because of any law, but because of who was already shopping there (selection). **The true treatment effect is zero.** The eco-label law is irrelevant. But this ongoing differential growth will trick DiD into finding one.

```{r pt-violation-dgp, fig.height=6.5, fig.width=10}
#| code-summary: "▶ Same violation: unmistakeable with 8 pre-periods, invisible with 2"
set.seed(2025)
n_dutch_v  <- 15; n_belgian_v <- 20; n_stores_v <- n_dutch_v + n_belgian_v
T_treat_v  <- 2022; n_post_v <- 3
diff_slope_v <- 0.15   # Dutch stores grow $0.15/yr faster throughout (selection, not treatment)
store_fes_v  <- rnorm(n_stores_v, 0, 0.6)

make_violation_panel <- function(yrs, true_eff = 0) {
  expand.grid(store = 1:n_stores_v, year = yrs) |> as_tibble() |>
    mutate(
      treated       = as.numeric(store <= n_dutch_v),
      country       = ifelse(treated == 1, "Netherlands", "Belgium"),
      post          = as.numeric(year >= T_treat_v),
      did_indicator = treated * post,
      store_fe      = store_fes_v[store],
      diff_trend    = treated * diff_slope_v * (year - T_treat_v),
      WTP           = 5.20 + store_fe + 0.08 * (year - T_treat_v) +
                      diff_trend + true_eff * did_indicator + rnorm(n(), 0, 0.40)
    )
}

panel_many_v <- make_violation_panel(2014:2024)   # 8 pre-periods
panel_few_v  <- make_violation_panel(2020:2024)   # 2 pre-periods

plot_pt_panel <- function(panel, subtitle) {
  panel |> group_by(country, year) |>
    summarise(WTP = mean(WTP), .groups = "drop") |>
    ggplot(aes(x = year, y = WTP, colour = country, group = country)) +
    geom_vline(xintercept = T_treat_v - 0.5, linetype = "dashed",
               colour = "grey50", linewidth = 0.9) +
    geom_line(linewidth = 1.2) + geom_point(size = 2.5) +
    scale_colour_manual(values = c("Netherlands" = clr_eco, "Belgium" = clr_ctrl)) +
    scale_x_continuous(breaks = unique(panel$year)) +
    labs(x = NULL, y = "Mean WTP ($)", colour = NULL, subtitle = subtitle) +
    theme_mod3()
}

p_many_v <- plot_pt_panel(panel_many_v,
  "8 pre-periods: the differential trend is unmistakeable — you would never trust DiD here")
p_few_v  <- plot_pt_panel(panel_few_v,
  "2 pre-periods: SAME DGP — the violation is invisible and DiD proceeds unchallenged")

(p_many_v / p_few_v) +
  plot_annotation(
    title    = "Parallel trends violation: Dutch WTP grows $0.15/yr faster than Belgian (true effect = $0)",
    subtitle = "How many pre-periods you happen to collect determines whether you can even see the problem",
    theme    = theme(plot.title    = element_text(size = 13, face = "bold"),
                     plot.subtitle = element_text(size = 11))
  )
```

With 8 pre-periods the violation is glaringly obvious — you would never run DiD here. With 2 pre-periods the same violation is invisible. Yet the DiD estimate is equally spurious in both cases:

```{r pt-violation-did}
#| code-summary: "▶ DiD on 2-pre-period data: statistically significant, entirely spurious"
did_viol <- lm_robust(WTP ~ did_indicator + factor(store) + factor(year),
                      data = panel_few_v, clusters = store)
dv <- tidy(did_viol) |> dplyr::filter(term == "did_indicator")
cat(sprintf(
  "True treatment effect  = $0.000\n
DiD estimate           = $%.3f  (SE = %.3f,  p = %.4f)\n95%% CI: [$%.3f, $%.3f]\n\nThis looks like a meaningful, significant eco-label effect.\nIt is entirely spurious. The violation was undetectable with 2 pre-periods.\n",
  dv$estimate, dv$std.error, dv$p.value, dv$conf.low, dv$conf.high
))
```

### Monte Carlo: Pre-Test Power and Type I Error {#sec-pt-mc}

We now formalise this with 300 simulations per condition. **True effect = \$0 throughout.** We vary the number of pre-treatment periods from 1 to 6 and track three quantities:

| Quantity | What it means |
|---|---|
| **Pre-test power** | How often does the pre-trend test detect the violation? (p < .05) |
| **Overall Type I error** | How often does DiD falsely reject H₀? |
| **Conditional Type I error** | Given the pre-test *passes* (p ≥ .05), how often does DiD still falsely reject? |

The conditional rate is what matters most: it is the false-positive rate faced by researchers who *did their due diligence* with an event study and got a clean-looking pre-trend.

```{r pt-simulation}
#| code-summary: "▶ Monte Carlo: 300 sims × 6 pre-period counts (true effect = 0)"
set.seed(2025)
N_SIM_PT <- 300

pt_sim_one <- function(n_pre) {
  years_s <- c(seq(T_treat_v - n_pre, T_treat_v - 1),
               T_treat_v:(T_treat_v + n_post_v - 1))
  n_yrs <- length(years_s)

  mat <- replicate(N_SIM_PT, {
    n_obs     <- n_stores_v * n_yrs
    store_s   <- rep(1:n_stores_v, each = n_yrs)
    year_s    <- rep(years_s, times = n_stores_v)
    treated_s <- as.numeric(store_s <= n_dutch_v)
    post_s    <- as.numeric(year_s >= T_treat_v)
    did_s     <- treated_s * post_s
    diff_s    <- treated_s * diff_slope_v * (year_s - T_treat_v)
    fe_s      <- rnorm(n_stores_v, 0, 0.6)[store_s]
    WTP_s     <- 5.20 + fe_s + 0.08 * (year_s - T_treat_v) + diff_s + rnorm(n_obs, 0, 0.40)

    df_s <- data.frame(
      WTP     = WTP_s,
      did     = did_s,
      store   = factor(store_s),
      year    = factor(year_s),
      treated = treated_s,
      post    = post_s,
      year_num= year_s
    )

    # DiD p-value with clustered SEs
    fit_d  <- lm(WTP ~ did + store + year, data = df_s)
    pval_d <- coeftest(fit_d, vcovCL(fit_d, cluster = ~store))["did", 4]

    # Pre-trend test: treated x linear year interaction in pre-period only
    if (n_pre >= 2) {
      pre_s      <- df_s[df_s$post == 0, ]
      pre_s$yr_c <- pre_s$year_num - mean(pre_s$year_num)
      fit_p  <- lm(WTP ~ treated * yr_c + store, data = pre_s)
      ct_p   <- coeftest(fit_p, vcovCL(fit_p, cluster = ~store))
      pval_p <- tryCatch(ct_p["treated:yr_c", 4], error = function(e) 1.0)
    } else {
      pval_p <- 1.0   # Cannot test with only 1 pre-period — always "passes"
    }

    c(pval_d, pval_p)
  })  # 2 × N_SIM_PT matrix

  data.frame(n_pre    = n_pre,
             did_pval = mat[1, ],
             pre_pval = mat[2, ])
}

sim_all_pt <- map_dfr(1:6, pt_sim_one)
cat(sprintf("Simulation complete: %d total runs across 6 pre-period conditions.\n",
            nrow(sim_all_pt)))
```

```{r pt-type1-plot, fig.height=5, fig.width=10, echo=FALSE}
type1_pt <- sim_all_pt |>
  group_by(n_pre) |>
  summarise(
    pwr_pretest = mean(pre_pval < 0.05, na.rm = TRUE),
    t1_overall  = mean(did_pval < 0.05, na.rm = TRUE),
    t1_cond     = {
      pass_idx <- pre_pval >= 0.05 | is.na(pre_pval)
      if (!any(pass_idx)) NA_real_ else mean(did_pval[pass_idx] < 0.05, na.rm = TRUE)
    },
    n_pass = sum(pre_pval >= 0.05 | is.na(pre_pval), na.rm = TRUE),
    .groups = "drop"
  )

type1_pt |>
  transmute(
    n_pre,
    `Pre-test power\n(detects violation)`          = pwr_pretest,
    `Overall DiD\nType I error`                    = t1_overall,
    `Conditional Type I\n(given pre-test passes)`  = t1_cond
  ) |>
  pivot_longer(-n_pre, names_to = "Metric", values_to = "Rate") |>
  mutate(Metric = factor(Metric, levels = c(
    "Pre-test power\n(detects violation)",
    "Overall DiD\nType I error",
    "Conditional Type I\n(given pre-test passes)"
  ))) |>
  ggplot(aes(x = n_pre, y = Rate, colour = Metric, shape = Metric)) +
  geom_hline(yintercept = 0.05, linetype = "dotted", colour = "grey40", linewidth = 0.9) +
  annotate("text", x = 1.15, y = 0.07, label = "Nominal 5%", size = 3.2,
           colour = "grey40", hjust = 0) +
  geom_line(linewidth = 1.2) + geom_point(size = 3.5) +
  scale_colour_manual(values = c(
    "Pre-test power\n(detects violation)"         = "#2563a8",
    "Overall DiD\nType I error"                   = "#e67e22",
    "Conditional Type I\n(given pre-test passes)" = "firebrick"
  )) +
  scale_shape_manual(values = c(
    "Pre-test power\n(detects violation)"         = 16,
    "Overall DiD\nType I error"                   = 17,
    "Conditional Type I\n(given pre-test passes)" = 18
  )) +
  scale_x_continuous(breaks = 1:6) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(
    x = "Number of pre-treatment periods collected",
    y = "Rate",
    colour = NULL, shape = NULL,
    title = "Type I error inflation under parallel trends violation (true treatment effect = $0)",
    subtitle = sprintf(
      "Dutch stores grow $0.15/yr faster than Belgian — entirely spurious; %d sims per condition",
      N_SIM_PT)
  ) +
  theme_mod3() +
  theme(legend.key.width = unit(2, "cm"))
```

```{r pt-summary-table, echo=FALSE}
type1_pt |>
  transmute(
    `Pre-periods`         = n_pre,
    `Pre-test power`      = sprintf("%.0f%%", 100 * pwr_pretest),
    `Sims passing`        = sprintf("%d / %d", n_pass, N_SIM_PT),
    `Overall Type I`      = sprintf("%.0f%%", 100 * t1_overall),
    `Conditional Type I`  = sprintf("%.0f%%", 100 * t1_cond)
  ) |>
  knitr::kable(
    caption = sprintf(
      "Monte Carlo summary (%d sims per condition, true effect = $0, differential trend = $0.15/yr). 'Conditional' = false positive rate among studies where the pre-trend test passed.",
      N_SIM_PT)
  )
```

::: {.callout-warning}
## What this means for practice

**The key pattern:**

- With **1 pre-period**: The test is impossible. The false-positive rate can be 40–70% — a researcher following standard practice would be nearly flipping a coin.
- With **2–3 pre-periods**: Pre-test power is low. Most violations slip through. Conditional Type I error is still far above the nominal 5%.
- With **5–6 pre-periods**: Power improves substantially. But violations that slip through the test still produce inflated Type I error.

**The fundamental problem is not the test — it is the data.** Researchers often have no choice but to use available archives, which may span only 1–3 pre-treatment periods. This is structurally similar to mediation: just as researchers rarely measure all possible confounders of the M→Y path, they rarely collect enough pre-treatment history to reliably test parallel trends.

**What to do:**

- Report the number of pre-treatment periods and be honest about the test's power
- Run placebo DiDs — apply the treatment label to earlier years or non-treated units and check that "effects" are near zero
- Use **Rambachan & Roth's (2023) `HonestDiD`** sensitivity analysis, which asks: *how large a pre-trend violation would be required to explain away your result?* This reframes the question from "is there a violation?" (often unknowable) to "how large a violation would matter?" (estimable and reportable)
- Treat a passing event study as weak evidence, not proof, of parallel trends
:::

---

# Part 5: Mediation — Does the Eco-Label Work *Through* Perceived Sustainability? {#sec-m3-part5}

Mediation analysis asks whether the eco-label affects WTP **directly** or **through a mechanism** — here, *perceived sustainability* (M). Three estimators are compared across three scenarios that each challenge a different method.

| Scenario | Key feature | Who wins |
|---|---|---|
| **1: Clean normal data** | No violations — a baseline | All methods perform well |
| **2: Large outliers** | Correlated high-leverage points inflate OLS b-path | **ROBMED** wins |
| **3: M–Y confounded** | Unobserved U causes both M and Y; true ACME = 0 | **Imai sensitivity** catches it |

::: {.callout-note}
## The mediation arithmetic — and where it can break

Baron–Kenny mediation decomposes X's total effect on Y into two paths:

$$\underbrace{X \xrightarrow{\;a\;} M \xrightarrow{\;b\;} Y}_{\text{indirect (ACME) = } a \times b} \quad + \quad \underbrace{X \xrightarrow{\;c'\;} Y}_{\text{direct (ADE)}}$$

The **ACME** (Average Causal Mediation Effect) is the product $a \times b$. It has two failure modes that the three scenarios below are specifically designed to reveal:

1. **The *b*-path can be estimated on the wrong data** — if outliers inflate the apparent slope between M and Y, $\hat{b}$ is too large, and so is $\hat{a} \times \hat{b}$. ROBMED addresses this.
2. **The *b*-path can be estimated on the wrong model** — if an unobserved variable $U^*$ drives both M and Y, the OLS $\hat{b}$ picks up the $U^*$ signal as if it were causal M→Y. No regression method alone can fix this; Imai's sensitivity analysis quantifies how severe the problem would need to be to explain away the result.
:::

## Causal Structures {#sec-med-dags}

```{r med-dag-scenarios, fig.height=3.2, fig.width=10, echo=FALSE}
#| code-summary: "▶ DAGs: true mediation (Scenarios 1 & 2) vs. confounded (Scenario 3)"
dA <- .dag(nodes = list(X=c(0,.5), M=c(.5,.9), Y=c(1,.5)),
           solid  = list(c("X","M"), c("M","Y"), c("X","Y")),
           title  = "Scenarios 1 & 2: True mediation\nX → M → Y  (+ direct path X → Y)")
dB <- .dag(nodes     = list(X=c(0,.5), M=c(.5,.9), Y=c(1,.5), U=c(.5,.1)),
           solid     = list(c("X","M"), c("X","Y"), c("U","M"), c("U","Y")),
           dashed    = list(c("M","Y")),
           unobserved= "U",
           title     = "Scenario 3: No causal M→Y\nU* confounds M and Y  (M→Y is spurious)")
dA | dB
```

## Setup: Three DGPs {#sec-med-dgp}

```{r med-dgp-setup}
#| code-summary: "▶ Three mediation DGPs: clean normal data, large outliers, M–Y confounded"
set.seed(2025)
N_med <- 350
X     <- rbinom(N_med, 1, 0.5)   # eco-label (1 = exposed to label)

# True parameters (Scenarios 1 & 2)
a_true    <- 0.50   # eco_label → perc_sust
b_true    <- 0.55   # perc_sust → WTP
c_prime   <- 0.30   # eco_label → WTP (direct path)
ACME_true <- a_true * b_true   # 0.275

# ── Scenario 1: Clean normal data ────────────────────────────────────────────
# No violations; both OLS and ROBMED should recover ACME ≈ 0.275
M_s1   <- a_true * X + rnorm(N_med, 0, 0.60)
Y_s1   <- c_prime * X + b_true * M_s1 + rnorm(N_med, 0, 0.60)
df_s1  <- data.frame(eco_label = X, perc_sust = M_s1, WTP = Y_s1)

# ── Scenario 2: Large correlated outliers (12% contamination) ───────────────
# Correlated shocks to M and Y residuals in the same direction: outlier obs
# suggest a much steeper M→Y slope → OLS b-path inflated → OLS ACME >> 0.275
# ROBMED MM-estimator downweights these high-leverage points → stays near truth
out_n   <- ceiling(N_med * 0.12)
out_idx <- sample(N_med, out_n)
out_sgn <- rep(1, out_n)
out_mag <- runif(out_n, 6, 10)   # 10–17× the baseline SD: strong leverage
eps_M2  <- rnorm(N_med, 0, 0.60)
eps_Y2  <- rnorm(N_med, 0, 0.60)
eps_M2[out_idx] <- eps_M2[out_idx] + out_sgn * out_mag          # M outlier
eps_Y2[out_idx] <- eps_Y2[out_idx] + out_sgn * out_mag * 0.90   # correlated Y outlier
M_s2   <- a_true * X + eps_M2
Y_s2   <- c_prime * X + b_true * M_s2 + eps_Y2
df_s2  <- data.frame(eco_label = X, perc_sust = M_s2, WTP = Y_s2)

# ── Scenario 3: M–Y confounded via unobserved U* (true ACME = 0) ─────────────
# U* drives both perc_sust and WTP; there is NO causal M → Y path
# OLS and ROBMED both find spurious ACME (Type I error) because they cannot
# separate causal M→Y from U-induced M–Y correlation
# Imai's medsens() reveals the finding's fragility via a small critical ρ*
U      <- rnorm(N_med, 0, 1)
M_s3   <- a_true * X + 0.60 * U + rnorm(N_med, 0, 0.55)
Y_s3   <- c_prime * X + 0.65 * U + rnorm(N_med, 0, 0.55)   # no M_s3 term!
df_s3  <- data.frame(eco_label = X, perc_sust = M_s3, WTP = Y_s3)
ACME_true_s3 <- 0   # no causal M → Y

cat(sprintf("N = %d  |  Treatment rate = %.0f%%\n", N_med, 100 * mean(X)))
cat(sprintf("Scenarios 1 & 2: true ACME = %.3f  (a=%.2f, b=%.2f, c'=%.2f)\n",
            ACME_true, a_true, b_true, c_prime))
cat(sprintf("Scenario 3:      true ACME = %.3f  (U* confounds both M and Y)\n",
            ACME_true_s3))
```

## Estimating All Three Methods {#sec-med-run}

```{r run-all-mediation, cache=FALSE}
#| code-summary: "▶ Run OLS delta method, ROBMED, and Imai on all three scenarios"
run_mediation_triple <- function(df, R = 500) {
  m_m <- lm(perc_sust ~ eco_label, data = df)
  m_y <- lm(WTP ~ eco_label + perc_sust, data = df)

  # 1. OLS: a × b with delta-method CI
  a_hat   <- coef(m_m)["eco_label"]
  b_hat   <- coef(m_y)["perc_sust"]
  ols_est <- a_hat * b_hat
  se_ab   <- sqrt(b_hat^2 * vcov(m_m)["eco_label","eco_label"] +
                  a_hat^2 * vcov(m_y)["perc_sust","perc_sust"])
  ols_lo  <- ols_est - 1.96 * se_ab
  ols_hi  <- ols_est + 1.96 * se_ab

  # 2. ROBMED: MM-robust regression (downweights high-leverage outliers)
  set.seed(99)
  med_rob  <- test_mediation(WTP ~ m(perc_sust) + eco_label,
                             data = df, method = "regression", robust = TRUE, R = R)
  rob      <- rob_extract(med_rob)

  # 3. Imai et al.: quasi-Bayesian (used for medsens() sensitivity analysis)
  set.seed(99)
  med_imai <- mediate(m_m, m_y, treat = "eco_label", mediator = "perc_sust",
                      boot = FALSE, sims = R)

  list(a_hat    = a_hat,          b_hat    = b_hat,
       ols_est  = ols_est,        ols_lo   = ols_lo,       ols_hi  = ols_hi,
       rob_est  = rob$est,        rob_lo   = rob$lo,       rob_hi  = rob$hi,
       imai_est = med_imai$d0,    imai_lo  = med_imai$d0.ci[1],
       imai_hi  = med_imai$d0.ci[2],
       med_imai = med_imai,       med_rob  = med_rob)
}

cat("Running three mediation scenarios (OLS + ROBMED + Imai, 500 sims each)...\n")
res_s1 <- run_mediation_triple(df_s1)
res_s2 <- run_mediation_triple(df_s2)
res_s3 <- run_mediation_triple(df_s3)
cat(sprintf("Scenario 1 — OLS: %.3f  ROBMED: %.3f  (truth = %.3f)\n",
            res_s1$ols_est, res_s1$rob_est, ACME_true))
cat(sprintf("Scenario 2 — OLS: %.3f  ROBMED: %.3f  (truth = %.3f)\n",
            res_s2$ols_est, res_s2$rob_est, ACME_true))
cat(sprintf("Scenario 3 — OLS: %.3f  ROBMED: %.3f  (truth = %.3f [no M→Y path])\n",
            res_s3$ols_est, res_s3$rob_est, ACME_true_s3))
```

## Scenario 1: Clean Normal Data — All Methods Should Work {#sec-med-clean}

With textbook normal errors and no confounding, OLS, ROBMED, and Imai all estimate ACME accurately. This is the baseline case each method was designed for.

```{r med-scenario1, echo=FALSE}
#| code-summary: "▶ Scenario 1 results: clean normal data"
tibble(
  Method   = c("OLS delta method", "ROBMED (MM-estimator)", "Imai (quasi-Bayes)"),
  Paths    = c(sprintf("a=%.3f, b=%.3f", res_s1$a_hat, res_s1$b_hat),
               "-- (robust MM)",
               "-- (uses OLS)"),
  Estimate = round(c(res_s1$ols_est, res_s1$rob_est, res_s1$imai_est), 3),
  CI95     = sprintf("[%.3f, %.3f]",
                     c(res_s1$ols_lo, res_s1$rob_lo, res_s1$imai_lo),
                     c(res_s1$ols_hi, res_s1$rob_hi, res_s1$imai_hi)),
  Bias     = sprintf("%+.3f",
                     c(res_s1$ols_est, res_s1$rob_est, res_s1$imai_est) - ACME_true),
  Verdict  = c("On target", "On target", "On target")
) |>
  knitr::kable(
    col.names = c("Method", "a & b paths", "Estimate", "95% CI", "Bias", "Verdict"),
    caption   = paste0("Scenario 1 \u2014 Clean normal data | true ACME = ",
                       round(ACME_true, 3), " | N = ", N_med),
    align     = c("l","l","r","c","r","c")
  )
```

## Scenario 2: Large Outliers — ROBMED's Advantage {#sec-med-outliers}

Twelve percent of participants show large, **correlated** deviations in both perceived sustainability (M) and WTP (Y) in the same direction. These high-leverage observations suggest a much steeper M→Y slope than the true 0.55 to OLS — inflating the b-path and thus ACME. The MM-estimator in ROBMED downweights points that deviate strongly from the main data cloud.

::: {.callout-note}
## Why correlated outliers inflate the *b*-path specifically

The key is *direction*: when participants who score unusually high on M also score unusually high on Y (same-sign shocks), they create data points that pull the OLS regression line steeply upward. To OLS, these are just informative extreme observations — it can't distinguish "genuinely high M causes genuinely high Y" from "an external shock hit both M and Y simultaneously."

**ROBMED's MM-estimator** assigns lower weight to observations whose residuals are large relative to the bulk of the data. The outlier participants — who are far from the M–Y regression cloud — receive close to zero weight, leaving the slope estimate dominated by the non-contaminated 88%.

**Why Imai doesn't help here**: `mediate()` calls the same `lm()` models internally. When those models are biased, the quasi-Bayesian sampling averages over biased posteriors. ROBMED is the right tool; Imai is not.
:::


```{r med-scenario2, echo=FALSE}
#| code-summary: "▶ Scenario 2 results: large outliers inflate OLS but not ROBMED"
tibble(
  Method   = c("OLS delta method", "ROBMED (MM-estimator)", "Imai (quasi-Bayes)"),
  `b-path` = sprintf("%.3f (true = %.2f)",
                     c(res_s2$b_hat, res_s2$rob_est / res_s2$a_hat, res_s2$b_hat),
                     rep(b_true, 3)),
  Estimate = round(c(res_s2$ols_est, res_s2$rob_est, res_s2$imai_est), 3),
  `95% CI` = sprintf("[%.3f, %.3f]",
                     c(res_s2$ols_lo, res_s2$rob_lo, res_s2$imai_lo),
                     c(res_s2$ols_hi, res_s2$rob_hi, res_s2$imai_hi)),
  Bias     = sprintf("%+.3f",
                     c(res_s2$ols_est, res_s2$rob_est, res_s2$imai_est) - ACME_true),
  Verdict  = c("\u2716 Inflated \u2014 outliers pull b-path up",
               "\u2713 Robust \u2014 outliers down-weighted",
               "\u2716 Inflated \u2014 uses OLS internally")
) |>
  knitr::kable(
    caption = sprintf(
      "Scenario 2 \u2014 12%% correlated outlier contamination | true ACME = %.3f | N = %d",
      ACME_true, N_med),
    align = c("l","l","r","c","r","l")
  )
```

```{r med-outlier-scatter, fig.height=3.5, fig.width=10, echo=FALSE}
#| code-summary: "▶ Scatter: clean vs contaminated data showing outlier leverage on OLS slope"
p_s1_sc <- ggplot(df_s1, aes(x = perc_sust, y = WTP)) +
  geom_point(alpha = 0.35, size = 1.5, colour = clr_ctrl) +
  geom_smooth(method = "lm", se = TRUE, colour = "firebrick", linewidth = 1.2) +
  labs(x = "Perceived sustainability (M)", y = "WTP ($)",
       title = "Scenario 1: Clean data",
       subtitle = sprintf("OLS b = %.3f  (true = %.2f)", res_s1$b_hat, b_true)) +
  theme_mod3()

df_s2_vis <- df_s2 |> mutate(outlier = seq_len(n()) %in% out_idx)
p_s2_sc <- ggplot(df_s2_vis, aes(x = perc_sust, y = WTP, colour = outlier)) +
  geom_point(alpha = 0.45, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, colour = "firebrick", linewidth = 1.2) +
  scale_colour_manual(values = c("FALSE" = "grey55", "TRUE" = "#e74c3c"),
                      labels = c("FALSE" = "Clean obs", "TRUE" = "Outlier (12%)"),
                      name = NULL) +
  labs(x = "Perceived sustainability (M)", y = "WTP ($)",
       title = "Scenario 2: Outliers (red) inflate the OLS slope",
       subtitle = sprintf("OLS b = %.3f  ROBMED b ≈ %.3f  (true = %.2f)",
                          res_s2$b_hat, res_s2$rob_est / res_s2$a_hat, b_true)) +
  theme_mod3()

p_s1_sc | p_s2_sc
```

## Scenario 3: M–Y Confounded — Regression-Based Methods Fail, Imai Diagnoses {#sec-med-confounded}

U* drives both perceived sustainability and WTP. **There is no causal M→Y path.** The true ACME = 0. Yet both OLS and ROBMED find a large, significant "mediation effect" because they cannot distinguish causal M→Y from U-induced M–Y correlation. This is a structural identification failure, not an estimation problem — no regression method alone can fix it. But Imai's sensitivity analysis reveals that the result is fragile.

```{r med-scenario3, echo=FALSE}
#| code-summary: "▶ Scenario 3: both OLS and ROBMED produce a Type I error"
tibble(
  Method   = c("OLS delta method", "ROBMED (MM-estimator)", "Imai (quasi-Bayes)"),
  `b-path (spurious)` = sprintf("%.3f (true = 0.00)", c(res_s3$b_hat, NA_real_, res_s3$b_hat)) |>
                        (\(x) { x[2] <- sprintf("%.3f (true = 0.00)", res_s3$rob_est / res_s3$a_hat); x })(),
  Estimate = round(c(res_s3$ols_est, res_s3$rob_est, res_s3$imai_est), 3),
  `95% CI` = sprintf("[%.3f, %.3f]",
                     c(res_s3$ols_lo, res_s3$rob_lo, res_s3$imai_lo),
                     c(res_s3$ols_hi, res_s3$rob_hi, res_s3$imai_hi)),
  Bias     = sprintf("%+.3f",
                     c(res_s3$ols_est, res_s3$rob_est, res_s3$imai_est) - ACME_true_s3),
  Verdict  = c("\u2716 Type I error \u2014 spurious mediation",
               "\u2716 Type I error \u2014 U* not outlier-based",
               "\u2716 Type I error \u2014 see \u03c1* below")
) |>
  knitr::kable(
    caption = sprintf(
      "Scenario 3 \u2014 M\u2013Y confounded via U* | true ACME = %.3f (no M\u2192Y path) | N = %d",
      ACME_true_s3, N_med),
    align = c("l","l","r","c","r","l")
  )
```

## Where Imai's Framework Shines: Sensitivity Analysis {#sec-med-sensitivity}

Imai et al.'s `medsens()` asks: *how large would the correlation between M-equation and Y-equation residuals (ρ) need to be to drive ACME to zero?* This **critical ρ\*** measures fragility:

- **Large ρ***: only substantial unmeasured confounding could explain away the result → **robust**
- **Small ρ***: even modest unmeasured confounding suffices → **fragile**

```{r med-sensitivity-compare, fig.height=4.5, fig.width=10}
#| code-summary: "▶ Sensitivity analysis: clean DGP (robust) vs. confounded DGP (fragile)"
set.seed(99)
sens_s1 <- medsens(res_s1$med_imai, rho.by = 0.05, effect.type = "indirect", sims = 500)
set.seed(99)
sens_s3 <- medsens(res_s3$med_imai, rho.by = 0.05, effect.type = "indirect", sims = 500)

safe_crit_rho <- function(s) {
  tryCatch({
    idx_neg <- which(s$d0 <= 0)
    if (length(idx_neg) == 0) NA_real_
    else round(min(abs(s$rho[idx_neg]), na.rm = TRUE), 2)
  }, error = function(e) NA_real_)
}
crit_rho_s1 <- safe_crit_rho(sens_s1)
crit_rho_s3 <- safe_crit_rho(sens_s3)

fmt_rho <- function(x) if (is.na(x)) "> 1.0" else sprintf("%.2f", x)
cat(sprintf("Scenario 1 (true ACME = 0.275): critical ρ* ≈ %s  → ROBUST finding\n",
            fmt_rho(crit_rho_s1)))
cat(sprintf("Scenario 3 (true ACME = 0):     critical ρ* ≈ %s  → FRAGILE finding\n",
            fmt_rho(crit_rho_s3)))
cat("\nSmaller ρ* = less confounding is needed to explain away the result.\n")

# Build tidy sensitivity curves using only the guaranteed $rho and $d0 fields
sens_df <- bind_rows(
  data.frame(rho      = sens_s1$rho,
             ACME     = sens_s1$d0,
             Scenario = "Scenario 1: Clean data (true ACME = 0.275)",
             stringsAsFactors = FALSE),
  data.frame(rho      = sens_s3$rho,
             ACME     = sens_s3$d0,
             Scenario = "Scenario 3: M\u2013Y confounded (true ACME = 0)",
             stringsAsFactors = FALSE)
)

ggplot(sens_df, aes(x = rho, y = ACME, colour = Scenario)) +
  geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.9) +
  geom_line(linewidth = 1.4) +
  { if (!is.na(crit_rho_s1))
      list(
        geom_vline(xintercept =  crit_rho_s1, linetype = "dashed",
                   colour = clr_eco, linewidth = 0.9),
        geom_vline(xintercept = -crit_rho_s1, linetype = "dashed",
                   colour = clr_eco, linewidth = 0.9),
        annotate("text", x = crit_rho_s1 + 0.03, y = 0.18,
                 label = sprintf("\u03c1*\u2248%s\n(ROBUST)", fmt_rho(crit_rho_s1)),
                 hjust = 0, size = 3.0, colour = clr_eco, fontface = "bold")
      ) } +
  { if (!is.na(crit_rho_s3))
      list(
        geom_vline(xintercept =  crit_rho_s3, linetype = "dashed",
                   colour = clr_ctrl, linewidth = 0.9),
        geom_vline(xintercept = -crit_rho_s3, linetype = "dashed",
                   colour = clr_ctrl, linewidth = 0.9),
        annotate("text", x = crit_rho_s3 + 0.03, y = -0.22,
                 label = sprintf("\u03c1*\u2248%s\n(FRAGILE)", fmt_rho(crit_rho_s3)),
                 hjust = 0, size = 3.0, colour = clr_ctrl, fontface = "bold")
      ) } +
  scale_colour_manual(
    values = c("Scenario 1: Clean data (true ACME = 0.275)" = clr_eco,
               "Scenario 3: M\u2013Y confounded (true ACME = 0)"  = clr_ctrl)) +
  labs(x       = expression(paste("Sensitivity parameter ", rho,
                                  "  (correlation between M- and Y-equation residuals)")),
       y       = "ACME (indirect effect)",
       colour  = NULL,
       title   = "Imai sensitivity: how much confounding would nullify the ACME?",
       subtitle = paste0("\u03c1* = critical value at which ACME crosses zero",
                         "  \u2014  smaller \u03c1* = more fragile result")) +
  theme_mod3() +
  theme(legend.position = "top")
```

::: {.callout-tip}
## Reading the sensitivity plot

Both Scenario 1 and Scenario 3 produce a *statistically significant* ACME from OLS — they look equally convincing at face value. The sensitivity analysis reveals very different levels of trust:

- **Scenario 1 (clean)**: ACME reaches zero only at $\rho^* \approx$ `r fmt_rho(crit_rho_s1)`. Even substantial unmeasured M–Y confounding cannot explain away the result. The finding is **robust**.
- **Scenario 3 (confounded)**: ACME crosses zero at $\rho^* \approx$ `r fmt_rho(crit_rho_s3)`. Only modest confounding is enough to nullify the result. The "significant" ACME is **fragile** — correctly so, since the true ACME is 0.

**The Imai advantage is not a better point estimate** (it uses the same linear models as OLS, so it inherits the same bias in Scenario 3). The advantage is the *sensitivity analysis itself* — an honest diagnostic that Baron–Kenny completely omits. A finding that reports both the ROBMED estimate and a large ρ* from `medsens()` is far more credible than one that does neither.
:::

## Comprehensive Bias Comparison Across All Scenarios {#sec-med-comparison}

```{r mediation-bias-comparison, fig.height=6.5, fig.width=9, echo=FALSE}
#| code-summary: "▶ Forest plot: OLS vs ROBMED vs Imai across all three scenarios"
comp_df <- tibble(
  Scenario = rep(c(
    sprintf("Scenario 1: Clean normal data\n(true ACME = %.3f)", ACME_true),
    sprintf("Scenario 2: Large outliers\n(true ACME = %.3f)", ACME_true),
    sprintf("Scenario 3: M\u2013Y confounded\n(true ACME = %.3f)", ACME_true_s3)
  ), each = 3),
  Method   = rep(c("OLS delta method", "ROBMED", "Imai (quasi-Bayes)"), 3),
  Truth    = rep(c(ACME_true, ACME_true, ACME_true_s3), each = 3),
  Estimate = c(res_s1$ols_est, res_s1$rob_est, res_s1$imai_est,
               res_s2$ols_est, res_s2$rob_est, res_s2$imai_est,
               res_s3$ols_est, res_s3$rob_est, res_s3$imai_est),
  CI_lo    = c(res_s1$ols_lo,  res_s1$rob_lo,  res_s1$imai_lo,
               res_s2$ols_lo,  res_s2$rob_lo,  res_s2$imai_lo,
               res_s3$ols_lo,  res_s3$rob_lo,  res_s3$imai_lo),
  CI_hi    = c(res_s1$ols_hi,  res_s1$rob_hi,  res_s1$imai_hi,
               res_s2$ols_hi,  res_s2$rob_hi,  res_s2$imai_hi,
               res_s3$ols_hi,  res_s3$rob_hi,  res_s3$imai_hi)
) |>
  mutate(
    Bias         = Estimate - Truth,
    Scenario     = fct_inorder(Scenario),
    Method       = factor(Method, levels = c("Imai (quasi-Bayes)", "ROBMED", "OLS delta method")),
    covers_truth = (CI_lo <= Truth & CI_hi >= Truth)
  )

method_colours <- c(
  "OLS delta method"  = clr_ctrl,
  "ROBMED"            = clr_eco,
  "Imai (quasi-Bayes)" = "#8e44ad"
)

ggplot(comp_df, aes(y = Method, x = Estimate, xmin = CI_lo, xmax = CI_hi,
                    colour = Method, shape = covers_truth)) +
  geom_vline(aes(xintercept = Truth), linetype = "dashed",
             colour = "grey30", linewidth = 1) +
  geom_pointrange(size = 0.7, linewidth = 1) +
  geom_text(aes(label = sprintf("bias=%+.3f", Bias), x = CI_hi + 0.012),
            hjust = 0, size = 2.8, colour = "grey20") +
  facet_wrap(~Scenario, ncol = 1, scales = "free_x") +
  scale_colour_manual(values = method_colours) +
  scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 4),
                     labels = c("TRUE" = "CI covers truth", "FALSE" = "CI misses truth"),
                     name   = NULL) +
  labs(x = "Estimated ACME (indirect effect)", y = NULL,
       colour = "Estimator",
       title  = "Bias comparison: OLS vs. ROBMED vs. Imai across three mediation scenarios",
       subtitle = paste0("Dashed line = true ACME  |  \u00d7 = CI misses truth  |  labels show signed bias\n",
                         "Scenario 3 note: all point estimates are spurious \u2014 ",
                         "Imai adds value via sensitivity analysis, not point estimates")) +
  theme_mod3() +
  theme(legend.position = "top",
        legend.box = "horizontal",
        strip.text = element_text(face = "bold", size = 9),
        legend.key.size = unit(0.7, "cm"))
```

::: {.callout-warning}
## Method × scenario decision guide

| Scenario | OLS | ROBMED | Imai `medsens()` |
|---|---|---|---|
| **Clean normal data** | ✔ Unbiased | ✔ Unbiased | ✔ Large ρ* confirms robustness |
| **Large outliers** | ✖ Biased upward | ✔ Downweights outliers | ✖ Inherits OLS bias |
| **M–Y confounded** | ✖ Type I error | ✖ Type I error | ✔ Small ρ* flags fragility |

**Key lesson**: ROBMED addresses *estimation* failures (non-normality, outliers). Imai's sensitivity analysis addresses *identification* failures (unmeasured confounding). They solve different problems and should be used together.

**In practice**: report ROBMED as your primary ACME estimate (robustness to outliers) *and* `medsens()` to quantify fragility to unmeasured confounding. A finding that survives both tests — ROBMED agrees with OLS, and ρ* is large — is the most credible mediation result you can report.
:::

## Why Confounding Causes Type I Errors: The 2×2 Design {#sec-med-2x2}

The three-scenario analysis above treats confounding as a special case. The 2×2 design below makes the mechanism explicit by crossing two independent dimensions:

- **Rows**: true mediation exists vs. does not exist (b = 0)
- **Columns**: M–Y confounded vs. unconfounded

The critical cell is **top-right** — no true mediation, but confounding is active. Both OLS and ROBMED report a large, significant indirect effect despite b = 0. This is a Type I error caused entirely by the unmeasured confounder U*, not by any true mediation.

```{r med-2x2-setup, echo=FALSE}
#| code-summary: "▶ 2×2 simulation setup"

# ── 2×2 parameters ────────────────────────────────────────────────────────────
p2_mx   <- 0.50   # X → M  (a path)
p2_my   <- 0.55   # M → Y  (b path, used in "true mediation" rows)
p2_xy   <- 0.30   # X → Y  (direct c' path)
p2_um   <- 0.55   # U → M  (confounding into mediator)
p2_uy   <- 0.60   # U → Y  (confounding into outcome)
p2_xu   <- 0.30   # X → U
N2      <- 350
R2      <- 300

set.seed(2024)

run_2x2 <- function(label, confound_m, confound_y, b_val) {
  p_um_act <- if (confound_m) p2_um else 0
  p_uy_act <- if (confound_y) p2_uy else 0
  w_a <- c(p2_mx, p_um_act, 0.30); w_a <- w_a / sum(w_a)
  w_c <- c(p2_xy, b_val,  p_uy_act, 0.30); w_c <- w_c / sum(w_c)
  w_u <- c(p2_xu, 0.30);  w_u <- w_u / sum(w_u)
  x   <- rbinom(N2, 1, 0.5)
  u   <- w_u[1]*x + w_u[2]*rnorm(N2)
  m   <- w_a[1]*x + w_a[2]*u + w_a[3]*rnorm(N2)
  y   <- w_c[1]*x + w_c[2]*m + w_c[3]*u + w_c[4]*rnorm(N2)
  df  <- data.frame(x = x, m = m, y = y)
  mm  <- lm(m ~ x, data = df)
  my  <- lm(y ~ x + m, data = df)
  med <- mediate(mm, my, treat = "x", mediator = "m",
                 boot = TRUE, sims = R2)
  true_indirect <- w_a[1] * w_c[2]
  data.frame(
    label    = label,
    confound = confound_m | confound_y,
    b_true   = b_val,
    true_ind = true_indirect,
    est      = med$d0,
    lo       = med$d0.ci[1],
    hi       = med$d0.ci[2],
    covers   = (med$d0.ci[1] <= true_indirect & med$d0.ci[2] >= true_indirect),
    stringsAsFactors = FALSE
  )
}

scenarios_2x2 <- list(
  list(label = "No true mediation\nUnconfounded",   cm = FALSE, co = FALSE, b = 0),
  list(label = "No true mediation\nConfounded (U*)", cm = TRUE,  co = TRUE,  b = 0),
  list(label = "True mediation\nUnconfounded",       cm = FALSE, co = FALSE, b = p2_my),
  list(label = "True mediation\nConfounded (U*)",    cm = TRUE,  co = TRUE,  b = p2_my)
)

res_2x2 <- do.call(rbind, lapply(scenarios_2x2, function(s) {
  run_2x2(s$label, s$cm, s$co, s$b)
}))
```

```{r med-2x2-dags, fig.height=5, fig.width=11, echo=FALSE}
#| code-summary: "▶ DAGs: the four 2×2 scenarios"

# ── DAG drawing helper (adapted from Mediation-Simulation.Rmd) ────────────────
draw_med_dag <- function(confound_m, confound_y, title = "", show_b = TRUE) {
  nodes <- data.frame(
    label = c("X", "M", "Y", "U*"),
    x     = c(0,   2,   4,   2),
    y     = c(0,   0,   0,   2)
  )
  b_etype <- if (show_b) "path" else "null_b"
  b_col   <- if (show_b) "grey30" else "grey70"
  b_lbl   <- if (show_b) "b" else "b\u00a0=\u00a00"

  edges <- rbind(
    data.frame(x=0, y=0, xend=2, yend=0, etype="path"),
    data.frame(x=2, y=0, xend=4, yend=0, etype=b_etype),
    data.frame(x=0, y=0, xend=2, yend=2, etype="neutral"),
    if (confound_m) data.frame(x=2, y=2, xend=2, yend=0.1, etype="confound") else NULL,
    if (confound_y) data.frame(x=2, y=2, xend=4, yend=0.1, etype="confound") else NULL
  )
  direct <- data.frame(x=0, y=0, xend=4, yend=0)

  ggplot() +
    geom_curve(data = direct, aes(x=x, y=y, xend=xend, yend=yend),
               curvature = -0.35, colour = "#2166ac", linewidth = 0.75,
               arrow = arrow(length = unit(0.16, "cm"), type = "open")) +
    annotate("text", x = 2, y = -0.85, colour = "#2166ac", size = 2.8,
             label = "c\u2019 (direct)", fontface = "italic") +
    geom_segment(data = edges,
                 aes(x=x, y=y, xend=xend, yend=yend,
                     colour=etype, linetype=etype),
                 linewidth = 0.85,
                 arrow = arrow(length = unit(0.16, "cm"), type = "open")) +
    annotate("text", x = 1,   y = 0.25, label = "a", size = 3.1,
             colour = "grey30", fontface = "italic") +
    annotate("text", x = 3,   y = 0.25, label = b_lbl, size = 3.1,
             colour = b_col, fontface = "italic") +
    geom_point(data = nodes, aes(x=x, y=y),
               shape = 21, size = 18, fill = "white", colour = "grey25", stroke = 1.3) +
    geom_text(data = nodes, aes(x=x, y=y, label=label), size = 4, fontface = "bold") +
    scale_colour_manual(guide = "none",
      values = c("path"="grey25", "null_b"="grey75",
                 "neutral"="grey60", "confound"="#d6604d")) +
    scale_linetype_manual(guide = "none",
      values = c("path"="solid", "null_b"="dotted",
                 "neutral"="solid", "confound"="dashed")) +
    coord_cartesian(xlim = c(-0.6, 4.6), ylim = c(-1.1, 2.5)) +
    labs(title = title) +
    theme_void(base_size = 11) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 9.5,
                                    margin = margin(b = 5)))
}

dag_list_2x2 <- list(
  draw_med_dag(FALSE, FALSE, "No true mediation \u00b7 Unconfounded\n\u2705 Correct null",  show_b = FALSE),
  draw_med_dag(TRUE,  TRUE,  "No true mediation \u00b7 Confounded\n\U0001F6A8 FALSE POSITIVE", show_b = FALSE),
  draw_med_dag(FALSE, FALSE, "True mediation \u00b7 Unconfounded\n\u2705 Unbiased",          show_b = TRUE),
  draw_med_dag(TRUE,  TRUE,  "True mediation \u00b7 Confounded\n\u26a0\ufe0f Biased upward",  show_b = TRUE)
)

(dag_list_2x2[[1]] | dag_list_2x2[[2]]) / (dag_list_2x2[[3]] | dag_list_2x2[[4]])
```

```{r med-2x2-plot, fig.height=5.5, fig.width=9, echo=FALSE}
#| code-summary: "▶ True vs. estimated ACME across all four cells"

cell_colours <- c(
  "No true mediation\nUnconfounded"    = "#4dac26",
  "No true mediation\nConfounded (U*)" = "#ca0020",
  "True mediation\nUnconfounded"       = "#0571b0",
  "True mediation\nConfounded (U*)"    = "#f4a582"
)

res_2x2$label <- factor(res_2x2$label, levels = names(cell_colours))

ggplot(res_2x2, aes(x = est, xmin = lo, xmax = hi,
                    y = "Indirect effect",
                    colour = label, shape = covers)) +
  geom_vline(aes(xintercept = true_ind), linetype = "dashed",
             colour = "grey30", linewidth = 0.9) +
  geom_pointrange(size = 0.9, linewidth = 1.2, position = position_dodge(width = 0.5)) +
  geom_point(aes(x = true_ind), shape = 17, size = 3.5,
             colour = "#333", position = position_dodge(width = 0.5)) +
  facet_wrap(~ label, ncol = 2, scales = "free_x") +
  scale_colour_manual(values = cell_colours, guide = "none") +
  scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 4),
                     labels = c("TRUE" = "CI covers truth", "FALSE" = "CI misses truth"),
                     name = NULL) +
  labs(x = "Estimated ACME (indirect effect)",
       y = NULL,
       title = "True vs. Estimated ACME \u2014 2\u00d72 Design",
       subtitle = paste0(
         "\u25b2 Black triangle = true DGP value  |  \u25cf Circle + bar = estimate \u00b1 95% CI\n",
         "\U0001F6A8 Top-right panel: b = 0 in truth, yet the model reports significant mediation \u2014 ",
         "a Type I error driven entirely by U*"
       )) +
  theme_mod3() +
  theme(
    strip.text         = element_text(face = "bold", size = 10, colour = "white",
                                      margin = margin(4, 4, 4, 4)),
    strip.background   = element_rect(fill = "grey30", colour = NA),
    panel.spacing      = unit(1.2, "lines"),
    legend.position    = "top",
    panel.grid.major.y = element_blank()
  )
```

::: {.callout-important}
## The Type I error mechanism in one sentence

When both U→M and U→Y are active, M and Y share a hidden common cause. The mediation model mistakes their spurious correlation for a real b path — and the bootstrap CI confidently narrows around a biased estimate, making the false positive look decisive.

**The fix is not a better estimator.** Imai's `medsens()` reveals how fragile the finding is; ROBMED guards against estimation errors from outliers. But neither can eliminate the Type I error if U* is real and unmeasured. The only solutions are design-based: measure U*, block the back-door path by design, or use an instrument for M.
:::

---

## Summary of Methods {#sec-summary-table}

```{r summary-table, echo=FALSE}
tibble(
  Method = c(
    "Randomised Experiment",
    "Linear Regression Adjustment",
    "Flexible Regression Adjustment",
    "IPW (stabilised)",
    "Entropy Balancing (WeightIt)",
    "Covariate Matching (Mahalanobis)",
    "Propensity Score Matching",
    "Doubly Robust (AIPW)",
    "Synthetic Control",
    "Regression Discontinuity",
    "Difference-in-Differences (TWFE)",
    "Synthetic DiD",
    "Mediation (Baron–Kenny OLS)",
    "Mediation (Imai et al.)",
    "ROBMED"
  ),
  `Key identifying assumption` = c(
    "Random assignment; SUTVA",
    "All confounders measured; linear functional form",
    "All confounders measured; correct non-linear specification",
    "All confounders measured; positivity (overlap)",
    "All confounders measured; moment balance sufficient",
    "All confounders measured; sufficient overlap",
    "All confounders measured; PS model correctly specified",
    "Either PS model OR outcome model is correctly specified",
    "Good pre-treatment fit; no interference",
    "Continuity at cutoff; no manipulation; no other discontinuities",
    "Parallel trends (in absence of treatment)",
    "Parallel trends after reweighting control units",
    "Sequential ignorability; no M–Y confounders",
    "Sequential ignorability; sensitivity analysis quantifies fragility",
    "Sequential ignorability; robust to outliers and non-normality"
  ),
  `Target estimand` = c(
    "ATE","ATE","ATE","ATE","ATE","ATT","ATT","ATE",
    "ATT (one treated unit)","LATE (near threshold)","ATT","ATT",
    "ACME","ACME","ACME"
  ),
  `When to use` = c(
    "Can randomise",
    "Observational; confounders measured; linear DGP",
    "Observational; confounders measured; non-linear DGP",
    "Observational; want ATE; good PS overlap",
    "Observational; want ATE; many covariates to balance",
    "Observational; want ATT; sufficient control units",
    "Observational; many covariates; report PS model sensitivity",
    "Observational; insurance against one misspecified model",
    "Few treated units; many pre-treatment periods",
    "Sharp threshold in a running variable",
    "Panel data; policy change in subset of units",
    "Panel data; parallel trends may not hold exactly",
    "Mediation hypothesis; well-controlled experiment",
    "Mediation with explicit assumptions + sensitivity analysis",
    "Mediation; non-normal WTP data or suspected outliers"
  )
) |>
  knitr::kable(caption="All methods in Module 3: assumptions, estimands, and use cases")
```

---

## Session Info {.unnumbered}

```{r session-info}
sessionInfo()
```
