---
title: "Module 1: Measurement"
subtitle: "Discriminant Validity, Omitted Variable Bias, Measurement Invariance, Latent Classes, Outliers, and Systematic Measurement Error"
code-tools: false
---

<iframe src="https://docs.google.com/presentation/d/e/2PACX-1vQQ3faOAAMPPcbScJcz6sh2J3xp0NsUmjdtyDl-0md-w5tMxPQprFUoz0rSWnq6Ug/pubembed?start=false&loop=false&delayms=3000" width="100%" height="380px" frameborder="0" allowfullscreen="true" mozallowfullscreen="true" webkitallowfullscreen="true" title="Module 1: Measurement Overview"></iframe>

## Overview

Good measurement is the foundation of good research, yet it is the step most often skipped. This module asks: *does your scale actually measure the construct you intend?* — and introduces the tools to find out.

We cover discriminant validity, omitted variable bias from measurement error, measurement invariance across groups, latent class structure, outlier detection, and systematic measurement error from bounded scales. Each topic is treated both conceptually and with live R code.

---

## Learning Goals

By the end of this module you should be able to:

- Distinguish construct validity from reliability, and explain why reliability alone is insufficient
- Diagnose discriminant validity problems using CFA and AVE/correlation comparisons
- Explain how classical measurement error biases regression coefficients and in which direction
- Test for measurement invariance across groups using configural, metric, and scalar CFA models
- Identify latent class structure using mixture models and interpret class profiles
- Detect and reason about outliers using Cook's D, DFFITS, and leverage diagnostics
- Explain how bounded scales structurally bias observed means through truncation, and diagnose this bias using skewness and Q-Q diagnostics on CTT residuals

---

## Recommended Reading

| Paper | Why it matters |
|---|---|
| [Flake & Fried (2020)](https://doi.org/10.1177/2515245920952393) — *Measurement Schmeasurement* | The best accessible critique of how measurement is treated in psychology and management |
| [Borsboom, Mellenbergh & van Heerden (2003)](https://doi.org/10.1037/0033-295X.110.2.203) | What does it actually mean for a test to measure a construct? |
| [Arkes & Tetlock (2004)](https://doi.org/10.1207/s15327965pli1504_01) | Challenges the construct validity of the Implicit Association Test — argues that IAT scores conflate racial preference with unrelated cognitive associations, making the construct unclear |
| [Vosgerau, Scopelliti & Huh (2020)](https://doi.org/10.1002/jcpy.1116) — *Exerting Self-Control ≠ Sacrificing Pleasure* | How reframing the DV changes conclusions |
| [Putnick & Bornstein (2016)](https://doi.org/10.1016/j.dr.2016.06.004) | Comprehensive tutorial on measurement invariance testing across groups and cultures |
| [Rousseeuw & Hubert (2011)](https://doi.org/10.1002/wics.196) | Overview of robust statistical methods and their advantages over classical estimators when outliers are present |
| [Zimek & Filzmoser (2018)](https://doi.org/10.1002/widm.1280) | Survey of outlier detection methods covering theoretical foundations and practical guidance |
| [Nylund, Asparouhov & Muthén (2007)](https://doi.org/10.1080/10705510701575396) | How to determine the number of latent classes; compares BIC, AIC, and likelihood ratio tests |
| [Liddell & Kruschke (2018)](https://doi.org/10.1016/j.jesp.2018.08.009) | Why treating ordinal Likert data as continuous distorts results, and how to handle it correctly |
| [Smithson & Verkuilen (2006)](https://doi.org/10.1037/1082-989X.11.1.54) | Introduction to beta regression for proportion and bounded-scale outcomes |
| [Cinelli & Hazlett (2020)](https://doi.org/10.1111/rssb.12348) | Sensitivity analysis for omitted variable bias — how large would an omitted variable need to be to overturn your result? |
| [Hodge & Austin (2004)](https://doi.org/10.1093/comjnl/47.1.85) | A readable survey of outlier and novelty detection methods across statistical and machine learning traditions |

**Useful Online Resources**

| Resource | What it covers |
|---|---|
| [R for Psychometrics](https://bookdown.org/sz_psyc490/r4psychometics/) | Introductory online textbook on psychometric modeling in R, including CFA and scale evaluation |
| [UCLA: Confirmatory Factor Analysis in R](https://stats.oarc.ucla.edu/r/seminars/rcfa/) | Step-by-step seminar on CFA using lavaan |
| [UCLA: Latent Growth Models in R](https://stats.oarc.ucla.edu/r/seminars/lgm/) | Tutorial on longitudinal SEM, measurement invariance testing, and growth curve models |

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 8,
  fig.height = 5,
  fig.path = "figs/",
  dev = "svg"
)

dir.create("figs", recursive = TRUE, showWarnings = FALSE)
```


## What This Tutorial Is About

There is a single idea running through everything in this tutorial:

> **Y is supposed to reflect X — but it actually reflects X *plus something else*.**

This problem shows up in six different parts that marketing researchers encounter regularly:

- **Part 1 — Discriminant Validity:** Your Green Purchase Intention scale is supposed to measure *purchase intention* (X) — but its items also pick up *environmental concern* (something else). The scale cannot be told apart from a different construct.

- **Part 2 — Omitted Variable Bias:** In your secondary data, observed *sales* (Y) is supposed to reflect *advertising spend* (X) — but sales also reflects *store quality* (something else, which you never measured). Your estimated advertising effect is inflated.

- **Part 3 — Measurement Invariance:** Your GPI items are supposed to reflect *green purchase intention* (X) across both experimental conditions — but two of the items shift upward in the treatment condition for reasons unrelated to the latent construct. The items don't mean the same thing in both groups.

- **Part 4 — Latent Subgroups:** Your sample contains two very different types of consumers — Green Champions and Price Skeptics — who respond to green marketing completely differently. When you aggregate across both groups, your outcome variable reflects a mixture of two different data-generating processes. The treatment effect you estimate doesn't apply cleanly to anyone.

- **Part 5 — Outliers:** Your dataset contains a cluster of luxury flagship stores that operate on completely different economics from regular stores. Including them distorts every coefficient in your model. Collective outliers are a version of the latent-subgroup problem; point outliers are a different statistical challenge. Both can be handled without deleting data.

- **Part 6 — Systematic Measurement Error:** Your outcome scale has hard bounds. Respondents near the ceiling cannot express upward variation; those near the floor cannot express downward variation. The observed mean is therefore a biased estimate of the true latent mean — not because of random noise, but because of the structure of the scale itself.

The statistical symptoms differ across these six parts, but the **underlying logic is identical**: your observed variable is contaminated by variance — or structurally constrained — in ways that prevent it from faithfully reflecting the latent construct or causal quantity you intend to study.

::: {.callout-note}
## Learning Objectives
By the end of this tutorial, you will be able to:

- Diagnose discriminant validity failures using HTMT and the DVI method (Pieters et al., 2025)
- Identify omitted variable bias through residual diagnostics and coefficient sensitivity analysis
- Test measurement invariance across groups using a configural → metric → scalar sequence in lavaan
- Use modification indices to pinpoint which specific items are non-invariant
- Use Gaussian mixture models (via `mclust`) to detect latent subgroups that distort your outcome
- Apply Local Outlier Factor (LOF) and DBSCAN to identify collective and point outliers
- Apply robust regression (MASS::rlm) as an alternative to deleting outliers
- Explain how bounded scales structurally bias observed means, and diagnose this using skewness tests on CTT residuals
- Explain the conceptual link between all six problems to a non-technical audience
:::


---


## Packages Needed

Install and load the required packages. If you haven't installed them before, run the `install.packages()` lines first.

```{r packages}
#| code-summary: "▶ Load required packages"
# Uncomment to install if needed:
# install.packages(c("lavaan", "semTools", "MASS", "ggplot2",
#                    "dplyr", "tidyr", "corrplot", "psych", "knitr", "lmtest",
#                    "mclust", "dbscan"))

library(lavaan)       # CFA and SEM (Parts 1 and 3)
library(semTools)     # htmt() and auxiliary SEM tools (Parts 1 and 3)
library(MASS)         # mvrnorm(): generate multivariate normal data (all parts)
library(ggplot2)      # Visualizations (all parts)
library(dplyr)        # Data manipulation (all parts)
library(tidyr)        # Data reshaping (Parts 2 and 3)
library(corrplot)     # Correlation heatmap (Part 1)
library(knitr)        # Nicely formatted tables (all parts)
library(lmtest)       # Breusch-Pagan heteroskedasticity test (Part 2)
library(mclust)       # Gaussian mixture models / latent class analysis (Part 4)
library(dbscan)       # Local outlier factor and DBSCAN clustering (Part 5)
```


---
