---
title: "Module 1 Code Tutorial"
subtitle: "Discriminant Validity, Omitted Variable Bias, Measurement Invariance, Latent Classes, and Outliers"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 8,
  fig.height = 5,
  fig.path = "figs/",
  dev = "svg"
)

dir.create("figs", recursive = TRUE, showWarnings = FALSE)
```

```{=html}
<style>
/* ── Floating bottom control bar ──────────────────────────────────────────── */
#global-output-toggle {
  position: fixed;
  bottom: 0;
  left: 0;
  right: 0;
  z-index: 9999;
  display: flex;
  flex-wrap: wrap;
  gap: 10px;
  justify-content: center;
  align-items: center;
  padding: 11px 24px;
  background: #1e3a5f;
  border-top: 3px solid #4a90d9;
  box-shadow: 0 -3px 16px rgba(0,0,0,0.35);
}
#global-output-toggle span {
  color: #c8d8ed;
  font-size: 0.85em;
  font-weight: 700;
  letter-spacing: 0.6px;
  text-transform: uppercase;
  margin-right: 6px;
}
/* Reserve space at page bottom so content never hides behind the bar */
body { padding-bottom: 70px !important; }

.gtbtn {
  padding: 9px 24px;
  color: #ffffff;
  border-radius: 7px;
  font-size: 1em;
  font-weight: 700;
  cursor: pointer;
  letter-spacing: 0.2px;
  transition: background 0.15s, border-color 0.15s;
  white-space: nowrap;
  border: 2px solid transparent;
}
/* Code button — blue */
.gtbtn.code-btn   { background: #2563a8; border-color: #5b9bd5; }
.gtbtn.code-btn:hover { background: #1a4a80; border-color: #90bfe8; }
/* Output button — teal/green */
.gtbtn.output-btn { background: #2d6a4f; border-color: #52b788; }
.gtbtn.output-btn:hover { background: #1b4332; border-color: #74c69d; }

/* ── Per-output collapsible wrappers ──────────────────────────────────────── */
details.out-fold { margin: 6px 0 14px 0; }
summary.out-fold-btn {
  display: inline-flex;
  align-items: center;
  gap: 7px;
  padding: 8px 20px;
  background: #eef4fb;
  border: 2px solid #4a90d9;
  border-radius: 7px;
  cursor: pointer;
  font-size: 0.94em;
  font-weight: 700;
  color: #1a3a6b;
  user-select: none;
  list-style: none;
  transition: background 0.12s;
}
summary.out-fold-btn::-webkit-details-marker { display: none; }
summary.out-fold-btn:hover { background: #cfe2f7; }
details[open].out-fold > summary.out-fold-btn { background: #d6ecff; }
summary.out-fold-btn .fold-icon { font-size: 1em; }
details[open] .fold-icon-closed { display: none; }
details:not([open]) .fold-icon-open { display: none; }
</style>

<script>
document.addEventListener('DOMContentLoaded', function () {

  /* ── 1. Wrap every cell output in a collapsible <details> ──────────────── */
  var outputSels = [
    '.cell-output-stdout',
    '.cell-output-stderr',
    '.cell-output-display'
  ];

  document.querySelectorAll(outputSels.join(',')).forEach(function (el) {
    if (el.closest('.callout') || el.closest('details.out-fold')) return;

    var hasImg = !!el.querySelector('img, svg, canvas');
    var hasTbl = !!el.querySelector('table');
    var label  = hasImg ? '📊  Show plot'
               : hasTbl ? '📋  Show table'
               :           '📄  Show output';

    var det = document.createElement('details');
    det.className = 'out-fold';

    var sum = document.createElement('summary');
    sum.className = 'out-fold-btn';
    sum.innerHTML =
      '<span class="fold-icon fold-icon-closed">▶</span>' +
      '<span class="fold-icon fold-icon-open">▼</span>' +
      label;

    el.parentNode.insertBefore(det, el);
    det.appendChild(sum);
    det.appendChild(el);
  });

  /* ── 2. Append the floating control bar to <body> ──────────────────────── */
  var bar = document.createElement('div');
  bar.id = 'global-output-toggle';
  bar.innerHTML =
    '<span>Quick controls:</span>' +
    '<button id="code-toggle-btn"   class="gtbtn code-btn"   onclick="toggleCode()">▼ Expand All Code</button>' +
    '<button id="output-toggle-btn" class="gtbtn output-btn" onclick="toggleOutput()">▼ Expand All Output</button>';

  document.body.appendChild(bar);
});

/* Track open/collapsed state separately for code and output */
var codeOpen   = false;
var outputOpen = false;

function toggleCode() {
  codeOpen = !codeOpen;
  document.querySelectorAll('details').forEach(function (d) {
    /* Code details contain a .sourceCode descendant; output folds carry class out-fold */
    if (!d.classList.contains('out-fold') && d.querySelector('.sourceCode')) {
      d.open = codeOpen;
    }
  });
  var btn = document.getElementById('code-toggle-btn');
  if (btn) btn.textContent = codeOpen ? '▲ Collapse All Code' : '▼ Expand All Code';
}

function toggleOutput() {
  outputOpen = !outputOpen;
  document.querySelectorAll('details.out-fold').forEach(function (d) {
    d.open = outputOpen;
  });
  var btn = document.getElementById('output-toggle-btn');
  if (btn) btn.textContent = outputOpen ? '▲ Collapse All Output' : '▼ Expand All Output';
}
</script>
```

## What This Tutorial Is About

There is a single idea running through everything in this tutorial:

> **Y is supposed to reflect X — but it actually reflects X *plus something else*.**

This problem shows up in five different guises that marketing researchers encounter regularly:

- **Case 1 — Discriminant Validity:** Your Green Purchase Intention scale is supposed to measure *purchase intention* (X) — but its items also pick up *environmental concern* (something else). The scale cannot be told apart from a different construct.

- **Case 2 — Omitted Variable Bias:** In your secondary data, observed *sales* (Y) is supposed to reflect *advertising spend* (X) — but sales also reflects *store quality* (something else, which you never measured). Your estimated advertising effect is inflated.

- **Case 3 — Measurement Non-Invariance:** Your GPI items are supposed to reflect *green purchase intention* (X) across both experimental conditions — but two of the items shift upward in the treatment condition for reasons unrelated to the latent construct. The items don't mean the same thing in both groups.

- **Case 4 — Latent Subgroups:** Your sample contains two very different types of consumers — Green Champions and Price Skeptics — who respond to green marketing completely differently. When you aggregate across both groups, your outcome variable reflects a mixture of two different data-generating processes. The treatment effect you estimate doesn't apply cleanly to anyone.

- **Case 5 — Outliers:** Your dataset contains a cluster of luxury flagship stores that operate on completely different economics from regular stores. Including them distorts every coefficient in your model. Collective outliers are a version of the latent-subgroup problem; point outliers are a different statistical challenge. Both can be handled without deleting data.

The statistical symptoms differ across these five cases, but the **underlying logic is identical**: your outcome variable is contaminated by variance that belongs to something other than the construct or cause you are trying to study.

::: {.callout-note}
## Learning Objectives
By the end of this tutorial, you will be able to:

- Diagnose discriminant validity failures using HTMT and the DVI method (Pieters et al., 2025)
- Identify omitted variable bias through residual diagnostics and coefficient sensitivity analysis
- Test measurement invariance across groups using a configural → metric → scalar sequence in lavaan
- Use modification indices to pinpoint which specific items are non-invariant
- Use Gaussian mixture models (via `mclust`) to detect latent subgroups that distort your outcome
- Apply Local Outlier Factor (LOF) and DBSCAN to identify collective and point outliers
- Apply robust regression (MASS::rlm) as an alternative to deleting outliers
- Explain the conceptual link between all five problems to a non-technical audience
:::


---


## Packages Needed

Install and load the required packages. If you haven't installed them before, run the `install.packages()` lines first.

```{r packages}
#| code-summary: "▶ Load required packages"
# Uncomment to install if needed:
# install.packages(c("lavaan", "semTools", "MASS", "ggplot2",
#                    "dplyr", "tidyr", "corrplot", "knitr", "lmtest",
#                    "mclust", "dbscan"))

library(lavaan)       # CFA and SEM (Cases 1 and 3)
library(semTools)     # htmt() and auxiliary SEM tools (Cases 1 and 3)
library(MASS)         # mvrnorm(): generate multivariate normal data (all cases)
library(ggplot2)      # Visualizations (all cases)
library(dplyr)        # Data manipulation (all cases)
library(tidyr)        # Data reshaping (Cases 2 and 3)
library(corrplot)     # Correlation heatmap (Case 1)
library(knitr)        # Nicely formatted tables (all cases)
library(lmtest)       # Breusch-Pagan heteroskedasticity test (Case 2)
library(mclust)       # Gaussian mixture models / latent class analysis (Case 4)
library(dbscan)       # Local outlier factor and DBSCAN clustering (Case 5)
```


---


# Case 1: Discriminant Validity {#sec-case1}

## The Study Scenario

We are simulating data from a **green marketing experiment** with 400 participants. The study was designed to answer: *does exposure to eco-friendly brand messaging increase consumers' green purchase intentions?*

**Experimental design:**

- Half the participants saw standard product advertising (**Control**)
- Half saw the same ads emphasising eco-friendly credentials (**Green Marketing**)

**Constructs measured (all on 7-point Likert scales):**

| Construct | Abbreviation | # Items | Role |
|---|---|---|---|
| Environmental Concern | EC | 4 | Covariate |
| Green Purchase Intention | GPI | 4 | Dependent Variable |
| Brand Attitude | BA | 3 | Covariate |

**The hidden problem we planted in the data:** The GPI scale was designed to measure purchase intention — but its items ended up being nearly indistinguishable from items measuring environmental concern. In real research, this often happens when constructs are conceptually close (caring about the environment vs. intending to buy green products). The two scales fail discriminant validity.

In all five cases in this tutorial, the core problem is the same: the **observed Y** you are working with is an imperfect, contaminated proxy for the **latent Y** you are trying to study.

::: {.callout-warning}
## What "failing discriminant validity" means here
The GPI (DV) scale picks up variation from *two* latent constructs — Green Purchase Intention *and* Environmental Concern — rather than cleanly measuring just one. This means any effect of the marketing campaign on GPI is hard to interpret: are participants more willing to buy, or just more environmentally concerned?
:::

::: {.callout-important}
## Remember: These methods only apply in specific situations

Discriminant validity is *always* a concern whenever you have two related constructs in your study — but the methods below (HTMT and DVI) only work in a specific setting:

- **You need multi-item scales** (e.g., 4-item Likert scales for GPI and EC)
- **You need at least 4 Likert-type items per construct** — fewer items and the statistics become unreliable
- **These methods do NOT apply to** single-item outcomes, willingness-to-pay (WTP), choice data, behavioral measures (e.g., actual purchases), or any non-Likert response format

**If your DV is WTP or a behavioral measure, you cannot use HTMT or DVI to check discriminant validity — but that doesn't mean discriminant validity isn't a problem! It just means the problem is harder to detect.**
:::


---


## Simulating the Data

We use `MASS::mvrnorm()` to generate correlated latent factor scores, then create item scores from those factors plus random measurement error. The items are rounded to a 7-point Likert scale.

**The key feature we're building in:** Environmental Concern (EC) and Green Purchase Intention (GPI) are correlated at **0.93** in the population — extremely high, and well above the 0.90 threshold that signals a discriminant validity problem.

```{r simulate-data}
#| code-summary: "▶ Simulate green marketing dataset (n=400)"
set.seed(2025)
n <- 400

# ── Step 1: Assign participants to conditions ──────────────────────────────────
treatment <- sample(c(0L, 1L), size = n, replace = TRUE)

# ── Step 2: Define the true factor correlation matrix ─────────────────────────
# EC and GPI are correlated at .93 — too high for discriminant validity
# EC–BA and GPI–BA are at more typical, moderate levels
phi_pop <- matrix(
  c(1.00, 0.93, 0.45,   # EC row
    0.93, 1.00, 0.50,   # GPI row
    0.45, 0.50, 1.00),  # BA row
  nrow = 3, byrow = TRUE,
  dimnames = list(c("EC", "GPI", "BA"), c("EC", "GPI", "BA"))
)

# ── Step 3: Generate latent factor scores ─────────────────────────────────────
latent_base <- MASS::mvrnorm(n = n, mu = c(0, 0, 0), Sigma = phi_pop)

EC_lat  <- latent_base[, 1]
GPI_lat <- latent_base[, 2] + 0.40 * treatment  # Green marketing raises GPI by .40 SD
BA_lat  <- latent_base[, 3]

# ── Step 4: Define item loadings ──────────────────────────────────────────────
# These represent how strongly each item reflects its latent construct
lambda_EC  <- c(0.78, 0.82, 0.74, 0.76)
lambda_GPI <- c(0.80, 0.76, 0.82, 0.78)
lambda_BA  <- c(0.72, 0.76, 0.70)

# ── Step 5: Generate continuous item scores (latent score + measurement error) ─
gen_items <- function(latent, loadings) {
  sapply(loadings, function(lam) {
    lam * latent + sqrt(1 - lam^2) * rnorm(length(latent))
  })
}

EC_cont  <- gen_items(EC_lat,  lambda_EC)
GPI_cont <- gen_items(GPI_lat, lambda_GPI)
BA_cont  <- gen_items(BA_lat,  lambda_BA)

# ── Step 6: Round to 7-point Likert scale ─────────────────────────────────────
# Cut the continuous distribution into 7 ordered categories
to_likert7 <- function(x) {
  z <- (x - mean(x)) / sd(x)   # standardise each item
  breaks <- c(-Inf, -1.5, -0.75, -0.25, 0.25, 0.75, 1.5, Inf)
  as.integer(cut(z, breaks = breaks, labels = 1:7))
}

EC_lik  <- apply(EC_cont,  2, to_likert7)
GPI_lik <- apply(GPI_cont, 2, to_likert7)
BA_lik  <- apply(BA_cont,  2, to_likert7)

# ── Step 7: Assemble the final data frame ─────────────────────────────────────
df <- data.frame(
  id        = 1:n,
  condition = factor(treatment, levels = c(0, 1),
                     labels = c("Control", "Green Marketing")),
  EC1  = EC_lik[, 1], EC2  = EC_lik[, 2],
  EC3  = EC_lik[, 3], EC4  = EC_lik[, 4],
  GPI1 = GPI_lik[, 1], GPI2 = GPI_lik[, 2],
  GPI3 = GPI_lik[, 3], GPI4 = GPI_lik[, 4],
  BA1  = BA_lik[, 1], BA2  = BA_lik[, 2], BA3  = BA_lik[, 3]
)

# Quick look at the data
head(df, 5)
```

```{r describe-data}
#| code-summary: "▶ Descriptive statistics by condition"
# Count participants per condition and compute scale means
df |>
  group_by(condition) |>
  summarise(
    n          = n(),
    Mean_EC    = round(rowMeans(across(EC1:EC4))  |> mean(), 2),
    Mean_GPI   = round(rowMeans(across(GPI1:GPI4)) |> mean(), 2),
    Mean_BA    = round(rowMeans(across(BA1:BA3))  |> mean(), 2)
  ) |>
  kable(col.names = c("Condition", "N", "Mean EC", "Mean GPI", "Mean BA"),
        caption = "Sample sizes and scale means by condition")
```

::: {.callout-note}
The green marketing manipulation works as intended: participants in the green marketing condition score higher on GPI (Green Purchase Intention) than those in the control condition.
:::


---


## First Look: Convergent and Discriminant Validity in the Raw Data

Before running any formal tests, let's look at the raw correlation structure of the 11 items. This heatmap tells you two things at once:

- **Convergent validity**: Items within the same scale should correlate *strongly* with each other. If they don't, the items aren't all measuring the same construct.
- **Discriminant validity**: Items from *different* scales should correlate *noticeably less* than items within the same scale. If they don't, the two scales cannot be told apart.

A healthy pattern: dark red within-scale blocks (strong convergent validity) and lighter between-scale blocks (good discriminant validity). The warning sign: when the between-scale block looks just as dark as the within-scale blocks.

```{r correlation-heatmap, fig.height = 7, fig.width = 7}
#| code-summary: "▶ Plot: item correlation heatmap"
# Extract just the item columns (no ID or condition)
items_df <- df |> select(EC1:BA3)

# Compute correlation matrix
item_cors <- cor(items_df)

# Color-coded heatmap
# Warm colors = high positive correlation; cool colors = low/negative
corrplot(item_cors,
         method   = "color",
         type     = "lower",
         order    = "original",     # keep original order so EC, GPI, BA stay grouped
         tl.col   = "black",
         tl.cex   = 0.85,
         addCoef.col = "white",     # print correlation values
         number.cex  = 0.65,
         col      = colorRampPalette(c("#313695", "#74add1", "#e0f3f8",
                                       "#fee090", "#f46d43", "#a50026"))(200),
         title    = "Item Correlation Matrix",
         mar      = c(0, 0, 1.5, 0))
```

::: {.callout-warning}
## What to look for in the heatmap
Compare the **red blocks** in the heatmap:

- The EC–EC block (top-left): high correlations ✓ *(items measure the same thing)*
- The GPI–GPI block (middle): high correlations ✓
- The **EC–GPI block** (the rectangle connecting EC and GPI items): *also* very high correlations ⚠️

When between-scale correlations (EC–GPI) are as strong as within-scale correlations (EC–EC, GPI–GPI), the two scales cannot be told apart. This is the discriminant validity problem in visual form.
:::

The table below makes the problem even clearer by summarizing the average within-scale versus between-scale correlations. Think of it this way:

- **Within-scale** (e.g., EC1 with EC2, EC3, EC4): These are correlations among items that are supposed to be measuring the *same* thing. They should be high — that's convergent validity working.
- **Between-scale** (e.g., EC1 with GPI1, GPI2, GPI3, GPI4): These are correlations among items from *different* constructs. They should be noticeably *lower* than the within-scale correlations. If they're not, the two scales are too similar to distinguish — that's a discriminant validity failure.

```{r htmt-illustration}
#| code-summary: "▶ Compute average within- and between-scale correlations"
# Compute average correlations within each scale and between scales
cors <- cor(items_df)

within_EC  <- mean(cors[1:4, 1:4][lower.tri(cors[1:4, 1:4])])
within_GPI <- mean(cors[5:8, 5:8][lower.tri(cors[5:8, 5:8])])
within_BA  <- mean(cors[9:11, 9:11][lower.tri(cors[9:11, 9:11])])

between_EC_GPI <- mean(abs(cors[1:4, 5:8]))
between_EC_BA  <- mean(abs(cors[1:4, 9:11]))
between_GPI_BA <- mean(abs(cors[5:8, 9:11]))

avg_cor_summary <- data.frame(
  Type  = c("Within EC",  "Within GPI", "Within BA",
            "Between EC–GPI ⚠️", "Between EC–BA", "Between GPI–BA"),
  Avg_r = round(c(within_EC, within_GPI, within_BA,
                  between_EC_GPI, between_EC_BA, between_GPI_BA), 3)
)
kable(avg_cor_summary, col.names = c("Correlation Type", "Average |r|"),
      caption = "Average within-scale and between-scale correlations")
```

::: {.callout-important}
## What the numbers are telling you

Look at the table and ask yourself: *How different are the within-scale and between-scale numbers for EC and GPI?*

Imagine you're a researcher who has no idea the data have a problem. You see:
- EC items correlate with each other at around, say, **0.68**
- GPI items correlate with each other at around **0.67**
- But EC items and GPI items correlate with each other at around **0.65**

Those three numbers are nearly identical. That means knowing someone's EC score tells you almost as much about their GPI items as their own GPI scores do. The two scales are effectively measuring the same thing, just with different item wording. That's the discriminant validity failure, visible in plain numbers before you run any formal test.

The HTMT ratio formalises exactly this comparison. If within-scale and between-scale correlations are similar, HTMT will be close to 1.0.
:::


---


## Method 1: HTMT Analysis

### What HTMT Measures

HTMT stands for **Heterotrait-Monotrait ratio of correlations** — which sounds technical, but the idea is simple:

- **Monotrait** = correlations between items from the *same* scale (within-scale)
- **Heterotrait** = correlations between items from *different* scales (between-scale)

The HTMT ratio asks: *"How large are the between-scale correlations relative to the within-scale correlations?"* If this ratio is close to 1.0, your scales are not distinguishable. If it's clearly below 1.0, the scales capture different things.

The threshold rules of thumb:

- **HTMT < 0.85** → discriminant validity supported (strict)
- **HTMT < 0.90** → discriminant validity supported (lenient)
- **HTMT ≥ 0.90** → discriminant validity **violated**

We will use **HTMT2**, the updated version of the index that uses the geometric mean of within-scale correlations. This version is more accurate and is now recommended over the original.

### Running the HTMT Analysis

::: {.callout-note}
## Data preparation for `htmt()`

**What `htmt()` needs:** A data frame containing *only the scale items* — no participant IDs, no condition variable, no demographics.

```r
items_df <- df |> select(EC1:BA3)   # drop id and condition
```

If you accidentally include non-item columns, the function will try to treat them as scale items and give you meaningless results.
:::

```{r htmt-analysis}
#| code-summary: "▶ Run HTMT2 analysis"
# ── Model specification ────────────────────────────────────────────────────────
# This tells htmt() which items belong to which construct.
# It uses the same syntax as lavaan's CFA model specification.
cfa_model <- '
  EC  =~ EC1 + EC2 + EC3 + EC4
  GPI =~ GPI1 + GPI2 + GPI3 + GPI4
  BA  =~ BA1  + BA2  + BA3
'

# ── Run the HTMT analysis ──────────────────────────────────────────────────────
htmt_result <- semTools::htmt(
  model    = cfa_model,   # factor structure: which items belong to which scale
  data     = items_df,    # item-level data ONLY (no id, no condition variable)
  htmt2    = TRUE,        # use HTMT2 (geometric mean) — recommended
  absolute = TRUE         # use absolute values of correlations
)

print(htmt_result)
```

::: {.callout-note}
## Key arguments to pay attention to

| Argument | What it does | What to set |
|---|---|---|
| `model` | Specifies which items belong to which construct | Same as your CFA model syntax |
| `data` | The raw item data | Items only — no ID or grouping variables |
| `htmt2` | Switches between original and updated HTMT formula | Set `TRUE` for HTMT2 (recommended) |
| `absolute` | Whether to take absolute values of correlations | Keep `TRUE` (the default) |
| `missing` | How to handle missing data | `"listwise"` is fine for most cases |
:::

### Visualizing the HTMT Results

A table of numbers is fine, but a chart makes it much easier to see which pairs are problematic. The bar chart below marks the HTMT thresholds for easy comparison.

```{r htmt-visualization, fig.height = 5}
#| code-summary: "▶ Plot: HTMT2 bar chart with thresholds"
# Extract the HTMT matrix and reshape for plotting
htmt_mat <- as.matrix(htmt_result)

# semTools fills only one triangle; we need a robust helper to extract a pair
get_htmt <- function(mat, r, c) {
  val <- mat[r, c]
  if (is.na(val)) val <- mat[c, r]   # try the other triangle
  as.numeric(val)
}

htmt_pairs <- data.frame(
  pair  = c("EC ↔ GPI", "EC ↔ BA", "GPI ↔ BA"),
  htmt  = c(get_htmt(htmt_mat, "EC", "GPI"),
            get_htmt(htmt_mat, "EC", "BA"),
            get_htmt(htmt_mat, "GPI", "BA")),
  label = c("Focal problem pair", "OK", "OK")
)

ggplot(htmt_pairs, aes(x = reorder(pair, -htmt), y = htmt, fill = label)) +
  geom_col(width = 0.55, colour = "white") +
  # Threshold lines
  geom_hline(yintercept = 0.90, linetype = "dashed",
             colour = "firebrick", linewidth = 0.9) +
  geom_hline(yintercept = 0.85, linetype = "dotted",
             colour = "darkorange", linewidth = 0.9) +
  # Value labels on bars
  geom_text(aes(label = round(htmt, 3)), vjust = -0.5,
            fontface = "bold", size = 4) +
  # Threshold annotations
  annotate("text", x = 3.45, y = 0.915,
           label = "Strict threshold (0.90)", colour = "firebrick",
           size = 3.3, hjust = 1) +
  annotate("text", x = 3.45, y = 0.865,
           label = "Lenient threshold (0.85)", colour = "darkorange",
           size = 3.3, hjust = 1) +
  scale_fill_manual(values = c("Focal problem pair" = "#d73027",
                               "OK"                 = "#4575b4"),
                    name = NULL) +
  scale_y_continuous(limits = c(0, 1.08), expand = c(0, 0)) +
  labs(
    x        = NULL,
    y        = "HTMT Value",
    title    = "HTMT2 Analysis: Discriminant Validity Check",
    subtitle = "Bars above the dashed line indicate discriminant validity concerns"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom",
        panel.grid.major.x = element_blank())
```

### Interpreting the HTMT Results

::: {.callout-warning}
## Reading the chart

- **EC ↔ GPI**: The HTMT value exceeds both thresholds (0.85 and 0.90). This is a clear discriminant validity violation. The GPI scale cannot be statistically distinguished from the EC scale.

- **EC ↔ BA** and **GPI ↔ BA**: Both well below the thresholds. Brand Attitude discriminates fine from the other two constructs.

**Conclusion from HTMT:** The GPI (our DV) and EC scales are too similar. Before drawing any conclusions about the marketing campaign's effect on GPI, we need to address this problem — or at least be transparent about it.
:::

**What HTMT doesn't tell you:** HTMT gives you a single number per pair and a rule-of-thumb threshold. It does not account for *how reliable* your scales are, and it doesn't provide a formal statistical decision with confidence intervals around the DVI estimate. That's where the Pieters et al. (2025) method comes in.


---


## Method 2: The DVI Method (Pieters et al., 2025)

### Why We Need Something More

The HTMT is a great screening tool, but it has two gaps:

1. **No formal inference**: The 0.85 and 0.90 thresholds are rules of thumb, not statistically principled tests.
2. **Ignores scale reliability**: A pair of highly reliable scales *should* be able to discriminate even when their factor correlation is somewhat high. HTMT doesn't account for this.

Pieters et al. (2025) propose the **Discriminant Validity Index (DVI)**, which directly tests whether a scale's reliability is high enough — relative to the factor correlation — to support discriminant validity. The method follows a clear two-step decision procedure.

### The Two-Step Logic

**Step 1 — The Phi Test:** Is the factor correlation (φ) meaningfully less than 1.0?

$$\text{DVI}_1 = 1 - |\phi|$$

If DVI₁ is significantly greater than zero (the confidence interval doesn't include zero), the two constructs aren't perfectly correlated and at least *some* distinction exists. If DVI₁ isn't significant, stop: discriminant validity has failed.

**Step 2 — The CR Test:** Even if φ < 1, is the scale reliability high enough to "rise above" the factor correlation?

$$\text{DVI}_{CR} = \sqrt{CR} - |\phi|$$

where CR is the **Congeneric Reliability** of the scale (sometimes called McDonald's omega). If √CR > |φ| (i.e., DVI_CR > 0 and its CI excludes zero), the scale has enough internal consistency to distinguish itself from the other construct.

::: {.callout-note}
## An intuitive way to think about Step 2
Imagine φ = 0.93 (constructs are highly correlated) and CR = 0.83 (pretty reliable scale). Then √CR ≈ 0.91. The question is: can √CR (0.91) rise above φ (0.93)? Here, 0.91 < 0.93 — so DVI_CR is *negative*. The scale isn't reliable enough to pull ahead of the factor correlation. Discriminant validity fails.

The intuition: reliability sets an upper bound on how distinctly a scale can behave. If that upper bound is itself lower than the factor correlation, the scale is fundamentally compromised.
:::

### Step 1: Specifying the CFA Model with Defined Parameters

The key insight from Pieters et al. (2025) is to define CR and DVI *inside* the lavaan model syntax using the `:=` operator. This lets lavaan automatically compute these values and their standard errors, giving us confidence intervals for free.

::: {.callout-note}
## Data preparation for the CFA

**Same as HTMT:** The CFA model should be fit on the item-level data only. The `condition` variable is not part of the measurement model.

Additionally, you **must label** every loading and every error variance in the model. These labels are needed to compute CR and DVI using `:=`. Without them, lavaan won't know which parameters to use in the formulas.
:::

```{r dvi-model-spec}
#| code-summary: "▶ Specify the labeled CFA/DVI model"
dvi_model <- '
  # Measurement model
  EC  =~ lam_EC1*EC1 + lam_EC2*EC2 + lam_EC3*EC3 + lam_EC4*EC4
  GPI =~ lam_GPI1*GPI1 + lam_GPI2*GPI2 + lam_GPI3*GPI3 + lam_GPI4*GPI4
  BA  =~ lam_BA1*BA1 + lam_BA2*BA2 + lam_BA3*BA3

  # Error variances
  EC1  ~~ th_EC1*EC1
  EC2  ~~ th_EC2*EC2
  EC3  ~~ th_EC3*EC3
  EC4  ~~ th_EC4*EC4

  GPI1 ~~ th_GPI1*GPI1
  GPI2 ~~ th_GPI2*GPI2
  GPI3 ~~ th_GPI3*GPI3
  GPI4 ~~ th_GPI4*GPI4

  BA1  ~~ th_BA1*BA1
  BA2  ~~ th_BA2*BA2
  BA3  ~~ th_BA3*BA3

  # Factor correlations
  EC  ~~ phi_EC_GPI*GPI
  EC  ~~ phi_EC_BA*BA
  GPI ~~ phi_GPI_BA*BA

  # Congeneric reliability
  CR_EC  := ((lam_EC1 + lam_EC2 + lam_EC3 + lam_EC4)^2) / (((lam_EC1 + lam_EC2 + lam_EC3 + lam_EC4)^2) + th_EC1 + th_EC2 + th_EC3 + th_EC4)
  CR_GPI := ((lam_GPI1 + lam_GPI2 + lam_GPI3 + lam_GPI4)^2) / (((lam_GPI1 + lam_GPI2 + lam_GPI3 + lam_GPI4)^2) + th_GPI1 + th_GPI2 + th_GPI3 + th_GPI4)
  CR_BA  := ((lam_BA1 + lam_BA2 + lam_BA3)^2) / (((lam_BA1 + lam_BA2 + lam_BA3)^2) + th_BA1 + th_BA2 + th_BA3)

  # DVI
  DVI_1      := 1 - sqrt(phi_EC_GPI^2)
  DVI_CR_EC  := sqrt(CR_EC) - sqrt(phi_EC_GPI^2)
  DVI_CR_GPI := sqrt(CR_GPI) - sqrt(phi_EC_GPI^2)
'
```

::: {.callout-note}
## Key features of the model specification to pay attention to

| Feature | Why it matters |
|---|---|
| **Labels on every loading** (`lam_EC1*EC1`) | Required to compute CR with `:=` |
| **Labels on every error variance** (`th_EC1*EC1`) | Required to compute CR with `:=` |
| **Factor correlations labelled** (`phi_EC_GPI*GPI`) | The phi label feeds into DVI formulas |
| **`sqrt(phi^2)` instead of `phi`** | Takes the absolute value (in case phi is negative) |
| **`:=` operator** | Defines derived parameters; lavaan computes their SEs automatically |
:::

### Step 2: Fitting the CFA

```{r cfa-fit}
#| code-summary: "▶ Fit CFA and get Wald confidence intervals"
# Fit the CFA using the labelled model
# std.lv = TRUE: fixes factor variances to 1, so phi values ARE correlations
fit_wald <- cfa(
  model  = dvi_model,
  data   = items_df,
  std.lv = TRUE      # ← IMPORTANT: without this, phi is not a correlation
)

# Check that the model converged and fits reasonably
summary(fit_wald, fit.measures = TRUE, standardized = TRUE)
```

::: {.callout-note}
## `std.lv = TRUE` — why it matters

When you set `std.lv = TRUE`, lavaan fixes the **variance of each latent factor to 1**. This means the factor covariance between EC and GPI is directly interpretable as a **correlation** (values between –1 and +1). Without this, the raw covariance would be in an arbitrary scale, and computing DVI from it would not be valid.

**Always use `std.lv = TRUE` when applying the DVI method.**
:::

### Step 3: Extracting and Interpreting the Wald Confidence Intervals

The `summary()` output above already contains the DVI values and their Wald confidence intervals. The code below pulls out just the parameters we care about.

```{r dvi-wald}
#| code-summary: "▶ Extract DVI estimates (Wald CIs)"
# Extract parameter estimates with 95% Wald confidence intervals
pe_wald <- parameterEstimates(fit_wald, level = 0.95)

# Filter to the derived parameters (DVI, CR, and the factor correlation)
key_params <- c("phi_EC_GPI", "CR_EC", "CR_GPI",
                "DVI_1", "DVI_CR_EC", "DVI_CR_GPI")

dvi_wald_table <- pe_wald |>
  filter(label %in% key_params) |>
  select(label, est, se, ci.lower, ci.upper, pvalue) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))

kable(
  dvi_wald_table,
  col.names = c("Parameter", "Estimate", "SE", "CI Lower", "CI Upper", "p-value"),
  caption   = "DVI Results with Wald 95% Confidence Intervals"
)
```

::: {.callout-note}
## How to read this table

Work through the rows in order:

- **phi_EC_GPI**: The estimated correlation between the EC and GPI latent factors. Our simulation set this at 0.93, so you should see something close to that. Values close to 1.0 are the core of the problem.

- **CR_EC / CR_GPI**: The reliability of each scale — essentially, how well the four items hang together. A value of, say, 0.84 means √CR ≈ 0.92. That 0.92 needs to exceed the factor correlation (0.93) for Step 2 to pass. When the factor correlation is that high, even a decent reliability of 0.84 isn't enough.

- **DVI_1**: Step 1. Estimate is `1 − |0.93| = 0.07`. The question is whether this tiny gap is statistically reliable (CI excludes zero). If the CI includes zero, the constructs are statistically indistinguishable from perfect overlap.

- **DVI_CR_EC / DVI_CR_GPI**: Step 2. Estimate is `√CR − 0.93`. If √CR is around 0.91 and the factor correlation is 0.93, you get `0.91 − 0.93 = −0.02` — a negative number, meaning the scale's reliability doesn't reach above the factor correlation. Discriminant validity fails.

A **positive** DVI value with a **CI that doesn't cross zero** = evidence for discriminant validity.
A **near-zero or negative** DVI value, or one whose **CI crosses zero** = discriminant validity is not supported.
:::

### Step 4: Bootstrap Confidence Intervals

Wald confidence intervals rely on asymptotic normality — they can be slightly off, especially for parameters that are constrained (like DVI_CR which can't go below –1). Bootstrap confidence intervals make no such assumptions and are more trustworthy.

::: {.callout-note}
## How bootstrap CIs work (in plain language)

Instead of a mathematical formula for the CI, we repeatedly resample the data (with replacement), refit the CFA each time, and compute DVI. After doing this thousands of times, we look at the 2.5th and 97.5th percentiles of the DVI values we got. Those percentiles form our 95% CI.

This is slower to compute but more honest about uncertainty.
:::

```{r dvi-bootstrap, cache=FALSE}
#| code-summary: "▶ Re-fit with bootstrap SEs (~60 sec)"
# Fit the model again with bootstrap standard errors
# bootstrap = 1000 is a reasonable minimum; use 2000+ for final reporting
set.seed(31415927)
fit_boot <- cfa(
  model     = dvi_model,
  data      = items_df,
  std.lv    = TRUE,
  se        = "bootstrap",   # switch from Wald to bootstrap
  bootstrap = 1000           # number of bootstrap resamples
)

# Extract parameter estimates with percentile bootstrap CIs
pe_boot <- parameterEstimates(
  fit_boot,
  boot.ci.type = "perc",   # percentile method (most commonly recommended)
  level        = 0.95
)

dvi_boot_table <- pe_boot |>
  filter(label %in% key_params) |>
  select(label, est, ci.lower, ci.upper) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))

kable(
  dvi_boot_table,
  col.names = c("Parameter", "Estimate", "Bootstrap CI Lower", "Bootstrap CI Upper"),
  caption   = "DVI Results with Percentile Bootstrap 95% Confidence Intervals (B = 1,000)"
)
```

::: {.callout-note}
## Bootstrap vs. Wald CIs — which to report?
In practice, report both. The Wald CIs are faster to compute and good for a first check. Use bootstrap CIs (with B ≥ 2,000) for the final analysis you report in a paper. If they tell the same story, you can be confident in your conclusions.
:::

### Step 5: Visualizing the DVI Results

A forest plot makes it easy to see all three DVI values and whether their CIs cross zero.

```{r dvi-forest-plot, fig.height = 5}
#| code-summary: "▶ Plot: DVI forest plot"
# Build a plotting data frame from the bootstrap results
dvi_plot_df <- dvi_boot_table |>
  filter(label %in% c("DVI_1", "DVI_CR_EC", "DVI_CR_GPI")) |>
  mutate(
    label_nice = case_when(
      label == "DVI_1"      ~ "DVI₁: Phi criterion\n(1 – |φ|)",
      label == "DVI_CR_EC"  ~ "DVI_CR (EC): CR criterion\n(√CR_EC – |φ|)",
      label == "DVI_CR_GPI" ~ "DVI_CR (GPI): CR criterion\n(√CR_GPI – |φ|)"
    ),
    supported = ci.lower > 0   # TRUE if entire CI is above zero
  )

ggplot(dvi_plot_df, aes(x = est, y = reorder(label_nice, est), colour = supported)) +
  # CI bar
  geom_segment(aes(x = ci.lower, xend = ci.upper,
                   y = reorder(label_nice, est), yend = reorder(label_nice, est)),
               linewidth = 1.5) +
  # Point estimate
  geom_point(size = 4) +
  # Zero line (the null: no discriminant validity)
  geom_vline(xintercept = 0, linetype = "dashed", colour = "black", linewidth = 0.8) +
  # Colour: green = supported, red = not supported
  scale_colour_manual(values = c("TRUE"  = "#2c7bb6",
                                 "FALSE" = "#d7191c"),
                      labels = c("TRUE"  = "DV supported (CI > 0)",
                                 "FALSE" = "DV NOT supported (CI ≤ 0)"),
                      name   = NULL) +
  labs(
    x        = "DVI Estimate (with 95% Bootstrap CI)",
    y        = NULL,
    title    = "DVI Forest Plot: EC ↔ GPI Pair",
    subtitle = "If the CI bar crosses the dashed line (0), discriminant validity is not supported"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank())
```

### Step 6: The Decision Procedure

Pieters et al. (2025) propose a structured two-step decision procedure. Work through it in order:

```{r decision-table}
#| code-summary: "▶ Two-step decision table"
# Determine outcomes based on bootstrap CIs
dvi_results <- dvi_boot_table |>
  filter(label %in% c("DVI_1", "DVI_CR_EC", "DVI_CR_GPI")) |>
  mutate(
    significant = ci.lower > 0,
    verdict     = if_else(significant, "Supported ✓", "Not supported ✗")
  ) |>
  select(label, est, ci.lower, ci.upper, verdict)

kable(
  dvi_results,
  col.names = c("DVI Metric", "Estimate", "95% CI Lower", "95% CI Upper", "Verdict"),
  caption   = "Two-Step Decision Summary for the EC–GPI Pair"
)
```

::: {.callout-warning}
## Applying the two-step decision rule

**Step 1 — DVI₁ (Phi criterion):** Is DVI₁ significantly > 0?

- If **YES**: The factor correlation is meaningfully less than 1.0. Some distinction exists. Proceed to Step 2.
- If **NO**: Discriminant validity has completely failed. EC and GPI are statistically indistinguishable — you cannot treat them as separate constructs in your analysis.

**Step 2 — DVI_CR (CR criterion):** Is DVI_CR significantly > 0 for BOTH scales?

- If DVI_CR > 0 for **both scales**: Discriminant validity is fully supported. The scales are reliable enough to rise above the factor correlation.
- If DVI_CR > 0 for **only one scale**: The problem is asymmetric. One scale is reliable enough to distinguish itself; the other isn't. That weaker scale is where your attention should focus — its items may be too conceptually similar to the other construct.
- If DVI_CR ≤ 0 for **both scales**: Even though the constructs aren't perfectly correlated (Step 1 passed), neither scale is reliable enough to pull ahead of the factor correlation. In practical terms: you cannot draw clean conclusions about which construct your DV is measuring.

**In our data:** Because we set the EC–GPI correlation at 0.93, you should see Step 1 barely passing (or failing), and Step 2 failing for at least one scale — likely both. The scales are simply too correlated to be treated as measuring distinct constructs, regardless of how the items were worded.
:::


---


## Comparing the Two Methods

### Side-by-Side Summary

```{r comparison-table}
#| code-summary: "▶ HTMT vs. DVI side-by-side summary"
# Assemble phi and HTMT values for the comparison table
phi_est     <- round(dvi_boot_table$est[dvi_boot_table$label == "phi_EC_GPI"], 3)
phi_ci_lo   <- round(dvi_boot_table$ci.lower[dvi_boot_table$label == "phi_EC_GPI"], 3)
phi_ci_hi   <- round(dvi_boot_table$ci.upper[dvi_boot_table$label == "phi_EC_GPI"], 3)
htmt_val    <- round(get_htmt(htmt_mat, "EC", "GPI"), 3)

comparison <- data.frame(
  Feature         = c("What it tests",
                      "Key output",
                      "Threshold / decision rule",
                      "Accounts for reliability?",
                      "Formal statistical test?",
                      "EC ↔ GPI result"),
  HTMT            = c("Between-scale vs. within-scale correlations",
                      paste0("HTMT2 = ", htmt_val),
                      "< 0.85 (strict) or < 0.90 (lenient)",
                      "No",
                      "No — rule of thumb only",
                      if_else(htmt_val >= 0.90, "FAILED ✗ (≥ 0.90)", "OK ✓")),
  `DVI (Pieters et al., 2025)` = c("Factor correlation and scale reliability",
                      paste0("φ = ", phi_est, " [", phi_ci_lo, ", ", phi_ci_hi, "]"),
                      "DVI > 0 with CI excluding zero",
                      "Yes — uses Congeneric Reliability",
                      "Yes — bootstrap CIs",
                      "See DVI table above")
)

kable(
  comparison,
  col.names = c("Feature", "HTMT", "DVI (Pieters et al., 2025)"),
  caption   = "Comparing the two discriminant validity methods"
)
```

### What to Do When Discriminant Validity Fails

If HTMT and DVI both flag the same pair of constructs, you have a genuine problem. Here are your options:

1. **Reconceptualise**: Are EC and GPI genuinely different constructs, or have you (like many sustainability researchers) been treating two facets of the same underlying construct as if they were distinct?

2. **Revise the scale**: Remove or rewrite GPI items that are too close in meaning to EC items. Re-collect data.

3. **Combine the scales**: If the constructs truly cannot be separated, consider treating them as a single broader construct (e.g., "pro-environmental orientation").

4. **Report transparently**: If none of the above is feasible, at minimum report the discriminant validity failure and discuss what it implies for interpreting your effects.

::: {.callout-important}
## The core take-home message

A marketing effect that appears on a GPI scale which fails discriminant validity from EC is not clearly interpretable. You cannot claim the campaign increased *purchase intention* if the scale measuring "purchase intention" is statistically indistinguishable from a scale measuring *environmental concern*. The effect could reflect either or both. Discriminant validity testing is not optional — it is a prerequisite for meaningful inference about relationships between constructs.
:::

## Other Methods for Assessing Discriminant Validity

The HTMT and DVI approaches covered here are the current best practice for multi-item Likert scales, but other methods exist:

- **Average Variance Extracted (AVE) criterion (Fornell & Larcker, 1981):** Compare the square root of each construct's AVE against its inter-construct correlations. Widely used but has known limitations — the criterion is often too lenient, and AVE itself is sensitive to the number of items. The Fornell–Larcker criterion was the dominant approach before HTMT.
- **Confirmatory Factor Analysis (CFA) model comparison:** Fit a constrained model where φ = 1 (constructs are the same) and compare fit to the unconstrained model using a likelihood ratio test. A significant difference supports discriminant validity.
- **Exploratory Factor Analysis (EFA):** In earlier-stage scale development, EFA can reveal whether items from two intended scales load on separate factors or cross-load, signaling discriminant validity problems before formal CFA.
- **Multitrait-multimethod (MTMM) analysis (Campbell & Fiske, 1959):** The classical approach — measure multiple traits using multiple methods and examine the pattern of convergent and discriminant correlations. Resource-intensive but gold standard for construct validation.
- **Network psychometrics:** Gaussian graphical models (GGMs) can reveal the partial correlation structure among items, making discriminant validity failures visible as dense cross-scale connections.


---


# Case 2: Omitted Variable Bias — A Related Problem in Observational Data {#sec-case2}

## Connecting to Case 1

In Case 1, the GPI scale was contaminated: instead of purely measuring *Green Purchase Intention*, it also picked up *Environmental Concern*. The scale measured **X + something else**.

Omitted variable bias (OVB) is a related — but distinct — problem. It occurs when you use observational (secondary) data to estimate the effect of one variable on another, but a third variable you didn't measure is influencing both.

Suppose you have sales figures and advertising spend for 500 stores. You want to know: *does more advertising lead to more sales?*

The danger is that **store quality** — the physical location, the staff, the store atmosphere — affects *both* how much stores spend on advertising (premium stores invest more) *and* how much they sell. If you leave store quality out of your model, your estimate of advertising's effect on sales is contaminated. The advertising coefficient picks up **advertising + store quality**.

::: {.callout-important}
## How OVB relates to discriminant validity

These are two different problems with an important connection:

- **Not all OVB is a discriminant validity violation.** OVB occurs in regression with any kind of outcome variable — sales figures, behavioral counts, stock prices. Discriminant validity failures specifically affect survey scales with multiple items. You can have OVB without any measurement issue at all.

- **But all OVB produces a discriminant validity violation — if your outcome is a scale.** If store quality is omitted from your model and your "sales" variable were a multi-item Likert scale (e.g., a customer satisfaction scale that is itself influenced by store quality), that scale would fail discriminant validity tests with the advertising construct. The omitted variable would inflate the latent correlation.

When OVB is present, your outcome variable Y effectively becomes a composite of two separate latent processes — the true causal effect you care about (advertising → sales) and the confounding pathway (quality → sales). This means Y no longer has a single, clean meaning. The fitted values Ŷ from your naive regression end up "measuring" a mixture of advertising effects and quality effects — just as a scale that fails discriminant validity measures a mixture of two constructs. That is why every case of OVB produces what would register as a discriminant validity violation if you tried to measure your outcome with a scale: the observed Y is a composite of two distinct latent data-generating processes, not a pure indicator of one.

| | What we think Y reflects | What Y actually reflects |
|---|---|---|
| **Case 1** | Green Purchase Intention (GPI) | GPI *and* Environmental Concern |
| **Case 2** | Effect of advertising on sales | Advertising *and* store quality |
:::

## The Scenario

A retail analytics team has secondary data on 500 stores: monthly advertising spend (in standardised units) and monthly sales (a standardised index). The research question is simple: *does advertising spend predict sales?*

What they cannot observe is **store quality** — a latent construct reflecting location desirability, staff quality, and store environment. Store quality correlates with advertising spend (premium stores can afford both) and strongly drives sales independently.

## Simulating the Observational Dataset

```{r case2-simulate}
#| code-summary: "▶ Simulate observational store dataset (n=500)"
set.seed(2025)
n_obs <- 500

# ── Store quality: continuous, unobserved ─────────────────────────────────────
# Positive = premium, negative = budget. Never in our dataset.
U_quality <- rnorm(n_obs)

# ── Advertising spend: strongly correlated with store quality ──────────────────
# Premium stores invest much more — this is the source of OVB.
X_ads <- 4 * U_quality + rnorm(n_obs, sd = 1.5)

# ── Sales: driven by advertising (small true effect) AND store quality (large) ─
# TRUE advertising coefficient: 0.20 (small)
# Store quality drives sales very strongly, AND creates a dramatic fan pattern
Y_sales <- 0.20 * X_ads + 5.00 * U_quality +
           rnorm(n_obs, sd = 0.3 + 1.5 * abs(U_quality))

# ── Store size: a rough proxy for quality we might observe in secondary data ──
store_size <- 1.5 * U_quality + rnorm(n_obs, sd = 1.8)

# ── Observed dataset (quality is hidden; store_size is sometimes available) ───
obs_df <- data.frame(
  advertising = X_ads,
  sales       = Y_sales,
  store_size  = store_size,
  quality     = U_quality    # kept here only for the diagnostic "reveal" plots
)
```

## Step 1: The Naive Analysis

We run the regression that most analysts would reach for first.

```{r case2-naive-reg}
#| code-summary: "▶ Naive OLS: advertising predicts sales"
# Naive regression — advertising is our only predictor
m_naive <- lm(sales ~ advertising, data = obs_df)
summary(m_naive)
```

::: {.callout-warning}
## What jumps out?
The advertising coefficient looks large and statistically highly significant. A researcher without knowledge of the data-generating process might conclude: "advertising has a strong effect on sales." But the **true** advertising coefficient is only **0.20**. The naive OLS estimate will be somewhere around **1.30** — more than six times the true value. Almost all of that estimated "effect" is actually store quality sneaking into the coefficient.

Think of it this way: in our simulation, premium stores happen to advertise more AND sell more — not because advertising drives sales, but because store quality drives both. OLS cannot tell the difference between "advertising caused the sales" and "quality-driven stores also happened to advertise more." The coefficient absorbs both.
:::

## Step 2: Diagnosing the Problem Through Residuals

A residual plot is one of the first places to look when you suspect something is wrong. If the model is correctly specified, the residuals should be scattered randomly around zero — no patterns, no fan shapes.

::: {.callout-important}
## Always plot your residuals — before anything else

This is not optional. Before running any formal diagnostic tests, plot your residuals against fitted values. This single habit will catch the vast majority of model misspecification problems — omitted variables, non-linearities, heteroskedasticity — that formal tests can miss or that you might not know to test for. Make residual plotting the first step in any regression workflow, not an afterthought.
:::

```{r case2-resid-plot-naive, fig.height = 4.5}
#| code-summary: "▶ Plot: residuals vs. fitted (naive model)"
obs_df <- obs_df |>
  mutate(fitted_naive = fitted(m_naive),
         resid_naive  = residuals(m_naive))

ggplot(obs_df, aes(x = fitted_naive, y = resid_naive)) +
  geom_point(alpha = 0.25, colour = "#4575b4") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "black") +
  geom_smooth(method = "loess", se = FALSE,
              colour = "firebrick", linewidth = 1) +
  labs(x     = "Fitted Values (predicted sales)",
       y     = "Residuals",
       title = "Residual Plot: Naive Model (Advertising Only)",
       subtitle = "A fanning pattern signals non-constant variance — something is missing") +
  theme_minimal(base_size = 12)
```

::: {.callout-warning}
## What the fan shape tells you
The residuals spread out dramatically as fitted values increase. This **heteroskedasticity** is a red flag: if the variance of errors is not constant, it usually means the model has left out something important that also varies with X.

Here, that missing variable is store quality. High-advertising stores tend to be premium stores, and premium stores have far more variable sales (luxury flagships have big swings; budget stores are consistently mediocre). The fan shape in the residual plot is the visual signature of that missing variable.

Connect this back to Case 1: just as the GPI scale was picking up Environmental Concern alongside purchase intention, our sales variable here is picking up store quality alongside advertising effectiveness. The residual plot is doing the job that HTMT and DVI did in Case 1 — telling you that Y reflects more than X.
:::

We can formally confirm the heteroskedasticity with a Breusch-Pagan test. A significant result means the variance of residuals is not constant — further evidence that something is systematically missing.

```{r case2-bp-test}
#| code-summary: "▶ Breusch-Pagan test for heteroskedasticity"
# Breusch-Pagan test: H0 = residuals have constant variance (homoskedasticity)
lmtest::bptest(m_naive)
```

Now let's do the "oracle reveal": colour the same residual plot by store quality to show where the heteroskedasticity comes from.

```{r case2-resid-reveal, fig.height = 4.5}
#| code-summary: "▶ Plot: residuals colored by true store quality"
# Cut quality into tertiles for colour legend clarity
obs_df <- obs_df |>
  mutate(quality_group = cut(quality,
                              breaks = quantile(quality, c(0, 1/3, 2/3, 1)),
                              labels = c("Low quality", "Medium", "High quality"),
                              include.lowest = TRUE))

ggplot(obs_df, aes(x = fitted_naive, y = resid_naive, colour = quality_group)) +
  geom_point(alpha = 0.35) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "black") +
  scale_colour_manual(
    values = c("Low quality"  = "#4575b4",
               "Medium"       = "#91bfdb",
               "High quality" = "#d73027"),
    name = "Store quality\n(hidden in real data)"
  ) +
  labs(x     = "Fitted Values",
       y     = "Residuals",
       title = "The Same Plot — Coloured by Store Quality",
       subtitle = "The fan pattern is entirely explained by the omitted variable U") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

::: {.callout-important}
## The detective logic
The fan pattern in the first plot told us *something* was wrong. The second plot (the "oracle reveal" — in practice you can't always do this) shows that the heteroskedasticity is driven entirely by the omitted variable. High-quality stores (red) cluster at the top right. Low-quality stores (blue) cluster at the bottom left. Store quality is confounding both advertising spend and sales.
:::

## Step 3: The Sensitivity Test

In real research you rarely have access to the true omitted variable. But you may have a **proxy** — a variable that correlates with the unobserved confounder even if it doesn't perfectly capture it. Here, store size is such a proxy.

The key diagnostic: **if adding a potential proxy to the model changes your main coefficient substantially, that proxy was correlated with both X and Y — and something in that proxy's space was biasing your original estimate.**

```{r case2-sensitivity}
#| code-summary: "▶ Coefficient sensitivity: adding a proxy control"
# Model with store size proxy
m_proxy  <- lm(sales ~ advertising + store_size, data = obs_df)

# Oracle model (impossible in real life — store quality is unobserved)
m_oracle <- lm(sales ~ advertising + quality, data = obs_df)

# Summary comparison table
coef_compare <- data.frame(
  Model = c("Naive (advertising only)",
            "With store-size proxy",
            "Oracle (true quality included)"),
  `Advertising coefficient` = round(c(coef(m_naive)["advertising"],
                                      coef(m_proxy)["advertising"],
                                      coef(m_oracle)["advertising"]), 3),
  `True value` = c(0.20, 0.20, 0.20)
)

kable(coef_compare,
      col.names = c("Model", "Advertising Coefficient", "True Value"),
      caption   = "Coefficient Sensitivity: How Much Does Adding a Proxy Change Things?")
```

::: {.callout-warning}
## How to interpret the sensitivity test
If your advertising coefficient **drops substantially** when you add a control variable, that control was correlated with both advertising and sales — a sign it was a partial proxy for an omitted confounder. The larger the drop, the more worried you should be about the original estimate.

If the coefficient barely moves, that control was probably not picking up a meaningful confounder (though it doesn't fully rule out other unmeasured variables).

**Rule of thumb:** A coefficient that changes by more than 10–15% when you add a theoretically motivated control deserves serious scrutiny. A coefficient that is cut in half or more (as in our simulation) is a clear warning of omitted variable bias.
:::

## Summary: What OVB Teaches Us

OVB and discriminant validity violations are different problems — OVB lives in the regression, DV violations live in the measurement model — but they share a common underlying logic. In both cases, the thing you are using as your outcome variable picks up more than one source of variance, and the contaminating source inflates the relationship you're trying to study.

Discriminant validity checks (Case 1) and residual/sensitivity diagnostics (Case 2) are both ways of asking the same fundamental question: *is my Y truly measuring what I think it's measuring, or is something else riding along with it?*

The key practical difference: if your Y is a multi-item scale and you have OVB (an omitted variable driving both X and Y), that omitted variable will inflate the latent correlation between your scales and cause them to fail discriminant validity tests too. So fixing the measurement model and fixing the regression model are not alternative strategies — you may need to address both.

The residual fan pattern you see here is also a hint that latent subgroups (Case 4) or collective outliers (Case 5) may be present — all three problems can produce heteroskedastic-looking residuals.

## Other Methods for Diagnosing Endogeneity and OVB

The residual diagnostics and proxy controls covered here are entry-level approaches. For more rigorous identification:

- **Instrumental variables (IV) / Two-Stage Least Squares (2SLS):** Find a variable that predicts X but does not directly affect Y (an "instrument"). Use it to isolate the exogenous variation in X. Powerful but requires strong theoretical justification for the instrument's validity.
- **Regression discontinuity design (RDD):** If assignment to X (or a confounding condition) is based on a threshold cutoff, the discontinuity creates a locally exogenous variation you can exploit. Requires the right data structure but provides clean causal identification.
- **Difference-in-differences (DiD):** Compare changes over time in treated vs. control units. Controls for time-invariant confounders. Requires parallel trends assumption.
- **Propensity score methods:** Model the probability of "treatment" (high advertising) as a function of observed covariates, then match or weight observations. Controls for observed confounders; does not handle unobserved ones.
- **Hausman test:** Formally tests whether OLS and IV estimates differ significantly. If they do, OLS is inconsistent (endogeneity is present). Requires a valid instrument.
- **Sensitivity analysis (Oster, 2019):** How large would the omitted variable's effect need to be (relative to included controls) to fully explain away the estimated coefficient? Quantifies the "how worried should I be?" question without requiring an instrument.
- **Panel data fixed effects:** If you have repeated observations per unit (stores over time), within-unit fixed effects remove all time-invariant confounders — including unobserved store quality.


---


# Case 3: Measurement Non-Invariance — A Special Case of Discriminant Validity {#sec-case3}

## Connecting to Cases 1 and 2

Cases 1 and 2 both showed how Y captures more than X. Case 3 is actually a **special case of the discriminant validity problem from Case 1** — it just appears in an experimental context.

Here's the connection: in Case 1, the GPI scale failed discriminant validity because it was contaminated by a *different construct* (Environmental Concern). In Case 3, the GPI scale can fail for a more subtle reason — the items behave differently across experimental groups. Two GPI items might score higher in the treatment condition not because participants genuinely intend to purchase more, but because the green marketing exposure makes those specific items feel more socially desirable or salient, independent of any real change in purchase intention.

When that happens, "GPI" no longer measures the same construct across groups. Within each group the scale may be fine; but across groups, the same item is measuring subtly different things. This is a discriminant validity failure in disguise — the measurement model is picking up "item-specific sensitivity to the marketing treatment" alongside "genuine purchase intention."

And if your sample contains latent subgroups (Case 4), the items may appear non-invariant simply because the two groups are at different levels of the construct — what looks like item shift might actually be group-level differences in the distribution of the latent construct itself.

::: {.callout-note}
## Two kinds of shift to distinguish

- **Item shift**: A specific item's baseline level changes across groups, independent of any change in the latent construct. Example: *"I always choose the most eco-friendly option available"* might score higher after green marketing exposure simply because it sounds more socially appropriate — not because purchase intention truly increased. This is a *measurement problem*. It's a bias in the ruler itself.

- **Construct shift**: The latent construct itself genuinely has a different mean across groups. This is what we *want* to detect with our experiment — a real change in Green Purchase Intention. This is a substantive finding.

**Item shift always accompanies construct shift when non-invariance is present.** That is, if your scale is non-invariant, any real construct shift will be mixed in with the spurious item shift — and you can't separate them without formal testing.
:::

| | What produces the group difference in Y | How it shows up |
|---|---|---|
| **Case 1** | GPI and EC overlap as constructs | High factor correlation; HTMT > 0.90 |
| **Case 2** | Omitted store quality confounds X and Y | Fan-shaped residuals; coefficient instability |
| **Case 3** | GPI items shift across conditions | Scalar model fails; mod. indices flag specific items |

## Two Scenarios to Compare

We will simulate **two versions** of the data to make the distinction crystal clear:

- **Scenario A — Item shift only**: The true GPI construct does NOT change due to the treatment. GPI1 and GPI2 score higher in the treatment group purely because of item-level response shift. Any apparent treatment effect in the data is entirely artifactual.

- **Scenario B — Item shift + construct shift**: The true GPI construct DOES change (by 0.40 SD). GPI1 and GPI2 also shift at the item level. The total observed difference in GPI items overestimates the real treatment effect.

Both scenarios show what non-invariance looks like. Scenario A is the extreme case where none of the observed treatment effect is real. Scenario B is more common in practice — there is a real effect, but it gets inflated.

## Simulating the Two Scenarios

```{r case3-simulate}
#| code-summary: "▶ Simulate two non-invariance scenarios (A: item shift only; B: item + construct shift)"
set.seed(2025)
n3     <- 400
treat3 <- sample(c(0L, 1L), n3, replace = TRUE)

# ── Shared factor structure (same phi_pop as Case 1) ──────────────────────────
latent3  <- MASS::mvrnorm(n3, mu = c(0, 0, 0), Sigma = phi_pop)
EC_lat3  <- latent3[, 1]
BA_lat3  <- latent3[, 3]

lam_GPI3  <- c(0.80, 0.76, 0.82, 0.78)
item_shift <- c(0.80, 0.80, 0.00, 0.00)   # GPI1 and GPI2 shift; GPI3/GPI4 do not

# ── EC and BA: fully invariant across scenarios ────────────────────────────────
EC3_items <- gen_items(EC_lat3, c(0.78, 0.82, 0.74, 0.76))
BA3_items <- gen_items(BA_lat3, c(0.72, 0.76, 0.70))

# ── Scenario A: item shift ONLY — true latent GPI is unchanged by treatment ───
GPI_lat3a <- latent3[, 2]           # NO construct shift
GPI3a <- sapply(seq_along(lam_GPI3), function(i) {
  lam_GPI3[i] * GPI_lat3a +
    sqrt(1 - lam_GPI3[i]^2) * rnorm(n3) +
    item_shift[i] * treat3
})

df3a <- data.frame(
  condition = factor(treat3, labels = c("Control", "Green Marketing")),
  EC1 = EC3_items[,1], EC2 = EC3_items[,2], EC3 = EC3_items[,3], EC4 = EC3_items[,4],
  GPI1 = GPI3a[,1], GPI2 = GPI3a[,2], GPI3 = GPI3a[,3], GPI4 = GPI3a[,4],
  BA1 = BA3_items[,1], BA2 = BA3_items[,2], BA3 = BA3_items[,3]
)

# ── Scenario B: item shift + construct shift — TRUE effect = 0.40 SD ──────────
GPI_lat3b <- latent3[, 2] + 0.40 * treat3   # construct shift
GPI3b <- sapply(seq_along(lam_GPI3), function(i) {
  lam_GPI3[i] * GPI_lat3b +
    sqrt(1 - lam_GPI3[i]^2) * rnorm(n3) +
    item_shift[i] * treat3
})

df3b <- data.frame(
  condition = factor(treat3, labels = c("Control", "Green Marketing")),
  EC1 = EC3_items[,1], EC2 = EC3_items[,2], EC3 = EC3_items[,3], EC4 = EC3_items[,4],
  GPI1 = GPI3b[,1], GPI2 = GPI3b[,2], GPI3 = GPI3b[,3], GPI4 = GPI3b[,4],
  BA1 = BA3_items[,1], BA2 = BA3_items[,2], BA3 = BA3_items[,3]
)
```

## Visualising Item Shift vs. Construct Shift

The key visual diagnostic: compare the mean difference (treatment minus control) for each GPI item across both scenarios. If the treatment effect is a **pure construct shift**, all four items should show roughly the same difference (proportional to their loadings). If some items show *disproportionately large* differences, that excess is item shift.

```{r case3-item-means, fig.height = 5.5, fig.width = 9}
#| code-summary: "▶ Plot: item mean differences across the two scenarios"
# Compute item mean differences for both scenarios
get_item_diffs <- function(df, scenario_label) {
  df |>
    group_by(condition) |>
    summarise(across(GPI1:GPI4, mean), .groups = "drop") |>
    pivot_longer(cols = GPI1:GPI4, names_to = "item", values_to = "mean_score") |>
    pivot_wider(names_from = condition, values_from = mean_score) |>
    mutate(
      difference = `Green Marketing` - Control,
      scenario   = scenario_label,
      type       = if_else(item %in% c("GPI1", "GPI2"),
                           "Non-invariant (item shift)",
                           "Invariant")
    )
}

diffs_a <- get_item_diffs(df3a, "Scenario A: Item shift only\n(true treatment effect = 0)")
diffs_b <- get_item_diffs(df3b, "Scenario B: Item shift + construct shift\n(true treatment effect = 0.40 SD)")
all_diffs <- bind_rows(diffs_a, diffs_b)

# Expected construct-only difference: 0.40 * loading for items 3 & 4 (scenario B)
expected_b <- 0.40 * mean(lam_GPI3[3:4])

ggplot(all_diffs, aes(x = item, y = difference, fill = type)) +
  geom_col(width = 0.6, colour = "white") +
  geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.5) +
  facet_wrap(~ scenario) +
  scale_fill_manual(
    values = c("Non-invariant (item shift)" = "#d73027",
               "Invariant"                  = "#4575b4"),
    name = NULL
  ) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  labs(
    x        = "GPI Item",
    y        = "Mean Difference (Treatment − Control)",
    title    = "Visualising Item Shift vs. Construct Shift",
    subtitle = "Blue bars show only real construct shift; red bars show extra item-level inflation"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold", size = 10))
```

::: {.callout-warning}
## What to look for across the two panels

**Scenario A (left panel):** GPI3 and GPI4 show near-zero differences — correct, because the true treatment effect is zero. GPI1 and GPI2 show substantial positive differences — entirely due to item shift. If you averaged all four items and compared group means, you would conclude there is a treatment effect. There isn't. The apparent effect is entirely measurement artifact.

**Scenario B (right panel):** GPI3 and GPI4 show moderate differences reflecting the real 0.40 SD construct shift. GPI1 and GPI2 show that same moderate shift *plus* the extra item shift on top. The total observed difference for GPI1/GPI2 is inflated well above what the latent shift would predict.

In both cases, item-shifted items (red) show disproportionately large treatment–control differences compared to invariant items (blue). This asymmetry is the fingerprint of measurement non-invariance.
:::

## Testing Measurement Invariance Step by Step

The formal test uses a sequence of nested CFA models, each adding one more constraint. We'll demonstrate using **Scenario B** (item shift + construct shift), which is the more common real-world situation.

::: {.callout-note}
## The three models and what they constrain — connected to item/construct shift

| Model | What is constrained | What it tests | Connection to shifts |
|---|---|---|---|
| **Configural** | Nothing (baseline) | Same factor structure in both groups? | If configural **fails**, the factor structure itself differs across groups — implying **both item shift and construct shift** are present (the most severe case) |
| **Metric** | Factor loadings equal | Do items relate to the construct with the same *sensitivity* across groups? | If metric **fails** (but configural holds), loadings differ — items have different sensitivity to the latent construct across groups — this is **item shift** (the ruler has different gradations) |
| **Scalar** | Loadings + item intercepts equal | Do items have the same *baseline level* across groups? | If scalar **fails** (but metric holds), intercepts differ — items have different baseline levels — this is **item baseline shift** (pure item shift, the cleanest case to diagnose) |

Failing configural invariance is the most severe outcome — it means the factor structure itself is not replicable across groups, implying both the constructs and the items are behaving differently. Failing metric invariance (while configural holds) means loadings differ — items have different sensitivity to the latent construct in each group. This is a form of item shift: the ruler itself has different gradations. Failing scalar invariance (while metric holds) means item intercepts differ — items have different baseline levels in each group. This is the classic item-level shift we built into our simulation. **You need scalar invariance at minimum to make valid comparisons of latent means across groups.**
:::

```{r case3-models, message=FALSE}
#| code-summary: "▶ Fit configural, metric, and scalar invariance models"
invar_model <- '
  EC  =~ EC1 + EC2 + EC3 + EC4
  GPI =~ GPI1 + GPI2 + GPI3 + GPI4
  BA  =~ BA1  + BA2  + BA3
'

# Fit the three-model sequence on Scenario B
fit_config <- cfa(invar_model, data = df3b, group = "condition", std.lv = TRUE)
fit_metric <- cfa(invar_model, data = df3b, group = "condition",
                  group.equal = "loadings", std.lv = TRUE)
fit_scalar <- cfa(invar_model, data = df3b, group = "condition",
                  group.equal = c("loadings", "intercepts"), std.lv = TRUE)
```

```{r case3-model-comparison}
#| code-summary: "▶ Likelihood ratio test sequence"
# Likelihood ratio test: significant Δχ² = adding the constraint worsens fit
lrt_result <- lavTestLRT(fit_config, fit_metric, fit_scalar)
print(lrt_result)
```

```{r case3-fit-table}
#| code-summary: "▶ Model fit indices table"
extract_fit <- function(fit, label) {
  fi <- fitmeasures(fit, c("chisq", "df", "cfi", "rmsea", "srmr"))
  data.frame(
    Model = label,
    Chisq = round(as.numeric(fi["chisq"]), 1),
    df    = as.integer(fi["df"]),
    CFI   = round(as.numeric(fi["cfi"]),   3),
    RMSEA = round(as.numeric(fi["rmsea"]), 3),
    SRMR  = round(as.numeric(fi["srmr"]),  3)
  )
}

fit_summary <- rbind(
  extract_fit(fit_config, "Configural"),
  extract_fit(fit_metric, "Metric"),
  extract_fit(fit_scalar, "Scalar (full)")
)
rownames(fit_summary) <- NULL

kable(fit_summary,
      col.names = c("Model", "χ²", "df", "CFI", "RMSEA", "SRMR"),
      caption   = "Model Fit Across the Invariance Sequence (Scenario B)")
```

::: {.callout-warning}
## Reading the results

- **Configural → Metric**: The χ² difference should be non-significant (p > .05). This means item sensitivity (loadings) is the same across groups. The items relate to GPI equally strongly in both conditions. ✓

- **Metric → Scalar**: The χ² difference **should be significant** (p < .05), because we built non-invariant intercepts into GPI1 and GPI2. Full scalar invariance fails. This is the formal evidence of item shift. ✗

When full scalar invariance fails, you cannot validly compare GPI scale means across conditions without first identifying *which* items are the problem.
:::

## Identifying the Non-Invariant Items

Modification indices tell you which constrained parameters, if freed, would most improve model fit. For a failed scalar model, focus on the **intercepts** (`op == "~1"`): the items with the largest modification indices are the non-invariant ones.

```{r case3-modindices}
#| code-summary: "▶ Modification indices — find non-invariant items"
mi_scalar     <- modindices(fit_scalar, sort. = TRUE)
mi_intercepts <- mi_scalar |>
  filter(op == "~1") |>
  arrange(desc(mi)) |>
  select(lhs, op, block, mi, epc)

kable(
  head(mi_intercepts, 10),
  col.names = c("Item", "Type", "Group block", "Mod. Index (MI)", "Expected Change"),
  digits    = 2,
  caption   = "Top Modification Indices for Item Intercepts (Scalar Model, Scenario B)"
)
```

::: {.callout-note}
## How to read the modification indices

Each row is a constrained parameter. The MI column tells you approximately how much the model χ² would drop if you freed that parameter. Large MI values for GPI item intercepts (op == "~1") point to items with different baseline levels across groups — i.e., items that have shifted.

GPI1 and GPI2 should appear at the top of this list. In real data, these are the items you would examine substantively: *why might these specific items behave differently after green marketing exposure, independently of any real change in purchase intention?*

::: {.callout-important}
## When can you drop non-invariant items?

If you **pre-registered your measurement invariance analysis** before collecting data, you may be able to drop the flagged items (GPI1, GPI2) and use a shortened scale — but only if at least two items remain invariant per factor and you pre-specified this decision rule. Dropping items *after* finding non-invariance without pre-registration is post-hoc and will raise reviewers' eyebrows.

If the analysis was not pre-registered, the better approach is partial invariance (see below).
:::
:::

## Fitting the Partial Invariance Model

Once you identify the non-invariant items, you free their intercepts and re-fit. This is called **partial scalar invariance**: most intercepts are constrained (equal across groups), but the problematic ones are freed.

```{r case3-partial}
#| code-summary: "▶ Fit partial scalar invariance model"
fit_partial <- cfa(invar_model, data = df3b, group = "condition",
                   group.equal   = c("loadings", "intercepts"),
                   group.partial = c("GPI1~1", "GPI2~1"),
                   std.lv = TRUE)

lrt_partial <- lavTestLRT(fit_metric, fit_partial, fit_scalar)
print(lrt_partial)
```

```{r case3-partial-fit}
#| code-summary: "▶ Full fit comparison (configural through full scalar)"
fit_summary2 <- rbind(
  extract_fit(fit_config,  "Configural"),
  extract_fit(fit_metric,  "Metric"),
  extract_fit(fit_partial, "Partial scalar (GPI1, GPI2 freed)"),
  extract_fit(fit_scalar,  "Full scalar")
)
rownames(fit_summary2) <- NULL
kable(fit_summary2,
      col.names = c("Model", "χ²", "df", "CFI", "RMSEA", "SRMR"),
      caption   = "Model Fit Including Partial Invariance Model (Scenario B)")
```

::: {.callout-note}
## What partial invariance means in practice
- **Partial vs. metric**: non-significant → freeing GPI1 and GPI2 intercepts is sufficient to restore acceptable fit. The measurement model is now accounting for item shift. ✓
- **Partial vs. full scalar**: significant → full invariance genuinely fails; GPI1 and GPI2 were necessary to free.

With partial scalar invariance, you *can* still compare latent means across groups — but you must be transparent about which items showed item shift and offer a theoretical explanation for why.
:::

## Separating Item Shift from Construct Shift

Here is the payoff. We compare the estimated latent GPI mean difference (treatment − control) under three models:

```{r case3-latent-means}
#| code-summary: "▶ Compare latent mean estimates across models"
get_lv_mean <- function(fit, lv = "GPI") {
  pe     <- parameterEstimates(fit)
  lv_row <- pe[pe$op == "~1" & pe$lhs == lv & pe$group == 2, ]
  if (nrow(lv_row) == 0) return(NA_real_)
  lv_row$est[1]
}

lv_scalar  <- get_lv_mean(fit_scalar)
lv_partial <- get_lv_mean(fit_partial)

latent_comparison <- data.frame(
  Model = c("Full scalar (ignores item shift — biased)",
            "Partial scalar (accounts for item shift — corrected)",
            "True value — known from Scenario B simulation"),
  GPI_latent_mean = round(c(lv_scalar, lv_partial, 0.40), 3)
)

kable(latent_comparison,
      col.names = c("Model", "Estimated GPI Latent Mean Difference (Treatment vs. Control)"),
      caption   = "How Non-Invariance Inflates the Estimated Treatment Effect")
```

::: {.callout-important}
## The core lesson: item shift inflates construct shift estimates

The full scalar model forces GPI1 and GPI2 to behave as if they're invariant. Since they're not, their elevated scores in the treatment group get attributed to a larger latent GPI shift — inflating the estimated treatment effect above 0.40.

The partial invariance model correctly absorbs the item-level shifts into the freed intercepts, and the estimated latent mean difference recovers to something close to the true 0.40.

**This is the discriminant validity connection:** the scale in the treatment group is partially measuring "item-specific response to green marketing" alongside "genuine purchase intention" — exactly the same "Y measures X + something else" problem as Case 1. Measurement invariance testing is how you detect it in an experimental setting.
:::

## Other Methods for Testing Measurement Invariance

The configural → metric → scalar sequence using lavaan is the standard approach for continuous (or near-continuous) items. Other options:

- **WLSMV estimator (lavaan):** For truly ordinal Likert data (5–7 points), use `estimator = "WLSMV"` in lavaan with polychoric correlations. The model comparison sequence is the same; the fit statistics and chi-square tests adjust for the ordinal scale.
- **Alignment optimization (Asparouhov & Muthén, 2014):** An alternative to the step-wise test sequence for large numbers of groups. Instead of testing constraints, it finds the optimal alignment of factor means and intercepts with minimum non-invariance. Useful when you have many groups (countries, time points) and expect some non-invariance.
- **Bayesian approximate measurement invariance (Muthén & Asparouhov, 2013):** Instead of exact equality constraints, allows parameters to be "approximately equal" across groups. Fits naturally in a Bayesian framework and gives more nuanced information about the degree of non-invariance.
- **Item Response Theory (IRT) / Differential Item Functioning (DIF):** In educational and clinical measurement, non-invariance is called differential item functioning. Logistic IRT models can flag specific items and parameter levels where DIF occurs. Particularly useful for binary or polytomous items.
- **Random intercept cross-lagged panel model (RI-CLPM):** For longitudinal data, this model separates within-person change from stable between-person differences, which is related to the question of whether your scale means the same thing for everyone.


---


# Case 4: Latent Subgroups — When Your Sample Contains Two Different Worlds {#sec-case4}

## Connecting to the Earlier Cases

All three previous cases involved a Y variable contaminated by something extra. Case 4 presents the same problem from a different angle: what if your sample itself is a mixture of two fundamentally different groups of people who respond to your study variables in completely different ways?

Imagine your 600 study participants come from two populations:

- **Green Champions** (~60% of the sample): Highly environmentally concerned, willing to pay premium prices for eco-friendly products. They respond strongly to the green marketing campaign.
- **Price Skeptics** (~40% of the sample): Price-driven, less concerned with environmental credentials. They barely respond to the same campaign.

If you analyse the full sample as if everyone is the same type of consumer, you get an estimate of the "average" treatment effect. But that average is the mathematical midpoint between two very different responses — it may not accurately describe *either* group. Your GPI outcome now measures a mixture of "how Green Champions responded" and "how Price Skeptics responded." That is, once again, Y reflects more than X.

This pooling-driven inflation of correlations is the same mechanism that produces apparent discriminant validity violations (Case 1): when you have two groups sitting in different positions on both EC and GPI, the pooled correlation between them will be much higher than within-group correlations — and may push you past the HTMT threshold.

::: {.callout-note}
## The connection to discriminant validity

When you pool two latent classes in a single analysis, the between-class variance inflates the apparent correlations among your variables. Green Champions are high on EC *and* high on GPI; Price Skeptics are low on both. Even if within each class EC and GPI were distinguishable, the pooled correlation may be much higher — and could push you into discriminant validity failure territory.

The latent class problem and the discriminant validity problem are therefore linked: latent subgroups can *cause* apparent discriminant validity violations in pooled data that would disappear if you analyzed each class separately.
:::

## Simulating the Two-Class Dataset

We need more variables than in earlier cases for the class-detection algorithm to work reliably. We include:

- **EC** — 3-item Environmental Concern scale
- **WTP** — 3-item Willingness to Pay Premium scale
- **PS** — 3-item Price Sensitivity scale (reversed relative to WTP)
- **GPI** — 4-item Green Purchase Intention (DV)

```{r case4-simulate}
#| code-summary: "▶ Simulate two-class consumer dataset (n=600)"
set.seed(2025)
n4   <- 600

# ── Assign latent class membership ────────────────────────────────────────────
# 60% Green Champions, 40% Price Skeptics
true_class <- sample(c("Green Champion", "Price Skeptic"),
                     size = n4, replace = TRUE,
                     prob = c(0.60, 0.40))

# ── Define class-specific latent factor means ─────────────────────────────────
# Green Champions: high EC, high WTP, low PS
# Price Skeptics: low EC, low WTP, high PS
# Centred within class (class differences are large)

class_means <- list(
  "Green Champion" = c(EC = 1.2,  WTP =  1.0, PS = -1.0, GPI =  1.0),
  "Price Skeptic"  = c(EC = -0.8, WTP = -0.7, PS =  0.8, GPI = -0.7)
)

# ── Generate latent scores: common factor structure, different means ───────────
lam_ec  <- c(0.78, 0.75, 0.80)
lam_wtp <- c(0.76, 0.80, 0.74)
lam_ps  <- c(0.72, 0.75, 0.70)
lam_gpi <- c(0.80, 0.76, 0.82, 0.78)

gen_class_items <- function(n, lam, class_mean, class_labels) {
  # latent scores with class-specific mean
  latent <- ifelse(class_labels == "Green Champion",
                   rnorm(n, mean = class_mean["Green Champion"], sd = 0.8),
                   rnorm(n, mean = class_mean["Price Skeptic"],  sd = 0.8))
  sapply(lam, function(l) l * latent + sqrt(1 - l^2) * rnorm(n))
}

EC_4  <- gen_class_items(n4, lam_ec,  c("Green Champion" = 1.2,  "Price Skeptic" = -0.8), true_class)
WTP_4 <- gen_class_items(n4, lam_wtp, c("Green Champion" = 1.0,  "Price Skeptic" = -0.7), true_class)
PS_4  <- gen_class_items(n4, lam_ps,  c("Green Champion" = -1.0, "Price Skeptic" =  0.8), true_class)
GPI_4 <- gen_class_items(n4, lam_gpi, c("Green Champion" = 1.0,  "Price Skeptic" = -0.7), true_class)

# ── Compute scale scores for LCA input ────────────────────────────────────────
df4 <- data.frame(
  true_class = true_class,
  EC_score   = rowMeans(EC_4),
  WTP_score  = rowMeans(WTP_4),
  PS_score   = rowMeans(PS_4),
  GPI_score  = rowMeans(GPI_4)
)
```

## Step 1: Seeing the Pooled vs. Within-Class Pattern

Before running any algorithm, let's look at what happens when you ignore class membership. We'll plot EC versus GPI, coloured first without class information (as a real researcher would see it), then revealing the true classes.

```{r case4-scatter, fig.height = 4.5, fig.width = 10}
#| code-summary: "▶ Plot: pooled scatterplot vs. oracle class-colored version"
p_blind <- ggplot(df4, aes(x = EC_score, y = GPI_score)) +
  geom_point(alpha = 0.3, colour = "grey40", size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, colour = "#d73027", linewidth = 1.1) +
  labs(
    x        = "Environmental Concern (scale score)",
    y        = "Green Purchase Intention (scale score)",
    title    = "What you see: one sample, one regression line",
    subtitle = "Strong positive association — but is it real?"
  ) +
  theme_minimal(base_size = 12)

p_reveal <- ggplot(df4, aes(x = EC_score, y = GPI_score, colour = true_class)) +
  geom_point(alpha = 0.35, size = 1.5) +
  geom_smooth(aes(group = true_class), method = "lm", se = FALSE, linewidth = 1.1) +
  scale_colour_manual(
    values = c("Green Champion" = "#2c7bb6", "Price Skeptic" = "#d73027"),
    name   = "True class"
  ) +
  labs(
    x        = "Environmental Concern (scale score)",
    y        = "Green Purchase Intention (scale score)",
    title    = "The oracle reveal: two classes, two worlds",
    subtitle = "Pooling inflates the EC–GPI correlation; within-class it is weaker"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

# Display side-by-side
gridExtra_available <- requireNamespace("gridExtra", quietly = TRUE)
if (gridExtra_available) {
  gridExtra::grid.arrange(p_blind, p_reveal, ncol = 2)
} else {
  print(p_blind)
  print(p_reveal)
}
```

::: {.callout-warning}
## Simpson's Paradox — when aggregation reverses the pattern

What you just saw in the two panels is a textbook example of **Simpson's Paradox**: a relationship that appears strong and consistent in the aggregate data reverses — or dissolves — when you condition on a hidden grouping variable.

Simpson's Paradox comes up whenever your data have a **nested structure** that you're ignoring. The classic marketing contexts:
- Survey data pooled across different market segments
- Panel data where unit-level effects differ from the aggregate trend
- Multi-country datasets where the country effect dominates individual-level effects
- Any dataset where "the sample" is actually a mixture of distinct populations

The high pooled EC–GPI correlation in the left panel does not mean EC and GPI are measuring the same thing within each group. It means the two groups sit in different positions on both variables — and when you pool them, that between-group variance inflates the correlation. This is *exactly* the mechanism that causes discriminant validity failures when latent subgroups are present (Case 1 connection).
:::

::: {.callout-warning}
## What the two panels show

**Left panel (what you see):** A single strong positive relationship between EC and GPI — suggesting the two constructs are highly correlated. This is exactly the discriminant validity warning sign from Case 1.

**Right panel (oracle reveal):** The relationship separates into two distinct clusters. Within each class, the EC–GPI correlation is much weaker. The high *pooled* correlation is largely a consequence of the two groups sitting at different positions on both axes — not because EC and GPI measure the same thing within a group.

This is called **Simpson's paradox** in statistics — aggregate patterns can completely reverse or disappear when you break the data down by a hidden grouping variable.
:::

## Step 2: The EM Algorithm — Finding the Hidden Groups

The Expectation-Maximisation (EM) algorithm is the engine behind Gaussian mixture models (the continuous-variable equivalent of latent class analysis). The idea is beautifully simple:

1. **Start with a guess**: Randomly assign each person to a class
2. **E step (Expectation)**: Given the current class estimates, compute the probability that each person belongs to each class
3. **M step (Maximization)**: Given those probabilities, update the class means and variances to best fit the data
4. **Repeat** until the assignments stop changing

::: {.callout-note}
## What the EM algorithm is actually doing — a concrete example

Imagine you are looking at a scatterplot of EC and GPI scores and you can see two rough blobs of points but you can't tell where one ends and the other begins.

**Starting guess (Step 0):** The algorithm randomly places two "cluster centers" somewhere in the data. Think of these as two starting hypotheses about where Green Champions and Price Skeptics tend to sit.

**E-step (Expectation):** For each person in the data, the algorithm computes: *"Given where the cluster centers currently are, what is the probability this person belongs to Class 1 vs. Class 2?"* Someone right at the edge between the two clusters gets, say, 55% Class 1 and 45% Class 2. Someone deep inside a cluster gets 99% vs 1%. These are **soft assignments** — everyone has a probability of belonging to each class, not a hard yes/no.

**M-step (Maximization):** Using those probabilities as weights, the algorithm recalculates the best-fitting cluster center for each class. A person who is 99% Class 1 contributes almost fully to the Class 1 center. A person who is 50/50 contributes half to each. The cluster centers shift to better fit the data.

**Repeat:** The algorithm alternates E and M steps until the cluster centers stop moving (converge). At that point, you have your estimated latent classes.

The key advantage over k-means (which you may have heard of): **EM produces probabilities**, not hard assignments. This is both more realistic (most people are not purely one type) and more statistically principled. It also naturally handles clusters of different shapes and sizes.
:::

```{r case4-lca, message=FALSE}
#| code-summary: "▶ Fit Gaussian Mixture Model (EM algorithm, G=2)"
# Scale the input variables for the GMM
# Use EC, WTP, PS, and GPI scores as indicator variables
scale_scores <- df4 |>
  select(EC_score, WTP_score, PS_score, GPI_score) |>
  scale()

# ── Fit Gaussian Mixture Model with 2 classes ─────────────────────────────────
# mclust tries different covariance structures automatically;
# G=2 specifies we want exactly 2 components (latent classes)
set.seed(2025)
gmm_fit <- mclust::Mclust(scale_scores, G = 2, verbose = FALSE)

# ── Extract classification results ────────────────────────────────────────────
df4$estimated_class <- factor(gmm_fit$classification,
                               levels = 1:2,
                               labels = c("Class 1", "Class 2"))

# ── Classification accuracy: how well did EM recover the true classes? ─────────
class_table <- table(True = df4$true_class,
                     Estimated = df4$estimated_class)
kable(class_table,
      caption = "True vs. Estimated Class Assignment (EM Algorithm)")
```

::: {.callout-note}
## Reading the classification table

Each row is a true class; each column is the EM-estimated class. The diagonal shows correct assignments. If the EM algorithm worked well, most Green Champions should land in one column and most Price Skeptics in the other. With four well-separated variables, you should see classification accuracy above 90%.

In real data, you won't have the "True" row — you'll only see the estimated classes. The table here is a diagnostic we can run because we simulated the data.
:::

## Step 3: Comparing Class Profiles

```{r case4-profiles, fig.height = 5}
#| code-summary: "▶ Plot: consumer segment profiles"
# Compute mean scale scores by estimated class
class_profiles <- df4 |>
  group_by(estimated_class) |>
  summarise(
    EC  = mean(EC_score),
    WTP = mean(WTP_score),
    PS  = mean(PS_score),
    GPI = mean(GPI_score),
    .groups = "drop"
  ) |>
  pivot_longer(cols = EC:GPI, names_to = "variable", values_to = "mean_score")

ggplot(class_profiles, aes(x = variable, y = mean_score,
                            fill = estimated_class, group = estimated_class)) +
  geom_col(position = position_dodge(0.7), width = 0.6, colour = "white") +
  geom_hline(yintercept = 0, colour = "grey40") +
  scale_fill_manual(
    values = c("Class 1" = "#2c7bb6", "Class 2" = "#d73027"),
    name   = "Estimated class"
  ) +
  scale_x_discrete(labels = c("EC" = "Environmental\nConcern",
                               "WTP" = "Willingness\nto Pay",
                               "PS"  = "Price\nSensitivity",
                               "GPI" = "Green Purchase\nIntention")) +
  labs(
    x        = NULL,
    y        = "Mean Standardised Score",
    title    = "Class Profiles: The Two Consumer Segments",
    subtitle = "Classes differ systematically across all four variables"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

::: {.callout-note}
## Interpreting the class profiles — reading the bar chart

Work through the chart one variable at a time:

**Environmental Concern (EC):** One class scores substantially above zero (positive standardized mean) and the other scores below zero. This tells you the two classes genuinely differ in how environmentally concerned they are — not a small difference, but a meaningful separation. Whichever class is high on EC is your candidate for "Green Champions."

**Willingness to Pay (WTP):** Does the high-EC class also score high on WTP? If yes, this is a coherent consumer segment: they care about the environment *and* are willing to pay for it. If you saw high EC but low WTP, that would suggest "virtue signaling" rather than true green consumption — a very different strategic implication.

**Price Sensitivity (PS):** Expect this to flip relative to EC and WTP. The environmentally-minded, WTP-high class should score *low* on price sensitivity. The other class should score high on price sensitivity — they prioritize cost over environmental credentials. When variables flip in opposite directions across classes, that confirms the classes are meaningfully distinct, not just statistical noise.

**Green Purchase Intention (GPI):** This is your DV. If the EC/WTP-high class also scores high on GPI, your treatment (green marketing) is likely to resonate with them. The other class's low GPI score tells you the campaign may have little traction there.

**The practical takeaway:** These profiles are the "face" of your latent classes. Before running class-specific analyses, you need to be able to give each class a substantive name and interpretation. If the profiles don't make sense theoretically, reconsider whether the number of classes you specified (G=2) is right — or whether the variables you used for the GMM are the right ones.
:::

## Step 4: What to Do Next — Class-Specific Analyses

```{r case4-class-effects}
#| code-summary: "▶ Treatment effects by latent class"
# Treatment effect of the green marketing campaign, estimated separately per class
# Recall: Green Champions have high GPI regardless; Skeptics barely move

# For illustration, simulate a treatment variable
set.seed(2025)
df4$treatment <- sample(c(0L, 1L), n4, replace = TRUE)

# Naive pooled model
m_pooled <- lm(GPI_score ~ treatment, data = df4)

# Class-specific models
m_class1 <- lm(GPI_score ~ treatment,
               data = filter(df4, estimated_class == "Class 1"))
m_class2 <- lm(GPI_score ~ treatment,
               data = filter(df4, estimated_class == "Class 2"))

class_effects <- data.frame(
  Model = c("Pooled (ignores classes)",
            "Class 1 (Green Champions)",
            "Class 2 (Price Skeptics)"),
  `Treatment effect on GPI` = round(c(coef(m_pooled)["treatment"],
                                      coef(m_class1)["treatment"],
                                      coef(m_class2)["treatment"]), 3),
  `N` = c(n4,
          sum(df4$estimated_class == "Class 1"),
          sum(df4$estimated_class == "Class 2"))
)

kable(class_effects,
      col.names = c("Model", "Treatment Effect on GPI", "N"),
      caption   = "Treatment Effect Estimates: Pooled vs. Class-Specific")
```

::: {.callout-important}
## What the class-specific estimates reveal

The pooled estimate is a weighted average of two very different effects. If Green Champions respond strongly to the campaign and Price Skeptics barely respond, the pooled estimate sits somewhere in the middle — understating the effect for Champions and overstating it for Skeptics. Neither number is particularly useful for a manager deciding how to allocate a marketing budget.

**Key takeaway for researchers**: If your sample contains meaningful subgroups with different response patterns, any single estimate of "the effect" is an artificial average. Always explore whether your findings are robust across identifiable segments. Latent class / Gaussian mixture models are one tool for discovering those segments when you don't know them in advance.
:::

## What to Look For in Your Own Data

::: {.callout-note}
## Checklist: signs that latent subgroups may be distorting your results

1. **Unexpectedly high HTMT / factor correlations** in pooled data that seem theoretically implausible — check whether subgroups could be inflating the pooled correlation (Case 1 connection)
2. **Bimodal or non-normal distributions** in your scale scores — histogram with two humps suggests a mixture
3. **Unstable regression coefficients** across subsamples (men vs. women, different markets, etc.)
4. **Large heterogeneity in individual responses** — residual variance is very high despite significant effects
5. **Substantive theory** predicting different consumer types — always the best starting point

When you suspect latent classes, fit GMMs with G=1, G=2, G=3, G=4 and compare BIC. In mclust, **higher BIC is better** — look for the number of classes where BIC peaks.
:::

```{r case4-bic}
#| code-summary: "▶ Plot: BIC model selection (G=1 to G=4)"
# Fit models for 1–4 classes and compare BIC
# mclust BIC: HIGHER is better (mclust uses a convention opposite to standard BIC)
# mclust maximizes BIC rather than minimizing it — the HIGHEST BIC = best fit
gmm_fits <- lapply(1:4, function(g) mclust::Mclust(scale_scores, G = g, verbose = FALSE))
bic_vals <- sapply(gmm_fits, function(m) max(m$BIC))   # mclust returns BIC matrix

bic_df <- data.frame(
  Classes = 1:4,
  BIC     = round(bic_vals, 1)
)

ggplot(bic_df, aes(x = factor(Classes), y = BIC, fill = factor(Classes == which.max(BIC)))) +
  geom_col(width = 0.55, colour = "white") +
  geom_text(aes(label = round(BIC, 0)), vjust = -0.5, fontface = "bold", size = 4) +
  scale_fill_manual(values = c("FALSE" = "#91bfdb", "TRUE" = "#2c7bb6"),
                    guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    x        = "Number of Latent Classes",
    y        = "BIC (mclust scale: higher = better fit)",
    title    = "Choosing the Number of Classes: BIC by Model",
    subtitle = "The highlighted bar is the best-fitting model — look for the peak then leveling off"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank())
```

::: {.callout-note}
## How to read the BIC chart

In mclust, **higher BIC is better** (mclust uses a flipped convention from textbook BIC). Look for the number of classes where BIC peaks — that is the model the data support. A clear peak at G=2 with lower values at G=1 and G=3 confirms a two-class solution. If BIC keeps rising to G=3 or G=4, you may need more classes — but always check that additional classes are substantively interpretable, not just statistical noise.
:::

## Other Methods for Detecting Latent Subgroups

Gaussian mixture models (mclust) are the continuous-variable analogue of latent class analysis. Related methods:

- **Latent Class Analysis (LCA) proper:** For categorical indicators (e.g., nominal responses, binary items), LCA fits discrete latent classes using the poLCA package in R. The logic is identical to the GMM shown here, but designed for non-continuous data.
- **Latent Profile Analysis (LPA):** When indicators are continuous (as in our scale scores), LCA applied to continuous data is technically LPA. The tidyLPA package provides a user-friendly interface and compares multiple model specifications.
- **K-means clustering:** Simpler and faster than GMM but makes stronger assumptions (spherical clusters of equal size). A good first pass but less principled for inference. Use it to explore; use GMM to confirm.
- **Hierarchical clustering:** Builds a dendrogram of observations. No need to specify the number of clusters in advance — the dendrogram lets you visually choose where to cut. Good for small-to-medium datasets; computationally expensive for large ones.
- **Moderated regression / interaction effects:** If you have a theoretically motivated segmentation variable (age, gender, market), a simpler approach than LCA is to test whether your key effects differ by segment using interaction terms.
- **Finite mixture models for regression (flexmix package in R):** Extends the GMM idea to regression contexts — fits separate regression lines for each latent class simultaneously. This is the most direct analog to "class-specific treatment effects" if you want a fully model-based approach.


---


# Case 5: Outlier Detection and Robust Estimation {#sec-case5}

## Connecting to the Earlier Cases

Case 4 showed what happens when your sample contains two fundamentally different types of respondents. Case 5 extends this logic to observational data — specifically, to the problem of **outliers**.

Not all outliers are the same. Two important distinctions:

- **Collective outliers** are a group of observations that form their own cluster, far from the rest of the data. They usually represent a *different population* — like luxury flagship stores in a dataset of regular convenience stores. These are a version of the latent-subgroup problem from Case 4: if you include them, your regression line tries to average over two different data-generating processes.

- **Point outliers** (or global outliers) are individual observations that are simply extreme on one or more variables. They may represent data errors, rare events, or genuinely unusual cases. They pull the regression line toward them and inflate standard errors.

Both types distort your results — but in different ways, and the solutions are different.

::: {.callout-note}
## The connection to discriminant validity and latent subgroups

Collective outliers are, structurally, the same problem as latent subgroups. A cluster of luxury flagship stores isn't "wrong data" — it's a different population with a different data-generating process. Including them in your pooled model is like combining Green Champions and Price Skeptics: the resulting coefficient doesn't accurately describe either group.

Point outliers, by contrast, are statistical problems rather than conceptual ones. A single data entry error or an unusually extreme store doesn't imply a hidden population — it's just a disruptive data point that needs to be handled carefully.
:::

## Simulating the Dataset

We create a store-level dataset with three groups:

- **Regular stores** (n=500): Standard advertising–sales relationship
- **Luxury flagships** (n=8): Collective outliers — much higher advertising and sales, forming their own cluster
- **Point outliers** (n=5): Individual stores with implausibly extreme values

```{r case5-simulate}
#| code-summary: "▶ Simulate store dataset with collective and point outliers"
set.seed(2025)

# ── Regular stores ─────────────────────────────────────────────────────────────
n_reg <- 500
adv_reg   <- rnorm(n_reg, mean = 5, sd = 1.5)
sales_reg <- 0.4 * adv_reg + rnorm(n_reg, sd = 1.2)

# ── Luxury flagship stores (collective outliers) ───────────────────────────────
# They advertise much more AND have much higher sales — different economic model
n_lux <- 8
adv_lux   <- rnorm(n_lux, mean = 18, sd = 1.2)
sales_lux <- 0.3 * adv_lux + rnorm(n_lux, sd = 0.8) + 10   # elevated baseline

# ── Point outliers (individual extreme observations) ───────────────────────────
n_pt  <- 5
adv_pt   <- c(22, 1, 3, 19, 2)          # extreme advertising values
sales_pt <- c(2, 18, -1, 25, 15)        # extreme / inconsistent sales

# ── Combine into single dataset ────────────────────────────────────────────────
df5 <- data.frame(
  advertising = c(adv_reg,    adv_lux,    adv_pt),
  sales       = c(sales_reg,  sales_lux,  sales_pt),
  store_type  = c(rep("Regular",   n_reg),
                  rep("Luxury flagship", n_lux),
                  rep("Point outlier",   n_pt))
)
```

## Step 1: Visualising the Three Types of Stores

```{r case5-scatter, fig.height = 5}
#| code-summary: "▶ Plot: stores colored by true type"
ggplot(df5, aes(x = advertising, y = sales, colour = store_type, shape = store_type)) +
  geom_point(aes(size = store_type), alpha = 0.7) +
  geom_smooth(data = filter(df5, store_type == "Regular"),
              method = "lm", se = TRUE, colour = "#4575b4",
              linewidth = 1.0, inherit.aes = FALSE,
              aes(x = advertising, y = sales)) +
  scale_colour_manual(
    values = c("Regular"         = "#4575b4",
               "Luxury flagship" = "#d73027",
               "Point outlier"   = "#1a9641"),
    name   = "Store type"
  ) +
  scale_shape_manual(
    values = c("Regular" = 16, "Luxury flagship" = 17, "Point outlier" = 8),
    name   = "Store type"
  ) +
  scale_size_manual(
    values = c("Regular" = 1.5, "Luxury flagship" = 3.5, "Point outlier" = 4),
    name   = "Store type"
  ) +
  labs(
    x        = "Advertising Spend (standardised)",
    y        = "Sales (standardised)",
    title    = "Three Types of Observations in the Dataset",
    subtitle = "Luxury flagships form their own cluster; point outliers are scattered extremes"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

::: {.callout-warning}
## What to notice

- **Regular stores** (blue): tight cluster with a clear advertising→sales relationship. This is the population you actually want to model.
- **Luxury flagships** (red triangles): a separate cluster in the upper right. They form a coherent group — this is the collective outlier problem. Including them will rotate your regression line toward theirs.
- **Point outliers** (green stars): scattered in unusual positions — very high advertising with low sales, or very low advertising with high sales. These are individually anomalous.
:::

## Step 2: Detecting Outliers with Local Outlier Factor (LOF)

**Local Outlier Factor (LOF)** is a density-based method: it compares how densely packed each observation's neighbourhood is relative to its neighbours' neighbourhoods. If a point is in a sparse region surrounded by denser regions, it gets a high LOF score.

- LOF ≈ 1: Normal point (density similar to neighbours)
- LOF > 2–3: Moderate outlier (noticeably sparser than surroundings)
- LOF >> 3: Strong outlier

```{r case5-lof, fig.height = 5}
#| code-summary: "▶ Compute LOF scores and visualize"
# Standardise the variables before computing LOF
df5_scaled <- data.frame(
  advertising = scale(df5$advertising),
  sales       = scale(df5$sales)
)

# Compute LOF scores (minPts = number of neighbours to consider)
df5$lof_score <- dbscan::lof(df5_scaled, minPts = 10)

# Visualise LOF scores
ggplot(df5, aes(x = advertising, y = sales,
                colour = lof_score, size = lof_score)) +
  geom_point(alpha = 0.8) +
  scale_colour_gradient2(low = "#4575b4", mid = "#ffffbf", high = "#d73027",
                         midpoint = 2, name = "LOF score") +
  scale_size_continuous(range = c(1, 5), name = "LOF score") +
  labs(
    x        = "Advertising Spend",
    y        = "Sales",
    title    = "Local Outlier Factor (LOF) Scores",
    subtitle = "Warmer colour and larger point = more outlier-like"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

```{r case5-lof-table}
#| code-summary: "▶ Table: top outliers by LOF score"
# Show the top outliers by LOF score
df5 |>
  arrange(desc(lof_score)) |>
  select(advertising, sales, store_type, lof_score) |>
  head(15) |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  kable(
    col.names = c("Advertising", "Sales", "Store Type", "LOF Score"),
    caption   = "Top 15 Observations by LOF Score (highest = most outlier-like)"
  )
```

::: {.callout-note}
## Reading the LOF scores — and what to do with them

**What LOF is telling you:** Each store's LOF score compares how densely packed its neighborhood is relative to its neighbors' neighborhoods. A score near 1.0 means the store is in a region as dense as its surroundings — it looks like a typical store. A score of 2 or 3 means the store is in a region roughly 2–3 times sparser than its neighbors' neighborhoods — it is noticeably isolated. Very high scores (5+) indicate severe outliers.

**The critical point about classification:** LOF scores do NOT give you clean "outlier" vs "not outlier" categories. There is no universal threshold that separates outliers from normal observations — you have to choose one, and that choice involves judgment. Common practice is to flag observations with LOF > 2 or LOF > 3 for investigation, but these cutoffs are conventions, not laws of nature.

**Practical steps for each high-LOF observation:**
1. Look at the observation. What are its actual values? Does anything look unusual?
2. Is it a data quality issue (entry error, unit mismatch)? Fix or remove.
3. Is it a legitimately extreme but real observation (a genuinely exceptional store)? Keep and investigate — it may be telling you something important about your data.
4. Does it form part of a coherent group of unusual observations? If yes, that's a collective outlier — treat it as a Case 4 problem (latent subgroup), not a Case 5 problem.

**Never delete observations just because they have high LOF scores.** LOF is a signal to investigate, not an automatic deletion criterion.
:::

## Step 3: DBSCAN Clustering to Find the Collective Outlier Cluster

**DBSCAN** (Density-Based Spatial Clustering of Applications with Noise) finds clusters of any shape in your data, and labels anything too sparse to belong to a cluster as **noise (cluster 0)**. This is useful for finding collective outliers: they form their own cluster (separate from the main group), while true point outliers get labelled as noise.

```{r case5-dbscan, fig.height = 5}
#| code-summary: "▶ DBSCAN clustering — find collective outlier cluster"
# Fit DBSCAN
# eps: maximum distance to be considered a "neighbour"
# minPts: minimum cluster size
dbscan_fit <- dbscan::dbscan(df5_scaled, eps = 0.4, minPts = 4)
df5$cluster <- factor(dbscan_fit$cluster)

ggplot(df5, aes(x = advertising, y = sales,
                colour = cluster, shape = cluster)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_colour_manual(
    values = c("0" = "#d73027",   # noise = red (point outliers)
               "1" = "#4575b4",   # main cluster = blue
               "2" = "#1a9641",   # luxury cluster = green
               "3" = "#fdae61"),  # any additional cluster
    labels = c("0" = "Noise / point outlier",
               "1" = "Main cluster (regular stores)",
               "2" = "Collective outlier cluster (luxury)",
               "3" = "Additional cluster"),
    name   = "DBSCAN cluster",
    drop   = FALSE
  ) +
  scale_shape_manual(
    values = c("0" = 8, "1" = 16, "2" = 17, "3" = 15),
    labels = c("0" = "Noise / point outlier",
               "1" = "Main cluster (regular stores)",
               "2" = "Collective outlier cluster (luxury)",
               "3" = "Additional cluster"),
    name   = "DBSCAN cluster",
    drop   = FALSE
  ) +
  labs(
    x        = "Advertising Spend",
    y        = "Sales",
    title    = "DBSCAN Clustering: Finding Collective and Point Outliers",
    subtitle = "Cluster 0 = noise (point outliers); separate clusters = collective outlier groups"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

::: {.callout-note}
## How to read the DBSCAN output — and the limits of inferred classification

**What the clusters mean:**
- **Cluster 0 (noise/red):** Points too isolated to belong to any cluster. These are candidates for point outliers — individually anomalous observations. Check each one.
- **Cluster 1 (main mass):** The main group of regular stores. This is your target population for most analyses.
- **Cluster 2+ (secondary clusters):** Coherent groups separated from the main mass — collective outliers. These are a latent subgroup problem (Case 4), not a statistical noise problem.

**Important: these are inferred, probabilistic classifications.** DBSCAN does not know "the truth" about which stores are outliers any more than you do. It finds density-based patterns in the data, but:
- The clusters depend on your choice of `eps` and `minPts` — different settings will produce different cluster assignments
- A store assigned to cluster 0 (noise) may simply be on the outskirts of a real cluster that DBSCAN missed with your parameter settings
- A store assigned to the main cluster may still be unusual in ways DBSCAN cannot see with just two variables

**The classifications are a starting point, not a verdict.** Use them to flag observations for further investigation. Do not treat cluster assignment as definitive.

**Choosing eps and minPts:** There is no single correct answer. A practical approach: plot the k-nearest-neighbor distances (sorted) and look for an "elbow" — the point where the distance curve bends sharply. That elbow value is a good starting point for eps. For minPts, a common default is 2 × (number of dimensions in your data).
:::

## Step 4: Robust Regression — No Need to Delete Outliers

The traditional advice — delete the outliers and refit — creates its own problems: it introduces researcher degrees of freedom (which ones do you delete?), it loses information, and it's post-hoc. A better approach is **robust regression**, which automatically downweights observations that don't fit the main data pattern.

`MASS::rlm()` uses **M-estimation**: instead of minimising the sum of squared residuals (which gives disproportionate influence to large residuals), it minimises a less sensitive function that penalises large residuals less severely.

```{r case5-robust, fig.height = 4.5}
#| code-summary: "▶ OLS vs. robust regression (rlm) coefficient comparison"
# ── Fit three regression models ───────────────────────────────────────────────
m_ols_all    <- lm(sales ~ advertising, data = df5)              # OLS with all data
m_ols_reg    <- lm(sales ~ advertising,                          # OLS on regular only
                   data = filter(df5, store_type == "Regular"))
m_robust     <- MASS::rlm(sales ~ advertising, data = df5)       # Robust on all data

# ── Compare coefficients ───────────────────────────────────────────────────────
coef_compare5 <- data.frame(
  Model = c("OLS — all stores (distorted)",
            "OLS — regular stores only (oracle)",
            "Robust regression (rlm) — all stores"),
  `Advertising coefficient` = round(c(coef(m_ols_all)["advertising"],
                                      coef(m_ols_reg)["advertising"],
                                      coef(m_robust)["advertising"]), 3)
)

kable(coef_compare5,
      col.names = c("Model", "Advertising Coefficient"),
      caption   = "OLS vs. Robust Regression: How Much Do Outliers Matter?")
```

```{r case5-comparison-plot, fig.height = 5}
#| code-summary: "▶ Plot: regression line comparison (OLS vs. robust)"
# Generate clean prediction lines for all three models
x_seq <- seq(min(df5$advertising), max(df5$advertising), length.out = 300)
newdat <- data.frame(advertising = x_seq)

pred_lines <- data.frame(
  advertising  = rep(x_seq, 3),
  sales        = c(predict(m_ols_all,  newdata = newdat),
                   predict(m_ols_reg,  newdata = newdat),
                   predict(m_robust,   newdata = newdat)),
  model        = rep(c("OLS — all stores (distorted)",
                       "OLS — regular only (benchmark)",
                       "Robust rlm — all stores"),
                     each = length(x_seq))
)

line_colors <- c(
  "OLS — all stores (distorted)"    = "#d73027",
  "OLS — regular only (benchmark)"  = "#4575b4",
  "Robust rlm — all stores"         = "#1a9641"
)
line_types  <- c(
  "OLS — all stores (distorted)"    = "dotted",
  "OLS — regular only (benchmark)"  = "solid",
  "Robust rlm — all stores"         = "dashed"
)

ggplot() +
  # Points: just color by store type, no size variation
  geom_point(data = df5,
             aes(x = advertising, y = sales, colour = store_type),
             alpha = 0.55, size = 2) +
  # Regression lines
  geom_line(data = pred_lines,
            aes(x = advertising, y = sales,
                colour = model, linetype = model),
            linewidth = 1.2) +
  scale_colour_manual(
    values = c(
      "Regular"                             = "grey60",
      "Luxury flagship"                     = "#e8998d",
      "Point outlier"                       = "#a8c5a0",
      "OLS — all stores (distorted)"        = "#d73027",
      "OLS — regular only (benchmark)"      = "#4575b4",
      "Robust rlm — all stores"             = "#1a9641"
    ),
    breaks = c("OLS — all stores (distorted)",
               "OLS — regular only (benchmark)",
               "Robust rlm — all stores"),
    name = "Regression line"
  ) +
  scale_linetype_manual(
    values = line_types,
    name   = "Regression line"
  ) +
  guides(
    colour   = guide_legend(order = 1, override.aes = list(linewidth = 1.5)),
    linetype = guide_legend(order = 1)
  ) +
  labs(
    x        = "Advertising Spend",
    y        = "Sales",
    title    = "OLS vs. Robust Regression: Effect of Outliers on Slope Estimates",
    subtitle = paste0(
      "Red dotted = OLS distorted by outliers  |  ",
      "Blue solid = benchmark (regular only)  |  ",
      "Green dashed = robust rlm"
    )
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position  = "bottom",
    legend.direction = "horizontal",
    legend.key.width = unit(1.8, "cm")
  )
```

::: {.callout-important}
## The core lesson from robust regression

**OLS (all stores):** Pulled toward the luxury cluster and point outliers, producing a distorted slope. If you use this coefficient to predict sales for a regular store, you'll be systematically wrong.

**OLS (regular only):** The "oracle" answer — what you'd get if you correctly identified and excluded the other groups. But in real research, you don't always know which stores to exclude.

**Robust regression (rlm, all stores):** Automatically downweights observations with large residuals (the outliers and luxury stores fit poorly under the regular-store model, so they get low weights). The resulting coefficient is close to the regular-only estimate — without having to decide in advance which observations to delete.

**Practical advice:**
1. Always plot your data and check for visual outliers before running any regression
2. Use LOF and/or DBSCAN to systematically detect outliers you might miss visually
3. For collective outliers (like luxury stores), investigate whether they represent a distinct subgroup — if so, treat them as a Case 4 problem
4. For point outliers, verify the data and report results with and without them as a robustness check
5. Use robust regression as your primary estimator when you suspect outliers but are not certain which observations to remove
:::

## Checking How Much Your Results Depend on Outliers

A simple sensitivity check: compare OLS and robust regression. If they give substantially different results, outliers are influencing your conclusions. If they agree, outliers aren't driving your findings.

```{r case5-sensitivity}
#| code-summary: "▶ Sensitivity table: OLS vs. rlm coefficients"
# Extract and compare key summary stats
sens_df <- data.frame(
  Method = c("OLS (all stores)", "Robust rlm (all stores)"),
  `Intercept` = round(c(coef(m_ols_all)[1], coef(m_robust)[1]), 3),
  `Advertising slope` = round(c(coef(m_ols_all)[2], coef(m_robust)[2]), 3)
)

kable(sens_df,
      col.names = c("Method", "Intercept", "Advertising Slope"),
      caption   = "Sensitivity Check: OLS vs. Robust Regression Coefficients")
```

::: {.callout-note}
## Reporting guidance

When you run a robust regression as a sensitivity check and it tells a *different* story from OLS:

- Report both estimates in your paper
- Investigate *which* observations are driving the OLS estimates (use LOF scores, DBSCAN clusters, or Cook's distance)
- Describe these observations substantively: are they a separate population? A data quality issue?
- Be transparent about the sensitivity of your conclusions to these observations

When OLS and robust regression agree: briefly note the robustness check as evidence that your results are not driven by a small number of influential observations.
:::

## Other Methods for Outlier Detection and Robust Estimation

LOF + DBSCAN and robust regression are practical entry points. The broader toolkit:

- **Mahalanobis distance:** A multivariate distance metric that accounts for the covariances among variables. Points far from the centroid in Mahalanobis distance space are multivariate outliers. Computationally simple; assumes roughly elliptical distributions.
- **Cook's distance:** In regression, measures how much each observation influences the fitted values. Observations with Cook's D > 4/n are conventionally flagged as influential. Complementary to LOF (which doesn't know about the regression).
- **Minimum Covariance Determinant (MCD) — robustcov / rrcov packages:** A robust covariance estimator that finds the subset of observations with the smallest covariance determinant, effectively excluding outliers from the covariance estimation. Use it to get robust Mahalanobis distances.
- **Isolation Forest (isotree package in R):** A tree-based outlier detection method. Anomalies are easier to isolate than normal points, so they end up in shorter trees. Scales well to large datasets. No distance computation needed.
- **One-class SVM:** Learns the boundary of "normal" observations and flags points outside it. Useful when you have a large clean training set and want to detect anomalies in new data.
- **Quantile regression (quantreg package):** Instead of modeling the mean, model multiple quantiles of Y. Outliers that pull the mean line don't affect quantile regression as severely. Useful when you care about the full distribution, not just the central tendency.
- **Sandwich standard errors (sandwich package):** If heteroskedasticity (the fan pattern from Case 2) is due to outliers or model misspecification, heteroskedasticity-consistent (HC) standard errors give valid inference without reweighting observations.


## References

Pieters, R., Srivastava, J., & Bagchi, R. (2025). Improving the discriminant validation of multi-item scales. *Journal of Marketing Research*. https://doi.org/10.1177/00222437251322089

Henseler, J., Ringle, C. M., & Sarstedt, M. (2015). A new criterion for assessing discriminant validity in variance-based structural equation modeling. *Journal of the Academy of Marketing Science*, 43(1), 115–135.

Millsap, R. E., & Kwok, O.-M. (2004). Evaluating the impact of partial factor loading and intercept invariance on selection in two populations. *Psychological Methods*, 9(1), 93–115.

Putnick, D. L., & Bornstein, M. H. (2016). Measurement invariance conventions and reporting: The state of the art and future directions for psychological research. *Developmental Review*, 41, 71–90.

Rosseel, Y. (2012). lavaan: An R package for structural equation modeling. *Journal of Statistical Software*, 48(2), 1–36.

Scrucca, L., Fop, M., Murphy, T. B., & Raftery, A. E. (2016). mclust 5: Clustering, classification and density estimation using Gaussian finite mixture models. *The R Journal*, 8(1), 289–317.

Breunig, M. M., Kriegel, H.-P., Ng, R. T., & Sander, J. (2000). LOF: Identifying density-based local outliers. *ACM SIGMOD Record*, 29(2), 93–104.

Ester, M., Kriegel, H.-P., Sander, J., & Xu, X. (1996). A density-based algorithm for discovering clusters in large spatial databases with noise. *Proceedings of the 2nd International Conference on Knowledge Discovery and Data Mining* (KDD-96), 226–231.

Venables, W. N., & Ripley, B. D. (2002). *Modern applied statistics with S* (4th ed.). Springer. [Source for MASS::rlm robust regression]

Wooldridge, J. M. (2010). *Econometric analysis of cross section and panel data* (2nd ed.). MIT Press.

---

*Tutorial prepared for the Advanced Methods Workshop · Module 1: Measurement · Rotterdam School of Management*
